---
title: "QC plots from AlignQC and internal priming analysis"
author: "Yupei"
date: '`r Sys.Date()`'
version: "v1.7"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
---

```{r setup_params_template, include=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
# Make params list for running outside of Rmarkdown
params <- list(
  random_seed = 2024,
  cache_dir = NULL,  # Set to a directory path to enable caching
  fig.path = NULL    # Set to a directory path to save figures, or NULL to not save
)
```

```{r setup_knitr_and_libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(cache = FALSE, cache.path = params$cache_dir)
knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),  # Enable caching only if cache_dir is provided
  fig.width = 15,                     # Set default figure width
  fig.height = 10,                    # Set default figure height
  message = FALSE,                    # Suppress messages in the output
  warning = FALSE,                    # Suppress warnings in the output
  fig.path = params$fig.path,         # Set figure path (NULL if not provided)
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL  # Set output devices if fig.path is provided
)

options(readr.show_col_types = FALSE)

# Set cache.path only if params$cache_dir is not NULL
if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache.path = params$cache_dir)
}


library(tidyr)
library(dplyr)
library(purrr)
library(readr)
library(tibble)
library(stringr)
library(ggplot2)

save_plot_data <- function(df, name) {
  if (is.null(params$fig.path)) {
    return(invisible(NULL))
  }
  out_dir <- params$fig.path
  if (!dir.exists(out_dir)) {
    dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)
  }
  readr::write_csv(df, file.path(out_dir, paste0(name, ".csv")))
  invisible(NULL)
}
```

```{r define_bulk_color_palette}

color_palette <- c(
  PacBio = "#df1995",
  PacBio_IP_filtered = "#a50026",
  ONT = "#00789b",
  ONT_1 = "#04476c",
  ONT_2 = "#24cdcd",
  Illumina = "#e88b20"
)
```
# BULK
## AlignQC
### Alignment stats
```{r bulk_alignqc_alignment_stats_load}
# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_bulk/result/qc/AlignQC/', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "alignment.stats.txt", full.names = TRUE)

# SR
dirs <- list.dirs("/vast/projects/LongBench/analysis/sr_bulk/result/qc/AlignQC/", recursive = TRUE, full.names = TRUE)
stats_files <- c(stats_files, list.files(dirs, pattern = "alignment.stats.txt", full.names = TRUE))


# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts) - 2]
  
  # Read file and convert to tibble
  stats <- read.table(file, sep = "\t", col.names = c("metric", "value"), fill = TRUE)
  
  # Add platform and type columns
  stats <- stats %>%
    mutate(platform = platform) %>%
    pivot_wider(names_from = metric, values_from = value)
  
  return(stats)
}

# Parse all files and combine into a single tibble
alignment.stats <- map_dfr(stats_files, parse_stats)

# split the platform column into two columns
alignment.stats <- alignment.stats %>%
  mutate(
    Platform = case_when(
      str_detect(platform, "pb.*IP_filtered") ~ "PacBio IP filtered",
      str_detect(platform, "dRNA.*IP_filtered") ~ "ONT dRNA IP filtered",
      str_detect(platform, "ont_bulk.*IP_filtered") ~ "ONT cDNA IP filtered",
      str_detect(platform, "dRNA") ~ "ONT dRNA",
      str_detect(platform, "ont_bulk") ~ "ONT cDNA",
      str_detect(platform, "pb") ~ "PacBio",
      TRUE ~ "Illumina",
    ),
    # remove *_
    Cell_line = gsub(".*_", "", platform)
  ) %>% select(-platform)


# Scale UNALIGNED_READS ALIGNED_READS SINGLE_ALIGN_READS GAPPED_ALIGN_READS CHIMERA_ALIGN_READS TRANSCHIMERA_ALIGN_READS SELFCHIMERA_ALIGN_READS by TOTAL_READS
alignment.stats <- alignment.stats %>%
  mutate_at(
    vars(UNALIGNED_READS:SELFCHIMERA_ALIGN_READS), ~ . / TOTAL_READS * 100
  ) %>%
  mutate_at(
    vars(UNALIGNED_BASES:SELFCHIMERA_ALIGN_BASES), ~ . / TOTAL_BASES * 100
  )

# remove TOTAL_READS and TOTAL_BASES
alignment.stats <- alignment.stats %>%
  select(-c(TOTAL_READS, TOTAL_BASES))

# Keep the IP filtered as a separate category tibble
alignment.stats.IP_filter <- alignment.stats %>% filter(str_detect(Platform, "IP filtered"))
alignment.stats <- alignment.stats %>% filter(!str_detect(Platform, "IP filtered"))
```


```{r bulk_alignqc_alignment_reads_plot, fig.height=6, fig.width=10}
# generate ggplot for each column, facet by different columns, x is dataset, y is value
library(ggplot2)
df_long <- alignment.stats %>%
  select(c(Platform,Cell_line, ALIGNED_READS, GAPPED_ALIGN_READS, CHIMERA_ALIGN_READS, TRANSCHIMERA_ALIGN_READS, SELFCHIMERA_ALIGN_READS)) %>%
  pivot_longer(cols = -c(Platform, Cell_line), names_to = "Metric", values_to = "Value")

df_summary <- df_long %>%
  group_by(Platform, Metric) %>%
  summarise(
    Mean_Value = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

save_plot_data(df_long, "bulk_alignqc_alignment_reads_raw_points")

ggplot() +
  # Bars for the mean
  geom_bar(data = df_summary, aes(x = Platform, y = Mean_Value, fill = Platform), 
           stat = "identity", position = "dodge", alpha = 0.7) +
  # Error bars (Mean ± SD)
  geom_errorbar(data = df_summary, aes(x = Platform, ymin = Mean_Value - SD, ymax = Mean_Value + SD), 
                width = 0.2, position = position_dodge(0.9)) +
  # Dots for individual values
  geom_point(data = df_long, aes(x = Platform, y = Value, color = Platform), 
             position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~Metric, scales = "free_y") +  # Facet by Metric
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(x = "Platform", y = "Percentage") +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname) +
  scale_color_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname)  # Match dot color to fill
```


```{r bulk_alignqc_alignment_bases_plot,fig.height=6, fig.width=10}
df_long <- alignment.stats %>% select(c(Platform,Cell_line,ALIGNED_BASES,GAPPED_ALIGN_BASES,CHIMERA_ALIGN_BASES,TRANSCHIMERA_ALIGN_BASES,SELFCHIMERA_ALIGN_BASES)) %>%
  pivot_longer(cols = -c(Platform, Cell_line), names_to = "Metric", values_to = "Value")
df_summary <- df_long %>%
  group_by(Platform, Metric) %>%
  summarise(
    Mean_Value = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

save_plot_data(df_long, "bulk_alignqc_alignment_bases_raw_points")

ggplot() +
  # Bars for the mean
  geom_bar(data = df_summary, aes(x = Platform, y = Mean_Value, fill = Platform), 
           stat = "identity", position = "dodge", alpha = 0.7) +
  # Error bars (Mean ± SD)
  geom_errorbar(data = df_summary, aes(x = Platform, ymin = Mean_Value - SD, ymax = Mean_Value + SD), 
                width = 0.2, position = position_dodge(0.9)) +
  # Dots for individual values
  geom_point(data = df_long, aes(x = Platform, y = Value, color = Platform), 
             position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~Metric, scales = "free_y") +  # Facet by Metric
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(x = "Platform", y = "Percentage") +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname) +
  scale_color_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname)  # Match dot color to fill

```

### Error state
```{r bulk_alignqc_error_state_plot}

# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_bulk/result/qc/AlignQC/', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "error_stats.txt", full.names = TRUE)

# SR
dirs <- list.dirs("/vast/projects/LongBench/analysis/sr_bulk/result/qc/AlignQC/", recursive = TRUE, full.names = TRUE)
stats_files <- c(stats_files, list.files(dirs, pattern = "error_stats.txt", full.names = TRUE))

# Parse all files and combine into a single tibble
error.stats <- map_dfr(stats_files, parse_stats)
# split the platform column into two columns
error.stats <- error.stats %>%
  mutate(
    Platform = case_when(
      str_detect(platform, "pb.*IP_filtered") ~ "PacBio IP filtered",
      str_detect(platform, "dRNA.*IP_filtered") ~ "ONT dRNA IP filtered",
      str_detect(platform, "ont_bulk.*IP_filtered") ~ "ONT cDNA IP filtered",
      str_detect(platform, "dRNA") ~ "ONT dRNA",
      str_detect(platform, "ont_bulk") ~ "ONT cDNA",
      str_detect(platform, "pb") ~ "PacBio",
      TRUE ~ "Illumina",
    ),
    # remove *_
    Cell_line = gsub(".*_", "", platform)
  ) %>% select(-platform) %>% filter(!str_detect(Platform, "IP filtered"))

error.stats <- error.stats %>%
  mutate(
    ANY_ERROR = ANY_ERROR / ALIGNMENT_BASES * 100,
    MISMATCHES = MISMATCHES / ALIGNMENT_BASES * 100,
    ANY_DELETION = ANY_DELETION / ALIGNMENT_BASES * 100,
    ANY_INSERTION = ANY_INSERTION / ALIGNMENT_BASES * 100
  )

df_long <- error.stats %>%
  select(c(
    Platform,
    Cell_line,
    ANY_ERROR,
    MISMATCHES,
    ANY_DELETION,
    ANY_INSERTION
  )) %>%
  pivot_longer(cols = -c(Platform, Cell_line), names_to = "Metric", values_to = "Value")

df_summary <- df_long %>%
  group_by(Platform, Metric) %>%
  summarise(
    Mean_Value = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

save_plot_data(df_long, "bulk_alignqc_error_state_raw_points")

ggplot() +
  # Bars for the mean
  geom_bar(data = df_summary, aes(x = Platform, y = Mean_Value, fill = Platform), 
           stat = "identity", position = "dodge", alpha = 0.7) +
  # Error bars (Mean ± SD)
  geom_errorbar(data = df_summary, aes(x = Platform, ymin = Mean_Value - SD, ymax = Mean_Value + SD), 
                width = 0.2, position = position_dodge(0.9)) +
  # Dots for individual values
  geom_point(data = df_long, aes(x = Platform, y = Value, color = Platform), 
             position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~Metric, scales = "free_y") +  # Facet by Metric
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(x = "Platform", y = "Percentage") +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname) +
  scale_color_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname)  # Match dot color to fill
```

### Junction shift
```{r bulk_alignqc_junction_shift_load}
# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_bulk/result/qc/AlignQC/', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "junvar.txt", full.names = TRUE)

# SR
dirs <- list.dirs("/vast/projects/LongBench/analysis/sr_bulk/result/qc/AlignQC/", recursive = TRUE, full.names = TRUE)
stats_files <- c(stats_files, list.files(dirs, pattern = "junvar.txt", full.names = TRUE))

# Parse all files and combine into a single tibble
junc.var <- map_dfr(stats_files, parse_stats)
# split the platform column into two columns
junc.var <- junc.var %>%
  mutate(
    Platform = case_when(
      str_detect(platform, "pb.*IP_filtered") ~ "PacBio IP filtered",
      str_detect(platform, "dRNA.*IP_filtered") ~ "ONT dRNA IP filtered",
      str_detect(platform, "ont_bulk.*IP_filtered") ~ "ONT cDNA IP filtered",
      str_detect(platform, "dRNA") ~ "ONT dRNA",
      str_detect(platform, "ont_bulk") ~ "ONT cDNA",
      str_detect(platform, "pb") ~ "PacBio",
      TRUE ~ "Illumina",
    ),
    # remove *_
    Cell_line = gsub(".*_", "", platform)
  ) %>%
  filter(!str_detect(Platform, "IP filtered")) %>%
  select(-platform)

# update all NA to 0
junc.var[is.na(junc.var)] <- 0

junc.var <- junc.var %>% mutate(
  # scales all integer columns by the sum of all integer columns per row
  across(where(is.integer), ~ . / rowSums(across(where(is.integer))) * 100)
)
```


```{r bulk_alignqc_perfect_junction_proportion_plot,fig.height=3, fig.width=4}
# Perfect junction
df_long <- junc.var %>%
  select(c(
    Platform,
    Cell_line,
    `0`
  ))

df_summary <- df_long %>%
  group_by(Platform) %>%
  summarise(
    Mean_Value = mean(`0`, na.rm = TRUE),
    SD = sd(`0`, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

save_plot_data(df_long, "bulk_alignqc_perfect_junction_raw_points")

ggplot() +
  # Bars for the mean
  geom_bar(
    data = df_summary, aes(x = Platform, y = Mean_Value, fill = Platform),
    stat = "identity", position = "dodge", alpha = 0.7
  ) +
  # Error bars (Mean ± SD)
  geom_errorbar(
    data = df_summary, aes(x = Platform, ymin = Mean_Value - SD, ymax = Mean_Value + SD),
    width = 0.2, position = position_dodge(0.9)
  ) +
  # Dots for individual values
  geom_point(
    data = df_long, aes(x = Platform, y = `0`, color = Platform),
    position = position_jitter(width = 0.2), size = 2
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels
  labs(x = "Platform", y = "Percentage") +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname()) +
  scale_color_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname()) + # Match dot color to fill    
  coord_cartesian(ylim = c(95, 100))  + 
  ggtitle("Percentage of annotated junction")
```


```{r bulk_alignqc_shifted_junction_proportion_plot}
# Other
df_long <- junc.var %>%
  select(-c(`0`)) %>%
  pivot_longer(cols = -c(Platform, Cell_line), names_to = "Metric", values_to = "Value") %>%
  mutate(Metric = as.numeric(Metric))

df_summary <- df_long %>%
  group_by(Platform, Metric) %>%
  summarise(
    Mean_Value = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

save_plot_data(df_long, "bulk_alignqc_shifted_junction_raw_points")

ggplot() +
  # Bars for the mean
  geom_bar(
    data = df_summary %>% filter(Metric != 0), aes(x = Metric, y = Mean_Value, fill = Platform),
    stat = "identity", position = "dodge", alpha = 0.7
  ) +
  # Error bars (Mean ± SD)
  geom_errorbar(
    data = df_summary, aes(x = Metric, ymin = Mean_Value - SD, ymax = Mean_Value + SD),
    width = 0.2, position = position_dodge(0.9)
  ) +
  # # Dots for individual values
  # geom_point(data = df_long, aes(x = Metric, y = Value, color = Platform),
  #            position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~Platform, scales = "free_y") + # Facet by Metric
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + # Rotate x-axis labels
  labs(x = "Shift", y = "Percentage") +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname()) +
  scale_color_manual(values = color_palette[c("Illumina", "ONT", "ONT_1", "PacBio")] %>% unname()) + # Match dot color to fill
  coord_cartesian(ylim = c(0, 0.2)) 

```


### Genomic Feature
```{r bulk_alignqc_genomic_feature_load}
# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_bulk/result/qc/AlignQC/', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "read_genomic_features.txt.gz", full.names = TRUE)

# SR
dirs <- list.dirs("/vast/projects/LongBench/analysis/sr_bulk/result/qc/AlignQC/", recursive = TRUE, full.names = TRUE)
stats_files <- c(stats_files, list.files(dirs, pattern = "read_genomic_features.txt.gz", full.names = TRUE))


# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts) - 2]
  
  # Read file and convert to tibble
  stats <- read_tsv(file, col_names = c("Read_id", "Feature", "exon_count", "Aligned_length")) %>%
    # change Aligned_length to bins 500, 1000, 1500, 2000, 2500, 3000, 3500, 4000, 4000+
    mutate(
      Aligned_length = cut(Aligned_length, breaks=c(seq(0, 4000, by=500), Inf), labels = c("0-500", "500-1000", "1000-1500", "1500-2000", "2000-2500", "2500-3000", "3000-3500", "3500-4000", "4000+")),
      Platform = case_when(
        str_detect(platform, "pb.*IP_filtered") ~ "PacBio IP filtered",
        str_detect(platform, "dRNA.*IP_filtered") ~ "ONT dRNA IP filtered",
        str_detect(platform, "ont_bulk.*IP_filtered") ~ "ONT cDNA IP filtered",
        str_detect(platform, "dRNA") ~ "ONT dRNA",
        str_detect(platform, "ont_bulk") ~ "ONT cDNA",
        str_detect(platform, "pb") ~ "PacBio",
        TRUE ~ "Illumina",
        ),
      Cell_line = gsub(".*_", "", platform)
    ) %>% 
    group_by(Aligned_length, Feature, Cell_line, Platform ) %>%
    summarise(
      count = n(),
      .groups = "drop"
    ) 
    
    
  

  # # Add platform and type columns
  # stats <- stats %>%
  #   pivot_wider(names_from = metric, values_from = value)
  # Get the proportion of each unique feature
  # stats <- stats %>%
  #   group_by(Feature) %>%
  #   summarise(
  #     count = n(),
  #     .groups = "drop"
  #   ) %>%
  #   mutate(
  #     proportion = count / sum(count) * 100,
  #     platform = platform
  #   ) %>%
  #   arrange(desc(proportion))
  return(stats)
}

# Parse all files and combine into a single tibble
genoemic.feature <- map_dfr(stats_files, parse_stats)
genoemic.feature$Feature <- factor(genoemic.feature$Feature, levels=c("intergenic", "intron","exon"))


```


```{r bulk_alignqc_genomic_feature_overall_proportion_plot,fig.height=3, fig.width=4}
# Compute average proportions per platform
summary_data <- genoemic.feature %>%
  group_by(Platform, Feature) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  group_by(Platform) %>%
  mutate(
      proportion = count / sum(count),
      IP_filtering = ifelse(str_detect(Platform, "IP filtered"), "IP filtered", "Not filtered"),
      # remove IP filtered from Platform for plotting
      Platform = gsub(" IP filtered", "", Platform)
  )

save_plot_data(summary_data, "bulk_alignqc_genomic_feature_overall_points")

# Create stacked bar plot
ggplot(summary_data, aes(x = Platform, y = proportion, fill = Feature)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  facet_wrap(~IP_filtering, scales = "free_y") +
  labs(x = "Platform", y = "Proportion", fill = "Feature") +
  scale_fill_brewer(palette = "Set2") + # Adjust color scheme
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



```{r bulk_alignqc_genomic_feature_by_aligned_length_plot}
# Compute average proportions per platform
summary_data <- genoemic.feature %>% 
  group_by(Platform) %>%
  mutate(platform_total = sum(count)) %>% 
  mutate(proportion=count/platform_total) 

save_plot_data(summary_data, "bulk_alignqc_genomic_feature_by_length_points")
        

ggplot(summary_data, aes(x = Aligned_length, y = proportion, fill = Feature)) +
  # Bars for the mean
    geom_bar(stat = "identity") +
  # Error bars (Mean ± SD)
  # geom_errorbar(data = df_summary, aes(x = Platform, ymin = Mean_Value - SD, ymax = Mean_Value + SD), 
  #               width = 0.2, position = position_dodge(0.9)) +
  # # Dots for individual values
  # geom_point(data = df_long, aes(x = Platform, y = Value, color = Platform), 
  #            position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~Platform, scales = "free_y") +  # Facet by Metric
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(x = "Aligned read length", y = "Proportion") +
  scale_fill_brewer(palette = "Set2") # Adjust color scheme
```


### Exon count and transcript completeness
```{r bulk_alignqc_exon_count_transcript_completeness_disabled, include=F, eval=F}
# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_bulk/result/qc/AlignQC/', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "annotbest.txt.gz", full.names = TRUE)

# SR
dirs <- list.dirs("/vast/projects/LongBench/analysis/sr_bulk/result/qc/AlignQC/", recursive = TRUE, full.names = TRUE)
stats_files <- c(stats_files, list.files(dirs, pattern = "annotbest.txt.gz", full.names = TRUE))


# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts) - 2]

  # Read file and convert to tibble
  stats <- read_tsv(file, col_names = c(
    "id",
    "Read_id",
    "tx_id",
    "g_id",
    "completeness",
    "exon_count_read1",
    "exon_count_read2",
    "exon_count_read3",
    "exon_count_anno",
    "align_length_read1",
    "align_length_read2",
    "align_length_anno",
    "tx_id_1",
    "tx_id_2",
    "unknown_number"
  )) %>%
    select(completeness, exon_count_anno) %>%
    mutate(
      Platform = case_when(
        str_detect(platform, "pb.*IP_filtered") ~ "PacBio IP filtered",
        str_detect(platform, "dRNA.*IP_filtered") ~ "ONT dRNA IP filtered",
        str_detect(platform, "ont_bulk.*IP_filtered") ~ "ONT cDNA IP filtered",
        str_detect(platform, "dRNA") ~ "ONT dRNA",
        str_detect(platform, "ont_bulk") ~ "ONT cDNA",
        str_detect(platform, "pb") ~ "PacBio",
        TRUE ~ "Illumina",
      ) %>% filter(!str_detect(., "IP filtered")),
      # remove *_
      Cell_line = gsub(".*_", "", platform)
    )

  # count the number of reads with different completeness and different exon count
  stats <- stats %>%
    group_by(Platform, Cell_line, completeness, exon_count_anno) %>%
    summarise(
      count = n(),
      .groups = "drop"
    )

  return(stats)
}

# Parse all files and combine into a single tibble
annotation.stat <- map_dfr(stats_files, parse_stats)

overall.completeness <- annotation.stat %>%
  group_by(Platform, Cell_line, completeness) %>%
  summarise(
    count = sum(count),
    .groups = "drop"
  )  %>% group_by(Platform, Cell_line) %>%
  mutate(
    proportion = count / sum(count) * 100
  ) %>% group_by(Platform,completeness) %>%
  summarise(avg_proportion = mean(proportion), .groups = "drop")

# Create stacked bar plot
ggplot(overall.completeness, aes(x = Platform, y = avg_proportion, fill = completeness)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Platform", y = "Average Proportion (%)", fill = "Feature") +
  scale_fill_brewer(palette = "Set2") # Adjust color scheme


```

## INTERNAL PRIMING
```{r bulk_internal_priming_proportion_plot, fig.height=3, fig.width=4}
dirs <- list.dirs("/vast/projects/LongBench/analysis/lr_bulk/result/int_prim_analysis", recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "*summary.txt", full.names = TRUE)


# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts)]
  
  # Read file and convert to tibble
  stats <- read.table(file, sep = ":", col.names = c("metric", "value"), fill = TRUE)
  
  # Add platform and type columns
  stats <- stats %>%
    mutate(platform = platform) %>%
    pivot_wider(names_from = metric, values_from = value)
  
  return(stats)
}

# Parse all files and combine into a single tibble
int.prim.stats <- map_dfr(stats_files, parse_stats)

# split the platform column into two columns
int.prim.stats <- int.prim.stats %>%
  mutate(
    Platform = case_when(
      str_detect(platform, "dRNA") ~ "ONT dRNA",
      str_detect(platform, "ont_bulk") ~ "ONT cDNA",
      str_detect(platform, "pb") ~ "PacBio",
      TRUE ~ "Illumina"
    ),
    # remove *_
    Cell_line = gsub(".*_", "", platform),
    `Internally primed reads percentage (%)` = `Proportion of internal priming reads` * 100
  ) %>% dplyr::select(-platform)


df_summary <- int.prim.stats %>%
  group_by(Platform) %>%
  summarise(
    Mean_Value = mean(`Internally primed reads percentage (%)`, na.rm = TRUE),
    SD = sd(`Internally primed reads percentage (%)`, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

save_plot_data(int.prim.stats, "bulk_internal_priming_raw_points")

ggplot() +
  # Bars for the mean
  geom_bar(data = df_summary, aes(x = Platform, y = Mean_Value, fill = Platform), 
           stat = "identity", position = "dodge", alpha = 0.7) +
  # Error bars (Mean ± SD)
  geom_errorbar(data = df_summary, aes(x = Platform, ymin = Mean_Value - SD, ymax = Mean_Value + SD), 
                width = 0.2, position = position_dodge(0.9)) +
  # Dots for individual values
  geom_point(data = int.prim.stats, aes(x = Platform, y = `Internally primed reads percentage (%)`, color = Platform), 
             position = position_jitter(width = 0.2), size = 2) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +  # Rotate x-axis labels
  labs(x = "Protocols", y = "Internally primed reads percentage (%)") +
  scale_fill_manual(values = color_palette[c("ONT", "ONT_1", "PacBio")] %>% unname) +
  scale_color_manual(values = color_palette[c("ONT", "ONT_1", "PacBio")] %>% unname) +  # Match dot color to fill + 
  theme_minimal()
```

## PICARD QC
```{r bulk_picard_base_distribution_load}
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_bulk/result/qc/coverage', recursive = TRUE, full.names = TRUE)
files <- list.files(dirs, pattern = ".picard.RNA_Metrics", full.names = TRUE)

extract_metrics <- function(file) {
  # Read the file and skip to the data table
  lines <- readLines(file)
  start <- grep("^PF_BASES", lines)  # header line index
  # Read the two lines of header + values into a temp text connection
  tab_text <- paste(lines[start:(start+1)], collapse = "\n")
  
  df <- read.delim(text = tab_text, stringsAsFactors = FALSE)
  
  # Select desired columns and convert to numeric
  cols <- c("PF_BASES", "PCT_CODING_BASES", "PCT_UTR_BASES",
            "PCT_INTRONIC_BASES", "PCT_INTERGENIC_BASES", "PCT_MRNA_BASES")
  df_sub <- df %>% select(all_of(cols))
  
  # Extract platform and cell line from filename
  fname <- basename(file)
  platform <- str_extract(fname, ".*(?=_bulk)")
  cell_line <- str_extract(fname, "(?<=bulk_)[^\\.]+")
  
  df_sub$platform <- platform
  df_sub$cell_line <- cell_line
  
  return(df_sub)
}



# Apply to all files
result_df <- do.call(rbind, lapply(files, extract_metrics))
```


```{r bulk_picard_base_distribution_plots, fig.width=4, fig.height=4}
# Gather metrics into long format for easier plotting
plot_df <- result_df %>%
  pivot_longer(cols = c(PCT_INTRONIC_BASES, PCT_INTERGENIC_BASES, PCT_MRNA_BASES),
               names_to = "metric", values_to = "value")

# Summarize mean and sd per platform & metric
summary_df <- plot_df %>%
  group_by(platform, metric) %>%
  summarise(mean_val = mean(value, na.rm=TRUE),
            sd_val = sd(value, na.rm=TRUE),
            .groups = 'drop')

save_plot_data(plot_df, "bulk_picard_base_distribution_raw_points")

# Barplot with error bars
ggplot(summary_df, aes(x=metric, y=mean_val, fill=platform)) +
  geom_bar(stat="identity", position=position_dodge(width=0.8)) +
  geom_errorbar(aes(ymin=mean_val-sd_val, ymax=mean_val+sd_val),
                position=position_dodge(width=0.8), width=0.2) +
  labs(y="Percentage", x="Metric", title="Mean % by Platform with SD") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1))

ggplot(summary_df, aes(x = platform, y = mean_val, fill = metric)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(y = "Proportion", x = "Metric", title = "Platform Contribution per Metric (Stacked Proportion)") +
  scale_y_continuous(labels = scales::percent_format()) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



```

## Longest aligned reads
```{r bulk_longest_aligned_reads_plot, fig.height=3,fig.width=3}
dirs <- list.dirs('/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/qc/TopAlignedRead', recursive = TRUE, full.names = TRUE)
files <- list.files(dirs, pattern = ".tsv", full.names = TRUE)

# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts)] %>% str_extract(".*(?=_bulk)")
  cell_line <- parts[length(parts)] %>% str_extract("(?<=bulk_)[^\\.]+") %>% str_remove("\\.tsv$")
  # Read file and convert to tibble
  stats <- read_tsv(file)
  
  # Add platform and type columns
  stats <- stats %>%
    mutate(platform = recode(platform,
                           "dRNA" = "ONT dRNA",
                           "pb"   = "PacBio",
                           "ont"  = "ONT cDNA"),
           cell_line = cell_line)
  return(stats)
}

color_palette <- c(
  PacBio = "#df1995",
  "ONT dRNA" = "#00789b",
  "ONT cDNA" = "#04476c",
  Illumina = "#e88b20"
)
# Apply to all files
result_df <- do.call(rbind, lapply(files, parse_stats))
save_plot_data(result_df, "bulk_top_aligned_reads_raw_points")
result_df %>%
  ggplot(aes(x = platform, y = aligned_length, colour = platform)) +
  geom_boxplot() +
  #scale_y_log10() +
  theme_minimal() +
  scale_color_manual(values = color_palette) + 
  theme(legend.position = "none") +
  labs(y = "Top 100 aligned length", x = "Sequencing protocols")

result_df %>%
  ggplot(aes(x = cell_line, y = aligned_length, colour = platform)) +
  geom_boxplot() +
  #scale_y_log10() +
  theme_minimal() +
  scale_color_manual(values = color_palette) + 
  theme(legend.position = "none") +
  labs(y = "Top 100 aligned length", x = "Cell lines")



result_df %>%
  ggplot(aes(x = cell_line, y = mean_quality, colour = platform)) +
  geom_boxplot() +
  #scale_y_log10() +
  theme_minimal() +
  scale_color_manual(values = color_palette) + 
  theme(legend.position = "none") +
  labs(y = "Mean quality of Top 100 aligned length", x = "Cell lines")

```

## NanoPlot: Quality vs Length Correlation
```{r bulk_nanoplot_qual_length_correlation}
nanoplot_dirs <- list.dirs(
  '/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/qc/NanoPlot',
  recursive = FALSE, full.names = TRUE
)
nanoplot_files <- file.path(nanoplot_dirs, "NanoPlot-data.tsv.gz")
nanoplot_files <- nanoplot_files[file.exists(nanoplot_files)]

nanoplot_cor <- lapply(nanoplot_files, function(f) {
  sample_name <- basename(dirname(f))
  df <- read_tsv(f, col_names = c("quals", "lengths"), skip = 1, show_col_types = FALSE)
  tibble(
    sample    = sample_name,
    platform  = str_extract(sample_name, "^[^_]+_[^_]+"),
    cell_line = str_extract(sample_name, "(?<=bulk_).+"),
    pearson   = cor(df$quals, df$lengths, method = "pearson"),
    spearman  = cor(df$quals, df$lengths, method = "spearman")
  )
}) |> bind_rows()

knitr::kable(nanoplot_cor, digits = 4,
             col.names = c("Sample", "Platform", "Cell line", "Pearson r", "Spearman rho"))
```

```{r bulk_nanoplot_qual_length_correlation_plot, fig.height=3, fig.width=5}
nanoplot_cor_long <- nanoplot_cor %>%
  pivot_longer(cols = c(pearson, spearman), names_to = "method", values_to = "correlation") %>%
  mutate(
    platform = recode(platform,
      "dRNA_bulk" = "ONT dRNA",
      "ont_bulk"  = "ONT cDNA",
      "pb_bulk"   = "PacBio"
    ),
    method = recode(method, pearson = "Pearson", spearman = "Spearman")
  )

color_palette_nano <- c(
  "ONT dRNA" = "#00789b",
  "ONT cDNA" = "#04476c",
  PacBio     = "#df1995"
)

ggplot(nanoplot_cor_long, aes(x = platform, y = correlation, colour = platform)) +
  geom_boxplot(outlier.shape = NA, width = 0.4) +
  geom_jitter(aes(shape = cell_line), width = 0.15, size = 2) +
  facet_wrap(~method) +
  scale_colour_manual(values = color_palette_nano) +
  theme_minimal() +
  ylim(c(-1, 1)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "right") +
  labs(x = "Platform", y = "Correlation", colour = "Platform", shape = "Cell line",
       title = "Quality vs Read Length Correlation (NanoPlot)")
```


# SC/SN
```{r define_sc_sn_color_palette}
color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b",
  ONT_1 = "#04476c",
  ONT_2 = "#24cdcd",
  Illumina = "#e88b20"
)
```

## Vireo QC
```{r sc_sn_vireo_doublet_unassigned_plot, fig.height=3, fig.width=6}
# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_sc_sn/result/vireo', recursive = TRUE, full.names = TRUE)
files <- list.files(dirs, pattern = "donor_ids.tsv", full.names = TRUE)

library(ggplot2)
library(dplyr)
library(stringr)
library(readr)
library(tibble)

# Read and summarize files
summary_df <- lapply(files, function(f) {
  df <- read_tsv(f, show_col_types = FALSE)
  total <- nrow(df)
  prop_table <- tibble(
    protocol = str_extract(f, "(ont|pb)_(sc|sn)"),
    doublet = mean(df$donor_id == "doublet"),
    unassigned = mean(df$donor_id == "unassigned")
  )
  pivot_longer(prop_table, cols = c(doublet, unassigned), names_to = "donor_id", values_to = "proportion")
}) %>% bind_rows()

# Add platform and type
summary_df <- summary_df %>%
  mutate(
    Platform = case_when(str_detect(protocol, "ont") ~ "ONT", TRUE ~ "PacBio"),
    Type = ifelse(str_detect(protocol, "_sc"), "SC", "SN")
  )

save_plot_data(summary_df, "sc_sn_vireo_doublet_unassigned_raw_points")

# Plot
ggplot(summary_df, aes(x = protocol, y = proportion*100, fill = Platform, alpha = Type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  facet_wrap(~donor_id) +
  scale_fill_manual(values = color_palette) +
  scale_alpha_manual(values = c(SC = 1, SN = 0.3)) +
  theme_minimal() +
  labs(x = "Protocol", y = "Percentage (%)", fill = "Platform", alpha = "Cell Type")
```

```{r sc_sn_cell_barcode_lists_load}
# get sc
BC_list.sc <- list.files('/vast/projects/LongBench/analysis/lr_sc_sn/result/misc/cell_line_bc_list/sc', pattern = "*BC_list.txt", full.names = TRUE, recursive = T) %>% 
  lapply(function(f) {
  df <- read_tsv(f,col_names=F) %>%
    rename(cell = X1) %>%
    mutate(cell_line = basename(f) %>% str_remove("_BC_list.txt"),
           Type = "SC") 
}) %>% bind_rows

BC_list.sn <- list.files('/vast/projects/LongBench/analysis/lr_sc_sn/result/misc/cell_line_bc_list/sn', pattern = "*BC_list.txt", full.names = TRUE, recursive = T) %>% 
  lapply(function(f) {
  df <- read_tsv(f,col_names=F) %>%
    rename(cell = X1) %>%
    mutate(cell_line = basename(f) %>% str_remove("_BC_list.txt"),
           Type = "SN") 
}) %>% bind_rows

BC_list <- bind_rows(BC_list.sc, BC_list.sn)
remove(BC_list.sc, BC_list.sn)
```

```{r sc_sn_vireo_ambient_load_and_assign}
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_sc_sn/result/vireo', recursive = TRUE, full.names = TRUE)
files <- list.files(dirs, pattern = "prop_ambient.tsv", full.names = TRUE)
# Read and summarize
ambient_df <- lapply(files, function(f) {
  df <- read_tsv(f, show_col_types = FALSE) %>%
    select(-logLik_ratio) %>%
    mutate(protocol = str_extract(f, "(ont|pb)_(sc|sn)"))
}) %>% bind_rows()

# Add metadata for plotting
ambient_df <- ambient_df %>%
  mutate(
    Platform = case_when(str_detect(protocol, "ont") ~ "ONT", TRUE ~ "PacBio"),
    Type = ifelse(str_detect(protocol, "_sc"), "SC", "SN")
  )

ambient_assigned <- ambient_df %>%
  inner_join(BC_list, by = c("cell", "Type")) %>%                                # Step 1: assign cell line
  rowwise() %>%
  mutate(prob_assigned = get(cell_line)) %>%                          # Step 2: extract value by column name
  ungroup()

```


```{r sc_sn_vireo_ambient_rna_proportion_plot, fig.height=3, fig.width=4}
# Color palette
color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b"
)

# Plot
save_plot_data(ambient_assigned, "sc_sn_vireo_ambient_rna_raw_points")

ggplot(ambient_assigned, aes(x = protocol, y = 1-prob_assigned, fill = Platform, alpha = Type)) +
  geom_boxplot(outlier.size = 0.5, width = 0.6) +
  scale_fill_manual(values = color_palette) +
  scale_alpha_manual(values = c(SC = 1, SN = 0.3)) +
  labs(x = "Protocol", y = "Ambient RNA read proportion", fill = "Platform", alpha = "Cell Type") +
  scale_y_log10() +
  theme_minimal()


ambient_assigned %>%
  mutate(ambient_prop = 1 - prob_assigned) %>%
  select(protocol, ambient_prop) %>%
  group_by(protocol) %>%
  summarise(
    median = median(ambient_prop),
    mean = mean(ambient_prop),
    q1 = quantile(ambient_prop, 0.25),
    q3 = quantile(ambient_prop, 0.75),
    min = min(ambient_prop),
    max = max(ambient_prop)
  )
```



## AlignQC
### Alignment stats (Skipped as no useful information for pseudobulk approach)

### Genomic Feature
```{r sc_sn_alignqc_genomic_feature_load}
# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkQC/AlignQC', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "read_genomic_features.txt.gz", full.names = TRUE)

# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts) - 2]
  
  # Read file and convert to tibble
  stats <- read_tsv(file, col_names = c("Read_id", "Feature", "exon_count", "Aligned_length")) %>%
    # change Aligned_length to bins 500, 1000, 1500, 2000, 2500, 3000, 3500, 4000, 4000+
    mutate(
      Aligned_length = cut(Aligned_length, breaks=c(seq(0, 4000, by=500), Inf), labels = c("0-500", "500-1000", "1000-1500", "1500-2000", "2000-2500", "2500-3000", "3000-3500", "3500-4000", "4000+")),
      Platform = case_when(
        str_detect(platform, "ont_sc") ~ "ONT SC",
        str_detect(platform, "ont_sn") ~ "ONT SN",
        str_detect(platform, "pb_sc") ~ "PacBio SC",
        str_detect(platform, "pb_sn") ~ "PacBio SN",

        ),
      Cell_line = gsub(".*_", "", platform)
    ) %>% 
    group_by(Aligned_length, Feature, Cell_line, Platform ) %>%
    summarise(
      count = n(),
      .groups = "drop"
    ) 
    
    
  

  # # Add platform and type columns
  # stats <- stats %>%
  #   pivot_wider(names_from = metric, values_from = value)
  # Get the proportion of each unique feature
  # stats <- stats %>%
  #   group_by(Feature) %>%
  #   summarise(
  #     count = n(),
  #     .groups = "drop"
  #   ) %>%
  #   mutate(
  #     proportion = count / sum(count) * 100,
  #     platform = platform
  #   ) %>%
  #   arrange(desc(proportion))
  return(stats)
}

# Parse all files and combine into a single tibble
genoemic.feature <- map_dfr(stats_files, parse_stats)
genoemic.feature$Feature <- factor(genoemic.feature$Feature, levels=c("intergenic", "intron","exon"))


```

```{r sc_sn_alignqc_genomic_feature_overall_proportion_plot,fig.height=3, fig.width=4}
# Compute average proportions per platform
summary_data <- genoemic.feature %>%
  group_by(Platform, Feature) %>%
  summarise(count = sum(count), .groups = "drop") %>%
  group_by(Platform) %>%
  mutate(
      proportion = count / sum(count),
  )

save_plot_data(summary_data, "sc_sn_alignqc_genomic_feature_overall_points")

# Create stacked bar plot
ggplot(summary_data, aes(x = Platform, y = proportion, fill = Feature)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(x = "Platform", y = "Proportion", fill = "Feature") +
  scale_fill_brewer(palette = "Set2") # Adjust color scheme
```




### Error state
```{r sc_sn_alignqc_error_state_disabled, include=F, eval=F}

# LR
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkQC/AlignQC', recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "error_stats.txt", full.names = TRUE)

# Parse all files and combine into a single tibble
error.stats <- map_dfr(stats_files, parse_stats)
# split the platform column into two columns
error.stats <- error.stats %>%
  separate(platform, into=c('Platform', 'Library Type','Cell_line'), sep = '_') %>%
  mutate(
    Platform = recode(
      Platform, ont = 'ONT', pb = 'PacBio'
    ),
    `Library Type` = recode(`Library Type`, sc='SC', sn='SN')
    )


error.stats <- error.stats %>%
  mutate(
    ANY_ERROR = ANY_ERROR / ALIGNMENT_BASES * 100,
    MISMATCHES = MISMATCHES / ALIGNMENT_BASES * 100,
    ANY_DELETION = ANY_DELETION / ALIGNMENT_BASES * 100,
    ANY_INSERTION = ANY_INSERTION / ALIGNMENT_BASES * 100
  )

df_long <- error.stats %>%
  select(c(
    Platform,
    `Library Type`,
    Cell_line,
    ANY_ERROR,
    MISMATCHES,
    ANY_DELETION,
    ANY_INSERTION
  )) %>%
  pivot_longer(cols = -c(Platform, Cell_line, `Library Type`), names_to = "Metric", values_to = "Value")

df_summary <- df_long %>%
  group_by(Platform, Metric, `Library Type`) %>%
  summarise(
    Mean_Value = mean(Value, na.rm = TRUE),
    SD = sd(Value, na.rm = TRUE),  # Standard deviation
    .groups = "drop"
  )

library(ggplot2)

# Add styling columns
df_summary <- df_summary %>%
  mutate(
    Platform_base = if_else(grepl("PacBio", Platform), "PacBio", "ONT"),
    Sample_Type = if_else(grepl("SC", `Library Type`), "SC", "SN"),
    fill_color = if_else(Platform_base == "PacBio", "#df1995", "#00789b"),
    border_color = fill_color,
    alpha_val = if_else(Sample_Type == "SC", 0.7, 0.3),
    Group = paste(Platform, `Library Type`, sep = " ")
  )

df_long <- df_long %>%
  mutate(
    Platform_base = if_else(grepl("PacBio", Platform), "PacBio", "ONT"),
    Sample_Type = if_else(grepl("SC", `Library Type`), "SC", "SN"),
    fill_color = if_else(Platform_base == "PacBio", "#df1995", "#00789b"),
    border_color = fill_color,
    alpha_val = if_else(Sample_Type == "SC", 1, 0.3),
    Group = paste(Platform, `Library Type`, sep = " ")
  )

# Plot
ggplot() +
  geom_bar(data = df_summary,
           aes(x = Group, y = Mean_Value, fill = fill_color, alpha = alpha_val),
           stat = "identity") +
  geom_errorbar(data = df_summary,
                aes(x = Group, ymin = Mean_Value - SD, ymax = Mean_Value + SD),
                width = 0.2) +
  geom_point(data = df_long,
             aes(x = Group, y = Value, color = border_color, alpha = alpha_val),
             position = position_jitter(width = 0.2), size = 2) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_identity() +
  scale_color_identity() +
  scale_alpha_identity() +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(x = "Platform + Library Type", y = "Percentage")
```

## INTERNAL PRIMING
```{r sc_sn_internal_priming_proportion_plot, fig.height=3, fig.width=3}
dirs <- list.dirs("/vast/projects/LongBench/analysis/lr_sc_sn/result/int_prim_analysis", recursive = TRUE, full.names = TRUE)
# Find all alignment stats files
stats_files <- list.files(dirs, pattern = "*summary.txt", full.names = TRUE)


# Function to parse each file
parse_stats <- function(file) {
  # Extract platform and type from file path
  parts <- strsplit(file, "/")[[1]]
  platform <- parts[length(parts)]
  
  # Read file and convert to tibble
  stats <- read.table(file, sep = ":", col.names = c("metric", "value"), fill = TRUE)
  
  # Add platform and type columns
  stats <- stats %>%
    mutate(platform = platform) %>%
    pivot_wider(names_from = metric, values_from = value)
  
  return(stats)
}

# Parse all files and combine into a single tibble
int.prim.stats <- map_dfr(stats_files, parse_stats)

# split the platform column into two columns
int.prim.stats <- int.prim.stats %>%
  mutate(
    Platform = case_when(
      str_detect(platform, "ont_sc") ~ "ONT SC",
      str_detect(platform, "ont_sn") ~ "ONT SN",
      str_detect(platform, "pb_sc") ~ "PacBio SC",
      str_detect(platform, "pb_sn") ~ "PacBio SN"
    ),
    `Internally primed reads percentage (%)` = `Proportion of internal priming reads` * 100
  ) %>% dplyr::select(-platform)

# Prepare data
df_summary <- int.prim.stats %>%
  mutate(
    Platform_base = if_else(grepl("PacBio", Platform), "PacBio", "ONT"),
    Sample_Type = if_else(grepl("SC", Platform), "SC", "SN"),
    fill_color = if_else(Platform_base == "PacBio", "#df1995", "#00789b"),
    border_color = fill_color,  # No need to repeat logic
    alpha_val = if_else(Sample_Type == "SC", 1, 0.3)
  )

save_plot_data(df_summary, "sc_sn_internal_priming_raw_points")

# Plot
ggplot(df_summary, aes(x = Platform, y = `Proportion of internal priming reads`*100)) +
  geom_bar(
    stat = "identity",
    aes(fill = fill_color, alpha = alpha_val),
    width = 0.7
  ) +
  scale_fill_identity() +
  scale_color_identity() +
  scale_alpha_identity() +
  labs(x = "Protocols", y = "Internally primed reads (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## PICARD QC
```{r sc_sn_picard_base_distribution_load}
dirs <- list.dirs('/vast/projects/LongBench/analysis/lr_sc_sn/result/qc/coverage', recursive = TRUE, full.names = TRUE)
files <- list.files(dirs, pattern = ".picard.RNA_Metrics", full.names = TRUE)

extract_metrics <- function(file) {
  # Read the file and skip to the data table
  lines <- readLines(file)
  start <- grep("^PF_BASES", lines)  # header line index
  # Read the two lines of header + values into a temp text connection
  tab_text <- paste(lines[start:(start+1)], collapse = "\n")
  
  df <- read.delim(text = tab_text, stringsAsFactors = FALSE)
  
  # Select desired columns and convert to numeric
  cols <- c("PCT_CODING_BASES", "PCT_UTR_BASES",
            "PCT_INTRONIC_BASES", "PCT_INTERGENIC_BASES", "PCT_MRNA_BASES")
  df_sub <- df %>% select(all_of(cols))
  
  # Extract platform and cell line from filename
  fname <- basename(file)
  platform <- str_extract(fname, ".*(?=.picard.)")
  
  df_sub$platform <- platform
  
  return(df_sub)
}



# Apply to all files
result_df <- do.call(rbind, lapply(files, extract_metrics))
```



```{r sc_sn_picard_base_distribution_plot, fig.width=4, fig.height=4}
# Gather metrics into long format for easier plotting
plot_df <- result_df %>%
  pivot_longer(cols = c(PCT_INTRONIC_BASES, PCT_INTERGENIC_BASES, PCT_MRNA_BASES),
               names_to = "metric", values_to = "value")

save_plot_data(plot_df, "sc_sn_picard_base_distribution_raw_points")

# Barplot with error bars
ggplot(plot_df, aes(x=platform, y=value, fill=metric)) +
  geom_bar(stat="identity", position = "stack") +
  labs(y="Percentage", x="Metric", title="Genominc feature % by Platform") +
  theme_bw() +
  theme(axis.text.x = element_text(angle=45, hjust=1))
```

# Read Count

```{r read_count_data_load}
read_count_df <- read.table(
  "/vast/projects/LongBench/analysis/figures/qc/read_count_summary.tsv",
  header = TRUE, sep = "\t"
) |>
  mutate(
    datatype = factor(datatype, levels = c("Illumina", "PacBio", "ONT dRNA", "ONT cDNA")),
    sample   = factor(sample, levels = sample)
  )

fill_scale <- scale_fill_manual(values = c(
  Illumina  = "#e88b20",
  PacBio    = "#df1995",
  `ONT dRNA` = "#00789b",
  `ONT cDNA` = "#04476c"
))
```

## Read counts and Gigabases
```{r read_count_barplots, fig.width=14, fig.height=5}
bar_theme <- theme_minimal() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))

p1 <- ggplot(read_count_df, aes(x = sample, y = read_count / 1e6, fill = datatype)) +
  geom_bar(stat = "identity") + bar_theme + fill_scale +
  labs(x = NULL, y = "Read counts (Millions)", fill = "Platform")

p2 <- ggplot(read_count_df, aes(x = sample, y = gb, fill = datatype)) +
  geom_bar(stat = "identity") + bar_theme + fill_scale +
  labs(x = "Sample", y = "Gigabases (Gb)", fill = "Platform")

gridExtra::grid.arrange(p1, p2, ncol = 1)
```

## Dual-axis (reads + Gb)
```{r read_count_dual_axis, fig.width=14, fig.height=5}
max_reads_M  <- max(read_count_df$read_count / 1e6, na.rm = TRUE)
max_gb       <- max(read_count_df$gb, na.rm = TRUE)
to_primary   <- function(x) x / max_gb * max_reads_M
from_primary <- function(x) x / max_reads_M * max_gb

ggplot(read_count_df, aes(x = sample)) +
  geom_bar(aes(y = read_count / 1e6, fill = datatype), stat = "identity") +
  geom_point(aes(y = to_primary(gb), color = datatype), size = 2.5) +
  geom_line(aes(y = to_primary(gb), group = 1), color = "grey40", linewidth = 0.4) +
  scale_y_continuous(
    name = "Read counts (Millions)",
    sec.axis = sec_axis(transform = from_primary, name = "Gigabases (Gb)")
  ) +
  fill_scale +
  scale_color_manual(values = c(
    Illumina = "#e88b20", PacBio = "#df1995",
    `ONT dRNA` = "#00789b", `ONT cDNA` = "#04476c"
  ), guide = "none") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 60, hjust = 1),
    axis.title.y.right = element_text(color = "grey40"),
    axis.text.y.right  = element_text(color = "grey40")
  ) +
  labs(x = "Sample", fill = "Platform")
```

# Read Length and Quality

```{r read_length_quality_data_load}
rlq <- read.table(
  "/vast/projects/LongBench/analysis/figures/qc/read_length_quality_summary.tsv",
  header = TRUE, sep = "\t"
) |>
  mutate(
    datatype = factor(datatype, levels = c("PacBio", "ONT dRNA", "ONT cDNA", "Illumina")),
    sample   = factor(sample, levels = sample)
  )

datatype_colors <- c(
  PacBio    = "#df1995",
  `ONT dRNA` = "#00789b",
  `ONT cDNA` = "#04476c",
  Illumina  = "#e88b20"
)
```

## Quality score distribution
```{r read_quality_boxplot, fig.width=20, fig.height=6}
ggplot(rlq, aes(x = sample, fill = datatype,
                ymin = qual_whisker_low, lower = qual_Q1, middle = qual_median,
                upper = qual_Q3, ymax = qual_whisker_high)) +
  geom_boxplot(stat = "identity") +
  theme_minimal() +
  labs(x = "Sample", y = "Quality Score", fill = "Data type") +
  scale_fill_manual(values = datatype_colors) +
  theme(text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Read length distribution
```{r read_length_boxplot, fig.width=20, fig.height=6}
ggplot(rlq, aes(x = sample, fill = datatype,
                ymin = length_whisker_low, lower = length_Q1, middle = length_median,
                upper = length_Q3, ymax = length_whisker_high)) +
  geom_boxplot(stat = "identity") +
  theme_minimal() +
  labs(x = "Sample", y = "Read Length (bp)", fill = "Data type") +
  scale_fill_manual(values = datatype_colors) +
  coord_cartesian(ylim = c(0, 5000)) +
  theme(text = element_text(size = 14),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Summary table
```{r read_length_quality_summary_table}
knitr::kable(rlq, digits = 2)
```

# Summary Table

```{r summary_table_build}
sc_sn_platforms <- c("ONT SC", "ONT SN", "PacBio SC", "PacBio SN",
                     "Illumina SC", "Illumina SN")

# Helper: parse a raw two-column stats file (metric / value) into a wide numeric tibble
parse_raw_stats <- function(file) {
  read.table(file, sep = "\t", col.names = c("metric", "value"), fill = TRUE) %>%
    pivot_wider(names_from = metric, values_from = value) %>%
    mutate(across(everything(), as.numeric))
}

# Helper: map SC/SN directory name to Platform label
sc_sn_dir_to_platform <- c(
  ont_sc = "ONT SC", ont_sn = "ONT SN",
  pb_sc  = "PacBio SC", pb_sn  = "PacBio SN"
)

# --- 1. Aligned reads % (bulk in memory + SC/SN from full-dataset AlignQC) ---
aligned_pct <- alignment.stats %>%
  select(Platform, Cell_line, pct_aligned = ALIGNED_READS)

sc_sn_align_files <- list.files(
  "/vast/projects/LongBench/analysis/lr_sc_sn/result/qc/AlignQC",
  pattern = "alignment_stats.txt", recursive = TRUE, full.names = TRUE
)
sc_sn_aligned_pct <- map_dfr(sc_sn_align_files, function(f) {
  dir_name <- strsplit(f, "/")[[1]] |> rev() %>% .[[3]]
  raw <- parse_raw_stats(f)
  tibble(Platform  = sc_sn_dir_to_platform[dir_name],
         Cell_line = NA_character_,
         pct_aligned = raw$ALIGNED_READS / raw$TOTAL_READS * 100)
})
aligned_pct <- bind_rows(aligned_pct, sc_sn_aligned_pct)

# --- 2. Error rate % (bulk in memory + SC/SN from full-dataset AlignQC) ---
error_pct <- error.stats %>%
  select(Platform, Cell_line, pct_any_error = ANY_ERROR)

sc_sn_error_files <- list.files(
  "/vast/projects/LongBench/analysis/lr_sc_sn/result/qc/AlignQC",
  pattern = "error_stats.txt", recursive = TRUE, full.names = TRUE
)
sc_sn_error_pct <- map_dfr(sc_sn_error_files, function(f) {
  dir_name <- strsplit(f, "/")[[1]] |> rev() %>% .[[3]]
  raw <- parse_raw_stats(f)
  tibble(Platform  = sc_sn_dir_to_platform[dir_name],
         Cell_line = NA_character_,
         pct_any_error = raw$ANY_ERROR / raw$ALIGNMENT_BASES * 100)
})
error_pct <- bind_rows(error_pct, sc_sn_error_pct)

# --- 3. Genomic features: LR bulk + SR bulk (re-read; variable overwritten by SC/SN chunk) ---
bulk_gf_files <- c(
  list.files(
    list.dirs("/vast/projects/LongBench/analysis/lr_bulk/result/qc/AlignQC/",
              recursive = TRUE, full.names = TRUE),
    pattern = "read_genomic_features.txt.gz", full.names = TRUE
  ),
  list.files(
    list.dirs("/vast/projects/LongBench/analysis/sr_bulk/result/qc/AlignQC/",
              recursive = TRUE, full.names = TRUE),
    pattern = "read_genomic_features.txt.gz", full.names = TRUE
  )
)

parse_bulk_gf <- function(file) {
  platform_str <- strsplit(file, "/")[[1]] |> rev() %>% .[[3]]
  read_tsv(file, col_names = c("Read_id", "Feature", "exon_count", "Aligned_length"),
           show_col_types = FALSE) %>%
    mutate(
      Platform = case_when(
        str_detect(platform_str, "pb.*IP_filtered")       ~ "PacBio IP filtered",
        str_detect(platform_str, "dRNA.*IP_filtered")     ~ "ONT dRNA IP filtered",
        str_detect(platform_str, "ont_bulk.*IP_filtered") ~ "ONT cDNA IP filtered",
        str_detect(platform_str, "dRNA")     ~ "ONT dRNA",
        str_detect(platform_str, "ont_bulk") ~ "ONT cDNA",
        str_detect(platform_str, "pb")       ~ "PacBio",
        TRUE ~ "Illumina"
      ),
      Cell_line = gsub(".*_", "", platform_str)
    ) %>%
    group_by(Platform, Cell_line, Feature) %>%
    summarise(count = n(), .groups = "drop")
}

bulk_gf_summary <- map_dfr(bulk_gf_files, parse_bulk_gf) %>%
  filter(!str_detect(Platform, "IP filtered")) %>%
  group_by(Platform, Cell_line) %>%
  mutate(pct = count / sum(count) * 100) %>%
  select(Platform, Cell_line, Feature, pct) %>%
  pivot_wider(names_from = Feature, values_from = pct, names_glue = "pct_{Feature}")

# --- 4. SC/SN genomic features from full-dataset AlignQC (one row per platform type) ---
sc_sn_gf_files <- list.files(
  "/vast/projects/LongBench/analysis/lr_sc_sn/result/qc/AlignQC",
  pattern = "read_genomic_features.txt.gz", recursive = TRUE, full.names = TRUE
)
sc_sn_gf_summary <- map_dfr(sc_sn_gf_files, function(f) {
  dir_name <- strsplit(f, "/")[[1]] |> rev() %>% .[[3]]
  read_tsv(f, col_names = c("Read_id", "Feature", "exon_count", "Aligned_length"),
           show_col_types = FALSE) %>%
    mutate(Platform = sc_sn_dir_to_platform[dir_name], Cell_line = NA_character_) %>%
    group_by(Platform, Cell_line, Feature) %>%
    summarise(count = n(), .groups = "drop")
}) %>%
  group_by(Platform, Cell_line) %>%
  mutate(pct = count / sum(count) * 100) %>%
  select(Platform, Cell_line, Feature, pct) %>%
  pivot_wider(names_from = Feature, values_from = pct, names_glue = "pct_{Feature}")

# --- 5. Combine ---
bulk_combined <- aligned_pct %>%
  filter(!Platform %in% sc_sn_platforms) %>%
  left_join(error_pct %>% filter(!Platform %in% sc_sn_platforms),
            by = c("Platform", "Cell_line")) %>%
  left_join(bulk_gf_summary, by = c("Platform", "Cell_line"))

# Skeleton ensures all SC/SN platforms appear even if AlignQC files are absent
all_sc_sn <- tibble(Platform = sc_sn_platforms, Cell_line = NA_character_)
sc_sn_combined <- all_sc_sn %>%
  left_join(sc_sn_gf_summary, by = c("Platform", "Cell_line")) %>%
  left_join(aligned_pct %>% filter(Platform %in% sc_sn_platforms),
            by = c("Platform", "Cell_line")) %>%
  left_join(error_pct %>% filter(Platform %in% sc_sn_platforms),
            by = c("Platform", "Cell_line"))

# --- 6. Join read counts/Gb and mean quality/length ---
sc_sn_sample_key <- c(
  "ONT SC"       = "ont_sc",
  "ONT SN"       = "ont_sn",
  "PacBio SC"    = "pb_sc",
  "PacBio SN"    = "pb_sn",
  "Illumina SC"  = "ill_sc",
  "Illumina SN"  = "ill_sn"
)
# Prefix for read_count_df lookup (Illumina bulk stored as "Illumina_H146")
rc_prefix_bulk <- c(
  "ONT cDNA" = "ont_bulk",
  "PacBio"   = "pb_bulk",
  "ONT dRNA" = "dRNA_bulk",
  "Illumina" = "Illumina"
)
# Prefix for rlq lookup (Illumina bulk stored as "ill_bulk_H146")
rlq_prefix_bulk <- c(
  "ONT cDNA" = "ont_bulk",
  "PacBio"   = "pb_bulk",
  "ONT dRNA" = "dRNA_bulk",
  "Illumina" = "ill_bulk"
)

read_counts_lookup <- read_count_df %>%
  select(sample, read_count, gb) %>%
  mutate(sample = as.character(sample))

rlq_lookup <- rlq %>%
  select(sample, qual_mean, length_median) %>%
  mutate(sample = as.character(sample))

qc_summary <- bind_rows(bulk_combined, sc_sn_combined) %>%
  mutate(
    rc_sample  = case_when(
      Platform %in% sc_sn_platforms ~ sc_sn_sample_key[Platform],
      TRUE ~ paste0(rc_prefix_bulk[Platform], "_", Cell_line)
    ),
    rlq_sample = case_when(
      Platform %in% sc_sn_platforms ~ sc_sn_sample_key[Platform],
      TRUE ~ paste0(rlq_prefix_bulk[Platform], "_", Cell_line)
    )
  ) %>%
  left_join(read_counts_lookup, by = c("rc_sample"  = "sample")) %>%
  left_join(rlq_lookup,         by = c("rlq_sample" = "sample")) %>%
  mutate(gb = gb / 1e9) %>%
  select(Platform, Cell_line, read_count, gb, pct_aligned, pct_any_error,
         pct_intergenic, pct_intron, pct_exon, qual_mean, length_median) %>%
  arrange(Platform, Cell_line)

# --- Write TSV ---
if (!is.null(params$fig.path)) {
  write_tsv(qc_summary, file.path(params$fig.path, "qc_summary_table.tsv"))
}

knitr::kable(qc_summary, digits = 2,
             col.names = c("Platform", "Cell line",
                           "# Passed reads", "Total bases (Gb)",
                           "Aligned reads (%)",
                           "Read-to-reference discordance (%)",
                           "Intergenic reads (%)", "Intronic reads (%)",
                           "Exonic reads (%)",
                           "Mean quality", "Median read length (bp)"))
```