---
title: "Bulk Gene & Transcript Identification"
author: "Yupei & Ash"
date: '`r Sys.Date()`'
version: "v1.0"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
params:
  random_seed: 2024
  input_DGEobject_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"
  bulk_meta: '/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt'
  cache_dir: NULL
  fig.path: 'figures/02_Bulk_Identification/'
  out.rds: '/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification.rds'
---

# Setup
```{r Input filename setup, include=FALSE, eval=F}
if (!exists("params")) {
  params <- list(
    random_seed = 2024,
    input_DGEobject_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
    bulk_meta = '/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt',
    cache_dir = NULL,
    fig.path = 'figures/02_Bulk_Identification/',
    out.rds = '/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification.rds'
  )
}
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 15,
  fig.height = 10,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir),
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"),
                      params$cache_dir))
```

```{r Libraries, include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ComplexUpset)
library(patchwork)
source("/home/users/allstaff/you.yu/LongBench/analysis/workflow/scripts/Rfunctions.R")

rename <- dplyr::rename
select <- dplyr::select

color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b",
  ONT_1 = "#04476c",
  ONT_2 = "#00789b",
  Illumina = "#e88b20"
)

output_list <- list()
```

```{r helper functions}
is.human <- function(dge) { grepl("^ENS", rownames(dge)) }

pad_dge <- function(dge, all_features) {
  missing <- setdiff(all_features, rownames(dge))
  if (length(missing) > 0) {
    pad <- matrix(0, nrow = length(missing), ncol = ncol(dge$counts),
                  dimnames = list(missing, colnames(dge$counts)))
    dge$counts <- rbind(dge$counts, pad)
  }
  dge[all_features, ]
}
```

# Load data
```{r Get the count data, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
list2env(readRDS(params$input_DGEobject_rds), .GlobalEnv)
```

# Gene identification
## Data preparation
```{r Gene filtering}
ill_bulk.gene.dge  <- ill_bulk.gene.dge[is.human(ill_bulk.gene.dge), ]
ont_bulk.gene.dge  <- ont_bulk.gene.dge[is.human(ont_bulk.gene.dge), ]
pb_bulk.gene.dge   <- pb_bulk.gene.dge[is.human(pb_bulk.gene.dge), ]
drna_bulk.gene.dge <- drna_bulk.gene.dge[is.human(drna_bulk.gene.dge), ]

all_genes <- Reduce(intersect, lapply(
  list(ill_bulk.gene.dge, ont_bulk.gene.dge, pb_bulk.gene.dge, drna_bulk.gene.dge),
  rownames
))

ill_bulk.gene.dge  <- pad_dge(ill_bulk.gene.dge, all_genes)
ont_bulk.gene.dge  <- pad_dge(ont_bulk.gene.dge, all_genes)
pb_bulk.gene.dge   <- pad_dge(pb_bulk.gene.dge, all_genes)
drna_bulk.gene.dge <- pad_dge(drna_bulk.gene.dge, all_genes)

rownames(ill_bulk.gene.dge)  <- rownames(ill_bulk.gene.dge)  %>% sub("\\..*", "", .)
rownames(ont_bulk.gene.dge)  <- rownames(ont_bulk.gene.dge)  %>% sub("\\..*", "", .)
rownames(pb_bulk.gene.dge)   <- rownames(pb_bulk.gene.dge)   %>% sub("\\..*", "", .)
rownames(drna_bulk.gene.dge) <- rownames(drna_bulk.gene.dge) %>% sub("\\..*", "", .)
```

```{r Gene discovery tibble}
gene_discovery <- tibble(
  gene_id    = all_genes,
  "Illumina" = filterByExpr(ill_bulk.gene.dge,  group = ill_bulk.gene.dge$samples$group),
  "ONT cDNA" = filterByExpr(ont_bulk.gene.dge,  group = ont_bulk.gene.dge$samples$group),
  "PacBio"   = filterByExpr(pb_bulk.gene.dge,   group = pb_bulk.gene.dge$samples$group),
  "ONT dRNA" = filterByExpr(drna_bulk.gene.dge, group = drna_bulk.gene.dge$samples$group)
) %>% filter(Illumina | `ONT cDNA` | PacBio | `ONT dRNA`)

gene_discovery$gene_id <- gene_discovery$gene_id %>% sub("\\..*", "", .)
```

```{r Add biotype and gene length}
tx_ref_df <- ill_bulk.tx.dge$genes %>%
  select(full_id, Length) %>%
  tibble() %>%
  mutate(
    biotype = strsplit(full_id, "\\|") %>% sapply(function(x) x[8]),
    gene_id = strsplit(full_id, "\\|") %>% sapply(function(x) x[2]),
    tx_id   = strsplit(full_id, "\\|") %>% sapply(function(x) x[1])
  )

gene_length_lookup <- tx_ref_df %>%
  group_by(gene_id) %>%
  summarise(mean_length = mean(Length, na.rm = TRUE), .groups = "drop") %>%
  mutate(gene_id = sub("\\..*", "", gene_id))

library(biomaRt)
biotypes_cache <- file.path(dirname(params$out.rds), "biotypes_cache.rds")
if (file.exists(biotypes_cache)) {
  biotypes <- readRDS(biotypes_cache)
} else {
  ensembl <- useEnsembl(biomart = "genes", dataset = "hsapiens_gene_ensembl")
  biotypes <- getBM(
    attributes = c("ensembl_gene_id", "gene_biotype"),
    filters    = "ensembl_gene_id",
    values     = gene_discovery$gene_id,
    mart       = ensembl
  )
  saveRDS(biotypes, biotypes_cache)
}

gene_discovery <- gene_discovery %>%
  left_join(gene_length_lookup, by = "gene_id") %>%
  rename(`Gene length` = mean_length) %>%
  left_join(biotypes, by = c("gene_id" = "ensembl_gene_id")) %>%
  rename(biotype = gene_biotype) %>%
  mutate(biotype = ifelse(biotype %in%
    names(sort(table(biotype), decreasing = TRUE)[1:6]),
    biotype, "Others"))

gene_discovery$biotype <- factor(
  gene_discovery$biotype,
  levels = c(sort(unique(gene_discovery$biotype)) %>% setdiff("Others"), "Others")
)
```

```{r Save gene_discovery}
output_list[["gene_discovery"]] <- gene_discovery
if (!is.null(params$fig.path)) {
  dir.create(params$fig.path, recursive = TRUE, showWarnings = FALSE)
  write_tsv(gene_discovery, file.path(params$fig.path, "gene_discovery.tsv"))
}
```

## GENE: Main Upset plot
```{r GENE: Main Upset plot, fig.height=4, fig.width=8}
ComplexUpset::upset(
  gene_discovery %>% mutate(
    biotype = case_when(
      biotype == "protein_coding" ~ "Protein-coding",
      TRUE ~ "Non-protein-coding"
    )
  ),
  intersect = c('Illumina', "ONT cDNA", "PacBio", "ONT dRNA"),
  annotations = list(
    "Gene length" = list(
      aes  = aes(x = intersection, y = `Gene length`),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 3, vjust = -0.2))
    ) + scale_fill_manual(values = c(
      "Non-protein-coding" = "#BAC5CB",
      "Protein-coding"     = '#B5C952'
    ))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom     = geom_bar(aes(fill = biotype), width = 0.8),
      position = 'right'
    ) + scale_fill_manual(values = c(
      "Non-protein-coding" = "#BAC5CB",
      "Protein-coding"     = '#B5C952'
    )) + theme(legend.position = "none")
  ),
  guides = 'over',
  wrap   = TRUE
)
```

## GENE: Mean overlap
We defined "Mean Overlap" to quantify the consistency of detected features across the four sequencing platforms. For any given platform, we calculated the proportion of its detected features also identified by at least two of the three remaining platforms, then averaged these proportions across all four platforms.

```{r GENE: Mean overlap}
calc_mean_overlap <- function(df, platforms = c('Illumina', 'ONT cDNA', 'PacBio', 'ONT dRNA')) {
  per_platform <- sapply(platforms, function(p) {
    others   <- setdiff(platforms, p)
    n_others <- rowSums(df[others])
    sum(df[[p]] & n_others >= 2) / sum(df[[p]])
  })
  mean(per_platform)
}

tibble(
  Biotype = c("Protein-coding", "Non-protein-coding", "All"),
  `Mean Overlap` = c(
    calc_mean_overlap(gene_discovery %>% filter(biotype == "protein_coding")),
    calc_mean_overlap(gene_discovery %>% filter(biotype != "protein_coding")),
    calc_mean_overlap(gene_discovery)
  )
) %>%
  mutate(`Mean Overlap` = round(`Mean Overlap`, 3)) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```

# Transcript identification
## Data preparation
```{r Transcript filtering}
ill_bulk.tx.dge  <- ill_bulk.tx.dge[is.human(ill_bulk.tx.dge), ]
ont_bulk.tx.dge  <- ont_bulk.tx.dge[is.human(ont_bulk.tx.dge), ]
pb_bulk.tx.dge   <- pb_bulk.tx.dge[is.human(pb_bulk.tx.dge), ]
drna_bulk.tx.dge <- drna_bulk.tx.dge[is.human(drna_bulk.tx.dge), ]

all_tx <- Reduce(intersect, lapply(
  list(ill_bulk.tx.dge, ont_bulk.tx.dge, pb_bulk.tx.dge, drna_bulk.tx.dge),
  rownames
))

ill_bulk.tx.dge  <- pad_dge(ill_bulk.tx.dge, all_tx)
ont_bulk.tx.dge  <- pad_dge(ont_bulk.tx.dge, all_tx)
pb_bulk.tx.dge   <- pad_dge(pb_bulk.tx.dge, all_tx)
drna_bulk.tx.dge <- pad_dge(drna_bulk.tx.dge, all_tx)

tx_discovery <- tibble(
  tx_id      = ill_bulk.tx.dge$genes$full_id,
  "Illumina" = filterByExpr(ill_bulk.tx.dge,  group = ill_bulk.tx.dge$samples$group,  min.count = 5),
  "ONT cDNA" = filterByExpr(ont_bulk.tx.dge,  group = ont_bulk.tx.dge$samples$group,  min.count = 5),
  "PacBio"   = filterByExpr(pb_bulk.tx.dge,   group = pb_bulk.tx.dge$samples$group,   min.count = 5),
  "ONT dRNA" = filterByExpr(drna_bulk.tx.dge, group = drna_bulk.tx.dge$samples$group, min.count = 5)
) %>% filter(Illumina | `ONT cDNA` | PacBio | `ONT dRNA`)

tx_discovery <- tx_discovery %>% mutate(
  biotype = strsplit(tx_id, "\\|") %>% sapply(function(x) x[8]),
  tx_length = strsplit(tx_id, "\\|") %>% sapply(function(x) x[7]) %>% as.integer(),
  gene_id = strsplit(tx_id, "\\|") %>% sapply(function(x) x[2]),
  tx_id   = strsplit(tx_id, "\\|") %>% sapply(function(x) x[1])
)

tx_discovery <- tx_discovery %>% mutate(
  biotype = ifelse(biotype %in%
    names(sort(table(biotype), decreasing = TRUE)[1:6]),
    biotype, "Others")
)
```

```{r Save tx_discovery}
output_list[["tx_discovery"]] <- tx_discovery
saveRDS(output_list, params$out.rds)
if (!is.null(params$fig.path)) {
  write_tsv(tx_discovery, file.path(params$fig.path, "tx_discovery.tsv"))
}
```

## Transcript: Main Upset plot
```{r Transcript: Main Upset plot, fig.height=4, fig.width=8}
ComplexUpset::upset(
  tx_discovery %>% mutate(
    biotype = case_when(
      biotype == "protein_coding" ~ "Protein-coding",
      TRUE ~ "Non-protein-coding"
    )
  ),
  intersect = c('Illumina', "ONT cDNA", "PacBio", "ONT dRNA"),
  annotations = list(
    "Transcript length" = list(
      aes  = aes(x = intersection, y = tx_length),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 3, vjust = -0.2))
    ) + scale_fill_manual(values = c(
      "Non-protein-coding" = "#BAC5CB",
      "Protein-coding"     = '#E3DEEE'
    ))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom     = geom_bar(aes(fill = biotype), width = 0.8),
      position = 'right'
    ) + scale_fill_manual(values = c(
      "Non-protein-coding" = "#BAC5CB",
      "Protein-coding"     = '#E3DEEE'
    )) + theme(legend.position = "none")
  ),
  guides = 'over',
  wrap   = TRUE
)
```

## Transcript: Mean overlap
Same metric as for genes, applied at the transcript level.

```{r Transcript: Mean overlap}
tibble(
  Biotype = c("Protein-coding", "Non-protein-coding", "All"),
  `Mean Overlap` = c(
    calc_mean_overlap(tx_discovery %>% filter(biotype == "protein_coding")),
    calc_mean_overlap(tx_discovery %>% filter(biotype != "protein_coding")),
    calc_mean_overlap(tx_discovery)
  )
) %>%
  mutate(`Mean Overlap` = round(`Mean Overlap`, 3)) %>%
  kable() %>%
  kable_styling(full_width = FALSE)
```


# Session Info
```{r Session Info}
sessionInfo()
```