---
title: "SC SN QC and preporcessing"
author: "Yupei"
date: '`r Sys.Date()`'
version: 'v1.5'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
    fig.path: NULL
    cache_dir: NULL
    filtered_so_rds_out: '/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/sc_sn_filtered_so.rds'
---
v1.0 The cellline annotation is after outlier removal and then filter using vireo result
v1.5 Add FACS median size correlation analysis


```{r setup, include=FALSE}
# this is a repeat of params for direct execution in non-rmarkdown environment
# check if params is defined
if (!exists("params")) {
  params <- list(
    fig.path = NULL,
    cache_dir = NULL,
    filtered_so_rds_out = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/sc_sn_filtered_so.rds"
  )
}

library(tidyverse)
source('/vast/projects/LongBench/analysis/workflow/sub_workflows/lr_sc_sn/rmarkdown/sc_long_preprocessing.R')

# Override BioMart-dependent loader with a timeout-safe version
get_gene_names_from_biomart <- function(gene_ids, dataset = "hsapiens_gene_ensembl") {
  mirrors <- c("useast", "uswest", "asia", "www")
  mart <- NULL

  for (m in mirrors) {
    mart <- tryCatch(
      biomaRt::useEnsembl(biomart = "genes", dataset = dataset, mirror = m),
      error = function(e) NULL
    )
    if (!is.null(mart)) break
  }

  if (is.null(mart)) {
    mart <- tryCatch(
      biomaRt::useEnsembl(biomart = "genes", dataset = dataset, host = "https://www.ensembl.org"),
      error = function(e) NULL
    )
  }

  if (is.null(mart)) {
    return(rep(NA_character_, length(gene_ids)))
  }

  bm <- tryCatch(
    biomaRt::getBM(
      attributes = c("ensembl_gene_id", "external_gene_name"),
      filters = "ensembl_gene_id",
      values = unique(gene_ids),
      mart = mart
    ),
    error = function(e) NULL
  )

  if (is.null(bm) || nrow(bm) == 0) {
    return(rep(NA_character_, length(gene_ids)))
  }

  bm <- bm[!is.na(bm$external_gene_name) & bm$external_gene_name != "", , drop = FALSE]
  bm <- bm[!duplicated(bm$ensembl_gene_id), , drop = FALSE]

  bm$external_gene_name[match(gene_ids, bm$ensembl_gene_id)]
}

load_flames_gene_quant <- function(fn_flames_gene_quant, project = "singlecell", min.cells = 3, min.features = 0, assay = "Gene_quant") {
  counts <- read.csv(fn_flames_gene_quant, row.names = 1)
  counts[is.na(counts)] <- 0

  gene_ids_all <- sub("\\..*$", "", rownames(counts))
  gene_names <- get_gene_names_from_biomart(gene_ids_all)

  if (all(is.na(gene_names) | gene_names == "")) {
    message("BioMart unavailable, using Ensembl gene IDs as feature names.")
    gene_names <- gene_ids_all
  }

  keep_idx <- !is.na(gene_names) & gene_names != ""
  keep_idx <- keep_idx & !duplicated(gene_names)

  counts <- counts[keep_idx, , drop = FALSE]
  gene_names <- gene_names[keep_idx]
  gene_ids <- gene_ids_all[keep_idx]
  rownames(counts) <- gene_names

  so <- Seurat::CreateSeuratObject(
    counts = counts,
    project = project,
    min.cells = min.cells,
    min.features = min.features,
    assay = assay
  )

  so[["percent.mt"]] <- Seurat::PercentageFeatureSet(so, features = grep("^MT", rownames(so), value = TRUE)) %>% round(2)
  so[["percent.ribo"]] <- Seurat::PercentageFeatureSet(so, features = grep("^RP[SL]", rownames(so), value = TRUE)) %>% round(2)

  so@misc$gene_info_map <- data.frame(gene_name = gene_names, gene_id = gene_ids)
  qc_fun <- get0("add_qc_state", mode = "function")
  if (!is.null(qc_fun)) {
    so <- qc_fun(so, "raw_gene_quant")
  }
  so
}

knitr::opts_chunk$set(
  cache = TRUE,
  cache.lazy = FALSE,
  fig.width = 15, 
  fig.height = 10, 
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir), 
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"), 
                      params$cache_dir))

rename <- dplyr::rename
select <- dplyr::select


```

```{r}
# Set up the params:
color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b",
  ONT_1 = "#04476c",
  ONT_2 = "#24cdcd",
  Illumina = "#e88b20"
)
options(future.globals.maxSize = 10 * 1024 * 1024 * 1024)
library(dplyr)
library(tidyverse)
library(Seurat)
library(patchwork)
library(Matrix)
library(ggplot2)
library(gridExtra)
library(stringr)
library(grid)
library(knitr)
library(DT)
library(glue)
library(kableExtra)

params_sc <- list(
  random_seed=2024,
  cache_dir=NULL,
  sc_long_preprocessing_script= '/vast/projects/LongBench/analysis/workflow/sub_workflows/lr_sc_sn/rmarkdown/sc_long_preprocessing.R',
  vireo_out_ont= '/vast/projects/LongBench/analysis/lr_sc_sn/result/vireo/ont_sc_vireo',
  vireo_out_pb='/vast/projects/LongBench/analysis/lr_sc_sn/result/vireo/pb_sc_vireo',
  vireo_out_ill='/vast/projects/LongBench/analysis/illumina_sc_sn/result/vireo/ill_sc_vireo',
  out_annotated_so_rds= NULL,
  fn_flames_gene_quant_ont= "/vast/projects/LongBench/analysis/lr_sc_sn/result/flames_out/ont_sc/gene_count.csv",
  fn_flames_gene_quant_pb="/vast/projects/LongBench/analysis/lr_sc_sn/result/flames_out/pb_sc/gene_count.csv",
  cell_ranger_out="/vast/projects/LongBench/analysis/illumina_sc_sn/result/cellranger/ill_sc/outs/filtered_feature_bc_matrix",
  output_figures="/vast/projects/LongBench/analysis/lr_sc_sn/result/figures/test_clustering_annotation.pdf"
)

params_sn <- list(
  random_seed=2024,
  cache_dir=NULL,
  sc_long_preprocessing_script= '/vast/projects/LongBench/analysis/workflow/sub_workflows/lr_sc_sn/rmarkdown/sc_long_preprocessing.R',
  vireo_out_ont= '/vast/projects/LongBench/analysis/lr_sc_sn/result/vireo/ont_sn_vireo',
  vireo_out_pb='/vast/projects/LongBench/analysis/lr_sc_sn/result/vireo/pb_sn_vireo',
  vireo_out_ill='/vast/projects/LongBench/analysis/illumina_sc_sn/result/vireo/ill_sn_vireo',
  out_annotated_so_rds= NULL,
  fn_flames_gene_quant_ont= "/vast/projects/LongBench/analysis/lr_sc_sn/result/flames_out/ont_sn/gene_count.csv",
  fn_flames_gene_quant_pb="/vast/projects/LongBench/analysis/lr_sc_sn/result/flames_out/pb_sn/gene_count.csv",
  cell_ranger_out="/vast/projects/LongBench/analysis/illumina_sc_sn/result/cellranger/ill_sn/outs/filtered_feature_bc_matrix",
  output_figures="/vast/projects/LongBench/analysis/lr_sc_sn/result/figures/test_clustering_annotation.pdf"
)

```


```{r}
# input file
so_ont <- load_flames_gene_quant(params_sc$fn_flames_gene_quant_ont, assay = "Gene_quant", project = "ont_sc")
so_pb <- load_flames_gene_quant(params_sc$fn_flames_gene_quant_pb, assay = "Gene_quant", project = "pb_sc")
data <- Read10X(data.dir = params_sc$cell_ranger_out)
so_ill <- CreateSeuratObject(counts = data, min.cells = 3, min.features = 0, assay = "Gene_quant", project = "ill_sc")
rm(data)
colnames(so_ill) <- gsub("-1", "", colnames(so_ill))
# Get mito and ribo gene percentage
so_ill[["percent.mt"]] <- PercentageFeatureSet(so_ill, features = grep("^MT", rownames(so_ill), value = T)) %>% round(2)
so_ill[["percent.ribo"]] <- PercentageFeatureSet(so_ill, features = grep("^RP[SL]", rownames(so_ill), value = T)) %>% round(2)

so_ont[["qc.pass"]] <- check.qc.pass(so_ont)
so_pb[["qc.pass"]] <- check.qc.pass(so_pb)
so_ill[["qc.pass"]] <- check.qc.pass(so_ill)

barcode_tibble <- readr::read_csv("/vast/projects/LongBench/analysis/figures/sc_sn_annotation/sc_cell_line_annotation_post_qc.csv")
so_ont$cell_line <- barcode_tibble$merged_cellline_anno[match(colnames(so_ont), barcode_tibble$barcode)]
so_pb$cell_line <- barcode_tibble$merged_cellline_anno[match(colnames(so_pb), barcode_tibble$barcode)]
so_ill$cell_line <- barcode_tibble$merged_cellline_anno[match(colnames(so_ill), barcode_tibble$barcode)]
high.conf.cells <- barcode_tibble %>%
  filter(PacBio & ONT & Illumina & !merged_cellline_anno %in% c("unassigned", "doublet")) %>%
  select(barcode, merged_cellline_anno) %>% pull(barcode)

so_ont.sc <- so_ont[,high.conf.cells]
so_pb.sc <- so_pb[,high.conf.cells]
so_ill.sc <- so_ill[,high.conf.cells]



# repeat for SN

# input file
so_ont <- load_flames_gene_quant(params_sn$fn_flames_gene_quant_ont, assay = "Gene_quant", project = "ont_sn")
so_pb <- load_flames_gene_quant(params_sn$fn_flames_gene_quant_pb, assay = "Gene_quant", project = "pb_sn")
data <- Read10X(data.dir = params_sn$cell_ranger_out)
so_ill <- CreateSeuratObject(counts = data, min.cells = 3, min.features = 0, assay = "Gene_quant", project = "ill_sn")
rm(data)
colnames(so_ill) <- gsub("-1", "", colnames(so_ill))
# Get mito and ribo gene percentage
so_ill[["percent.mt"]] <- PercentageFeatureSet(so_ill, features = grep("^MT", rownames(so_ill), value = T)) %>% round(2)
so_ill[["percent.ribo"]] <- PercentageFeatureSet(so_ill, features = grep("^RP[SL]", rownames(so_ill), value = T)) %>% round(2)

so_ont[["qc.pass"]] <- check.qc.pass(so_ont)
so_pb[["qc.pass"]] <- check.qc.pass(so_pb)
so_ill[["qc.pass"]] <- check.qc.pass(so_ill)


barcode_tibble <- readr::read_csv("/vast/projects/LongBench/analysis/figures/sc_sn_annotation/sn_cell_line_annotation_post_qc.csv")
so_ont$cell_line <- barcode_tibble$merged_cellline_anno[match(colnames(so_ont), barcode_tibble$barcode)]
so_pb$cell_line <- barcode_tibble$merged_cellline_anno[match(colnames(so_pb), barcode_tibble$barcode)]
so_ill$cell_line <- barcode_tibble$merged_cellline_anno[match(colnames(so_ill), barcode_tibble$barcode)]
high.conf.cells <- barcode_tibble %>%
  filter(PacBio & ONT & Illumina & !merged_cellline_anno %in% c("unassigned", "doublet")) %>%
  select(barcode, merged_cellline_anno) %>% pull(barcode)
so_ont.sn <- so_ont[,high.conf.cells]
so_pb.sn <- so_pb[,high.conf.cells]
so_ill.sn <- so_ill[,high.conf.cells]

rm(so_ont,  so_pb, so_ill)

```


# Plot1: Read counts and feature counts distribution of filtered objects
```{r, fig.width=8, fig.height=6}
# Feature plot
plot_d <- bind_rows(
  lapply(c(so_ont.sc, so_pb.sc, so_ill.sc, so_ont.sn, so_pb.sn, so_ill.sn), function(so) {
    FetchData(so, c(
      "orig.ident",
      "nFeature_Gene_quant",
      "nCount_Gene_quant",
      "percent.mt",
      "percent.ribo"
    ))
  })
) %>% 
  separate(orig.ident, into=c("Sequencer", "Library.Type")) %>%
  mutate(Sequencer = factor(Sequencer, levels = c("ill", 'ont', 'pb')),
         Library.Type = factor(Library.Type)
  )

# violin plot
library(ggplot2)
plot_d$log10_counts <- log10(plot_d$nCount_Gene_quant)
plots <- map(
  c("nFeature_Gene_quant", "log10_counts", "percent.mt", "percent.ribo"),
  function(x){
    ggplot(plot_d, aes(x = Library.Type, y = !!sym(x))) +  
      geom_boxplot(aes(fill = Sequencer), width = 0.5, position = position_dodge(0.9), 
                   outlier.size = 0.5) +  # Set smaller outlier dots
      scale_fill_manual(values = color_palette[c("Illumina", "ONT", "PacBio")] %>% unname) + 
      theme_minimal() +
      labs(title = "Box Plot", x = "Protocols", y = x)
  }
)

wrap_plots(plots, guides = "collect")
plot_d %>%
  group_by(Sequencer, Library.Type) %>%
  summarise(
    median_nFeature = median(nFeature_Gene_quant, na.rm = TRUE),
    Q1_nFeature = quantile(nFeature_Gene_quant, 0.25, na.rm = TRUE),
    Q3_nFeature = quantile(nFeature_Gene_quant, 0.75, na.rm = TRUE),
    median_nCount = median(nCount_Gene_quant, na.rm = TRUE),
    Q1_nCount = quantile(nCount_Gene_quant, 0.25, na.rm = TRUE),
    Q3_nCount = quantile(nCount_Gene_quant, 0.75, na.rm = TRUE),
    median_percent_mt = median(percent.mt, na.rm = TRUE),
    Q1_percent_mt = quantile(percent.mt, 0.25, na.rm = TRUE),
    Q3_percent_mt = quantile(percent.mt, 0.75, na.rm = TRUE),
    median_percent_ribo = median(percent.ribo, na.rm = TRUE),
    Q1_percent_ribo = quantile(percent.ribo, 0.25, na.rm = TRUE),
    Q3_percent_ribo = quantile(percent.ribo, 0.75, na.rm = TRUE),
    .groups = "drop"
  ) 

```


# UMAP
```{r, echo=F}
options(future.globals.maxSize = 5 * 1024 * 1024 * 1024)
so_ont.sc <- so_ont.sc %>% get_clusters(ndim = 30, res = 0.5, seed = params$random_seed)
so_ont.sn <- so_ont.sn %>% get_clusters(ndim = 30, res = 0.5, seed = params$random_seed)
so_pb.sc <- so_pb.sc %>% get_clusters(ndim = 30, res = 0.5, seed = params$random_seed)
so_pb.sn <- so_pb.sn %>% get_clusters(ndim = 30, res = 0.5, seed = params$random_seed)
so_ill.sc <- so_ill.sc %>% get_clusters(ndim = 30, res = 0.5, seed = params$random_seed)
so_ill.sn <- so_ill.sn %>% get_clusters(ndim = 30, res = 0.5, seed = params$random_seed)
```


```{r, echo=F, cache=FALSE}
# Define custom colors
# Define custom colors
cell_line_color_palette <- c(
  "H146" = "#004d1a", "H69" = "#1cad1c", "SHP77" = "#a3d830",  # sclc_a (Greens)
  "H526" = "#a08544", "H211" = "#efca11",                      # sclc_p (Browns)
  "H1975" = "#5d4075", "H2228" = "#5c39c4", "HCC827" = "#dd0de4"  # luad (Purple-Gold)
)
# Plotting with dynamic titles
so_list <- list(
  'Illumina SC' =so_ill.sc, 
  'PacBio SC' =so_pb.sc, 
  'ONT SC' = so_ont.sc, 
  'Illumina SN' =so_ill.sn,
  'PacBio SN' =so_pb.sn, 
  'ONT SN' =so_ont.sn)
# Plotting with dynamic titles


cell_line_plots <- map(names(so_list), ~ {
  DimPlot(
    so_list[[.x]],
    group.by = "cell_line",
    reduction = "umap.Gene_quant",
    combine = FALSE,  # Returns a list
    label.size = 5,
    label = FALSE,
    cols = cell_line_color_palette
  )[[1]] + ggtitle(.x)
})

cell_line_plots %>% patchwork::wrap_plots(plots, guides = "collect")


so_list <- list(
  'Illumina SC' =so_ill.sc,
  'PacBio SC' =so_pb.sc, 
  'ONT SC' = so_ont.sc,
  'Illumina SN' =so_ill.sn,
  'PacBio SN' =so_pb.sn,
  'ONT SN' =so_ont.sn
  )

# save the seurat objects
saveRDS(so_list, file = params$filtered_so_rds_out,compress = FALSE)

# The UMAP 1 and 2 Coordinates 

umap_coords_df <- purrr::imap_dfr(so_list, function(so, protocol_name) {
  umap_coords <- Embeddings(so, reduction = "umap.Gene_quant")
  data.frame(
    cell_barcode = rownames(umap_coords),
    UMAP_1 = umap_coords[, 1],
    UMAP_2 = umap_coords[, 2],
    protocol = protocol_name,
    cell_line = so$cell_line[match(rownames(umap_coords), colnames(so))]
  )
})

umap_coords_df
```

With the so_list, plot the boxplot of library size for each cell line and protocol.
```{r, fig.width=8, fig.height=6}
if (!exists("so_list")) {
  so_list <- readRDS(params$filtered_so_rds_out)
}

library_size_df <- purrr::imap_dfr(so_list, function(so, protocol_name) {
  tmp <- FetchData(
    so,
    vars = c("cell_line", "nCount_Gene_quant")
  )
  tmp$protocol <- protocol_name
  tmp
}) %>%
  tidyr::separate(protocol, into = c("Sequencer", "Library.Type"), sep = " ", remove = FALSE) %>%
  dplyr::filter(!is.na(cell_line), cell_line != "", !cell_line %in% c("unassigned", "doublet")) %>%
  dplyr::mutate(
    protocol = factor(protocol, levels = names(so_list)),
    Sequencer = factor(Sequencer, levels = c("Illumina", "ONT", "PacBio")),
    Library.Type = factor(Library.Type, levels = c("SC", "SN"))
  )

ggplot(library_size_df, aes(x = cell_line, y = nCount_Gene_quant, fill = protocol)) +
  geom_boxplot(outlier.size = 0.2, width = 0.7) +
  facet_wrap(~Library.Type, scales = "free_y") +
  scale_fill_manual(values = c(
    "Illumina SC" = "#e88b20",
    "PacBio SC" = "#df1995",
    "ONT SC" = "#00789b",
    "Illumina SN" = "#f1b86e",
    "PacBio SN" = "#e66db9",
    "ONT SN" = "#4ba4bf"
  )) +
  theme_minimal(base_size = 12) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.position = "bottom"
  ) +
  labs(x = "Cell line", y = "Library size (nCount_Gene_quant)", fill = "Protocol")
```

Correlation of median library size with FACS median size metadata.
```{r, fig.width=10, fig.height=6}
facs_xlsx <- "/vast/projects/LongBench/meta_data/FACS_Median_size.xlsx"

if (!file.exists(facs_xlsx)) {
  stop(glue::glue("FACS metadata file not found: {facs_xlsx}"))
}

facs_df_raw <- readxl::read_excel(facs_xlsx)

# infer columns robustly
cellline_col <- names(facs_df_raw)[stringr::str_detect(tolower(names(facs_df_raw)), "cell.?line|cellline")][1]
sample_col <- names(facs_df_raw)[stringr::str_detect(tolower(names(facs_df_raw)), "sample")][1]
numeric_cols <- names(facs_df_raw)[vapply(facs_df_raw, is.numeric, logical(1))]

# prefer FSC-A, then SSC-A, then any numeric median/size-like column
facs_candidates <- names(facs_df_raw)[stringr::str_detect(tolower(names(facs_df_raw)), "fsc|ssc|facs|median|size")]
facs_candidates <- facs_candidates[facs_candidates %in% numeric_cols]
facs_candidates <- c(
  facs_candidates[stringr::str_detect(tolower(facs_candidates), "fsc")],
  facs_candidates[stringr::str_detect(tolower(facs_candidates), "ssc")],
  facs_candidates
) %>% unique()
facs_col <- facs_candidates[1]

if (is.na(facs_col)) {
  if (length(numeric_cols) == 0) {
    stop("Could not find a numeric FACS median-size column in FACS_Median_size.xlsx")
  }
  facs_col <- numeric_cols[1]
}

known_cell_lines <- sort(unique(library_size_df$cell_line))
known_cell_lines_upper <- toupper(known_cell_lines)

extract_cell_line <- function(x) {
  if (is.na(x) || x == "") return(NA_character_)
  xu <- toupper(as.character(x))

  # Alias used in FACS sample names: HCC -> HCC827
  if ("HCC827" %in% known_cell_lines && stringr::str_detect(xu, "(^|[^A-Z0-9])HCC([^A-Z0-9]|$)")) {
    return("HCC827")
  }

  hit <- known_cell_lines[stringr::str_detect(xu, paste0("(^|[^A-Z0-9])", known_cell_lines_upper, "([^A-Z0-9]|$)"))]
  if (length(hit) == 0) return(NA_character_)
  hit[1]
}

if (!is.na(cellline_col)) {
  facs_df <- facs_df_raw %>%
    dplyr::transmute(
      cell_line = as.character(.data[[cellline_col]]),
      facs_median_size = as.numeric(.data[[facs_col]])
    )
} else if (!is.na(sample_col)) {
  facs_df <- facs_df_raw %>%
    dplyr::transmute(
      sample_name = as.character(.data[[sample_col]]),
      cell_line = purrr::map_chr(sample_name, extract_cell_line),
      facs_median_size = as.numeric(.data[[facs_col]])
    )
} else {
  stop("Could not infer cell-line information: no cell-line column and no sample column found in FACS_Median_size.xlsx")
}

facs_df <- facs_df %>%
  dplyr::filter(!is.na(cell_line), cell_line != "", !is.na(facs_median_size))

median_libsize_df <- library_size_df %>%
  dplyr::group_by(protocol, Sequencer, Library.Type, cell_line) %>%
  dplyr::summarise(median_library_size = median(nCount_Gene_quant, na.rm = TRUE), .groups = "drop") %>%
  dplyr::left_join(facs_df, by = "cell_line") %>%
  dplyr::filter(!is.na(facs_median_size))

cor_stats <- median_libsize_df %>%
  dplyr::group_by(protocol) %>%
  dplyr::summarise(
    n = dplyr::n(),
    pearson_r = cor(median_library_size, facs_median_size, method = "pearson", use = "complete.obs"),
    spearman_rho = cor(median_library_size, facs_median_size, method = "spearman", use = "complete.obs"),
    .groups = "drop"
  )

print(cor_stats)

p <- ggplot(median_libsize_df, aes(x = facs_median_size, y = median_library_size, color = Sequencer)) +
  geom_point(size = 2.5, alpha = 0.9) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.5, color = "grey30") +
  facet_wrap(~protocol, scales = "free_y") +
  scale_color_manual(values = c("Illumina" = "#e88b20", "ONT" = "#00789b", "PacBio" = "#df1995")) +
  theme_minimal(base_size = 12) +
  labs(
    x = glue::glue("FACS median size ({facs_col})"),
    y = "Median library size (nCount_Gene_quant)",
    color = "Sequencer"
  )

if (requireNamespace("ggrepel", quietly = TRUE)) {
  p <- p + ggrepel::geom_text_repel(
    aes(label = cell_line),
    size = 3,
    box.padding = 0.35,
    point.padding = 0.25,
    max.overlaps = Inf,
    show.legend = FALSE
  )
} else {
  message("Package 'ggrepel' not installed; using geom_text with overlap checking.")
  p <- p + geom_text(
    aes(label = cell_line),
    size = 3,
    vjust = -0.6,
    check_overlap = TRUE,
    show.legend = FALSE
  )
}

p
```

# Session Info
```{r}
sessionInfo()
```