---

title: "Bulk gene/transcript identification: Oarfish vs IsoQuant (no_novel)"
author: "Yupei"
date: '`r Sys.Date()`'
version: "v1.1"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  # Oarfish + Illumina DGE bundle
  oarfish_dge_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"
  # IsoQuant (no_novel) DGE bundles
  isoquant_ont_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds"
  isoquant_pb_rds:  "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds"
  isoquant_drna_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds"
  cache_dir: NULL
  fig.path: 'figures/'
  out.rds: '/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification_tool_comparison.rds'
---

# Setup
```{r Input filename setup, include=FALSE, eval=F}
# This chunk is used only when running the file chunk by chunk manually
params <- list(
  random_seed = 2024,
  cache_dir = NULL,
  oarfish_dge_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
  isoquant_ont_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds",
  isoquant_pb_rds  = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds",
  isoquant_drna_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds",
  out.rds = '/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification_tool_comparison.rds'
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 15,
  fig.height = 10,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir),
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"),
                      params$cache_dir))
```

```{r Libraries, include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(ComplexUpset)
library(ComplexHeatmap)
source("/home/users/allstaff/you.yu/LongBench/analysis/workflow/scripts/Rfunctions.R")

rename <- dplyr::rename
select <- dplyr::select

color_palette <- c(
  PacBio    = "#df1995",
  ONT       = "#00789b",
  ONT_1     = "#04476c",
  ONT_2     = "#00789b",
  Illumina  = "#e88b20"
)

# Tool colors for cross-tool comparison plots
tool_colors <- c(
  Oarfish  = "#2166ac",
  IsoQuant = "#d6604d"
)

output_list <- list()
```

```{r helpers}
is.human <- function(dge) { grepl("^ENS", rownames(dge)) }

# Strip pipe-separated annotations and version numbers from rownames
normalize_rownames <- function(dge) {
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}

# Pad missing features with zeros so two DGE objects cover the same feature universe
pad_dge <- function(dge, all_features) {
  missing <- setdiff(all_features, rownames(dge))
  if (length(missing) > 0) {
    pad <- matrix(0, nrow = length(missing), ncol = ncol(dge$counts),
                  dimnames = list(missing, colnames(dge$counts)))
    dge$counts <- rbind(dge$counts, pad)
  }
  dge[all_features, ]
}
```


# Load DGE objects

```{r Load Oarfish DGE}
# ---- Oarfish + Illumina (TX and Gene level) ----
oarfish_obj <- readRDS(params$oarfish_dge_rds)
ill_bulk.tx.dge    <- oarfish_obj[["ill_bulk.tx.dge"]]
ill_bulk.gene.dge  <- oarfish_obj[["ill_bulk.gene.dge"]]

# Oarfish long-read TX-level (rownames may be pipe-separated in legacy RDS)
oar.ont_bulk.tx.dge   <- oarfish_obj[["ont_bulk.tx.dge"]]
oar.pb_bulk.tx.dge    <- oarfish_obj[["pb_bulk.tx.dge"]]
oar.drna_bulk.tx.dge  <- oarfish_obj[["drna_bulk.tx.dge"]]

# Oarfish long-read Gene-level
oar.ont_bulk.gene.dge  <- oarfish_obj[["ont_bulk.gene.dge"]]
oar.pb_bulk.gene.dge   <- oarfish_obj[["pb_bulk.gene.dge"]]
oar.drna_bulk.gene.dge <- oarfish_obj[["drna_bulk.gene.dge"]]

rm(oarfish_obj)
```

```{r Load IsoQuant DGE}
# ---- IsoQuant (no_novel) bundles ----
isoquant_ont_obj  <- readRDS(params$isoquant_ont_rds)
isoquant_pb_obj   <- readRDS(params$isoquant_pb_rds)
isoquant_drna_obj <- readRDS(params$isoquant_drna_rds)

iq.ont_bulk.tx.dge   <- isoquant_ont_obj[["ont_bulk.tx.dge"]]
iq.pb_bulk.tx.dge    <- isoquant_pb_obj[["pb_bulk.tx.dge"]]
iq.drna_bulk.tx.dge  <- isoquant_drna_obj[["dRNA_bulk.tx.dge"]]

iq.ont_bulk.gene.dge  <- isoquant_ont_obj[["ont_bulk.gene.dge"]]
iq.pb_bulk.gene.dge   <- isoquant_pb_obj[["pb_bulk.gene.dge"]]
iq.drna_bulk.gene.dge <- isoquant_drna_obj[["dRNA_bulk.gene.dge"]]

rm(isoquant_ont_obj, isoquant_pb_obj, isoquant_drna_obj)
```

```{r Normalize IDs}
# IsoQuant rownames: strip pipe annotations and version numbers
iq.ont_bulk.tx.dge   <- normalize_rownames(iq.ont_bulk.tx.dge)
iq.pb_bulk.tx.dge    <- normalize_rownames(iq.pb_bulk.tx.dge)
iq.drna_bulk.tx.dge  <- normalize_rownames(iq.drna_bulk.tx.dge)
iq.ont_bulk.gene.dge  <- normalize_rownames(iq.ont_bulk.gene.dge)
iq.pb_bulk.gene.dge   <- normalize_rownames(iq.pb_bulk.gene.dge)
iq.drna_bulk.gene.dge <- normalize_rownames(iq.drna_bulk.gene.dge)

# Illumina gene IDs: strip version numbers
rownames(ill_bulk.gene.dge) <- sub("\\..*", "", rownames(ill_bulk.gene.dge))

# Oarfish gene IDs: strip version numbers
rownames(oar.ont_bulk.gene.dge)  <- sub("\\..*", "", rownames(oar.ont_bulk.gene.dge))
rownames(oar.pb_bulk.gene.dge)   <- sub("\\..*", "", rownames(oar.pb_bulk.gene.dge))
rownames(oar.drna_bulk.gene.dge) <- sub("\\..*", "", rownames(oar.drna_bulk.gene.dge))
```

```{r Build TX annotation reference}
# Build a reference table: tx_id -> gene_id, biotype, length
# Using the Illumina $genes slot which carries the original pipe-separated full_id
# (or rownames if full_id is not present)
get_full_id_col <- function(dge) {
  if ("full_id" %in% colnames(dge$genes)) {
    dge$genes$full_id
  } else {
    # Fall back to rownames of the genes data frame
    rownames(dge$genes)
  }
}

full_ids <- get_full_id_col(ill_bulk.tx.dge)

tx_annotation <- tibble(full_id = full_ids) %>%
  mutate(
    tx_id   = strsplit(full_id, "\\|") %>% sapply(function(x) x[1]),
    gene_id = strsplit(full_id, "\\|") %>% sapply(function(x) x[2]),
    length  = strsplit(full_id, "\\|") %>% sapply(function(x) x[7]) %>% as.integer(),
    biotype = strsplit(full_id, "\\|") %>% sapply(function(x) x[8])
  ) %>%
  mutate(
    tx_id   = sub("\\..*", "", tx_id),
    gene_id = sub("\\..*", "", gene_id)
  ) %>%
  select(tx_id, gene_id, length, biotype) %>%
  distinct(tx_id, .keep_all = TRUE)
```

```{r Data inventory, results='asis'}
inventory <- tibble(
  tool    = c(rep("Oarfish", 3), rep("IsoQuant", 3)),
  dataset = rep(c("ONT cDNA", "PacBio", "ONT dRNA"), 2),
  tx_features = c(
    nrow(oar.ont_bulk.tx.dge), nrow(oar.pb_bulk.tx.dge), nrow(oar.drna_bulk.tx.dge),
    nrow(iq.ont_bulk.tx.dge),  nrow(iq.pb_bulk.tx.dge),  nrow(iq.drna_bulk.tx.dge)
  ),
  gene_features = c(
    nrow(oar.ont_bulk.gene.dge), nrow(oar.pb_bulk.gene.dge), nrow(oar.drna_bulk.gene.dge),
    nrow(iq.ont_bulk.gene.dge),  nrow(iq.pb_bulk.gene.dge),  nrow(iq.drna_bulk.gene.dge)
  )
)
kable(inventory, caption = "Feature counts per tool and dataset (before human filtering)")
```


# Gene discovery analysis

## Build `gene_discovery` per tool

```{r Gene: Build per-tool gene_discovery}
# ---- Filter to human genes only ----
ill_bulk.gene.dge      <- ill_bulk.gene.dge[is.human(ill_bulk.gene.dge), ]
oar.ont_bulk.gene.dge  <- oar.ont_bulk.gene.dge[is.human(oar.ont_bulk.gene.dge), ]
oar.pb_bulk.gene.dge   <- oar.pb_bulk.gene.dge[is.human(oar.pb_bulk.gene.dge), ]
oar.drna_bulk.gene.dge <- oar.drna_bulk.gene.dge[is.human(oar.drna_bulk.gene.dge), ]
iq.ont_bulk.gene.dge   <- iq.ont_bulk.gene.dge[is.human(iq.ont_bulk.gene.dge), ]
iq.pb_bulk.gene.dge    <- iq.pb_bulk.gene.dge[is.human(iq.pb_bulk.gene.dge), ]
iq.drna_bulk.gene.dge  <- iq.drna_bulk.gene.dge[is.human(iq.drna_bulk.gene.dge), ]

# ---- Compute union gene universe and pad all objects ----
all_genes <- Reduce(union, lapply(
  list(ill_bulk.gene.dge,
       oar.ont_bulk.gene.dge, oar.pb_bulk.gene.dge, oar.drna_bulk.gene.dge,
       iq.ont_bulk.gene.dge,  iq.pb_bulk.gene.dge,  iq.drna_bulk.gene.dge),
  rownames
))

ill_bulk.gene.dge      <- pad_dge(ill_bulk.gene.dge,      all_genes)
oar.ont_bulk.gene.dge  <- pad_dge(oar.ont_bulk.gene.dge,  all_genes)
oar.pb_bulk.gene.dge   <- pad_dge(oar.pb_bulk.gene.dge,   all_genes)
oar.drna_bulk.gene.dge <- pad_dge(oar.drna_bulk.gene.dge, all_genes)
iq.ont_bulk.gene.dge   <- pad_dge(iq.ont_bulk.gene.dge,   all_genes)
iq.pb_bulk.gene.dge    <- pad_dge(iq.pb_bulk.gene.dge,    all_genes)
iq.drna_bulk.gene.dge  <- pad_dge(iq.drna_bulk.gene.dge,  all_genes)

# Strip any remaining version numbers
strip_version <- function(dge) { rownames(dge) <- sub("\\..*", "", rownames(dge)); dge }
ill_bulk.gene.dge      <- strip_version(ill_bulk.gene.dge)
oar.ont_bulk.gene.dge  <- strip_version(oar.ont_bulk.gene.dge)
oar.pb_bulk.gene.dge   <- strip_version(oar.pb_bulk.gene.dge)
oar.drna_bulk.gene.dge <- strip_version(oar.drna_bulk.gene.dge)

# ---- Apply filterByExpr for each dataset and tool ----
gene_discovery <- tibble(
  gene_id               = sub("\\..*", "", all_genes),
  Illumina              = filterByExpr(ill_bulk.gene.dge,      group = ill_bulk.gene.dge$samples$group),
  `ONT cDNA (Oarfish)`  = filterByExpr(oar.ont_bulk.gene.dge,  group = oar.ont_bulk.gene.dge$samples$group),
  `PacBio (Oarfish)`    = filterByExpr(oar.pb_bulk.gene.dge,   group = oar.pb_bulk.gene.dge$samples$group),
  `ONT dRNA (Oarfish)`  = filterByExpr(oar.drna_bulk.gene.dge, group = oar.drna_bulk.gene.dge$samples$group),
  `ONT cDNA (IsoQuant)` = filterByExpr(iq.ont_bulk.gene.dge,   group = iq.ont_bulk.gene.dge$samples$group),
  `PacBio (IsoQuant)`   = filterByExpr(iq.pb_bulk.gene.dge,    group = iq.pb_bulk.gene.dge$samples$group),
  `ONT dRNA (IsoQuant)` = filterByExpr(iq.drna_bulk.gene.dge,  group = iq.drna_bulk.gene.dge$samples$group)
) %>% filter(
  Illumina | `ONT cDNA (Oarfish)` | `PacBio (Oarfish)` | `ONT dRNA (Oarfish)` |
  `ONT cDNA (IsoQuant)` | `PacBio (IsoQuant)` | `ONT dRNA (IsoQuant)`
)
```

## Add biotype and gene length

```{r Gene: Add biotype and gene length}
gene_length_lookup <- tx_annotation %>%
  group_by(gene_id) %>%
  summarise(mean_length = mean(length, na.rm = TRUE), .groups = "drop")

# Reuse biotype annotations already fetched by Bulk_identification.Rmd
existing_id_rds <- "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification_IP_filtered.rds"
biotype_lookup <- readRDS(existing_id_rds)[["gene_discovery"]] %>%
  select(gene_id, biotype) %>%
  distinct(gene_id, .keep_all = TRUE)

gene_discovery <- gene_discovery %>%
  left_join(gene_length_lookup, by = "gene_id") %>%
  rename(`Gene length` = mean_length) %>%
  left_join(biotype_lookup, by = "gene_id") %>%
  mutate(biotype = ifelse(
    !is.na(biotype) & biotype %in% names(sort(table(biotype), decreasing = TRUE)[1:6]),
    biotype, "Others"
  ))

gene_discovery$biotype <- factor(
  gene_discovery$biotype,
  levels = c(sort(unique(gene_discovery$biotype)) %>% setdiff("Others"), "Others")
)
```

## Add intronic gene info

```{r Gene: Add intronic gene info}
rds_list        <- readRDS("/home/users/allstaff/you.yu/LongBench/analysis/workflow/rmarkdown/RDS/intronic_gene_and_exon_count.rds")
overlaps_tibble <- rds_list$exon_intron_overlaps %>%
  mutate(exon_gene  = sub("\\..*", "", exon_gene),
         intron_gene = sub("\\..*", "", intron_gene))
exons_count <- rds_list$exons_count %>% mutate(Gene = sub("\\..*", "", Gene))
rm(rds_list)

gene_discovery <- gene_discovery %>%
  mutate(
    `Intronic Gene` = gene_id %in% overlaps_tibble[overlaps_tibble$intron_gene %in% gene_id, ]$exon_gene
  ) %>%
  left_join(exons_count, by = c("gene_id" = "Gene"))

output_list[["gene_discovery"]] <- gene_discovery
```

## Gene detection plots

### Upset plot: Oarfish platforms
```{r GENE: Upset Oarfish, fig.height=5, fig.width=10}
ComplexUpset::upset(
  gene_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = c("Illumina", "ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)"),
  annotations = list(
    "Gene length" = list(
      aes  = aes(x = intersection, y = `Gene length`),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) + ggtitle("Gene detection — Oarfish")
```

### Upset plot: IsoQuant platforms
```{r GENE: Upset IsoQuant, fig.height=5, fig.width=10}
ComplexUpset::upset(
  gene_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = c("Illumina", "ONT cDNA (IsoQuant)", "PacBio (IsoQuant)", "ONT dRNA (IsoQuant)"),
  annotations = list(
    "Gene length" = list(
      aes  = aes(x = intersection, y = `Gene length`),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) + ggtitle("Gene detection — IsoQuant (no_novel)")
```

### Upset plot: All platforms and tools (including Illumina)
```{r GENE: Upset combined, fig.height=6, fig.width=16}
lr_cols <- c("ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)",
             "ONT cDNA (IsoQuant)", "PacBio (IsoQuant)", "ONT dRNA (IsoQuant)")
all_cols_upset <- c("Illumina", lr_cols)

# Genes detected by Illumina + all 3 Oarfish platforms
oar_ill_mask <- gene_discovery$Illumina &
                gene_discovery$`ONT cDNA (Oarfish)` &
                gene_discovery$`PacBio (Oarfish)` &
                gene_discovery$`ONT dRNA (Oarfish)`
n_oar_ill    <- sum(oar_ill_mask)
# Of those, also confirmed by all 3 IsoQuant platforms
n_all7_gene  <- sum(oar_ill_mask &
                    gene_discovery$`ONT cDNA (IsoQuant)` &
                    gene_discovery$`PacBio (IsoQuant)` &
                    gene_discovery$`ONT dRNA (IsoQuant)`)
# Of those, no IsoQuant support at all
n_no_iq_gene <- sum(oar_ill_mask &
                    !gene_discovery$`ONT cDNA (IsoQuant)` &
                    !gene_discovery$`PacBio (IsoQuant)` &
                    !gene_discovery$`ONT dRNA (IsoQuant)`)

ComplexUpset::upset(
  gene_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = all_cols_upset,
  n_intersections = 30,
  annotations = list(
    "Gene length" = list(
      aes  = aes(x = intersection, y = `Gene length`),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) +
  ggtitle("Gene detection — Illumina + Oarfish + IsoQuant (all platforms)") +
  patchwork::plot_annotation(
    caption = glue::glue(
      "Illumina ∩ all Oarfish: {scales::comma(n_oar_ill)} genes  |  ",
      "Also in all IsoQuant: {scales::comma(n_all7_gene)} ({scales::percent(n_all7_gene / n_oar_ill, accuracy = 0.1)})  |  ",
      "No IsoQuant support: {scales::comma(n_no_iq_gene)} ({scales::percent(n_no_iq_gene / n_oar_ill, accuracy = 0.1)})"
    ),
    theme = theme(plot.caption = element_text(size = 11, hjust = 0.5, face = "bold"))
  )
```

### Tool comparison per platform
For each long-read platform: how many genes are detected by Oarfish only, IsoQuant only, or both?
```{r GENE: Per-platform tool comparison, fig.height=3, fig.width=9}
platform_pairs <- list(
  `ONT cDNA`  = c("ONT cDNA (Oarfish)",  "ONT cDNA (IsoQuant)"),
  `PacBio`    = c("PacBio (Oarfish)",    "PacBio (IsoQuant)"),
  `ONT dRNA`  = c("ONT dRNA (Oarfish)",  "ONT dRNA (IsoQuant)")
)

tool_comparison_gene <- map_dfr(names(platform_pairs), function(platform) {
  oar_col <- platform_pairs[[platform]][1]
  iq_col  <- platform_pairs[[platform]][2]
  gene_discovery %>%
    mutate(
      category = case_when(
        .data[[oar_col]] & .data[[iq_col]]   ~ "Both",
        .data[[oar_col]] & !.data[[iq_col]]  ~ "Oarfish only",
        !.data[[oar_col]] & .data[[iq_col]]  ~ "IsoQuant only",
        TRUE                                  ~ "Neither"
      )
    ) %>%
    filter(category != "Neither") %>%
    count(category) %>%
    mutate(platform = platform)
})

tool_comparison_gene$category <- factor(
  tool_comparison_gene$category,
  levels = c("Oarfish only", "Both", "IsoQuant only")
)

ggplot(tool_comparison_gene, aes(x = platform, y = n, fill = category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
    "Oarfish only"  = tool_colors["Oarfish"],
    "Both"          = "#b2b2b2",
    "IsoQuant only" = tool_colors["IsoQuant"]
  )) +
  theme_minimal() +
  labs(title = "Gene detection: tool overlap per platform",
       y = "Number of genes", x = NULL, fill = "Category") +
  theme(axis.text.x = element_text(size = 11))
```

```{r GENE: Per-platform tool comparison by biotype, fig.height=4, fig.width=12}
tool_comparison_gene_biotype <- map_dfr(names(platform_pairs), function(platform) {
  oar_col <- platform_pairs[[platform]][1]
  iq_col  <- platform_pairs[[platform]][2]
  gene_discovery %>%
    mutate(
      category = case_when(
        .data[[oar_col]] & .data[[iq_col]]   ~ "Both",
        .data[[oar_col]] & !.data[[iq_col]]  ~ "Oarfish only",
        !.data[[oar_col]] & .data[[iq_col]]  ~ "IsoQuant only",
        TRUE                                  ~ "Neither"
      ),
      biotype2 = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")
    ) %>%
    filter(category != "Neither") %>%
    count(category, biotype2) %>%
    mutate(platform = platform)
})

tool_comparison_gene_biotype$category <- factor(
  tool_comparison_gene_biotype$category,
  levels = c("Oarfish only", "Both", "IsoQuant only")
)

ggplot(tool_comparison_gene_biotype, aes(x = category, y = n, fill = biotype2)) +
  geom_col(position = "fill") +
  facet_wrap(~platform) +
  scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
  theme_minimal() +
  labs(title = "Biotype composition by tool-specific vs shared gene detection",
       y = "Proportion", x = NULL, fill = "Biotype") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

``` 

### Tool overlap summary
```{r GENE: Tool overlap summary, results='asis'}
gene_overlap_summary <- tool_comparison_gene %>%
  group_by(platform) %>%
  summarise(
    Oarfish_only = sum(n[category == "Oarfish only"], na.rm = TRUE),
    Both         = sum(n[category == "Both"], na.rm = TRUE),
    IsoQuant_only = sum(n[category == "IsoQuant only"], na.rm = TRUE),
    Total_detected = sum(n, na.rm = TRUE),
    `Shared fraction` = Both / Total_detected,
    `Oarfish fraction` = (Oarfish_only + Both) / Total_detected,
    `IsoQuant fraction` = (IsoQuant_only + Both) / Total_detected,
    .groups = "drop"
  )

kable(
  gene_overlap_summary,
  digits = 3,
  caption = "Gene detection overlap summary: Oarfish vs IsoQuant by platform"
)
```

### Jaccard similarity (all 7 platform-tool combinations)
```{r GENE: Jaccard similarity, fig.height=4, fig.width=5}
all_cols <- c("Illumina", lr_cols)
binary_matrix <- gene_discovery %>%
  select(all_of(all_cols)) %>%
  mutate(across(everything(), as.integer))

jaccard_sim <- function(x, y) { sum(x & y) / sum(x | y) }
jaccard_matrix <- outer(all_cols, all_cols, Vectorize(function(i, j) {
  jaccard_sim(binary_matrix[[i]], binary_matrix[[j]])
}))
colnames(jaccard_matrix) <- rownames(jaccard_matrix) <- all_cols

Heatmap(jaccard_matrix,
        name = "Jaccard\nSimilarity",
        col  = colorRampPalette(c("grey90", "#06317F"))(50),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(round(jaccard_matrix[i, j], 2), x, y,
                    gp = gpar(col = ifelse(jaccard_matrix[i, j] > 0.5, "white", "black"), fontsize = 9))
        },
        row_names_gp    = gpar(fontsize = 9),
        column_names_gp = gpar(fontsize = 9))
```

### Detected gene length distribution
```{r GENE: Length distribution, fig.height=4, fig.width=10}
gene_discovery$length_bin <- cut(
  gene_discovery$`Gene length`,
  breaks = c(0, 500, 1000, 1500, 2000, 2500, Inf),
  labels = c("0-0.5k", "0.5-1k", "1-1.5k", "1.5-2k", "2-2.5k", "2.5k+")
)

bg_dist <- gene_discovery %>%
  filter(!is.na(length_bin)) %>%
  count(length_bin) %>%
  mutate(label = "All annotated genes")

gene_discovery %>%
  pivot_longer(cols = all_of(lr_cols), names_to = "tool_platform", values_to = "keep") %>%
  filter(keep, !is.na(length_bin)) %>%
  mutate(
    tool = if_else(grepl("Oarfish", tool_platform), "Oarfish", "IsoQuant"),
    platform = sub(" \\(.*\\)", "", tool_platform)
  ) %>%
  count(length_bin, tool, platform) %>%
  ggplot(aes(x = length_bin, y = n, color = tool, group = tool)) +
  geom_line(data = bg_dist, aes(x = length_bin, y = n, group = 1),
            color = "grey60", linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~platform, ncol = 3) +
  scale_color_manual(values = tool_colors) +
  theme_minimal() +
  labs(title = "Detected genes by length bin",
       x = "Gene length (avg transcript length)", y = "# of genes", color = "Tool")
```


# Transcript discovery analysis

## Build `tx_discovery` per tool

```{r TX: Filter to human}
ill_bulk.tx.dge       <- ill_bulk.tx.dge[is.human(ill_bulk.tx.dge), ]
oar.ont_bulk.tx.dge   <- oar.ont_bulk.tx.dge[is.human(oar.ont_bulk.tx.dge), ]
oar.pb_bulk.tx.dge    <- oar.pb_bulk.tx.dge[is.human(oar.pb_bulk.tx.dge), ]
oar.drna_bulk.tx.dge  <- oar.drna_bulk.tx.dge[is.human(oar.drna_bulk.tx.dge), ]
iq.ont_bulk.tx.dge    <- iq.ont_bulk.tx.dge[is.human(iq.ont_bulk.tx.dge), ]
iq.pb_bulk.tx.dge     <- iq.pb_bulk.tx.dge[is.human(iq.pb_bulk.tx.dge), ]
iq.drna_bulk.tx.dge   <- iq.drna_bulk.tx.dge[is.human(iq.drna_bulk.tx.dge), ]
```

```{r TX: Normalize Oarfish TX rownames}
# Oarfish TX rownames may still be pipe-separated; strip to ENST IDs.
# Store pipe-separated full_id before stripping, so we can recover annotations later.
save_full_id <- function(dge) {
  if (!"full_id" %in% colnames(dge$genes)) {
    dge$genes$full_id <- rownames(dge)
  }
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}

ill_bulk.tx.dge      <- save_full_id(ill_bulk.tx.dge)
oar.ont_bulk.tx.dge  <- save_full_id(oar.ont_bulk.tx.dge)
oar.pb_bulk.tx.dge   <- save_full_id(oar.pb_bulk.tx.dge)
oar.drna_bulk.tx.dge <- save_full_id(oar.drna_bulk.tx.dge)
# IsoQuant TX already normalized (done above); ensure no pipe remains
iq.ont_bulk.tx.dge   <- save_full_id(iq.ont_bulk.tx.dge)
iq.pb_bulk.tx.dge    <- save_full_id(iq.pb_bulk.tx.dge)
iq.drna_bulk.tx.dge  <- save_full_id(iq.drna_bulk.tx.dge)
```

```{r TX: Build tx_discovery}
all_tx <- Reduce(union, lapply(
  list(ill_bulk.tx.dge,
       oar.ont_bulk.tx.dge, oar.pb_bulk.tx.dge, oar.drna_bulk.tx.dge,
       iq.ont_bulk.tx.dge,  iq.pb_bulk.tx.dge,  iq.drna_bulk.tx.dge),
  rownames
))

ill_bulk.tx.dge      <- pad_dge(ill_bulk.tx.dge,      all_tx)
oar.ont_bulk.tx.dge  <- pad_dge(oar.ont_bulk.tx.dge,  all_tx)
oar.pb_bulk.tx.dge   <- pad_dge(oar.pb_bulk.tx.dge,   all_tx)
oar.drna_bulk.tx.dge <- pad_dge(oar.drna_bulk.tx.dge, all_tx)
iq.ont_bulk.tx.dge   <- pad_dge(iq.ont_bulk.tx.dge,   all_tx)
iq.pb_bulk.tx.dge    <- pad_dge(iq.pb_bulk.tx.dge,    all_tx)
iq.drna_bulk.tx.dge  <- pad_dge(iq.drna_bulk.tx.dge,  all_tx)

tx_discovery <- tibble(
  tx_id                 = all_tx,
  Illumina              = filterByExpr(ill_bulk.tx.dge,      group = ill_bulk.tx.dge$samples$group,      min.count = 5),
  `ONT cDNA (Oarfish)`  = filterByExpr(oar.ont_bulk.tx.dge,  group = oar.ont_bulk.tx.dge$samples$group,  min.count = 5),
  `PacBio (Oarfish)`    = filterByExpr(oar.pb_bulk.tx.dge,   group = oar.pb_bulk.tx.dge$samples$group,   min.count = 5),
  `ONT dRNA (Oarfish)`  = filterByExpr(oar.drna_bulk.tx.dge, group = oar.drna_bulk.tx.dge$samples$group, min.count = 5),
  `ONT cDNA (IsoQuant)` = filterByExpr(iq.ont_bulk.tx.dge,   group = iq.ont_bulk.tx.dge$samples$group,   min.count = 5),
  `PacBio (IsoQuant)`   = filterByExpr(iq.pb_bulk.tx.dge,    group = iq.pb_bulk.tx.dge$samples$group,    min.count = 5),
  `ONT dRNA (IsoQuant)` = filterByExpr(iq.drna_bulk.tx.dge,  group = iq.drna_bulk.tx.dge$samples$group,  min.count = 5)
) %>% filter(
  Illumina | `ONT cDNA (Oarfish)` | `PacBio (Oarfish)` | `ONT dRNA (Oarfish)` |
  `ONT cDNA (IsoQuant)` | `PacBio (IsoQuant)` | `ONT dRNA (IsoQuant)`
)
```

## Add biotype and length

```{r TX: Add annotation}
# Join biotype/length/gene_id from the reference table built from Illumina annotations
tx_discovery <- tx_discovery %>%
  left_join(tx_annotation, by = "tx_id") %>%
  mutate(biotype = ifelse(
    biotype %in% names(sort(table(biotype), decreasing = TRUE)[1:6]),
    biotype, "Others"
  ))

output_list[["tx_discovery"]] <- tx_discovery
saveRDS(output_list, params$out.rds)
```

## Tx detection plots

### Upset plot: Oarfish platforms
```{r TX: Upset Oarfish, fig.height=5, fig.width=10}
ComplexUpset::upset(
  tx_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = c("Illumina", "ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)"),
  annotations = list(
    "Transcript length" = list(
      aes  = aes(x = intersection, y = length),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) + ggtitle("Transcript detection — Oarfish")
```

### Upset plot: IsoQuant platforms
```{r TX: Upset IsoQuant, fig.height=5, fig.width=10}
ComplexUpset::upset(
  tx_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = c("Illumina", "ONT cDNA (IsoQuant)", "PacBio (IsoQuant)", "ONT dRNA (IsoQuant)"),
  annotations = list(
    "Transcript length" = list(
      aes  = aes(x = intersection, y = length),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) + ggtitle("Transcript detection — IsoQuant (no_novel)")
```

### Upset plot: All platforms and tools (including Illumina)
```{r TX: Upset combined, fig.height=6, fig.width=16}
# Transcripts detected by Illumina + all 3 Oarfish platforms
oar_ill_mask_tx <- tx_discovery$Illumina &
                   tx_discovery$`ONT cDNA (Oarfish)` &
                   tx_discovery$`PacBio (Oarfish)` &
                   tx_discovery$`ONT dRNA (Oarfish)`
n_oar_ill_tx    <- sum(oar_ill_mask_tx)
# Of those, also confirmed by all 3 IsoQuant platforms
n_all7_tx       <- sum(oar_ill_mask_tx &
                       tx_discovery$`ONT cDNA (IsoQuant)` &
                       tx_discovery$`PacBio (IsoQuant)` &
                       tx_discovery$`ONT dRNA (IsoQuant)`)
# Of those, no IsoQuant support at all
n_no_iq_tx      <- sum(oar_ill_mask_tx &
                       !tx_discovery$`ONT cDNA (IsoQuant)` &
                       !tx_discovery$`PacBio (IsoQuant)` &
                       !tx_discovery$`ONT dRNA (IsoQuant)`)

ComplexUpset::upset(
  tx_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = all_cols_upset,
  n_intersections = 30,
  annotations = list(
    "Transcript length" = list(
      aes  = aes(x = intersection, y = length),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) +
  ggtitle("Transcript detection — Illumina + Oarfish + IsoQuant (all platforms)") +
  patchwork::plot_annotation(
    caption = glue::glue(
      "Illumina ∩ all Oarfish: {scales::comma(n_oar_ill_tx)} transcripts  |  ",
      "Also in all IsoQuant: {scales::comma(n_all7_tx)} ({scales::percent(n_all7_tx / n_oar_ill_tx, accuracy = 0.1)})  |  ",
      "No IsoQuant support: {scales::comma(n_no_iq_tx)} ({scales::percent(n_no_iq_tx / n_oar_ill_tx, accuracy = 0.1)})"
    ),
    theme = theme(plot.caption = element_text(size = 11, hjust = 0.5, face = "bold"))
  )
```

### Tool comparison per platform
```{r TX: Per-platform tool comparison, fig.height=3, fig.width=9}
tool_comparison_tx <- map_dfr(names(platform_pairs), function(platform) {
  oar_col <- platform_pairs[[platform]][1]
  iq_col  <- platform_pairs[[platform]][2]
  tx_discovery %>%
    mutate(
      category = case_when(
        .data[[oar_col]] & .data[[iq_col]]   ~ "Both",
        .data[[oar_col]] & !.data[[iq_col]]  ~ "Oarfish only",
        !.data[[oar_col]] & .data[[iq_col]]  ~ "IsoQuant only",
        TRUE                                  ~ "Neither"
      )
    ) %>%
    filter(category != "Neither") %>%
    count(category) %>%
    mutate(platform = platform)
})

tool_comparison_tx$category <- factor(
  tool_comparison_tx$category,
  levels = c("Oarfish only", "Both", "IsoQuant only")
)

ggplot(tool_comparison_tx, aes(x = platform, y = n, fill = category)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c(
    "Oarfish only"  = tool_colors["Oarfish"],
    "Both"          = "#b2b2b2",
    "IsoQuant only" = tool_colors["IsoQuant"]
  )) +
  theme_minimal() +
  labs(title = "Transcript detection: tool overlap per platform",
       y = "Number of transcripts", x = NULL, fill = "Category") +
  theme(axis.text.x = element_text(size = 11))
```

```{r TX: Per-platform tool comparison by biotype, fig.height=4, fig.width=12}
tool_comparison_tx_biotype <- map_dfr(names(platform_pairs), function(platform) {
  oar_col <- platform_pairs[[platform]][1]
  iq_col  <- platform_pairs[[platform]][2]
  tx_discovery %>%
    mutate(
      category = case_when(
        .data[[oar_col]] & .data[[iq_col]]   ~ "Both",
        .data[[oar_col]] & !.data[[iq_col]]  ~ "Oarfish only",
        !.data[[oar_col]] & .data[[iq_col]]  ~ "IsoQuant only",
        TRUE                                  ~ "Neither"
      ),
      biotype2 = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")
    ) %>%
    filter(category != "Neither") %>%
    count(category, biotype2) %>%
    mutate(platform = platform)
})

tool_comparison_tx_biotype$category <- factor(
  tool_comparison_tx_biotype$category,
  levels = c("Oarfish only", "Both", "IsoQuant only")
)

ggplot(tool_comparison_tx_biotype, aes(x = category, y = n, fill = biotype2)) +
  geom_col(position = "fill") +
  facet_wrap(~platform) +
  scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
  theme_minimal() +
  labs(title = "Biotype composition by tool-specific vs shared transcript detection",
       y = "Proportion", x = NULL, fill = "Biotype") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

``` 

### Tool overlap summary
```{r TX: Tool overlap summary, results='asis'}
tx_overlap_summary <- tool_comparison_tx %>%
  group_by(platform) %>%
  summarise(
    Oarfish_only = sum(n[category == "Oarfish only"], na.rm = TRUE),
    Both         = sum(n[category == "Both"], na.rm = TRUE),
    IsoQuant_only = sum(n[category == "IsoQuant only"], na.rm = TRUE),
    Total_detected = sum(n, na.rm = TRUE),
    `Shared fraction` = Both / Total_detected,
    `Oarfish fraction` = (Oarfish_only + Both) / Total_detected,
    `IsoQuant fraction` = (IsoQuant_only + Both) / Total_detected,
    .groups = "drop"
  )

kable(
  tx_overlap_summary,
  digits = 3,
  caption = "Transcript detection overlap summary: Oarfish vs IsoQuant by platform"
)
```

### Jaccard similarity
```{r TX: Jaccard similarity, fig.height=4, fig.width=5}
binary_matrix_tx <- tx_discovery %>%
  select(all_of(all_cols)) %>%
  mutate(across(everything(), as.integer))

jaccard_matrix_tx <- outer(all_cols, all_cols, Vectorize(function(i, j) {
  jaccard_sim(binary_matrix_tx[[i]], binary_matrix_tx[[j]])
}))
colnames(jaccard_matrix_tx) <- rownames(jaccard_matrix_tx) <- all_cols

Heatmap(jaccard_matrix_tx,
        name = "Jaccard\nSimilarity",
        col  = colorRampPalette(c("grey90", "#06317F"))(50),
        cell_fun = function(j, i, x, y, width, height, fill) {
          grid.text(round(jaccard_matrix_tx[i, j], 2), x, y,
                    gp = gpar(col = ifelse(jaccard_matrix_tx[i, j] > 0.5, "white", "black"), fontsize = 9))
        },
        row_names_gp    = gpar(fontsize = 9),
        column_names_gp = gpar(fontsize = 9))
```

### Discovered transcript length distribution
```{r TX: Length distribution, fig.height=4, fig.width=10}
tx_discovery$length_bin <- cut(
  tx_discovery$length,
  breaks = c(0, 500, 1000, 1500, 2000, 2500, Inf),
  labels = c("0-0.5k", "0.5-1k", "1-1.5k", "1.5-2k", "2-2.5k", "2.5k+")
)

bg_dist_tx <- tx_discovery %>%
  filter(!is.na(length_bin)) %>%
  count(length_bin) %>%
  mutate(label = "All annotated transcripts")

tx_discovery %>%
  pivot_longer(cols = all_of(lr_cols), names_to = "tool_platform", values_to = "keep") %>%
  filter(keep, !is.na(length_bin)) %>%
  mutate(
    tool     = if_else(grepl("Oarfish", tool_platform), "Oarfish", "IsoQuant"),
    platform = sub(" \\(.*\\)", "", tool_platform)
  ) %>%
  count(length_bin, tool, platform) %>%
  ggplot(aes(x = length_bin, y = n, color = tool, group = tool)) +
  geom_line(data = bg_dist_tx, aes(x = length_bin, y = n, group = 1),
            color = "grey60", linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~platform, ncol = 3) +
  scale_color_manual(values = tool_colors) +
  theme_minimal() +
  labs(title = "Detected transcripts by length bin",
       x = "Transcript length", y = "# of transcripts", color = "Tool",
       caption = "Dashed grey line: all annotated transcripts")
```

```{r TX: Length distribution protein-coding, fig.height=4, fig.width=10}
bg_dist_tx_pc <- tx_discovery %>%
  filter(!is.na(length_bin), biotype == "protein_coding") %>%
  count(length_bin)

tx_discovery %>%
  filter(biotype == "protein_coding") %>%
  pivot_longer(cols = all_of(lr_cols), names_to = "tool_platform", values_to = "keep") %>%
  filter(keep, !is.na(length_bin)) %>%
  mutate(
    tool     = if_else(grepl("Oarfish", tool_platform), "Oarfish", "IsoQuant"),
    platform = sub(" \\(.*\\)", "", tool_platform)
  ) %>%
  count(length_bin, tool, platform) %>%
  ggplot(aes(x = length_bin, y = n, color = tool, group = tool)) +
  geom_line(data = bg_dist_tx_pc, aes(x = length_bin, y = n, group = 1),
            color = "grey60", linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~platform, ncol = 3) +
  scale_color_manual(values = tool_colors) +
  theme_minimal() +
  labs(title = "Detected protein-coding transcripts by length bin",
       x = "Transcript length", y = "# of transcripts", color = "Tool",
       caption = "Dashed grey line: all annotated protein-coding transcripts")
```
