---
title: "Platform-specific DE transcript diagnosis"
author: "Yupei"
date: '`r Sys.Date()`'
version: 'v1.0'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
  bulk_DE_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds"
  rarefaction_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE_rarefaction.rds"
  log2FC_cutoff: 0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  fig.width = 15,
  fig.height = 10,
  warning = FALSE,
  message = FALSE,
  cache = FALSE
)

if (!exists("params")) {
  params <- list(
    random_seed = 2024,
    cache_dir = NULL,
    fig.path = NULL,
    bulk_DE_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds",
    rarefaction_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE_rarefaction.rds",
    log2FC_cutoff = 0
  )
}

knitr::opts_chunk$set(
  fig.path = params$fig.path,
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL
)
if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache.path = params$cache_dir)
}
```

```{r Libraries, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(purrr)
library(stringr)
library(ggpubr)
library(edgeR)
```

# Setup

```{r load data}
bulk_DE_rst <- readRDS(params$bulk_DE_rds)
dge_lists   <- bulk_DE_rst$dge_lists
dte_lists   <- bulk_DE_rst$dte_lists

name_map <- c(ill = "Illumina", ont = "ONT cDNA", pb = "PacBio", drna = "ONT dRNA")
names(dge_lists) <- name_map[names(dge_lists)]
names(dte_lists) <- name_map[names(dte_lists)]

platforms   <- c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA")
comparisons <- c("AvsL")
```

For each platform, we take transcripts that are DE (FDR < 0.05) **exclusively** in that platform and not in any other. We then ask:

1. **False positive?** Does the other platform have sufficient counts (logCPM > 0, i.e. CPM > 1) but still fail to detect DE (FDR > 0.05)? If the transcript is well expressed in the other platform but not called DE, the specific-platform call may be a false positive.

2. **Detection bias?** Is the transcript lowly expressed (logCPM ≤ 0) in the other platform? This suggests the other platform simply did not capture this transcript, and the DE call is likely real.

Each scatter plot shows logCPM in the specific platform (x-axis) vs logCPM in the other platform (y-axis) for platform-exclusive DE transcripts. Points are coloured by logFC direction in the other platform. The horizontal dashed line marks logCPM = 0 (CPM = 1): points **above** this line are Q1 candidates (expressed but missed); points **below** are Q2 candidates (not captured).

```{r helpers}
color_palette <- c(
  PacBio     = "#df1995",
  "ONT dRNA" = "#00789b",
  "ONT cDNA" = "#04476c",
  Illumina   = "#e88b20"
)

get_exclusive_tx_ids <- function(comp, specific_plt) {
  de_sets <- purrr::map(platforms, function(plt) {
    tbl <- dte_lists[[plt]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id") %>%
      mutate(FDR = p.adjust(PValue, method = "BH"))
    tbl$transcript_id[tbl$FDR < 0.05]
  })
  names(de_sets) <- platforms
  other_ids <- purrr::reduce(de_sets[setdiff(platforms, specific_plt)], union)
  setdiff(de_sets[[specific_plt]], other_ids)
}
```

# Platform-missed DE transcripts and length bias

For each platform we identify transcripts that are **not DE** (FDR ≥ 0.05, or absent from that platform's test) but are **DE in at least one other platform** (FDR < 0.05). This captures all missed DE calls regardless of whether the miss was due to absence or low power. We compare the length distribution of these missed transcripts across platforms, and against a background of transcripts DE in all four platforms.

```{r platform absence setup}
de_sets <- purrr::map(platforms, function(plt) {
  dte_lists[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH")) %>%
    filter(FDR < 0.05) %>%
    pull(transcript_id)
})
names(de_sets) <- platforms

tx_len_bio <- dte_lists[["Illumina"]][["AvsL.DE"]]$genes %>%
  tibble::rownames_to_column("transcript_id") %>%
  mutate(biotype = sapply(strsplit(full_id, "\\|"), `[`, 8),
         biotype_simple = case_when(
           biotype == "protein_coding"  ~ "protein_coding",
           biotype == "lncRNA"          ~ "lncRNA",
           grepl("pseudogene", biotype) ~ "pseudogene",
           TRUE                         ~ "other_ncRNA"
         )) %>%
  select(transcript_id, Length, biotype_simple)

# Universal DE: DE in all 4 platforms (background reference)
universal_de_ids <- purrr::reduce(de_sets, intersect)

# DE in at least one platform
any_de_ids <- purrr::reduce(de_sets, union)

# Missed DE: DE in ≥1 other platform, not DE in this platform
missed_df <- purrr::map_dfr(platforms, function(plt) {
  de_in_any_other <- purrr::reduce(de_sets[setdiff(platforms, plt)], union)
  tibble(
    transcript_id = setdiff(de_in_any_other, de_sets[[plt]]),
    platform = plt
  )
}) %>%
  left_join(tx_len_bio, by = "transcript_id") %>%
  filter(!is.na(Length)) %>%
  mutate(platform = factor(platform, levels = platforms))

bg_universal <- tx_len_bio %>%
  filter(transcript_id %in% universal_de_ids)
```

```{r platform absence plot, fig.height=5, fig.width=11}
med_missed <- missed_df %>%
  group_by(platform) %>%
  summarise(med = median(Length, na.rm = TRUE), n = n(), .groups = "drop")

med_bg <- tibble(
  med = median(bg_universal$Length, na.rm = TRUE),
  n   = nrow(bg_universal)
)

ggplot(missed_df, aes(x = platform, y = Length, fill = platform)) +
  geom_hline(yintercept = med_bg$med, linetype = "dashed", color = "gray40") +
  geom_violin(trim = TRUE, alpha = 0.75, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
  geom_text(data = med_missed,
            aes(x = platform, y = med,
                label = paste0(round(med), "\nn=", n)),
            vjust = -0.3, size = 2.5, color = "black", inherit.aes = FALSE) +
  scale_fill_manual(values = color_palette) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Length of platform-missed DE transcripts (not DE here, DE in >=1 other platform)",
    subtitle = paste0("Dashed line = universal DE (all 4 platforms) median: ",
                      round(med_bg$med), " bp  (n=", med_bg$n, ")"),
    x = NULL, y = "Transcript length (bp, log10 scale)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 15, hjust = 1))
```

# DTE stratified by transcript length

Transcripts are divided into four length bins: **<1 kb**, **1–3 kb**, **3–10 kb**, and **>10 kb**. Within each bin we compute per platform: (1) the number of expressed transcripts (logCPM > 0), and (2) the DE rate — the fraction of expressed transcripts that are called DE (FDR < 0.05). The two panels together separate capture bias (fewer expressed transcripts in a bin) from statistical-power bias (expressed but not called DE). If PacBio size-selection depletes short transcripts, its expressed count should drop in the <1 kb bin; if ONT cDNA processivity limits long-transcript coverage, its expressed count and DE rate should drop in the >10 kb bin.

```{r dte by length bin, fig.height=8, fig.width=11}
length_breaks <- c(0, 1000, 3000, 10000, Inf)
length_labels <- c("<1 kb", "1-3 kb", "3-10 kb", ">10 kb")

tx_binned <- tx_len_bio %>%
  mutate(length_bin = cut(Length, breaks = length_breaks,
                          labels = length_labels, include.lowest = TRUE))

# Common expressed: intersection of all 4 platforms' filterByExpr sets
common_ids <- purrr::map(platforms, function(plt)
  rownames(dte_lists[[plt]][["AvsL.DE"]]$table)) %>%
  purrr::reduce(intersect)

strat_len <- purrr::map_dfr(platforms, function(plt) {
  tbl <- dte_lists[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH"))

  p1 <- tbl %>%
    filter(transcript_id %in% common_ids) %>%
    left_join(tx_binned %>% select(transcript_id, length_bin), by = "transcript_id") %>%
    filter(!is.na(length_bin)) %>%
    group_by(length_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Common expressed\n(all 4 platforms)", .groups = "drop")

  p2 <- tbl %>%
    left_join(tx_binned %>% select(transcript_id, length_bin), by = "transcript_id") %>%
    filter(!is.na(length_bin)) %>%
    group_by(length_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Platform-specific\nfilterByExpr", .groups = "drop")

  bind_rows(p1, p2)
}) %>%
  mutate(length_bin = factor(length_bin, levels = length_labels),
         platform   = factor(platform, levels = platforms),
         type       = factor(type, levels = c("Common expressed\n(all 4 platforms)",
                                              "Platform-specific\nfilterByExpr")))

p_len_n <- ggplot(strat_len,
       aes(x = length_bin, y = n, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = color_palette) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~type) +
  labs(title = "Number of expressed transcripts by length bin",
       x = NULL, y = "n transcripts (log10)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_len_rate <- ggplot(strat_len,
       aes(x = length_bin, y = de_rate * 100, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = color_palette) +
  scale_size_continuous(range = c(1.5, 6), labels = scales::comma, name = "n transcripts") +
  facet_wrap(~type) +
  labs(title = "DE rate (%) by length bin",
       x = "Transcript length bin", y = "% DE (FDR < 0.05)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

ggarrange(p_len_n, p_len_rate, nrow = 2, align = "v")
```

The same analysis on the **20M downsampled** data shows how reduced sequencing depth interacts with length bias — shorter transcripts may lose the most power when depth drops.

```{r dte by length bin 20M, fig.height=8, fig.width=11}
dte_20m <- readRDS(params$rarefaction_rds)[["20M"]]$dte

# Reuse tx_binned (length bins built from full-data annotation)
# 20M genes slot has Length but no full_id; join by transcript_id
tx_binned_20m <- dte_20m[["Illumina"]][["AvsL.DE"]]$genes %>%
  tibble::rownames_to_column("transcript_id") %>%
  select(transcript_id, Length) %>%
  mutate(length_bin = cut(Length, breaks = length_breaks,
                          labels = length_labels, include.lowest = TRUE))

common_ids_20m <- purrr::map(platforms, function(plt)
  rownames(dte_20m[[plt]][["AvsL.DE"]]$table)) %>%
  purrr::reduce(intersect)

strat_len_20m <- purrr::map_dfr(platforms, function(plt) {
  tbl <- dte_20m[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH"))

  p1 <- tbl %>%
    filter(transcript_id %in% common_ids_20m) %>%
    left_join(tx_binned_20m %>% select(transcript_id, length_bin), by = "transcript_id") %>%
    filter(!is.na(length_bin)) %>%
    group_by(length_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Common expressed\n(all 4 platforms)", .groups = "drop")

  p2 <- tbl %>%
    left_join(tx_binned_20m %>% select(transcript_id, length_bin), by = "transcript_id") %>%
    filter(!is.na(length_bin)) %>%
    group_by(length_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Platform-specific\nfilterByExpr", .groups = "drop")

  bind_rows(p1, p2)
}) %>%
  mutate(length_bin = factor(length_bin, levels = length_labels),
         platform   = factor(platform, levels = platforms),
         type       = factor(type, levels = c("Common expressed\n(all 4 platforms)",
                                              "Platform-specific\nfilterByExpr")))

p_len_n_20m <- ggplot(strat_len_20m,
       aes(x = length_bin, y = n, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = color_palette) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~type) +
  labs(title = "Number of expressed transcripts by length bin (20M reads)",
       x = NULL, y = "n transcripts (log10)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_len_rate_20m <- ggplot(strat_len_20m,
       aes(x = length_bin, y = de_rate * 100, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = color_palette) +
  scale_size_continuous(range = c(1.5, 6), labels = scales::comma, name = "n transcripts") +
  facet_wrap(~type) +
  labs(title = "DE rate (%) by length bin (20M reads)",
       x = "Transcript length bin", y = "% DE (FDR < 0.05)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

ggarrange(p_len_n_20m, p_len_rate_20m, nrow = 2, align = "v")
```

# DTE stratified by number of isoforms per gene

Genes with more isoforms are harder to quantify accurately — short-read platforms struggle to assign reads to the correct isoform when many share similar exon content. We expect higher-isoform-count genes to show greater platform divergence in DE calls. Transcripts are grouped by the number of isoforms annotated for their parent gene (from the Illumina DTE annotation): **1**, **2**, **3–5**, **6–10**, and **>10** isoforms.

```{r dte by isoform count, fig.height=8, fig.width=11}
isoform_count <- dte_lists[["Illumina"]][["AvsL.DE"]]$genes %>%
  tibble::rownames_to_column("transcript_id") %>%
  mutate(gene_id = sapply(strsplit(full_id, "\\|"), `[`, 6)) %>%
  group_by(gene_id) %>%
  mutate(n_isoforms = n()) %>%
  ungroup() %>%
  mutate(isoform_bin = cut(n_isoforms,
                           breaks = c(0, 1, 2, 5, 10, Inf),
                           labels = c("1", "2", "3-5", "6-10", ">10"),
                           include.lowest = TRUE)) %>%
  select(transcript_id, gene_id, n_isoforms, isoform_bin)

strat_iso <- purrr::map_dfr(platforms, function(plt) {
  tbl <- dte_lists[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH"))

  p1 <- tbl %>%
    filter(transcript_id %in% common_ids) %>%
    left_join(isoform_count %>% select(transcript_id, isoform_bin),
              by = "transcript_id") %>%
    filter(!is.na(isoform_bin)) %>%
    group_by(isoform_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Common expressed\n(all 4 platforms)",
              .groups = "drop")

  p2 <- tbl %>%
    left_join(isoform_count %>% select(transcript_id, isoform_bin),
              by = "transcript_id") %>%
    filter(!is.na(isoform_bin)) %>%
    group_by(isoform_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Platform-specific\nfilterByExpr",
              .groups = "drop")

  bind_rows(p1, p2)
}) %>%
  mutate(isoform_bin = factor(isoform_bin, levels = c("1", "2", "3-5", "6-10", ">10")),
         platform    = factor(platform, levels = platforms),
         type        = factor(type, levels = c("Common expressed\n(all 4 platforms)",
                                               "Platform-specific\nfilterByExpr")))

p_iso_n <- ggplot(strat_iso,
       aes(x = isoform_bin, y = n, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = color_palette) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~type) +
  labs(title = "Number of expressed transcripts by isoform count (full depth)",
       x = NULL, y = "n transcripts (log10)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_iso_rate <- ggplot(strat_iso,
       aes(x = isoform_bin, y = de_rate * 100, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = color_palette) +
  scale_size_continuous(range = c(1.5, 6), labels = scales::comma,
                        name = "n transcripts") +
  facet_wrap(~type) +
  labs(title = "DE rate (%) by isoform count (full depth)",
       x = "Number of isoforms per gene", y = "% DE (FDR < 0.05)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

ggarrange(p_iso_n, p_iso_rate, nrow = 2, align = "v")
```

```{r dte by isoform count 20M, fig.height=8, fig.width=11}
strat_iso_20m <- purrr::map_dfr(platforms, function(plt) {
  tbl <- dte_20m[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH"))

  p1 <- tbl %>%
    filter(transcript_id %in% common_ids_20m) %>%
    left_join(isoform_count %>% select(transcript_id, isoform_bin),
              by = "transcript_id") %>%
    filter(!is.na(isoform_bin)) %>%
    group_by(isoform_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Common expressed\n(all 4 platforms)",
              .groups = "drop")

  p2 <- tbl %>%
    left_join(isoform_count %>% select(transcript_id, isoform_bin),
              by = "transcript_id") %>%
    filter(!is.na(isoform_bin)) %>%
    group_by(isoform_bin) %>%
    summarise(n = n(), n_de = sum(FDR < 0.05), de_rate = n_de / n,
              platform = plt, type = "Platform-specific\nfilterByExpr",
              .groups = "drop")

  bind_rows(p1, p2)
}) %>%
  mutate(isoform_bin = factor(isoform_bin, levels = c("1", "2", "3-5", "6-10", ">10")),
         platform    = factor(platform, levels = platforms),
         type        = factor(type, levels = c("Common expressed\n(all 4 platforms)",
                                               "Platform-specific\nfilterByExpr")))

p_iso_n_20m <- ggplot(strat_iso_20m,
       aes(x = isoform_bin, y = n, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_manual(values = color_palette) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~type) +
  labs(title = "Number of expressed transcripts by isoform count (20M reads)",
       x = NULL, y = "n transcripts (log10)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

p_iso_rate_20m <- ggplot(strat_iso_20m,
       aes(x = isoform_bin, y = de_rate * 100, color = platform, group = platform)) +
  geom_line(linewidth = 0.8) +
  geom_point(aes(size = n)) +
  scale_color_manual(values = color_palette) +
  scale_size_continuous(range = c(1.5, 6), labels = scales::comma,
                        name = "n transcripts") +
  facet_wrap(~type) +
  labs(title = "DE rate (%) by isoform count (20M reads)",
       x = "Number of isoforms per gene", y = "% DE (FDR < 0.05)", color = NULL) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "top")

ggarrange(p_iso_n_20m, p_iso_rate_20m, nrow = 2, align = "v")
```

# Diagnosis plots per platform

```{r diagnosis, fig.height=5, fig.width=14, results='asis'}
for (specific_plt in platforms) {
  cat("\n\n##", specific_plt, "-specific transcripts\n\n")

  other_plts <- setdiff(platforms, specific_plt)

  plotlist <- purrr::map(other_plts, function(other_plt) {
    all_data <- purrr::map_dfr(comparisons, function(comp) {
      excl_ids <- get_exclusive_tx_ids(comp, specific_plt)
      if (length(excl_ids) == 0) return(NULL)

      spec_tbl <- dte_lists[[specific_plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("transcript_id") %>%
        filter(transcript_id %in% excl_ids) %>%
        select(transcript_id, logCPM_spec = logCPM, logFC_spec = logFC)

      other_tbl <- dte_lists[[other_plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("transcript_id") %>%
        mutate(FDR_other = p.adjust(PValue, method = "BH")) %>%
        filter(transcript_id %in% excl_ids) %>%
        select(transcript_id, logCPM_other = logCPM, logFC_other = logFC, FDR_other)

      inner_join(spec_tbl, other_tbl, by = "transcript_id") %>%
        mutate(
          comparison = comp,
          category   = if_else(logCPM_other > 0,
                               "Expressed in other (Q1: potential FP)",
                               "Low in other (Q2: detection bias)")
        )
    })

    if (nrow(all_data) == 0) return(NULL)

    n_q1 <- sum(all_data$category == "Expressed in other (Q1: potential FP)")
    n_q2 <- sum(all_data$category == "Low in other (Q2: detection bias)")

    ggplot(all_data, aes(x = logCPM_spec, y = logCPM_other, color = logFC_other)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
      geom_point(alpha = 0.6, size = 0.8) +
      scale_color_gradient2(low = "blue", mid = "gray80", high = "red", midpoint = 0,
                            name = paste0("logFC\n(", other_plt, ")")) +
      facet_wrap(~comparison, ncol = 3) +
      annotate("text", x = -Inf, y = Inf,
               label = paste0("Q1: ", n_q1, "  Q2: ", n_q2),
               hjust = -0.1, vjust = 1.5, size = 3, color = "black") +
      labs(
        title = paste0(specific_plt, "-specific | other: ", other_plt),
        x     = paste0("logCPM (", specific_plt, ")"),
        y     = paste0("logCPM (", other_plt, ")")
      ) +
      theme_minimal(base_size = 8)
  }) %>% purrr::compact()

  if (length(plotlist) > 0) {
    p <- ggarrange(plotlist = plotlist, ncol = 3, nrow = 1)
    print(p)
  }
}
```

# LogFC consistency for expressed-but-missed transcripts (Q1)

For transcripts classified as Q1 (logCPM > 0 in the other platform but not DE), check whether the logFC direction is consistent across platforms. Consistent logFC with non-significant FDR in the other platform would suggest an underpowered call rather than a true false positive.

# Transcript length distribution of non-universal DE transcripts

Transcript lengths are compared for transcripts that are DE (FDR < 0.05) in each platform but **not** detected by all four platforms (i.e., commonly identified transcripts are excluded). Known technical biases predict: PacBio size-selection depletes short transcripts → PacBio-enriched DE transcripts should skew longer; ONT cDNA has poor processivity for long molecules → cDNA-enriched DE transcripts should skew shorter. The background distribution (all expressed transcripts in Illumina, logCPM > 0) is shown for reference.

```{r tx length setup}
tx_length_lookup <- dte_lists[["Illumina"]][["AvsL.DE"]]$genes %>%
  tibble::rownames_to_column("transcript_id") %>%
  select(transcript_id, Length)

# Universal DE: transcripts DE (FDR < 0.05) in ALL four platforms
universal_de <- purrr::map(platforms, function(plt) {
  tbl <- dte_lists[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH"))
  tbl$transcript_id[tbl$FDR < 0.05]
}) %>%
  purrr::reduce(intersect)

# Background: all expressed transcripts (logCPM > 0 in Illumina)
bg_ids <- dte_lists[["Illumina"]][["AvsL.DE"]]$table %>%
  tibble::rownames_to_column("transcript_id") %>%
  filter(logCPM > 0) %>%
  pull(transcript_id)

bg_lengths <- tx_length_lookup %>%
  filter(transcript_id %in% bg_ids) %>%
  mutate(platform = "Background (expressed)")

# Non-universal DE: DE in this platform but not in all 4
excl_lengths <- purrr::map_dfr(platforms, function(plt) {
  tbl <- dte_lists[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH"))
  de_ids <- tbl$transcript_id[tbl$FDR < 0.05]
  tibble(transcript_id = setdiff(de_ids, universal_de), platform = plt)
}) %>%
  left_join(tx_length_lookup, by = "transcript_id") %>%
  filter(!is.na(Length))
```

```{r tx length violin, fig.height=5, fig.width=10}
plt_order <- c("Background (expressed)", platforms)
bg_plt_colors <- c("Background (expressed)" = "gray70", color_palette)

plot_df <- bind_rows(bg_lengths, excl_lengths) %>%
  mutate(platform = factor(platform, levels = plt_order))

medians <- plot_df %>%
  group_by(platform) %>%
  summarise(med = median(Length, na.rm = TRUE), .groups = "drop")

ns <- plot_df %>% count(platform)

ggplot(plot_df, aes(x = platform, y = Length, fill = platform)) +
  geom_violin(trim = TRUE, alpha = 0.7, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
  geom_text(data = medians, aes(x = platform, y = med, label = round(med)),
            vjust = -0.5, size = 2.8, color = "black", inherit.aes = FALSE) +
  geom_text(data = ns, aes(x = platform, y = 0.9, label = paste0("n=", n)),
            vjust = 1, size = 2.5, color = "gray40", inherit.aes = FALSE) +
  scale_fill_manual(values = bg_plt_colors) +
  scale_y_log10(labels = scales::comma) +
  labs(
    title = "Transcript length — non-universal DE transcripts (AvsL)",
    x     = NULL,
    y     = "Transcript length (bp, log10 scale)"
  ) +
  theme_minimal(base_size = 11) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 20, hjust = 1))
```

```{r tx length density, fig.height=4, fig.width=9}
ggplot(excl_lengths, aes(x = Length, color = platform, fill = platform)) +
  geom_density(alpha = 0.15, linewidth = 0.8) +
  geom_density(data = bg_lengths, aes(x = Length),
               color = "gray50", fill = NA, linetype = "dashed",
               linewidth = 0.7, inherit.aes = FALSE) +
  scale_x_log10(labels = scales::comma) +
  scale_color_manual(values = color_palette) +
  scale_fill_manual(values  = color_palette) +
  annotate("text", x = Inf, y = Inf, label = "dashed = background",
           hjust = 1.1, vjust = 1.5, size = 3, color = "gray50") +
  labs(
    title = "Transcript length density — non-universal DE vs background",
    x     = "Transcript length (bp, log10 scale)",
    y     = "Density",
    color = "Platform", fill = "Platform"
  ) +
  theme_minimal(base_size = 11)
```

# Transcript length, expression level, and biotype by DE category

We classify each expressed transcript (logCPM > 0) per platform into **Universal DE** (FDR < 0.05 in all four platforms), **Platform-enriched DE** (FDR < 0.05 here but not all four), and **Not DE**. We then examine transcript length, expression level (logCPM), and biotype across these classes to understand what drives platform-specific DE calls.

```{r per platform class setup}
# Biotype lookup from full_id (field 8 of pipe-separated annotation)
bio_lookup <- dte_lists[["Illumina"]][["AvsL.DE"]]$genes %>%
  tibble::rownames_to_column("transcript_id") %>%
  mutate(biotype = sapply(strsplit(full_id, "\\|"), `[`, 8),
         biotype_simple = case_when(
           biotype == "protein_coding" ~ "protein_coding",
           biotype == "lncRNA"         ~ "lncRNA",
           grepl("pseudogene", biotype) ~ "pseudogene",
           TRUE                        ~ "other_ncRNA"
         )) %>%
  select(transcript_id, Length, biotype_simple)

class_colors <- c(
  "Not DE"               = "gray75",
  "Platform-enriched DE" = "#FC8D59",
  "Universal DE"         = "#91BFDB"
)

per_platform_df <- purrr::map_dfr(platforms, function(plt) {
  dte_lists[[plt]][["AvsL.DE"]]$table %>%
    tibble::rownames_to_column("transcript_id") %>%
    mutate(FDR = p.adjust(PValue, method = "BH")) %>%
    filter(logCPM > 0) %>%
    left_join(bio_lookup, by = "transcript_id") %>%
    mutate(
      platform = plt,
      de_class = case_when(
        transcript_id %in% universal_de ~ "Universal DE",
        FDR < 0.05                      ~ "Platform-enriched DE",
        TRUE                            ~ "Not DE"
      ),
      de_class = factor(de_class, levels = c("Not DE", "Platform-enriched DE", "Universal DE"))
    )
})
```

## Transcript length

No strong length bias for PacBio. ONT cDNA platform-enriched DE transcripts trend shorter (median ~1730 bp vs ~2137 bp for Not DE), which is linked to their lncRNA enrichment (see biotype section below).

```{r length capture bias violin, fig.height=5, fig.width=13}
medians2 <- per_platform_df %>%
  filter(!is.na(Length)) %>%
  group_by(platform, de_class) %>%
  summarise(med = median(Length, na.rm = TRUE), .groups = "drop")

ns2 <- per_platform_df %>% filter(!is.na(Length)) %>% count(platform, de_class)

ggplot(per_platform_df %>% filter(!is.na(Length)),
       aes(x = de_class, y = Length, fill = de_class)) +
  geom_violin(trim = TRUE, alpha = 0.75, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
  geom_text(data = medians2, aes(x = de_class, y = med, label = round(med)),
            vjust = -0.5, size = 2.3, color = "black", inherit.aes = FALSE) +
  geom_text(data = ns2, aes(x = de_class, y = 0.9, label = paste0("n=", n)),
            vjust = 1, size = 2.2, color = "gray40", inherit.aes = FALSE) +
  scale_fill_manual(values = class_colors) +
  scale_y_log10(labels = scales::comma) +
  facet_wrap(~platform, nrow = 1) +
  labs(title = "Transcript length by DE category (AvsL, expressed transcripts)",
       x = NULL, y = "Transcript length (bp, log10 scale)") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 25, hjust = 1))
```

## Expression level (logCPM)

Platform-enriched DE transcripts have **lower logCPM** than both Universal DE and Not DE across all platforms. These are lowly-expressed transcripts sitting at the detection margin — a platform misses them not because they are absent, but because its count depth for that transcript is insufficient for statistical power.

```{r logcpm by class, fig.height=5, fig.width=13}
med_cpm <- per_platform_df %>%
  group_by(platform, de_class) %>%
  summarise(med = median(logCPM, na.rm = TRUE), .groups = "drop")

ns_cpm <- per_platform_df %>% count(platform, de_class)

ggplot(per_platform_df, aes(x = de_class, y = logCPM, fill = de_class)) +
  geom_violin(trim = TRUE, alpha = 0.75, scale = "width") +
  geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
  geom_text(data = med_cpm, aes(x = de_class, y = med, label = round(med, 1)),
            vjust = -0.5, size = 2.3, color = "black", inherit.aes = FALSE) +
  scale_fill_manual(values = class_colors) +
  facet_wrap(~platform, nrow = 1) +
  labs(title = "logCPM by DE category (AvsL, expressed transcripts)",
       x = NULL, y = "logCPM") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none", axis.text.x = element_text(angle = 25, hjust = 1))
```

## Biotype composition

**lncRNA enrichment is a universal property of platform-enriched DE calls, not specific to any one platform.** In both FLAMES and IsoQuant, every platform shows higher lncRNA fraction in platform-enriched DE than in its Not DE background (FLAMES: Illumina 11→16%, ONT cDNA 11→23%, PacBio 11→16%, ONT dRNA 10→11%; IsoQuant: all platforms ~6→10%). The interpretation is that lncRNAs are inherently harder to detect consistently across platforms — lower expression, more variable structure — so they tend to accumulate in platform-specific rather than universal DE calls. Conversely, `other_ncRNA` (retained intron, NMD transcripts) is **depleted** in platform-enriched DE: these transcripts are expressed but not differentially expressed, and the depletion holds across all platforms. Universal DE is strongly protein-coding biased, as expected for robustly detectable, highly expressed transcripts.

```{r biotype composition, fig.height=5, fig.width=13}
biotype_pct <- per_platform_df %>%
  count(platform, de_class, biotype_simple) %>%
  group_by(platform, de_class) %>%
  mutate(pct = 100 * n / sum(n)) %>%
  ungroup() %>%
  mutate(biotype_simple = factor(biotype_simple,
                                 levels = c("protein_coding", "lncRNA",
                                            "other_ncRNA", "pseudogene")))

bio_colors <- c(
  protein_coding = "#4E79A7",
  lncRNA         = "#F28E2B",
  other_ncRNA    = "#76B7B2",
  pseudogene     = "#B07AA1"
)

ggplot(biotype_pct, aes(x = de_class, y = pct, fill = biotype_simple)) +
  geom_bar(stat = "identity", width = 0.7) +
  scale_fill_manual(values = bio_colors, name = "Biotype") +
  facet_wrap(~platform, nrow = 1) +
  labs(title = "Biotype composition by DE category (AvsL, expressed transcripts)",
       x = NULL, y = "% of transcripts") +
  theme_minimal(base_size = 10) +
  theme(axis.text.x = element_text(angle = 25, hjust = 1))
```

# LogFC consistency for expressed-but-missed transcripts (Q1)

For transcripts classified as Q1 (logCPM > 0 in the other platform but not DE), check whether the logFC direction is consistent across platforms. Consistent logFC with non-significant FDR in the other platform would suggest an underpowered call rather than a true false positive.

```{r q1 logfc consistency, fig.height=5, fig.width=14, results='asis'}
for (specific_plt in platforms) {
  cat("\n\n##", specific_plt, "-specific (Q1 only)\n\n")

  other_plts <- setdiff(platforms, specific_plt)

  plotlist <- purrr::map(other_plts, function(other_plt) {
    all_data <- purrr::map_dfr(comparisons, function(comp) {
      excl_ids <- get_exclusive_tx_ids(comp, specific_plt)
      if (length(excl_ids) == 0) return(NULL)

      spec_tbl <- dte_lists[[specific_plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("transcript_id") %>%
        filter(transcript_id %in% excl_ids) %>%
        select(transcript_id, logCPM_spec = logCPM, logFC_spec = logFC)

      other_tbl <- dte_lists[[other_plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("transcript_id") %>%
        mutate(FDR_other = p.adjust(PValue, method = "BH")) %>%
        filter(transcript_id %in% excl_ids) %>%
        select(transcript_id, logCPM_other = logCPM, logFC_other = logFC, FDR_other)

      inner_join(spec_tbl, other_tbl, by = "transcript_id") %>%
        filter(logCPM_other > 0) %>%   # Q1 only
        mutate(comparison = comp,
               same_direction = sign(logFC_spec) == sign(logFC_other))
    })

    if (nrow(all_data) == 0) return(NULL)

    pct_same <- round(100 * mean(all_data$same_direction, na.rm = TRUE), 1)

    ggplot(all_data, aes(x = logFC_spec, y = logFC_other, color = same_direction)) +
      geom_hline(yintercept = 0, linetype = "dashed", color = "gray70") +
      geom_vline(xintercept = 0, linetype = "dashed", color = "gray70") +
      geom_point(alpha = 0.6, size = 0.8) +
      scale_color_manual(values = c("TRUE" = "#2166ac", "FALSE" = "#d6604d"),
                         name = "Same\ndirection") +
      geom_abline(slope = 1, intercept = 0, linetype = "dotted", color = "black") +
      facet_wrap(~comparison, ncol = 3) +
      labs(
        title    = paste0(specific_plt, "-specific Q1 | other: ", other_plt),
        subtitle = paste0(pct_same, "% same logFC direction"),
        x        = paste0("logFC (", specific_plt, ")"),
        y        = paste0("logFC (", other_plt, ")")
      ) +
      theme_minimal(base_size = 8)
  }) %>% purrr::compact()

  if (length(plotlist) > 0) {
    p <- ggarrange(plotlist = plotlist, ncol = 3, nrow = 1)
    print(p)
  }
}
```
