---
title: "Cross-platform logFC correlation across rarefaction depths"
author: "Yupei"
date: '`r Sys.Date()`'
version: 'v1.0'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
params:
  cache_dir: NULL
  fig.path: NULL
  bulk_de_rds:    "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds"
  checkpoint_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE_rarefaction.rds"
---

```{r setup, include=FALSE}
if (!exists("params")) params <- list(
  cache_dir = NULL,
  fig.path = NULL,
  bulk_de_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds",
  checkpoint_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE_rarefaction.rds"
)

knitr::opts_chunk$set(
  echo = TRUE, fig.height = 5, fig.width = 12,
  warning = FALSE, message = FALSE, cache = FALSE
)
if (!is.null(params$cache_dir))
  knitr::opts_chunk$set(cache.path = params$cache_dir)
if (!is.null(params$fig.path))
  knitr::opts_chunk$set(fig.path = params$fig.path, dev = c("png", "svg"))
```

```{r libraries, include=FALSE}
library(edgeR)
library(tidyverse)
select    <- dplyr::select
filter    <- dplyr::filter
intersect <- base::intersect
```

# Load data

```{r load}
platforms   <- c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA")
comparisons <- c("AvsL", "PvsL", "AvsP")
depths_show <- c("1M", "5M", "10M", "15M", "20M", "full")
depth_order <- factor(depths_show, levels = depths_show)

# Checkpoint from the DE loop (bulk_DE_rarefaction.rds)
de_results <- readRDS(params$checkpoint_rds)
depths_done <- intersect(depths_show, names(de_results))

# Reference feature sets from full-depth DE analysis (bulk_DE.rds)
full_de <- readRDS(params$bulk_de_rds)
name_map <- c(ill = "Illumina", ont = "ONT cDNA", pb = "PacBio", drna = "ONT dRNA")
full_dge_lists <- full_de$dge_lists
full_dte_lists <- full_de$dte_lists
names(full_dge_lists) <- name_map[names(full_dge_lists)]
names(full_dte_lists) <- name_map[names(full_dte_lists)]

ref_gene_ids <- lapply(full_dge_lists, function(plat) {
  lapply(setNames(comparisons, comparisons), function(cmp)
    rownames(plat[[paste0(cmp, ".DE")]]$table))
})

ref_tx_ids <- lapply(full_dte_lists, function(plat) {
  lapply(setNames(comparisons, comparisons), function(cmp) {
    ids <- rownames(plat[[paste0(cmp, ".DE")]]$table)
    sub("\\..*", "", sub("\\|.*", "", ids))   # strip pipe then version
  })
})
```

# Checkpoint summary

```{r checkpoint-summary}
feature_counts <- bind_rows(lapply(depths_done, function(depth) {
  bind_rows(lapply(c(gene = "dge", transcript = "dte"), function(slot) {
    bind_rows(lapply(platforms, function(pl) {
      bind_rows(lapply(comparisons, function(cmp) {
        tbl <- tryCatch(
          de_results[[depth]][[slot]][[pl]][[paste0(cmp, ".DE")]]$table,
          error = function(e) NULL)
        tibble(
          depth      = depth,
          platform   = pl,
          contrast   = cmp,
          n_features = if (is.null(tbl)) NA_integer_ else nrow(tbl)
        )
      }))
    }))
  }), .id = "level")
})) %>%
  mutate(depth = factor(depth, levels = depths_show))

feature_counts %>%
  group_by(depth, level, platform) %>%
  summarise(n_features = sum(n_features, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = level, values_from = n_features) %>%
  arrange(depth, platform) %>%
  knitr::kable(caption = "Features tested per depth and platform (summed across contrasts)")
```

# Build logFC tables

```{r logfc-tables}
get_logfc <- function(depth, pl, cmp, level = "gene") {
  tbl <- tryCatch(
    de_results[[depth]][[if (level == "gene") "dge" else "dte"]][[pl]][[paste0(cmp, ".DE")]]$table,
    error = function(e) NULL)
  if (is.null(tbl)) return(NULL)
  id_col <- if (level == "gene") "gene_id" else "tx_id"
  tbl <- tbl %>% rownames_to_column(id_col) %>% select(all_of(c(id_col, "logFC")))
  if (level == "transcript")
    tbl <- tbl %>% mutate(tx_id = sub("\\..*", "", sub("\\|.*", "", tx_id)))
  tbl
}

platform_pairs <- combn(platforms, 2, simplify = FALSE)
pair_labels    <- sapply(platform_pairs, paste, collapse = " vs ")
```

# Plot C: Cross-platform logFC correlation vs depth

Pearson r on a stable feature set: the intersection of features testable in both platforms at every depth (per platform pair × contrast curve), further restricted to features used in the full-depth analysis (`bulk_DE.rds`). This ensures `n` is constant across the depth axis for each curve.

```{r compute-cor}
# Intersection of reference features across ALL platforms for each level/contrast
# All tested features at full depth
common_gene_ids <- setNames(lapply(comparisons, function(cmp)
  Reduce(intersect, lapply(platforms, function(pl) ref_gene_ids[[pl]][[cmp]]))),
  comparisons)

common_tx_ids <- setNames(lapply(comparisons, function(cmp)
  Reduce(intersect, lapply(platforms, function(pl) ref_tx_ids[[pl]][[cmp]]))),
  comparisons)

# P < 0.05 significant features at full depth (union across platforms per contrast)
ref_gene_ids_sig <- lapply(full_dge_lists, function(plat) {
  lapply(setNames(comparisons, comparisons), function(cmp) {
    tbl <- plat[[paste0(cmp, ".DE")]]$table
    rownames(tbl)[tbl$PValue < 0.05]
  })
})
ref_tx_ids_sig <- lapply(full_dte_lists, function(plat) {
  lapply(setNames(comparisons, comparisons), function(cmp) {
    tbl <- plat[[paste0(cmp, ".DE")]]$table
    ids <- rownames(tbl)[tbl$PValue < 0.05]
    sub("\\..*", "", sub("\\|.*", "", ids))
  })
})
common_gene_ids_sig <- setNames(lapply(comparisons, function(cmp)
  Reduce(intersect, lapply(platforms, function(pl) ref_gene_ids_sig[[pl]][[cmp]]))),
  comparisons)
common_tx_ids_sig <- setNames(lapply(comparisons, function(cmp)
  Reduce(intersect, lapply(platforms, function(pl) ref_tx_ids_sig[[pl]][[cmp]]))),
  comparisons)

message(sprintf("All-feature intersection: gene %s | transcript %s",
  paste(sapply(common_gene_ids,     length), collapse = "/"),
  paste(sapply(common_tx_ids,       length), collapse = "/")))
message(sprintf("Sig-feature intersection (P<0.05 full depth): gene %s | transcript %s",
  paste(sapply(common_gene_ids_sig, length), collapse = "/"),
  paste(sapply(common_tx_ids_sig,   length), collapse = "/")))

compute_cor <- function(level, depths = depths_done,
                        common_ids_gene = common_gene_ids,
                        common_ids_tx   = common_tx_ids) {
  common_ids <- if (level == "gene") common_ids_gene else common_ids_tx
  id_col     <- if (level == "gene") "gene_id"       else "tx_id"

  bind_rows(lapply(seq_along(platform_pairs), function(pi) {
    p1 <- platform_pairs[[pi]][1]
    p2 <- platform_pairs[[pi]][2]
    bind_rows(lapply(comparisons, function(cmp) {
      per_depth_ids <- lapply(depths, function(depth) {
        d1 <- get_logfc(depth, p1, cmp, level)
        d2 <- get_logfc(depth, p2, cmp, level)
        if (is.null(d1) || is.null(d2)) return(character(0))
        intersect(d1[[id_col]], d2[[id_col]])
      })
      ref <- Reduce(intersect, c(per_depth_ids, list(common_ids[[cmp]])))

      bind_rows(lapply(depths, function(depth) {
        d1 <- get_logfc(depth, p1, cmp, level)
        d2 <- get_logfc(depth, p2, cmp, level)
        if (is.null(d1) || is.null(d2) || length(ref) == 0)
          return(tibble(depth = depth, pair = pair_labels[[pi]], contrast = cmp, r = NA_real_, n = 0L))
        joined <- inner_join(
          d1 %>% filter(.data[[id_col]] %in% ref),
          d2 %>% filter(.data[[id_col]] %in% ref),
          by = id_col, suffix = c("_p1", "_p2")) %>%
          drop_na(logFC_p1, logFC_p2)
        r <- if (nrow(joined) >= 5) cor(joined$logFC_p1, joined$logFC_p2) else NA_real_
        tibble(depth = depth, pair = pair_labels[[pi]], contrast = cmp,
               r = r, n = nrow(joined))
      }))
    }))
  }))
}

depths_10M <- intersect(c("10M", "15M", "20M", "full"), depths_done)

# All features (no significance filter)
logfc_cor         <- compute_cor("gene")
logfc_cor_tx      <- compute_cor("transcript")
logfc_cor_10M     <- compute_cor("gene",       depths = depths_10M)
logfc_cor_tx_10M  <- compute_cor("transcript", depths = depths_10M)

# P < 0.05 significant features at full depth
logfc_cor_sig        <- compute_cor("gene",       common_ids_gene = common_gene_ids_sig, common_ids_tx = common_tx_ids_sig)
logfc_cor_tx_sig     <- compute_cor("transcript", common_ids_gene = common_gene_ids_sig, common_ids_tx = common_tx_ids_sig)
logfc_cor_10M_sig    <- compute_cor("gene",       depths = depths_10M, common_ids_gene = common_gene_ids_sig, common_ids_tx = common_tx_ids_sig)
logfc_cor_tx_10M_sig <- compute_cor("transcript", depths = depths_10M, common_ids_gene = common_gene_ids_sig, common_ids_tx = common_tx_ids_sig)
```

```{r plot-cor, fig.height=5, fig.width=12}
contrast_labels <- c(AvsL = "SCLC-A vs LUAD", PvsL = "SCLC-P vs LUAD", AvsP = "SCLC-A vs SCLC-P")

make_cor_plot <- function(cor_gene, cor_tx, title_suffix, subtitle_suffix) {
  bind_rows(
    cor_gene %>% mutate(level = "Gene"),
    cor_tx   %>% mutate(level = "Transcript")
  ) %>%
    mutate(depth = factor(depth, levels = depths_show),
           level = factor(level, levels = c("Gene", "Transcript"))) %>%
    ggplot(aes(x = depth, y = r, color = pair, linetype = level,
               group = interaction(pair, level))) +
    geom_line(linewidth = 0.7, na.rm = TRUE) +
    geom_point(size = 1.8, na.rm = TRUE) +
    geom_text(aes(label = n), size = 2.2, vjust = -0.8, show.legend = FALSE, na.rm = TRUE) +
    facet_wrap(~contrast, ncol = 3, labeller = labeller(contrast = contrast_labels)) +
    coord_cartesian(ylim = c(0.7, 1)) +
    labs(title = paste("Cross-platform logFC correlation vs sequencing depth", title_suffix),
         subtitle = subtitle_suffix,
         x = "Depth", y = "Pearson r (logFC)",
         color = "Platform pair", linetype = "Level") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          legend.position = "bottom")
}

sub_all <- "Fixed feature set: testable at all depths in both platforms (per curve)"
sub_10M <- "Fixed feature set: testable at 10M+ in both platforms (per curve)"
sub_sig_all <- "Fixed feature set: P<0.05 at full depth, testable at all depths (per curve)"
sub_sig_10M <- "Fixed feature set: P<0.05 at full depth, testable at 10M+ (per curve)"

make_cor_plot(logfc_cor,        logfc_cor_tx,        "(all depths, all features)",  sub_all)
make_cor_plot(logfc_cor_10M,    logfc_cor_tx_10M,    "(10M+, all features)",        sub_10M)
make_cor_plot(logfc_cor_sig,    logfc_cor_tx_sig,    "(all depths, P<0.05)",        sub_sig_all)
make_cor_plot(logfc_cor_10M_sig,logfc_cor_tx_10M_sig,"(10M+, P<0.05)",             sub_sig_10M)
```
