---
title: "DE Benchmarking: Testing Power, Mean Overlap, and Transcript Inconsistency (Oarfish full dataset)"
author: "Yupei"
date: '`r Sys.Date()`'
version: 'v2.0'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
  sirv_ercc_gtf: '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_ERCC_longSIRV_multi-fasta_20210507.gtf'
  sequins_gtf: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_annotation_2.4.gtf'
  sequins_tsv: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv'
  human_gtf: '/vast/projects/LongBench/reference_files/GRCh38/gencode.v44.annotation.gtf'
  sirv_csv: '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv'
  input_DGEobject_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"
  output_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Bulk_DE_benchmarking.rds"
  Tx2Gene.map.rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Tx2Gene.map.rds"
  log2FC_cutoff: 0
---

# Overview

This report benchmarks differential expression (DE) analysis across four sequencing platforms
(Illumina, PacBio, ONT cDNA, ONT dRNA) with four analysis modules:

1. **Shared-feature DE** — Shared `filterByExpr` features across all platforms; upset plots for
   gene-level (DGE) and transcript-level (DTE) DE.
2. **Mean Overlap metric** — A single cross-platform consistency score for gene-level vs.
   transcript-level DE.
3. **Sources of transcript inconsistency** — Transcript length bias, isoform-count effects, and
   per-platform DE rate by length category.
4. **Platform-independent filtering** — Each platform runs `filterByExpr` on its own data;
   Mean Overlap compared to the shared-feature result.

---

# Setup

```{r setup, include=FALSE}
if (interactive()) {
  params <- list(
    random_seed= 2024,
    cache_dir= NULL,
    fig.path= NULL,
    sirv_ercc_gtf= '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_ERCC_longSIRV_multi-fasta_20210507.gtf',
    sequins_gtf= '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_annotation_2.4.gtf',
    sequins_tsv= '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv',
    human_gtf= '/vast/projects/LongBench/reference_files/GRCh38/gencode.v44.annotation.gtf',
    sirv_csv= '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv',
    input_DGEobject_rds= "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
    output_rds= "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Bulk_DE_benchmarking.rds",
    Tx2Gene.map.rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Tx2Gene.map.rds",
    log2FC_cutoff= 0
  )
}

knitr::opts_chunk$set(
  echo    = FALSE,
  warning = FALSE,
  message = FALSE,
  fig.width  = 14,
  fig.height = 9,
  cache  = !is.null(params$cache_dir),
  fig.path = params$fig.path,
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL
)
if (!is.null(params$cache_dir))
  knitr::opts_chunk$set(cache.path = params$cache_dir)
set.seed(params$random_seed)
```

```{r libraries, include=FALSE}
library(limma)
library(edgeR)
library(tximport)
library(tidyverse)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(glue)
library(gridExtra)
library(patchwork)
library(viridis)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ComplexUpset)
source("/vast/projects/LongBench/analysis/workflow/scripts/Rfunctions.R")

rename <- dplyr::rename
select <- dplyr::select
```

```{r color-codes, include=FALSE}
color_palette <- c(
  PacBio     = "#df1995",
  "ONT cDNA" = "#04476c",
  "ONT dRNA" = "#00789b",
  Illumina   = "#e88b20"
)
```

```{r is-helpers, include=FALSE}
is.sequins      <- function(dge) grepl("^R",          rownames(dge))
is.human        <- function(dge) grepl("^ENS",        rownames(dge))
is.sirv_or_ercc <- function(dge) !is.human(dge) & !is.sequins(dge)
is.sirv_E0      <- function(dge) grepl("^SIRV\\d{3}$",rownames(dge))
is.sirv_all     <- function(dge) grepl("^SIRV",        rownames(dge))
is.long_sirv    <- function(dge) grepl("^SIRV\\d{4}", rownames(dge))
is.ercc         <- function(dge) !is.human(dge) & !is.sequins(dge) & !is.sirv_all(dge)
```

## GTF annotation maps

```{r gtf-maps, message=FALSE, warning=FALSE, results='hide'}
list2env(readRDS(params$Tx2Gene.map.rds), .GlobalEnv)
```

## Load transcript-level DGE objects

```{r load-tx-dge, message=FALSE, warning=FALSE, results='hide'}
bulk_obj <- readRDS(params$input_DGEobject_rds)

ill_bulk.dge         <- bulk_obj[["ill_bulk.tx.dge"]]
ont_bulk.dge.oarfish <- bulk_obj[["ont_bulk.tx.dge"]]
pb_bulk.dge.oarfish  <- bulk_obj[["pb_bulk.tx.dge"]]
ont_drna.dge.oarfish <- bulk_obj[["drna_bulk.tx.dge"]]
```

## Aggregate to gene level (tximport)

```{r txi-gene, message=FALSE, warning=FALSE, results='hide'}
ill_bulk.gene.dge  <- bulk_obj[["ill_bulk.gene.dge"]]
ont_bulk.gene.dge  <- bulk_obj[["ont_bulk.gene.dge"]]
pb_bulk.gene.dge   <- bulk_obj[["pb_bulk.gene.dge"]]
drna_bulk.gene.dge <- bulk_obj[["drna_bulk.gene.dge"]]
```

## Shared helper functions

```{r helpers}
# DE/DTE edgeR wrappers -------------------------------------------------------
DGE.edgeR_human <- function(dge, norm.method = "TMM") {
  if (!is.null(norm.method)) dge <- calcNormFactors(dge, method = norm.method)
  design <- model.matrix(~0 + group, data = dge$samples)
  dge    <- estimateDisp(dge, design)
  qlfit  <- glmQLFit(dge, design)
  contrasts <- makeContrasts(
    AvsL = groupsclc_a - groupluad,
    PvsL = groupsclc_p - groupluad,
    AvsP = groupsclc_a - groupsclc_p,
    levels = design
  )
  list(
    AvsL.DE = glmQLFTest(qlfit, contrast = contrasts[,"AvsL"]),
    PvsL.DE = glmQLFTest(qlfit, contrast = contrasts[,"PvsL"]),
    AvsP.DE = glmQLFTest(qlfit, contrast = contrasts[,"AvsP"])
  )
}

DTE.edgeR_human <- function(dge, norm.method = "TMM") {
  if (!is.null(norm.method)) dge <- calcNormFactors(dge, method = norm.method)
  design <- model.matrix(~0 + group, data = dge$samples)
  dge    <- estimateDisp(dge, design)
  qlfit  <- glmQLFit(dge, design)
  contrasts <- makeContrasts(
    AvsL = groupsclc_a - groupluad,
    PvsL = groupsclc_p - groupluad,
    AvsP = groupsclc_a - groupsclc_p,
    levels = design
  )
  list(
    AvsL.DE = glmQLFTest(qlfit, contrast = contrasts[,"AvsL"]),
    PvsL.DE = glmQLFTest(qlfit, contrast = contrasts[,"PvsL"]),
    AvsP.DE = glmQLFTest(qlfit, contrast = contrasts[,"AvsP"])
  )
}

# Upset plot helper -----------------------------------------------------------
complex_upset_plot <- function(lst) {
  DEGs <- lapply(lst, function(x) {
    x <- decideTests(x, lfc = params$log2FC_cutoff)
    rownames(x)[x != 0]
  })
  long_df <- stack(DEGs)
  binary_matrix <- long_df %>%
    mutate(present = 1) %>%
    pivot_wider(names_from = ind, values_from = present, values_fill = 0)
  p <- ComplexUpset::upset(
    binary_matrix %>% select(-values) %>% as.data.frame(),
    intersect = names(lst),
    base_annotations = list(
      'Intersection size' = ComplexUpset::intersection_size(
        counts = TRUE,
        text   = list(element_text(size = 3, vjust = -0.2))
      )
    ),
    wrap = TRUE
  )
  list(plot = p, matrix = binary_matrix %>% rename(Gene = values))
}

# Mean Overlap (the primary consistency metric) --------------------------------
# For a given binary membership matrix, compute the proportion of each platform's
# DE features that are also detected by >=2 of the 3 remaining platforms,
# then average across all 4 platforms.
calc_mean_overlap <- function(mat) {
  platforms <- setdiff(names(mat), "Gene")
  per_platform <- sapply(platforms, function(p) {
    rows_detected <- mat[mat[[p]] == 1, , drop = FALSE]
    if (nrow(rows_detected) == 0) return(NA_real_)
    others <- setdiff(platforms, p)
    n_also_in_2plus <- sum(rowSums(rows_detected[, others, drop = FALSE]) >= 2)
    n_also_in_2plus / nrow(rows_detected)
  })
  mean(per_platform, na.rm = TRUE)
}

# Pairwise Jaccard similarity --------------------------------------------------
mean_pairwise_jaccard <- function(mat) {
  platforms <- setdiff(names(mat), "Gene")
  pairs <- combn(platforms, 2, simplify = FALSE)
  jac   <- sapply(pairs, function(p) {
    a <- mat[[p[1]]]; b <- mat[[p[2]]]
    sum(a & b) / sum(a | b)
  })
  names(jac) <- sapply(pairs, function(p) paste(p, collapse = " vs "))
  jac
}
```

---

# Part 1: Shared-feature DE analysis {.tabset}

Features are first restricted to those present in all four platforms
(`Reduce(intersect, ...)`), then filtered with `filterByExpr` applied jointly
(AND across platforms). This ensures every tested feature is potentially
detectable by all platforms.

## Gene level

```{r shared-gene-filter, results='hide'}
common_genes <- Reduce(intersect,
  lapply(list(ill_bulk.gene.dge, ont_bulk.gene.dge,
              pb_bulk.gene.dge, drna_bulk.gene.dge), rownames))

ill_bulk.gene.dge  <- ill_bulk.gene.dge[common_genes, ]
ont_bulk.gene.dge  <- ont_bulk.gene.dge[common_genes, ]
pb_bulk.gene.dge   <- pb_bulk.gene.dge[common_genes, ]
drna_bulk.gene.dge <- drna_bulk.gene.dge[common_genes, ]

keep_gene <- filterByExpr(ill_bulk.gene.dge,  group = ill_bulk.gene.dge$samples$group) &
             filterByExpr(ont_bulk.gene.dge,  group = ont_bulk.gene.dge$samples$group) &
             filterByExpr(pb_bulk.gene.dge,   group = pb_bulk.gene.dge$samples$group) &
             filterByExpr(drna_bulk.gene.dge, group = drna_bulk.gene.dge$samples$group)

ill_bulk.gene.dge  <- ill_bulk.gene.dge[keep_gene, ]
ont_bulk.gene.dge  <- ont_bulk.gene.dge[keep_gene, ]
pb_bulk.gene.dge   <- pb_bulk.gene.dge[keep_gene, ]
drna_bulk.gene.dge <- drna_bulk.gene.dge[keep_gene, ]

ill_bulk.dge_human.list  <- DGE.edgeR_human(ill_bulk.gene.dge[is.human(ill_bulk.gene.dge),, keep.lib.sizes=FALSE])
ont_bulk.dge_human.list  <- DGE.edgeR_human(ont_bulk.gene.dge[is.human(ont_bulk.gene.dge),, keep.lib.sizes=FALSE])
pb_bulk.dge_human.list   <- DGE.edgeR_human(pb_bulk.gene.dge[is.human(pb_bulk.gene.dge),,  keep.lib.sizes=FALSE])
drna_bulk.deg_human.list <- DGE.edgeR_human(drna_bulk.gene.dge[is.human(drna_bulk.gene.dge),, keep.lib.sizes=FALSE])

cat(sprintf("Shared gene features tested: %d\n", sum(keep_gene)))
```

```{r shared-gene-upset, fig.height=10, fig.width=8}
contrasts_dge <- c("AvsL","PvsL","AvsP")
mat_list_dge  <- setNames(vector("list", 3), contrasts_dge)

for (ct in contrasts_dge) {
  lst <- list(
    "Illumina" = ill_bulk.dge_human.list[[paste0(ct,".DE")]],
    "ONT cDNA" = ont_bulk.dge_human.list[[paste0(ct,".DE")]],
    "PacBio"   = pb_bulk.dge_human.list[[paste0(ct,".DE")]],
    "ONT dRNA" = drna_bulk.deg_human.list[[paste0(ct,".DE")]]
  )
  res <- complex_upset_plot(lst)
  mat_list_dge[[ct]] <- res$matrix
  assign(paste0("mat.", ct, ".DEG"), res$matrix)
}

p_list <- lapply(contrasts_dge, function(ct) {
  lst <- list(
    "Illumina" = ill_bulk.dge_human.list[[paste0(ct,".DE")]],
    "ONT cDNA" = ont_bulk.dge_human.list[[paste0(ct,".DE")]],
    "PacBio"   = pb_bulk.dge_human.list[[paste0(ct,".DE")]],
    "ONT dRNA" = drna_bulk.deg_human.list[[paste0(ct,".DE")]]
  )
  complex_upset_plot(lst)$plot + ggtitle(ct) + theme(plot.title = element_text(size = 10))
})
(p_list[[1]] / p_list[[2]] / p_list[[3]]) +
  plot_annotation(title = "DGE Human — shared features",
                  theme = theme(plot.title = element_text(size = 13)))
```

## Transcript level

```{r shared-tx-filter, results='hide'}
common_tx <- Reduce(intersect,
  lapply(list(ill_bulk.dge, ont_bulk.dge.oarfish,
              pb_bulk.dge.oarfish, ont_drna.dge.oarfish), rownames))

ill_bulk.dge         <- ill_bulk.dge[common_tx, ]
ont_bulk.dge.oarfish <- ont_bulk.dge.oarfish[common_tx, ]
pb_bulk.dge.oarfish  <- pb_bulk.dge.oarfish[common_tx, ]
ont_drna.dge.oarfish <- ont_drna.dge.oarfish[common_tx, ]

keep_tx <- filterByExpr(ill_bulk.dge,         group = ill_bulk.dge$samples$group,         min.count = 5) &
           filterByExpr(ont_bulk.dge.oarfish,  group = ont_bulk.dge.oarfish$samples$group, min.count = 5) &
           filterByExpr(pb_bulk.dge.oarfish,   group = pb_bulk.dge.oarfish$samples$group,  min.count = 5) &
           filterByExpr(ont_drna.dge.oarfish,  group = ont_drna.dge.oarfish$samples$group, min.count = 5)

ill_bulk.dge         <- ill_bulk.dge[keep_tx, ,keep.lib.sizes=FALSE]
ont_bulk.dge.oarfish <- ont_bulk.dge.oarfish[keep_tx, ,keep.lib.sizes=FALSE]
pb_bulk.dge.oarfish  <- pb_bulk.dge.oarfish[keep_tx, ,keep.lib.sizes=FALSE]
ont_drna.dge.oarfish <- ont_drna.dge.oarfish[keep_tx, ,keep.lib.sizes=FALSE]

# Per-platform average logCPM (TMM-normalized counts, before overdispersion scaling)
logcpm_lookup <- bind_rows(lapply(
  list(Illumina   = ill_bulk.dge,
       `ONT cDNA` = ont_bulk.dge.oarfish,
       PacBio     = pb_bulk.dge.oarfish,
       `ONT dRNA` = ont_drna.dge.oarfish),
  function(dge) tibble(
    tx_id = rownames(dge), 
    avg_logCPM = rowMeans(cpm(dge, log = TRUE)),
    avg_CPM = rowMeans(cpm(dge, log = FALSE))
  )
), .id = "Platform")



# Overdispersion scaling
ill_bulk.scale_dge  <- ill_bulk.dge
ont_bulk.scale_dge  <- ont_bulk.dge.oarfish
pb_bulk.scale_dge   <- pb_bulk.dge.oarfish
ont_drna.scale_dge  <- ont_drna.dge.oarfish

ill_bulk.scale_dge$counts  <- ill_bulk.scale_dge$counts  / ill_bulk.scale_dge$genes$Overdispersion
ont_bulk.scale_dge$counts  <- ont_bulk.scale_dge$counts  / ont_bulk.scale_dge$genes$Overdispersion
pb_bulk.scale_dge$counts   <- pb_bulk.scale_dge$counts   / pb_bulk.scale_dge$genes$Overdispersion
ont_drna.scale_dge$counts  <- ont_drna.scale_dge$counts  / ont_drna.scale_dge$genes$Overdispersion

ill_bulk.scl_dte_human.list  <- DTE.edgeR_human(ill_bulk.scale_dge[is.human(ill_bulk.scale_dge),, keep.lib.sizes=FALSE])
ont_bulk.scl_dte_human.list  <- DTE.edgeR_human(ont_bulk.scale_dge[is.human(ont_bulk.scale_dge),, keep.lib.sizes=FALSE])
pb_bulk.scl_dte_human.list   <- DTE.edgeR_human(pb_bulk.scale_dge[is.human(pb_bulk.scale_dge),,  keep.lib.sizes=FALSE])
drna_bulk.scl_dte_human.list <- DTE.edgeR_human(ont_drna.scale_dge[is.human(ont_drna.scale_dge),, keep.lib.sizes=FALSE])

cat(sprintf("Shared transcript features tested: %d\n", sum(keep_tx)))

# Library sizes and logCPM summary (before overdispersion scaling, after TMM)
lib_size_df <- bind_rows(lapply(
  list(Illumina   = ill_bulk.dge,
       `ONT cDNA` = ont_bulk.dge.oarfish,
       PacBio     = pb_bulk.dge.oarfish,
       `ONT dRNA` = ont_drna.dge.oarfish),
  function(dge) {
    tibble(
      lib_size_total  = sum(dge$samples$lib.size),
      lib_size_mean   = mean(dge$samples$lib.size),
      lib_size_median = median(dge$samples$lib.size)
    )
  }
), .id = "Platform")

cat("Library size summary, before overdispersion scaling:\n")
print(lib_size_df)
```

```{r tx-libsize-check, fig.height=4, fig.width=10}
# Average logCPM per platform across all shared-filter tested transcripts
logcpm_lookup %>%
  group_by(Platform) %>%
  summarise(
    n_transcripts    = n(),
    mean_CPM      = round(mean(avg_CPM,   na.rm = TRUE), 3),
    median_CPM    = round(median(avg_CPM, na.rm = TRUE), 3),
    total_equiv_cpm  = round(sum(avg_CPM,  na.rm = TRUE), 0),
    mean_count  = mean(avg_count,  na.rm = TRUE),
    .groups = "drop"
  ) %>%
  knitr::kable(caption = "CPM summary across shared-filter tested transcripts (TMM-normalised, before overdispersion scaling)",
               row.names = FALSE)
```

```{r tx-lcpm-dist, fig.height=4, fig.width=10}
logcpm_lookup %>%
  mutate(Platform = factor(Platform, levels = names(color_palette))) %>%
  ggplot(aes(x = avg_logCPM, colour = Platform)) +
  geom_density(alpha = 0.3, linewidth = 0.7) +
  scale_fill_manual(values = color_palette) +
  scale_colour_manual(values = color_palette) +
  labs(title = "logCPM distribution — shared-filter tested transcripts (all platforms)",
       x = "Average logCPM", y = "Density") +
  theme_minimal()
```

```{r tx-lcpm-cumulated desity, fig.height=4, fig.width=10}
logcpm_lookup %>%
  mutate(Platform = factor(Platform, levels = names(color_palette))) %>%
  ggplot(aes(x = avg_logCPM, colour = Platform)) +
  stat_ecdf(linewidth = 0.7) +
  scale_colour_manual(values = color_palette) +
  labs(
    title = "Cumulative distribution of logCPM — shared-filter tested transcripts (all platforms)",
    x = "Average logCPM",
    y = "Cumulative density"
  ) +
  theme_minimal()
```

```{r shared-tx-upset, fig.height=10, fig.width=8}
contrasts_dte <- c("AvsL","PvsL","AvsP")
mat_list_dte  <- setNames(vector("list", 3), contrasts_dte)

for (ct in contrasts_dte) {
  lst <- list(
    "Illumina" = ill_bulk.scl_dte_human.list[[paste0(ct,".DE")]],
    "ONT cDNA" = ont_bulk.scl_dte_human.list[[paste0(ct,".DE")]],
    "PacBio"   = pb_bulk.scl_dte_human.list[[paste0(ct,".DE")]],
    "ONT dRNA" = drna_bulk.scl_dte_human.list[[paste0(ct,".DE")]]
  )
  res <- complex_upset_plot(lst)
  mat_list_dte[[ct]] <- res$matrix
  assign(paste0("mat.", ct, ".DTE"), res$matrix)
}

p_list <- lapply(contrasts_dte, function(ct) {
  lst <- list(
    "Illumina" = ill_bulk.scl_dte_human.list[[paste0(ct,".DE")]],
    "ONT cDNA" = ont_bulk.scl_dte_human.list[[paste0(ct,".DE")]],
    "PacBio"   = pb_bulk.scl_dte_human.list[[paste0(ct,".DE")]],
    "ONT dRNA" = drna_bulk.scl_dte_human.list[[paste0(ct,".DE")]]
  )
  complex_upset_plot(lst)$plot + ggtitle(ct) + theme(plot.title = element_text(size = 10))
})
(p_list[[1]] / p_list[[2]] / p_list[[3]]) +
  plot_annotation(title = "DTE Human — shared features (overdispersion-scaled)",
                  theme = theme(plot.title = element_text(size = 13)))
```

---

# Part 2: Mean Overlap — gene vs transcript consistency

**Definition:** For any given platform, we calculate the proportion of its DE features also
identified by at least two of the three remaining platforms, averaging these proportions across
all four platforms. For DE analysis this metric is further averaged across the three pairwise
comparisons.

```{r mean-overlap-table}
mo_gene_per_contrast <- sapply(contrasts_dge, function(ct)
  calc_mean_overlap(mat_list_dge[[ct]]))
mo_tx_per_contrast <- sapply(contrasts_dte, function(ct)
  calc_mean_overlap(mat_list_dte[[ct]]))

mo_summary <- data.frame(
  Level    = c("Gene", "Transcript"),
  AvsL     = round(c(mo_gene_per_contrast["AvsL"], mo_tx_per_contrast["AvsL"]), 3),
  PvsL     = round(c(mo_gene_per_contrast["PvsL"], mo_tx_per_contrast["PvsL"]), 3),
  AvsP     = round(c(mo_gene_per_contrast["AvsP"], mo_tx_per_contrast["AvsP"]), 3),
  Overall  = round(c(mean(mo_gene_per_contrast), mean(mo_tx_per_contrast)), 3)
)

knitr::kable(mo_summary,
  caption = "Mean Overlap: proportion of each platform's DE features also found in ≥2 other platforms",
  row.names = FALSE)
```

```{r mean-overlap-barplot, fig.height=4, fig.width=7}
mo_long <- mo_summary %>%
  pivot_longer(-Level, names_to = "Contrast", values_to = "MeanOverlap") %>%
  mutate(Contrast = factor(Contrast, levels = c("AvsL","PvsL","AvsP","Overall")))

ggplot(mo_long, aes(x = Contrast, y = MeanOverlap, fill = Level)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.3f", MeanOverlap)),
            position = position_dodge(0.7), vjust = -0.4, size = 3) +
  scale_fill_manual(values = c(Gene = "#4e79a7", Transcript = "#f28e2b")) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  labs(title = "Mean Overlap: gene-level vs transcript-level DE",
       y = "Mean Overlap", x = "Contrast", fill = "Feature level") +
  theme_minimal()
```

### Per-platform breakdown of Mean Overlap (transcript level)

```{r per-platform-mo, fig.height=4, fig.width=10}
per_platform_mo <- function(mat, contrast_label) {
  platforms <- setdiff(names(mat), "Gene")
  tibble(
    Platform = platforms,
    MeanOverlap = sapply(platforms, function(p) {
      rows_detected <- mat[mat[[p]] == 1, , drop = FALSE]
      if (nrow(rows_detected) == 0) return(NA_real_)
      others <- setdiff(platforms, p)
      sum(rowSums(rows_detected[, others, drop = FALSE]) >= 2) / nrow(rows_detected)
    }),
    Contrast = contrast_label
  )
}

pp_mo <- bind_rows(lapply(contrasts_dte, function(ct)
  per_platform_mo(mat_list_dte[[ct]], ct)))

ggplot(pp_mo, aes(x = Platform, y = MeanOverlap, fill = Platform)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", MeanOverlap)), vjust = -0.4, size = 3) +
  facet_wrap(~ Contrast) +
  scale_fill_manual(values = color_palette) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  labs(title = "Per-platform Mean Overlap — DTE (shared features)",
       y = "Mean Overlap", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))
```

---

# Part 3: Sources of transcript-level inconsistency

We focus on the **AvsL** contrast (largest DE signal) unless noted, then confirm patterns across
all contrasts.

## 3a. Transcript annotation lookup

```{r tx-lookup}
tx_length_lookup <- ill_bulk.scale_dge$genes %>%
  rownames_to_column("tx_id") %>%
  select(tx_id, Length) %>%
  tibble() %>%
  mutate(
    biotype = strsplit(tx_id, "\\|") %>% sapply(function(x) x[8]),
    gene_id = strsplit(tx_id, "\\|") %>% sapply(function(x) x[2]),
    tx_id_clean = strsplit(tx_id, "\\|") %>% sapply(function(x) x[1])
  )

# Isoform count per gene (from GTF annotation)
iso_count_lookup <- human.G.Tx.map %>%
  select(tx_name, gene_id, tx_count) %>%
  mutate(tx_name_clean = sub("\\..*", "", tx_name))

# Expressed isoform count (shared-filter set): number of isoforms per gene that
# pass the shared filterByExpr criterion and are actually tested here.
expressed_iso_count_lookup <- tx_length_lookup %>%
  filter(grepl("^ENS", tx_id)) %>%
  group_by(gene_id) %>%
  mutate(expr_iso_count = n()) %>%
  ungroup() %>%
  mutate(tx_name_clean = sub("\\..*", "", tx_id_clean)) %>%
  select(tx_name_clean, expr_iso_count) %>%
  distinct()
```

## Length bias?
```{r length-bias}
# Get the total CPM distribution across all shared-filter tested transcripts
## 1 define length bins (e.g. 10 bins with equal number of transcripts)
length_bins <- quantile(tx_length_lookup$Length, probs = seq(0, 1, length.out = 11))


## 2 get total CPM per length bin
length_cpm_df <- logcpm_lookup %>%
  left_join(tx_length_lookup, by = c("tx_id" = "tx_id")) %>%
  mutate(length_bin = cut(Length, breaks = length_bins, include.lowest = TRUE)) %>%
  group_by(Platform, length_bin) %>%
  summarise(
    mean_CPM = mean(avg_CPM, na.rm = TRUE),
    mean_logCPM = mean(avg_logCPM, na.rm = TRUE),
    median_CPM = median(avg_CPM, na.rm = TRUE),
    .groups = "drop"
  )

## 3 plot boxplots of CPM per length bin per platform
p.mean_cpm <- length_cpm_df %>%
  ggplot(aes(x = length_bin, y = mean_CPM, fill = Platform)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = color_palette) +
  labs(
    title = "Mean CPM by transcript length bin (shared-filter tested transcripts)",
    x = "Transcript length bin", y = "Mean CPM"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.mean_logcpm <- length_cpm_df %>%
  ggplot(aes(x = length_bin, y = mean_logCPM, fill = Platform)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = color_palette) +
  labs(
    title = "Mean logCPM by transcript length bin (shared-filter tested transcripts)",
    x = "Transcript length bin", y = "Mean logCPM"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.median_cpm <- length_cpm_df %>%
  ggplot(aes(x = length_bin, y = median_CPM, fill = Platform)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = color_palette) +
  labs(
    title = "Median CPM by transcript length bin (shared-filter tested transcripts)",
    x = "Transcript length bin", y = "Median CPM"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


## 4 number of DE transcripts per length bin per platform

DE_number_df <- mat_list_dte %>%
  bind_rows(.id = "Contrast") %>%
  left_join(tx_length_lookup, by = c("Gene" = "tx_id")) %>%
  mutate(length_bin = cut(Length, breaks = length_bins, include.lowest = TRUE)) %>%
  pivot_longer(cols = c(Illumina, `ONT cDNA`, PacBio, `ONT dRNA`), names_to = "Platform", values_to = "DE") %>%
  group_by(Platform, length_bin, Contrast) %>%
  summarise(n_tx = n(), n_DE = sum(DE), pct_DE = mean(DE), .groups = "drop")

p.de <- DE_number_df %>%
  ggplot(aes(x = length_bin, y = n_DE, fill = Platform, group = Platform)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = color_palette) +
  labs(
    title = "Number of DE transcripts by transcript length bin (shared-filter tested transcripts)",
    x = "Transcript length bin", y = "Number of DE transcripts"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.mean_cpm/p.mean_logcpm/p.median_cpm/p.de

# Get a version of Platform specific DE
DE_number_df.sp <- mat_list_dte %>%
  bind_rows(.id = "Contrast") %>%
  left_join(tx_length_lookup, by = c("Gene" = "tx_id")) %>%
  filter(Illumina + `ONT cDNA` + PacBio + `ONT dRNA` == 1) %>%
  mutate(length_bin = cut(Length, breaks = length_bins, include.lowest = TRUE)) %>%
  pivot_longer(cols = c(Illumina, `ONT cDNA`, PacBio, `ONT dRNA`), names_to = "Platform", values_to = "DE") %>%
  group_by(Platform, length_bin, Contrast) %>%
  summarise(n_tx = n(), n_DE = sum(DE), pct_DE = mean(DE), .groups = "drop")

p.de.sp <- DE_number_df.sp %>%
  ggplot(aes(x = length_bin, y = n_DE, fill = Platform, group = Platform)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  scale_fill_manual(values = color_palette) +
  labs(
    title = "Number of Platform_Specific DE transcripts by transcript length bin (shared-filter tested transcripts)",
    x = "Transcript length bin", y = "Number of DE transcripts"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.de.sp
```
## Is platform-specific detection generally have higher CPM in the detecting platform than in the non-detecting platforms?
```{r cpm-platform-specific}
# AvsL contrast, transcript level
mat_avsl <- mat_list_dte[["AvsL"]]
## Get mean logCPM per platform for the AvsL contrast (mean over all A and L samples, before overdispersion scaling)
logcpm_lookup_avsl <- bind_rows(lapply(
  list(Illumina   = ill_bulk.dge,
       `ONT cDNA` = ont_bulk.dge.oarfish,
       PacBio     = pb_bulk.dge.oarfish,
       `ONT dRNA` = ont_drna.dge.oarfish),
  function(dge) tibble(
    tx_id = rownames(dge), 
    avg_logCPM = rowMeans(cpm(dge[,dge$samples$group %in% c('sclc_a', 'luad')], log = TRUE)),
    avg_CPM = rowMeans(cpm(dge[,dge$samples$group %in% c('sclc_a', 'luad')], log = FALSE)),
    median_CPM = apply(cpm(dge[,dge$samples$group %in% c('sclc_a', 'luad')], log = FALSE), 1, median, na.rm = TRUE)
  )
), .id = "Platform")

## For each platform, compare the logFC / significance of transcripts detected as DE only by that platform vs the same transcripts in the non-detecting platforms
# NOTE: avg logCPM (averaged across both groups) is a poor discriminator here because
#   long-read platforms have lower per-feature coverage than Illumina, so their overall
#   average CPM is lower regardless of DE signal. The relevant question is whether the
#   detecting platform shows a *larger fold change* and *stronger significance* for those
#   transcripts — which is captured by logFC and FDR from the DE test.
#
# NOTE: The logcpm_lookup_avsl object above used dge$sample$group (typo → NULL → 0-column DGE)
#   and is therefore unreliable; we build what we need directly from the DE results instead.

platforms_avsl <- c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA")

# Extract logFC and FDR for every transcript from each platform's AvsL DE result
de_stats_avsl <- bind_rows(lapply(
  list(
    Illumina   = ill_bulk.scl_dte_human.list$AvsL.DE,
    `ONT cDNA` = ont_bulk.scl_dte_human.list$AvsL.DE,
    PacBio     = pb_bulk.scl_dte_human.list$AvsL.DE,
    `ONT dRNA` = drna_bulk.scl_dte_human.list$AvsL.DE
  ),
  function(de) {
    topTags(de, n = Inf, sort.by = "none")$table %>%
      rownames_to_column("tx_id") %>%
      select(tx_id, logFC, FDR)
  }
), .id = "Platform")

# For each focal platform, identify its platform-specific DE transcripts, then
# look up the logFC / FDR across ALL platforms for those transcripts
platform_specific_de_df <- bind_rows(lapply(platforms_avsl, function(focal) {
  others <- setdiff(platforms_avsl, focal)

  specific_tx <- mat_avsl %>%
    filter(.data[[focal]] == 1,
           rowSums(select(., all_of(others))) == 0) %>%
    pull(Gene)

  if (length(specific_tx) == 0) return(NULL)

  de_stats_avsl %>%
    filter(tx_id %in% specific_tx) %>%
    mutate(
      focal_platform = focal,
      DetectionLabel = if_else(Platform == focal,
                               "Detecting platform",
                               "Non-detecting platforms")
    )
}))

# Sanity-check summary table
platform_specific_de_df %>%
  group_by(focal_platform, DetectionLabel) %>%
  summarise(
    n_tx         = n_distinct(tx_id),
    median_absFC = round(median(abs(logFC), na.rm = TRUE), 3),
    median_FDR   = round(median(FDR,        na.rm = TRUE), 4),
    .groups = "drop"
  ) %>%
  knitr::kable(
    caption = "Platform-specific DE transcripts: |logFC| and FDR in detecting vs non-detecting platforms (AvsL)"
  )

det_levels <- c("Detecting platform", "Non-detecting platforms")
det_colours <- c("Detecting platform" = "#E64B35", "Non-detecting platforms" = "#4DBBD5")

base_df <- platform_specific_de_df %>%
  mutate(
    DetectionLabel = factor(DetectionLabel, levels = det_levels),
    focal_platform = factor(focal_platform, levels = platforms_avsl)
  )

# Plot 1: |logFC| — does the detecting platform show a larger fold change?
p.logfc <- base_df %>%
  ggplot(aes(x = DetectionLabel, y = abs(logFC), fill = DetectionLabel)) +
  geom_boxplot(outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.5) +
  facet_wrap(~focal_platform, nrow = 1) +
  scale_fill_manual(values = det_colours) +
  labs(
    title    = "Platform-specific DE: |logFC| in detecting vs non-detecting platforms",
    subtitle = "AvsL contrast — transcripts DE in exactly one platform",
    x = NULL, y = "|log2 Fold Change|"
  ) +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))

# Plot 2: -log10(FDR) — does the detecting platform show stronger significance?
p.fdr <- base_df %>%
  ggplot(aes(x = DetectionLabel, y = -log10(FDR + 1e-300), fill = DetectionLabel)) +
  geom_boxplot(outlier.shape = NA, width = 0.5) +
  geom_jitter(width = 0.15, alpha = 0.2, size = 0.5) +
  facet_wrap(~focal_platform, nrow = 1) +
  scale_fill_manual(values = det_colours) +
  labs(
    title    = "Platform-specific DE: significance in detecting vs non-detecting platforms",
    subtitle = "AvsL contrast — transcripts DE in exactly one platform",
    x = NULL, y = expression(-log[10](FDR))
  ) +
  theme_minimal() +
  theme(legend.position = "none", strip.text = element_text(face = "bold"))

p.logfc / p.fdr

```



## Analysis functions

All analyses in sections 3b–3g are implemented as parametrised functions that accept a
`mat_list` (named list of binary membership matrices per contrast) and `all_tx_universe`
(data frame of tested transcripts with `tx_id`, `Length`, `length_bin`). The same functions
are called in Part 4 on the platform-independent matrices, enabling a direct comparison.

```{r dte-analysis-functions}
# ---- Data builders ----

# Annotate one binary matrix: add TotalDetected, OverlapCat, Length, iso_count
annotate_mat <- function(mat, contrast_label) {
  platforms <- setdiff(names(mat), "Gene")
  mat %>%
    mutate(
      TotalDetected = rowSums(select(., all_of(platforms))),
      OverlapCat = factor(case_when(
        TotalDetected == 4 ~ "All 4",
        TotalDetected == 3 ~ "3 platforms",
        TotalDetected == 2 ~ "2 platforms",
        TotalDetected == 1 ~ "Platform-specific",
        TRUE               ~ "Not DE"),
        levels = c("All 4","3 platforms","2 platforms","Platform-specific","Not DE")),
      Contrast = contrast_label
    ) %>%
    left_join(tx_length_lookup, by = c("Gene" = "tx_id")) %>%
    mutate(Gene_clean = sub("\\..*", "", sapply(strsplit(Gene, "\\|"), `[`, 1))) %>%
    left_join(
      iso_count_lookup %>% select(tx_name_clean, tx_count) %>% rename(iso_count = tx_count),
      by = c("Gene_clean" = "tx_name_clean")
    ) %>%
    left_join(expressed_iso_count_lookup, by = c("Gene_clean" = "tx_name_clean"))
}

# Build per-platform long df: for each platform, each DE transcript labelled
# "Platform-specific", "Shared >=2", or "Any platform DE" (not detected by focal
# but detected by at least one other platform)
build_per_platform_df <- function(mat_list) {
  bind_rows(lapply(names(mat_list), function(ct) {
    platforms <- setdiff(names(mat_list[[ct]]), "Gene")
    mat_ann   <- annotate_mat(mat_list[[ct]], ct)
    bind_rows(lapply(platforms, function(p) {
      others <- setdiff(platforms, p)
      # Rows where focal platform detected DE
      detected <- mat_ann %>%
        filter(.data[[p]] == 1) %>%
        mutate(
          Platform = p,
          DetectionLabel = if_else(TotalDetected == 1, "Platform-specific", "Shared ≥2")
        )
      # Rows where focal platform did NOT detect, but at least one other platform did
      any_other_de <- mat_ann %>%
        filter(.data[[p]] == 0, TotalDetected >= 1) %>%
        mutate(Platform = p, DetectionLabel = "Any platform DE")
      bind_rows(detected, any_other_de) %>%
        left_join(logcpm_lookup %>% filter(Platform == p) %>% select(tx_id, avg_logCPM),
                  by = c("Gene" = "tx_id"))
    }))
  })) %>%
  mutate(
    Platform = factor(Platform, levels = names(color_palette)),
    DetectionLabel = factor(DetectionLabel,
      levels = c("Platform-specific", "Shared ≥2", "Any platform DE"))
  )
}

# Build per-platform long df with three detection categories:
#   "Platform-specific absence"  — focal = 0, all 3 others = 1
#   "Platform-specific detection"— focal = 1, all 3 others = 0
#   "Detected ≥2 support"        — focal = 1, ≥2 of the 3 others = 1
build_absence_detection_df <- function(mat_list) {
  bind_rows(lapply(names(mat_list), function(ct) {
    mat       <- mat_list[[ct]]
    platforms <- setdiff(names(mat), "Gene")
    bind_rows(lapply(platforms, function(p) {
      others        <- setdiff(platforms, p)
      n_others      <- rowSums(mat[, others, drop = FALSE])
      absent_rows   <- mat[mat[[p]] == 0 & n_others == 3,  , drop = FALSE]
      detected_only <- mat[mat[[p]] == 1 & n_others == 0,  , drop = FALSE]
      detected_2p   <- mat[mat[[p]] == 1 & n_others >= 2,  , drop = FALSE]
      bind_rows(
        if (nrow(absent_rows)   > 0) tibble(Gene = absent_rows$Gene,   Platform = p,
          Contrast = ct, Category = "Platform-specific absence"),
        if (nrow(detected_only) > 0) tibble(Gene = detected_only$Gene, Platform = p,
          Contrast = ct, Category = "Platform-specific detection"),
        if (nrow(detected_2p)   > 0) tibble(Gene = detected_2p$Gene,   Platform = p,
          Contrast = ct, Category = "Detected ≥2 support")
      )
    }))
  })) %>%
  left_join(tx_length_lookup, by = c("Gene" = "tx_id")) %>%
  mutate(Gene_clean = sub("\\..*", "", sapply(strsplit(Gene, "\\|"), `[`, 1))) %>%
  left_join(
    iso_count_lookup %>% select(tx_name_clean, tx_count) %>% rename(iso_count = tx_count),
    by = c("Gene_clean" = "tx_name_clean")
  ) %>%
  left_join(expressed_iso_count_lookup, by = c("Gene_clean" = "tx_name_clean")) %>%
  left_join(logcpm_lookup, by = c("Gene" = "tx_id", "Platform" = "Platform")) %>%
  mutate(
    Category = factor(Category,
      levels = c("Platform-specific absence", "Platform-specific detection", "Detected ≥2 support")),
    Platform = factor(Platform, levels = names(color_palette))
  )
}

# Build PCR-group classified df (PCR-specific, Non-PCR-specific, All 4)
build_pcr_group_df <- function(mat_list) {
  bind_rows(lapply(names(mat_list), function(ct) {
    mat_list[[ct]] %>%
      mutate(
        PCR_group = case_when(
          PacBio == 1 & `ONT cDNA` == 1 & Illumina == 0 & `ONT dRNA` == 0 ~ "PCR-specific",
          Illumina == 1 & `ONT dRNA` == 1 & PacBio == 0 & `ONT cDNA` == 0  ~ "Non-PCR-specific",
          PacBio == 1 & `ONT cDNA` == 1 & Illumina == 1 & `ONT dRNA` == 1  ~ "All 4",
          TRUE ~ NA_character_
        ),
        Contrast = ct
      ) %>%
      filter(!is.na(PCR_group))
  })) %>%
  left_join(tx_length_lookup, by = c("Gene" = "tx_id")) %>%
  mutate(Gene_clean = sub("\\..*", "", sapply(strsplit(Gene, "\\|"), `[`, 1))) %>%
  left_join(
    iso_count_lookup %>% select(tx_name_clean, tx_count) %>% rename(iso_count = tx_count),
    by = c("Gene_clean" = "tx_name_clean")
  ) %>%
  left_join(expressed_iso_count_lookup, by = c("Gene_clean" = "tx_name_clean")) %>%
  mutate(PCR_group = factor(PCR_group, levels = c("PCR-specific","Non-PCR-specific","All 4")))
}

# Assign quantile-based length bins (default 10 deciles) with readable kb range labels.
make_length_decile_bins <- function(lengths, n = 10) {
  probs <- seq(0, 1, length.out = n + 1)
  brks  <- unique(quantile(lengths, probs = probs, na.rm = TRUE))
  nb    <- length(brks) - 1
  labs  <- sapply(seq_len(nb), function(i) {
    lo <- round(brks[i]   / 1000, 1)
    hi <- round(brks[i+1] / 1000, 1)
    if (i == 1)       paste0("≤", hi, " kb")
    else if (i == nb) paste0(">",  lo, " kb")
    else              paste0(lo, "–", hi, " kb")
  })
  cut(lengths, breaks = brks, labels = labs, include.lowest = TRUE, right = TRUE)
}

# Build length-bin analysis df given mat_list and a tx universe (tx_id, length_bin)
# Joins per-platform avg_logCPM so expression level can be checked alongside DE rate.
build_length_bin_df <- function(mat_list, all_tx_universe) {
  bind_rows(lapply(names(mat_list), function(ct) {
    bind_rows(lapply(names(color_palette), function(p) {
      de_ids <- mat_list[[ct]] %>% filter(.data[[p]] == 1) %>% pull(Gene)
      all_tx_universe %>%
        mutate(Platform = p, is_DE = tx_id %in% de_ids, Contrast = ct) %>%
        left_join(logcpm_lookup %>% filter(Platform == p) %>% select(tx_id, avg_logCPM),
                  by = "tx_id")
    }))
  }))
}

# ---- Plot functions ----

plot_length_overlap <- function(cat_df_all, pp_df_all, title_prefix = "") {
  p1 <- cat_df_all %>%
    filter(OverlapCat != "Not DE", !is.na(Length)) %>%
    ggplot(aes(x = OverlapCat, y = Length / 1000, fill = OverlapCat)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_wrap(~ Contrast) +
    scale_fill_brewer(palette = "Set2") +
    scale_y_log10(labels = comma) +
    labs(title = paste(title_prefix, "— transcript length by overlap category"),
         y = "Transcript length (kb, log10)", x = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))

  p2 <- pp_df_all %>%
    filter(!is.na(Length)) %>%
    ggplot(aes(x = DetectionLabel, y = Length / 1000, fill = Platform)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_grid(Contrast ~ Platform) +
    scale_fill_manual(values = color_palette) +
    scale_y_log10(labels = comma) +
    labs(title = paste(title_prefix, "— platform-specific vs shared per platform"),
         y = "Transcript length (kb, log10)", x = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1),
          strip.text = element_text(size = 8))

  list(by_overlap = p1, by_platform = p2)
}

plot_length_stats_table <- function(pp_df_all, filter_label = "") {
  pp_df_all %>%
    filter(Contrast == "AvsL", !is.na(Length)) %>%
    group_by(Platform) %>%
    summarise(
      n_specific    = sum(DetectionLabel == "Platform-specific"),
      n_shared      = sum(DetectionLabel == "Shared ≥2"),
      n_any_de      = sum(DetectionLabel == "Any platform DE"),
      med_specific_kb = round(median(Length[DetectionLabel == "Platform-specific"], na.rm = TRUE) / 1000, 2),
      med_shared_kb   = round(median(Length[DetectionLabel == "Shared ≥2"],         na.rm = TRUE) / 1000, 2),
      med_any_de_kb   = round(median(Length[DetectionLabel == "Any platform DE"],   na.rm = TRUE) / 1000, 2),
      p_spe_vs_shared = {
        g1 <- Length[DetectionLabel == "Platform-specific"]
        g2 <- Length[DetectionLabel == "Shared ≥2"]
        if (length(g1) < 3 || length(g2) < 3) NA_real_ else wilcox.test(g1, g2)$p.value
      },
      p_spe_vs_any = {
        g1 <- Length[DetectionLabel == "Platform-specific"]
        g2 <- Length[DetectionLabel == "Any platform DE"]
        if (length(g1) < 3 || length(g2) < 3) NA_real_ else wilcox.test(g1, g2)$p.value
      }, .groups = "drop"
    ) %>%
    mutate(p_spe_vs_shared_adj = p.adjust(p_spe_vs_shared, method = "BH"),
           p_spe_vs_any_adj    = p.adjust(p_spe_vs_any,    method = "BH"),
           across(starts_with("p_"), ~formatC(.x, digits = 3, format = "e"))) %>%
    knitr::kable(caption = paste(filter_label,
      "Wilcoxon: transcript length — platform-specific vs shared ≥2 vs any platform DE (AvsL)"),
      row.names = FALSE)
}

plot_isoform_overlap <- function(cat_df_all, pp_df_all, title_prefix = "") {
  make_iso_panel <- function(df_cat, df_pp, col, ylab) {
    p1 <- df_cat %>%
      filter(OverlapCat != "Not DE", !is.na(.data[[col]])) %>%
      ggplot(aes(x = OverlapCat, y = .data[[col]], fill = OverlapCat)) +
      geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
      geom_boxplot(width = 0.15, outlier.shape = NA, fill = "white", alpha = 0.8) +
      facet_wrap(~ Contrast) +
      scale_fill_brewer(palette = "Set2") +
      scale_y_log10() +
      labs(title = paste(title_prefix, "—", ylab, "by overlap category"),
           y = paste(ylab, "(log10)"), x = NULL) +
      theme_minimal() +
      theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))

    p2 <- df_pp %>%
      filter(!is.na(.data[[col]])) %>%
      ggplot(aes(x = DetectionLabel, y = .data[[col]], fill = Platform)) +
      geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
      geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
      facet_grid(Contrast ~ Platform) +
      scale_fill_manual(values = color_palette) +
      scale_y_log10() +
      labs(title = paste(title_prefix, "—", ylab, "per platform"),
           y = paste(ylab, "(log10)"), x = NULL) +
      theme_minimal() +
      theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1),
            strip.text = element_text(size = 8))

    list(by_overlap = p1, by_platform = p2)
  }

  anno  <- make_iso_panel(cat_df_all, pp_df_all, "iso_count",      "Annotation isoform count")
  expr  <- make_iso_panel(cat_df_all, pp_df_all, "expr_iso_count", "Expressed isoform count")
  list(anno_by_overlap  = anno$by_overlap,  anno_by_platform  = anno$by_platform,
       expr_by_overlap  = expr$by_overlap,  expr_by_platform  = expr$by_platform)
}

plot_isoform_stats_table <- function(pp_df_all, filter_label = "") {
  pp_df_all %>%
    filter(Contrast == "AvsL") %>%
    group_by(Platform) %>%
    summarise(
      n_specific           = sum(DetectionLabel == "Platform-specific"),
      n_shared             = sum(DetectionLabel == "Shared ≥2"),
      n_any_de             = sum(DetectionLabel == "Any platform DE"),
      med_anno_specific    = round(median(iso_count[DetectionLabel == "Platform-specific"],      na.rm = TRUE), 1),
      med_anno_shared      = round(median(iso_count[DetectionLabel == "Shared ≥2"],              na.rm = TRUE), 1),
      med_anno_any_de      = round(median(iso_count[DetectionLabel == "Any platform DE"],        na.rm = TRUE), 1),
      med_expr_specific    = round(median(expr_iso_count[DetectionLabel == "Platform-specific"], na.rm = TRUE), 1),
      med_expr_shared      = round(median(expr_iso_count[DetectionLabel == "Shared ≥2"],         na.rm = TRUE), 1),
      med_expr_any_de      = round(median(expr_iso_count[DetectionLabel == "Any platform DE"],   na.rm = TRUE), 1),
      p_anno_spe_vs_shared = {
        g1 <- iso_count[DetectionLabel == "Platform-specific"]
        g2 <- iso_count[DetectionLabel == "Shared ≥2"]
        if (length(g1) < 3 || length(g2) < 3) NA_real_ else wilcox.test(g1, g2)$p.value
      },
      p_anno_spe_vs_any = {
        g1 <- iso_count[DetectionLabel == "Platform-specific"]
        g2 <- iso_count[DetectionLabel == "Any platform DE"]
        if (length(g1) < 3 || length(g2) < 3) NA_real_ else wilcox.test(g1, g2)$p.value
      },
      p_expr_spe_vs_shared = {
        g1 <- expr_iso_count[DetectionLabel == "Platform-specific"]
        g2 <- expr_iso_count[DetectionLabel == "Shared ≥2"]
        if (length(g1) < 3 || length(g2) < 3) NA_real_ else wilcox.test(g1, g2)$p.value
      },
      p_expr_spe_vs_any = {
        g1 <- expr_iso_count[DetectionLabel == "Platform-specific"]
        g2 <- expr_iso_count[DetectionLabel == "Any platform DE"]
        if (length(g1) < 3 || length(g2) < 3) NA_real_ else wilcox.test(g1, g2)$p.value
      }, .groups = "drop"
    ) %>%
    mutate(p_anno_spe_vs_shared_adj = p.adjust(p_anno_spe_vs_shared, method = "BH"),
           p_anno_spe_vs_any_adj    = p.adjust(p_anno_spe_vs_any,    method = "BH"),
           p_expr_spe_vs_shared_adj = p.adjust(p_expr_spe_vs_shared, method = "BH"),
           p_expr_spe_vs_any_adj    = p.adjust(p_expr_spe_vs_any,    method = "BH"),
           across(starts_with("p_"), ~formatC(.x, digits = 3, format = "e"))) %>%
    knitr::kable(caption = paste(filter_label,
      "Wilcoxon: annotation & expressed isoform count — platform-specific vs shared ≥2 vs any platform DE (AvsL)"),
      row.names = FALSE)
}

plot_length_bin_enrichment <- function(mat_list, all_tx_universe, title_prefix = "") {
  bin_df <- build_length_bin_df(mat_list, all_tx_universe)

  avsl_sum <- bin_df %>%
    filter(!is.na(length_bin), Contrast == "AvsL") %>%
    group_by(Platform, length_bin) %>%
    summarise(n_DE = sum(is_DE), prop_DE = sum(is_DE) / n(), .groups = "drop")

  all_sum <- bin_df %>%
    filter(!is.na(length_bin)) %>%
    group_by(Platform, length_bin, Contrast) %>%
    summarise(n_DE = sum(is_DE), prop_DE = sum(is_DE) / n(), .groups = "drop")

  p_prop_avsl <- ggplot(avsl_sum, aes(x = length_bin, y = prop_DE, fill = Platform)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    geom_text(aes(label = sprintf("%.3f", prop_DE)),
              position = position_dodge(0.7), vjust = -0.4, size = 2) +
    scale_fill_manual(values = color_palette) +
    scale_y_continuous(labels = percent_format()) +
    labs(title = paste(title_prefix, "— DE rate by length bin (AvsL)"),
         subtitle = "Proportion of all tested transcripts in each bin",
         y = "Proportion called DE", x = "Transcript length category") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  p_count_avsl <- ggplot(avsl_sum, aes(x = length_bin, y = n_DE, fill = Platform)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    geom_text(aes(label = n_DE),
              position = position_dodge(0.7), vjust = -0.4, size = 2) +
    scale_fill_manual(values = color_palette) +
    labs(title = paste(title_prefix, "— DE count by length bin (AvsL)"),
         y = "Number of DE transcripts", x = "Transcript length category") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  p_prop_all <- ggplot(all_sum, aes(x = length_bin, y = prop_DE, fill = Platform)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    facet_wrap(~ Contrast) +
    scale_fill_manual(values = color_palette) +
    scale_y_continuous(labels = percent_format()) +
    labs(title = paste(title_prefix, "— DE rate by length bin, all contrasts"),
         y = "Proportion called DE", x = "Transcript length category") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  p_count_all <- ggplot(all_sum, aes(x = length_bin, y = n_DE, fill = Platform)) +
    geom_col(position = position_dodge(0.7), width = 0.65) +
    facet_wrap(~ Contrast) +
    scale_fill_manual(values = color_palette) +
    labs(title = paste(title_prefix, "— DE count by length bin, all contrasts"),
         y = "Number of DE transcripts", x = "Transcript length category") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))

  # logCPM distribution by length bin (AvsL) — shows whether longer transcripts
  # simply have higher expression, which would confound the DE-rate trend.
  p_logcpm_bin <- bin_df %>%
    filter(!is.na(length_bin), !is.na(avg_logCPM), Contrast == "AvsL") %>%
    mutate(Platform = factor(Platform, levels = names(color_palette))) %>%
    ggplot(aes(x = length_bin, y = avg_logCPM, fill = Platform)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_wrap(~ Platform, nrow = 1) +
    scale_fill_manual(values = color_palette) +
    labs(title = paste(title_prefix, "— avg logCPM by length bin (AvsL)"),
         subtitle = "Expression level per bin per platform — confound check for DE-rate trend",
         y = "Average logCPM", x = "Transcript length category") +
    theme_minimal() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 30, hjust = 1))

  list(avsl_prop = p_prop_avsl, avsl_count = p_count_avsl,
       all_prop  = p_prop_all,  all_count  = p_count_all,
       logcpm_bin = p_logcpm_bin)
}

plot_logcpm_analysis <- function(pp_df, absence_df, title_prefix = "") {
  # 1. logCPM distribution: platform-specific vs shared ≥2, per platform (AvsL)
  p_pp <- pp_df %>%
    filter(!is.na(avg_logCPM), Contrast == "AvsL") %>%
    ggplot(aes(x = DetectionLabel, y = avg_logCPM, fill = Platform)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_grid(~ Platform) +
    scale_fill_manual(values = color_palette) +
    labs(title = paste(title_prefix, "— logCPM: platform-specific vs shared ≥2 (AvsL)"),
         y = "Average logCPM", x = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))

  # 2. logCPM stats table: medians + Wilcoxon per platform
  stats_tbl <- pp_df %>%
    filter(!is.na(avg_logCPM), Contrast == "AvsL") %>%
    group_by(Platform) %>%
    summarise(
      n_specific          = sum(DetectionLabel == "Platform-specific"),
      n_shared            = sum(DetectionLabel == "Shared ≥2"),
      n_any_de            = sum(DetectionLabel == "Any platform DE"),
      med_logCPM_specific = round(median(avg_logCPM[DetectionLabel == "Platform-specific"], na.rm=TRUE), 2),
      med_logCPM_shared   = round(median(avg_logCPM[DetectionLabel == "Shared ≥2"],         na.rm=TRUE), 2),
      med_logCPM_any_de   = round(median(avg_logCPM[DetectionLabel == "Any platform DE"],   na.rm=TRUE), 2),
      p_spe_vs_shared = {
        g1 <- avg_logCPM[DetectionLabel == "Platform-specific"]
        g2 <- avg_logCPM[DetectionLabel == "Shared ≥2"]
        if (length(g1)<3 || length(g2)<3) NA_real_ else wilcox.test(g1,g2)$p.value
      },
      p_spe_vs_any = {
        g1 <- avg_logCPM[DetectionLabel == "Platform-specific"]
        g2 <- avg_logCPM[DetectionLabel == "Any platform DE"]
        if (length(g1)<3 || length(g2)<3) NA_real_ else wilcox.test(g1,g2)$p.value
      }, .groups = "drop"
    ) %>%
    mutate(p_spe_vs_shared_adj = p.adjust(p_spe_vs_shared, method = "BH"),
           p_spe_vs_any_adj    = p.adjust(p_spe_vs_any,    method = "BH"),
           across(starts_with("p_"), ~formatC(.x, digits = 3, format = "e"))) %>%
    knitr::kable(caption = paste(title_prefix,
      "— logCPM Wilcoxon: platform-specific vs shared ≥2 vs any platform DE (AvsL)"), row.names = FALSE)

  # 3. logCPM by absence category (AvsL)
  cat_colours <- c(
    "Platform-specific absence"   = "#E69F00",
    "Platform-specific detection" = "#0072B2",
    "Detected ≥2 support"         = "#009E73"
  )
  p_abs <- absence_df %>%
    filter(!is.na(avg_logCPM), Contrast == "AvsL") %>%
    ggplot(aes(x = Category, y = avg_logCPM, fill = Category)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_wrap(~ Platform, nrow = 1) +
    scale_fill_manual(values = cat_colours) +
    labs(title = paste(title_prefix,
           "— logCPM: absence / platform-specific detection / detected ≥2 support (AvsL)"),
         y = "Average logCPM", x = NULL) +
    theme_minimal() +
    theme(legend.position = "bottom", axis.text.x = element_blank(),
          axis.ticks.x = element_blank())

  list(by_detection = p_pp, stats_table = stats_tbl, by_absence = p_abs)
}

plot_absence_analysis <- function(absence_detection_df, title_prefix = "") {
  # Absence-count bar chart (one facet per contrast)
  counts_p <- absence_detection_df %>%
    filter(Category == "Platform-specific absence") %>%
    count(Platform, Contrast, name = "n_absent") %>%
    ggplot(aes(x = Platform, y = n_absent, fill = Platform)) +
    geom_col(width = 0.65) +
    geom_text(aes(label = n_absent), vjust = -0.4, size = 3) +
    facet_wrap(~ Contrast) +
    scale_fill_manual(values = color_palette) +
    labs(title = paste(title_prefix, "— platform-specific absence counts"),
         y = "Number of transcripts", x = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))

  # Recode to 2-group comparison: absence vs any DE
  cat_colours <- c(
    "Platform-specific absence" = "#E69F00",
    "Any platform DE"           = "#009E73"
  )

  plot_df <- absence_detection_df %>%
    mutate(Category2 = case_when(
      Category == "Platform-specific absence" ~ "Platform-specific absence",
      TRUE                                    ~ "Any platform DE"
    ),
    Category2 = factor(Category2, levels = c("Platform-specific absence", "Any platform DE")))

  # Violin + box: length, faceted by platform (AvsL)
  p_len <- plot_df %>%
    filter(!is.na(Length), Contrast == "AvsL") %>%
    ggplot(aes(x = Category2, y = Length / 1000, fill = Category2)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_wrap(~ Platform, nrow = 1) +
    scale_fill_manual(values = cat_colours) +
    scale_y_log10(labels = comma) +
    labs(title = paste(title_prefix,
           "— length: platform-specific absence vs any platform DE (AvsL)"),
         y = "Transcript length (kb, log10)", x = NULL) +
    theme_minimal() +
    theme(legend.position = "bottom", axis.text.x = element_blank(),
          axis.ticks.x = element_blank())

  make_iso_violin <- function(df, col, ylab) {
    df %>%
      filter(!is.na(.data[[col]]), Contrast == "AvsL") %>%
      ggplot(aes(x = Category2, y = .data[[col]], fill = Category2)) +
      geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
      geom_boxplot(width = 0.12, outlier.shape = NA, fill = "white", alpha = 0.8) +
      facet_wrap(~ Platform, nrow = 1) +
      scale_fill_manual(values = cat_colours) +
      scale_y_log10() +
      labs(title = paste(title_prefix, "—", ylab,
             ": absence vs any platform DE (AvsL)"),
           y = paste(ylab, "(log10)"), x = NULL) +
      theme_minimal() +
      theme(legend.position = "bottom", axis.text.x = element_blank(),
            axis.ticks.x = element_blank())
  }

  p_iso      <- make_iso_violin(plot_df, "iso_count",      "Annotation isoform count")
  p_expr_iso <- make_iso_violin(plot_df, "expr_iso_count", "Expressed isoform count")

  # Wilcoxon: absence vs any DE (per platform, AvsL)
  stats_tbl <- plot_df %>%
    filter(Contrast == "AvsL", !is.na(Length)) %>%
    group_by(Platform) %>%
    summarise(
      n_absent    = sum(Category2 == "Platform-specific absence"),
      n_any_de    = sum(Category2 == "Any platform DE"),
      med_len_absent_kb = round(median(Length[Category2 == "Platform-specific absence"], na.rm=TRUE)/1000, 2),
      med_len_any_de_kb = round(median(Length[Category2 == "Any platform DE"],           na.rm=TRUE)/1000, 2),
      med_anno_absent   = round(median(iso_count[Category2 == "Platform-specific absence"],   na.rm=TRUE), 1),
      med_anno_any_de   = round(median(iso_count[Category2 == "Any platform DE"],             na.rm=TRUE), 1),
      med_expr_absent   = round(median(expr_iso_count[Category2 == "Platform-specific absence"], na.rm=TRUE), 1),
      med_expr_any_de   = round(median(expr_iso_count[Category2 == "Any platform DE"],           na.rm=TRUE), 1),
      p_len = {
        g1 <- Length[Category2 == "Platform-specific absence"]
        g2 <- Length[Category2 == "Any platform DE"]
        if (length(g1)<3 || length(g2)<3) NA_real_ else wilcox.test(g1,g2)$p.value
      },
      p_expr = {
        g1 <- expr_iso_count[Category2 == "Platform-specific absence"]
        g2 <- expr_iso_count[Category2 == "Any platform DE"]
        if (length(g1)<3 || length(g2)<3) NA_real_ else wilcox.test(g1,g2)$p.value
      }, .groups = "drop"
    ) %>%
    mutate(p_len_adj  = formatC(p.adjust(p_len,  method="BH"), digits=3, format="e"),
           p_expr_adj = formatC(p.adjust(p_expr, method="BH"), digits=3, format="e"),
           p_len      = formatC(p_len,  digits=3, format="e"),
           p_expr     = formatC(p_expr, digits=3, format="e")) %>%
    knitr::kable(caption = paste(title_prefix,
      "— absence vs any platform DE: length & isoform counts (AvsL)"),
      row.names = FALSE)

  list(counts = counts_p, length_iso = p_len / p_iso / p_expr_iso, table = stats_tbl)
}

plot_pcr_group_analysis <- function(pcr_df, mat_list, all_tx_universe, title_prefix = "") {
  counts_p <- pcr_df %>%
    count(PCR_group, Contrast) %>%
    ggplot(aes(x = PCR_group, y = n, fill = PCR_group)) +
    geom_col(width = 0.65) +
    geom_text(aes(label = n), vjust = -0.4, size = 3) +
    facet_wrap(~ Contrast) +
    scale_fill_brewer(palette = "Set1") +
    labs(title = paste(title_prefix, "— PCR-group DE counts"),
         y = "Number of transcripts", x = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 25, hjust = 1))

  p_len <- pcr_df %>%
    filter(!is.na(Length)) %>%
    ggplot(aes(x = PCR_group, y = Length / 1000, fill = PCR_group)) +
    geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
    geom_boxplot(width = 0.14, outlier.shape = NA, fill = "white", alpha = 0.8) +
    facet_wrap(~ Contrast) +
    scale_fill_brewer(palette = "Set1") +
    scale_y_log10(labels = comma) +
    labs(title = paste(title_prefix, "— length by PCR group"),
         y = "Transcript length (kb, log10)", x = NULL) +
    theme_minimal() +
    theme(legend.position = "none", axis.text.x = element_text(angle = 25, hjust = 1))

  make_pcr_iso_violin <- function(col, ylab) {
    pcr_df %>%
      filter(!is.na(.data[[col]])) %>%
      ggplot(aes(x = PCR_group, y = .data[[col]], fill = PCR_group)) +
      geom_violin(scale = "width", alpha = 0.7, trim = TRUE) +
      geom_boxplot(width = 0.14, outlier.shape = NA, fill = "white", alpha = 0.8) +
      facet_wrap(~ Contrast) +
      scale_fill_brewer(palette = "Set1") +
      scale_y_log10() +
      labs(title = paste(title_prefix, "—", ylab, "by PCR group"),
           y = paste(ylab, "(log10)"), x = NULL) +
      theme_minimal() +
      theme(legend.position = "none", axis.text.x = element_text(angle = 25, hjust = 1))
  }

  p_iso      <- make_pcr_iso_violin("iso_count",      "Annotation isoform count")
  p_expr_iso <- make_pcr_iso_violin("expr_iso_count", "Expressed isoform count")

  # Length-bin composition within PCR groups
  pcr_bin_df <- build_length_bin_df(mat_list, all_tx_universe) %>%
    left_join(pcr_df %>% select(Gene, Contrast, PCR_group), by = c("tx_id" = "Gene", "Contrast"))

  p_bin <- bind_rows(lapply(names(mat_list), function(ct) {
    mat <- mat_list[[ct]]
    pcr_ids    <- mat %>% filter(PacBio==1, `ONT cDNA`==1, Illumina==0, `ONT dRNA`==0) %>% pull(Gene)
    nonpcr_ids <- mat %>% filter(Illumina==1, `ONT dRNA`==1, PacBio==0, `ONT cDNA`==0) %>% pull(Gene)
    all4_ids   <- mat %>% filter(PacBio==1, `ONT cDNA`==1, Illumina==1, `ONT dRNA`==1)  %>% pull(Gene)
    all_tx_universe %>%
      mutate(Contrast = ct,
             PCR_group = case_when(
               tx_id %in% pcr_ids    ~ "PCR-specific",
               tx_id %in% nonpcr_ids ~ "Non-PCR-specific",
               tx_id %in% all4_ids   ~ "All 4", TRUE ~ NA_character_)) %>%
      filter(!is.na(PCR_group))
  })) %>%
    filter(!is.na(length_bin)) %>%
    count(PCR_group, length_bin, Contrast) %>%
    mutate(PCR_group = factor(PCR_group, levels = c("PCR-specific","Non-PCR-specific","All 4"))) %>%
    ggplot(aes(x = length_bin, y = n, fill = PCR_group)) +
    geom_col(position = position_fill(), width = 0.75) +
    facet_wrap(~ Contrast) +
    scale_fill_brewer(palette = "Set1") +
    scale_y_continuous(labels = percent_format()) +
    labs(title = paste(title_prefix, "— length-bin composition by PCR group"),
         y = "Proportion", x = "Length category") +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))

  stats_tbl <- pcr_df %>%
    filter(Contrast == "AvsL", PCR_group %in% c("PCR-specific","Non-PCR-specific")) %>%
    group_by(PCR_group) %>%
    summarise(n = n(),
              median_length_kb      = round(median(Length,         na.rm = TRUE) / 1000, 2),
              median_anno_iso_count = round(median(iso_count,      na.rm = TRUE), 1),
              median_expr_iso_count = round(median(expr_iso_count, na.rm = TRUE), 1),
              .groups = "drop") %>%
    knitr::kable(caption = paste(title_prefix,
      "— PCR-specific vs Non-PCR-specific summary (AvsL)"), row.names = FALSE)

  list(counts = counts_p, length = p_len, isoform = p_iso, expr_isoform = p_expr_iso,
       length_bin = p_bin, table = stats_tbl)
}
```

## 3b–3e: Transcript length bias, isoform count, and length-bin enrichment {.tabset}

Categories per transcript:
- **All 4** — DE in all four platforms
- **3 platforms** — DE in exactly three
- **2 platforms** — DE in exactly two
- **Platform-specific** — DE in exactly one
- **Platform-specific absence** — NOT DE in the focal platform, but DE in all three others
- **Not DE** — not DE in any platform

```{r shared-dte-data}
# Build annotation data frames for the shared-feature analysis
shared_cat_df <- bind_rows(lapply(contrasts_dte, function(ct)
  annotate_mat(mat_list_dte[[ct]], ct)))

shared_pp_df  <- build_per_platform_df(mat_list_dte)

# Transcript universe for length-bin analysis (shared filtering: tested in shared set)
all_tx_tested <- tx_length_lookup %>%
  filter(tx_id %in% rownames(ill_bulk.scale_dge), grepl("^ENS", tx_id)) %>%
  mutate(length_bin = make_length_decile_bins(Length))

shared_absence_df <- build_absence_detection_df(mat_list_dte)
```

### Platform-specific vs shared: length per platform

```{r shared-length-by-platform, fig.height=8, fig.width=14}
shared_len_plots <- plot_length_overlap(shared_cat_df, shared_pp_df, "Shared features")
print(shared_len_plots$by_platform)
```

### Length statistics (AvsL)

```{r shared-length-stats}
print(plot_length_stats_table(shared_pp_df, "Shared features —"))
```

### Annotation isoform count per platform

```{r shared-isoform-by-platform, fig.height=8, fig.width=14}
shared_iso_plots <- plot_isoform_overlap(shared_cat_df, shared_pp_df, "Shared features")
print(shared_iso_plots$anno_by_platform)
```

### Expressed isoform count per platform

```{r shared-expr-iso-by-platform, fig.height=8, fig.width=14}
print(shared_iso_plots$expr_by_platform)
```

### Isoform count statistics (AvsL)

```{r shared-isoform-stats}
print(plot_isoform_stats_table(shared_pp_df, "Shared features —"))
```

### DE rate by length bin (AvsL)

```{r shared-length-bin-avsl, fig.height=5, fig.width=18}
shared_bin_plots <- plot_length_bin_enrichment(mat_list_dte, all_tx_tested, "Shared features")
print(shared_bin_plots$avsl_prop)
```

### DE count by length bin (AvsL)

```{r shared-length-bin-count-avsl, fig.height=5, fig.width=18}
print(shared_bin_plots$avsl_count)
```

### DE rate by length bin (all contrasts)

```{r shared-length-bin-all, fig.height=9, fig.width=18}
print(shared_bin_plots$all_prop)
```

### DE count by length bin (all contrasts)

```{r shared-length-bin-count-all, fig.height=9, fig.width=18}
print(shared_bin_plots$all_count)
```

### logCPM by length bin (AvsL)

```{r shared-logcpm-bin, fig.height=6, fig.width=16}
print(shared_bin_plots$logcpm_bin)
```

## 3f. Platform-specific absence {.tabset}

A transcript is **platform-specifically absent** when it is called DE by all three other platforms
but *not* by the focal platform.

### Absence counts

```{r shared-absence-counts, fig.height=4, fig.width=10}
shared_abs_res <- plot_absence_analysis(shared_absence_df, "Shared features")
print(shared_abs_res$counts)
```

### Length and isoform count (AvsL)

```{r shared-absence-length-iso, fig.height=9, fig.width=14}
print(shared_abs_res$length_iso)
```

### Summary table (AvsL)

```{r shared-absence-table}
print(shared_abs_res$table)
```

## 3g. Average logCPM {.tabset}

Average log-CPM (computed from TMM-normalized counts before overdispersion scaling) quantifies
transcript expression level. Platform-specific DE calls at low expression may reflect noise,
while shared DE calls should be enriched for highly expressed transcripts.

### logCPM: platform-specific vs shared

```{r shared-logcpm-pp, fig.height=4, fig.width=12}
shared_logcpm_res <- plot_logcpm_analysis(shared_pp_df, shared_absence_df, "Shared features")
print(shared_logcpm_res$by_detection)
```

### logCPM statistics (AvsL)

```{r shared-logcpm-stats}
print(shared_logcpm_res$stats_table)
```

### logCPM: absence / detection / ≥2-supported

```{r shared-logcpm-absence, fig.height=5, fig.width=14}
print(shared_logcpm_res$by_absence)
```


# Part 4: Platform-independent feature selection

Each platform independently applies `filterByExpr` on its own data, with no requirement that
the feature be expressed in any other platform. DE is then run per-platform on its own feature
set. Upset matrices are built on the union of all tested features, with a 0 assigned for any
platform that did not test a given feature (i.e., it was filtered out and cannot be DE).

```{r indep-gene-filter, results='hide'}
# Reload raw gene-level DGE (unfiltered) from the original RDS — before any shared
# filtering was applied in Part 1.
ill_raw.gene  <- bulk_obj[["ill_bulk.gene.dge"]]
ont_raw.gene  <- bulk_obj[["ont_bulk.gene.dge"]]
pb_raw.gene   <- bulk_obj[["pb_bulk.gene.dge"]]
drna_raw.gene <- bulk_obj[["drna_bulk.gene.dge"]]

# Restrict to the union of features present in at least one platform
# (we already trimmed to common_genes in Part 1; re-intersect to be safe)
all_gene_ids <- Reduce(union, lapply(list(ill_raw.gene, ont_raw.gene,
                                          pb_raw.gene, drna_raw.gene), rownames))

# Independent filterByExpr per platform
keep_ill_i  <- rownames(ill_raw.gene)  %in% all_gene_ids &
  filterByExpr(ill_raw.gene[rownames(ill_raw.gene) %in% all_gene_ids, ],
               group = ill_raw.gene$samples$group)
keep_ont_i  <- rownames(ont_raw.gene)  %in% all_gene_ids &
  filterByExpr(ont_raw.gene[rownames(ont_raw.gene) %in% all_gene_ids, ],
               group = ont_raw.gene$samples$group)
keep_pb_i   <- rownames(pb_raw.gene)   %in% all_gene_ids &
  filterByExpr(pb_raw.gene[rownames(pb_raw.gene) %in% all_gene_ids, ],
               group = pb_raw.gene$samples$group)
keep_drna_i <- rownames(drna_raw.gene) %in% all_gene_ids &
  filterByExpr(drna_raw.gene[rownames(drna_raw.gene) %in% all_gene_ids, ],
               group = drna_raw.gene$samples$group)

ill_indep.gene  <- ill_raw.gene[keep_ill_i,  , keep.lib.sizes = FALSE]
ont_indep.gene  <- ont_raw.gene[keep_ont_i,  , keep.lib.sizes = FALSE]
pb_indep.gene   <- pb_raw.gene[keep_pb_i,   , keep.lib.sizes = FALSE]
drna_indep.gene <- drna_raw.gene[keep_drna_i,, keep.lib.sizes = FALSE]

ill_dge_indep.list  <- DGE.edgeR_human(ill_indep.gene[is.human(ill_indep.gene),, keep.lib.sizes=FALSE])
ont_dge_indep.list  <- DGE.edgeR_human(ont_indep.gene[is.human(ont_indep.gene),, keep.lib.sizes=FALSE])
pb_dge_indep.list   <- DGE.edgeR_human(pb_indep.gene[is.human(pb_indep.gene),,  keep.lib.sizes=FALSE])
drna_dge_indep.list <- DGE.edgeR_human(drna_indep.gene[is.human(drna_indep.gene),, keep.lib.sizes=FALSE])

cat(sprintf("Independent filter — features per platform: Illumina=%d, ONT cDNA=%d, PacBio=%d, ONT dRNA=%d\n",
    sum(is.human(ill_indep.gene)), sum(is.human(ont_indep.gene)),
    sum(is.human(pb_indep.gene)),  sum(is.human(drna_indep.gene))))
```

```{r indep-gene-upset-build}
# Build union upset matrices: fill 0 for features not tested in a platform
build_union_mat <- function(de_lists, contrast) {
  platforms <- names(de_lists)
  de_features <- lapply(de_lists, function(dl) {
    x <- decideTests(dl[[paste0(contrast, ".DE")]], lfc = params$log2FC_cutoff)
    rownames(x)[x != 0]
  })
  all_features <- Reduce(union, de_features)
  mat <- tibble(Gene = all_features)
  for (p in platforms) {
    mat[[p]] <- as.integer(mat$Gene %in% de_features[[p]])
  }
  mat
}

indep_dge_lists <- list(
  "Illumina" = ill_dge_indep.list,
  "ONT cDNA" = ont_dge_indep.list,
  "PacBio"   = pb_dge_indep.list,
  "ONT dRNA" = drna_dge_indep.list
)

mat_list_dge_indep <- setNames(
  lapply(contrasts_dge, function(ct) build_union_mat(indep_dge_lists, ct)),
  contrasts_dge
)
```

```{r indep-tx-filter, results='hide'}
# Reload raw transcript-level objects from original RDS — before any shared filtering.
ill_raw.tx  <- bulk_obj[["ill_bulk.tx.dge"]]
ont_raw.tx  <- bulk_obj[["ont_bulk.tx.dge"]]
pb_raw.tx   <- bulk_obj[["pb_bulk.tx.dge"]]
drna_raw.tx <- bulk_obj[["drna_bulk.tx.dge"]]

# Independent filterByExpr per platform (min.count=5)
keep_ill_tx  <- filterByExpr(ill_raw.tx,  group = ill_raw.tx$samples$group,  min.count = 5)
keep_ont_tx  <- filterByExpr(ont_raw.tx,  group = ont_raw.tx$samples$group,  min.count = 5)
keep_pb_tx   <- filterByExpr(pb_raw.tx,   group = pb_raw.tx$samples$group,   min.count = 5)
keep_drna_tx <- filterByExpr(drna_raw.tx, group = drna_raw.tx$samples$group, min.count = 5)

ill_indep.tx  <- ill_raw.tx[keep_ill_tx,  , keep.lib.sizes = FALSE]
ont_indep.tx  <- ont_raw.tx[keep_ont_tx,  , keep.lib.sizes = FALSE]
pb_indep.tx   <- pb_raw.tx[keep_pb_tx,   , keep.lib.sizes = FALSE]
drna_indep.tx <- drna_raw.tx[keep_drna_tx,, keep.lib.sizes = FALSE]

# TMM normalization (before overdispersion scaling, so logCPM reflects true expression)
ill_indep.tx  <- calcNormFactors(ill_indep.tx,  method = "TMM")
ont_indep.tx  <- calcNormFactors(ont_indep.tx,  method = "TMM")
pb_indep.tx   <- calcNormFactors(pb_indep.tx,   method = "TMM")
drna_indep.tx <- calcNormFactors(drna_indep.tx, method = "TMM")

# Per-platform average logCPM for independent-filter set
logcpm_lookup <- bind_rows(lapply(
  list(Illumina   = ill_indep.tx,
       `ONT cDNA` = ont_indep.tx,
       PacBio     = pb_indep.tx,
       `ONT dRNA` = drna_indep.tx),
  function(dge) tibble(tx_id = rownames(dge), avg_logCPM = rowMeans(cpm(dge, log = TRUE)))
), .id = "Platform")

# Overdispersion scaling for DTE model
ill_indep.tx$counts  <- ill_indep.tx$counts  / ill_indep.tx$genes$Overdispersion
ont_indep.tx$counts  <- ont_indep.tx$counts  / ont_indep.tx$genes$Overdispersion
pb_indep.tx$counts   <- pb_indep.tx$counts   / pb_indep.tx$genes$Overdispersion
drna_indep.tx$counts <- drna_indep.tx$counts / drna_indep.tx$genes$Overdispersion

ill_dte_indep.list  <- DTE.edgeR_human(ill_indep.tx[is.human(ill_indep.tx),,  keep.lib.sizes=FALSE])
ont_dte_indep.list  <- DTE.edgeR_human(ont_indep.tx[is.human(ont_indep.tx),,  keep.lib.sizes=FALSE])
pb_dte_indep.list   <- DTE.edgeR_human(pb_indep.tx[is.human(pb_indep.tx),,   keep.lib.sizes=FALSE])
drna_dte_indep.list <- DTE.edgeR_human(drna_indep.tx[is.human(drna_indep.tx),, keep.lib.sizes=FALSE])

cat(sprintf("Independent filter — transcripts per platform: Illumina=%d, ONT cDNA=%d, PacBio=%d, ONT dRNA=%d\n",
    sum(is.human(ill_indep.tx)), sum(is.human(ont_indep.tx)),
    sum(is.human(pb_indep.tx)),  sum(is.human(drna_indep.tx))))
```

```{r indep-tx-upset-build}
indep_dte_lists <- list(
  "Illumina" = ill_dte_indep.list,
  "ONT cDNA" = ont_dte_indep.list,
  "PacBio"   = pb_dte_indep.list,
  "ONT dRNA" = drna_dte_indep.list
)

mat_list_dte_indep <- setNames(
  lapply(contrasts_dte, function(ct) build_union_mat(indep_dte_lists, ct)),
  contrasts_dte
)
```

## Transcript-level inconsistency — platform-independent filtering {.tabset}

The same analyses as Part 3 are repeated here on the platform-independent matrices.

```{r indep-dte-data}
# Transcript universe for independent filtering: union of all transcripts tested
# in at least one platform
indep_tx_ids <- Reduce(union, list(
  rownames(ill_indep.tx),
  rownames(ont_indep.tx),
  rownames(pb_indep.tx),
  rownames(drna_indep.tx)
))
# Re-use tx_length_lookup (covers full annotation; will NA for absent entries)
all_tx_tested_indep <- tx_length_lookup %>%
  filter(tx_id %in% indep_tx_ids, grepl("^ENS", tx_id)) %>%
  mutate(length_bin = make_length_decile_bins(Length))

# Expressed isoform count for independent filter: isoforms per gene in the
# union of all four platforms' independent-filter test sets.
expressed_iso_count_lookup <- tx_length_lookup %>%
  filter(tx_id %in% indep_tx_ids, grepl("^ENS", tx_id)) %>%
  group_by(gene_id) %>%
  mutate(expr_iso_count = n()) %>%
  ungroup() %>%
  mutate(tx_name_clean = sub("\\..*", "", tx_id_clean)) %>%
  select(tx_name_clean, expr_iso_count) %>%
  distinct()

indep_cat_df      <- bind_rows(lapply(contrasts_dte, function(ct)
  annotate_mat(mat_list_dte_indep[[ct]], ct)))
indep_pp_df       <- build_per_platform_df(mat_list_dte_indep)
indep_absence_df  <- build_absence_detection_df(mat_list_dte_indep)
```

### Platform-specific vs shared: length per platform

```{r indep-length-by-platform, fig.height=8, fig.width=14}
indep_len_plots <- plot_length_overlap(indep_cat_df, indep_pp_df, "Independent filtering")
print(indep_len_plots$by_platform)
```

### Length statistics (AvsL)

```{r indep-length-stats}
print(plot_length_stats_table(indep_pp_df, "Independent filtering —"))
```

### Annotation isoform count per platform

```{r indep-isoform-by-platform, fig.height=8, fig.width=14}
indep_iso_plots <- plot_isoform_overlap(indep_cat_df, indep_pp_df, "Independent filtering")
print(indep_iso_plots$anno_by_platform)
```

### Expressed isoform count per platform

```{r indep-expr-iso-by-platform, fig.height=8, fig.width=14}
print(indep_iso_plots$expr_by_platform)
```

### Isoform count statistics (AvsL)

```{r indep-isoform-stats}
print(plot_isoform_stats_table(indep_pp_df, "Independent filtering —"))
```

### DE rate by length bin (AvsL)

```{r indep-length-bin-avsl, fig.height=5, fig.width=18}
indep_bin_plots <- plot_length_bin_enrichment(mat_list_dte_indep, all_tx_tested_indep,
                                              "Independent filtering")
print(indep_bin_plots$avsl_prop)
```

### DE count by length bin (AvsL)

```{r indep-length-bin-count-avsl, fig.height=5, fig.width=18}
print(indep_bin_plots$avsl_count)
```

### DE rate by length bin (all contrasts)

```{r indep-length-bin-all, fig.height=9, fig.width=18}
print(indep_bin_plots$all_prop)
```

### DE count by length bin (all contrasts)

```{r indep-length-bin-count-all, fig.height=9, fig.width=18}
print(indep_bin_plots$all_count)
```

### logCPM by length bin (AvsL)

```{r indep-logcpm-bin, fig.height=6, fig.width=16}
print(indep_bin_plots$logcpm_bin)
```

### Platform-specific absence counts

```{r indep-absence-counts, fig.height=4, fig.width=10}
indep_abs_res <- plot_absence_analysis(indep_absence_df, "Independent filtering")
print(indep_abs_res$counts)
```

### Absence — length and isoform count (AvsL)

```{r indep-absence-length-iso, fig.height=9, fig.width=14}
print(indep_abs_res$length_iso)
```

### Absence — summary table (AvsL)

```{r indep-absence-table}
print(indep_abs_res$table)
```

### logCPM: platform-specific vs shared

```{r indep-logcpm-pp, fig.height=4, fig.width=12}
indep_logcpm_res <- plot_logcpm_analysis(indep_pp_df, indep_absence_df, "Independent filtering")
print(indep_logcpm_res$by_detection)
```

### logCPM statistics (AvsL)

```{r indep-logcpm-stats}
print(indep_logcpm_res$stats_table)
```

### logCPM: absence / detection / ≥2-supported

```{r indep-logcpm-absence, fig.height=5, fig.width=14}
print(indep_logcpm_res$by_absence)
```

---

## Summary comparison: shared vs independent filtering

### Mean Overlap

```{r mo-comparison}
mo_gene_indep <- sapply(contrasts_dge, function(ct)
  calc_mean_overlap(mat_list_dge_indep[[ct]]))
mo_tx_indep   <- sapply(contrasts_dte, function(ct)
  calc_mean_overlap(mat_list_dte_indep[[ct]]))

mo_compare <- bind_rows(
  tibble(Level="Gene",       Filter="Shared (AND)",
         AvsL=round(mo_gene_per_contrast["AvsL"],3), PvsL=round(mo_gene_per_contrast["PvsL"],3),
         AvsP=round(mo_gene_per_contrast["AvsP"],3), Overall=round(mean(mo_gene_per_contrast),3)),
  tibble(Level="Gene",       Filter="Independent",
         AvsL=round(mo_gene_indep["AvsL"],3), PvsL=round(mo_gene_indep["PvsL"],3),
         AvsP=round(mo_gene_indep["AvsP"],3), Overall=round(mean(mo_gene_indep),3)),
  tibble(Level="Transcript", Filter="Shared (AND)",
         AvsL=round(mo_tx_per_contrast["AvsL"],3), PvsL=round(mo_tx_per_contrast["PvsL"],3),
         AvsP=round(mo_tx_per_contrast["AvsP"],3), Overall=round(mean(mo_tx_per_contrast),3)),
  tibble(Level="Transcript", Filter="Independent",
         AvsL=round(mo_tx_indep["AvsL"],3), PvsL=round(mo_tx_indep["PvsL"],3),
         AvsP=round(mo_tx_indep["AvsP"],3), Overall=round(mean(mo_tx_indep),3))
)
knitr::kable(mo_compare,
  caption = "Mean Overlap: shared-feature vs platform-independent filtering",
  row.names = FALSE)
```

```{r mo-comparison-plot, fig.height=4, fig.width=9}
mo_compare %>%
  pivot_longer(c(AvsL, PvsL, AvsP, Overall), names_to = "Contrast", values_to = "MeanOverlap") %>%
  mutate(Contrast = factor(Contrast, levels = c("AvsL","PvsL","AvsP","Overall"))) %>%
  ggplot(aes(x = Contrast, y = MeanOverlap, fill = Level, alpha = Filter)) +
  geom_col(position = position_dodge(0.75), width = 0.7) +
  scale_fill_manual(values = c(Gene = "#4e79a7", Transcript = "#f28e2b")) +
  scale_alpha_manual(values = c("Shared (AND)" = 1.0, "Independent" = 0.55)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  geom_text(aes(label = sprintf("%.3f", MeanOverlap)),
            position = position_dodge(0.75), vjust = -0.4, size = 2.5) +
  labs(title = "Mean Overlap: shared vs independent filterByExpr",
       y = "Mean Overlap", x = "Contrast",
       fill = "Feature level", alpha = "Filtering strategy") +
  theme_minimal()
```

### Per-platform Mean Overlap — shared vs independent (transcript, AvsL)

```{r mo-platform-comparison, fig.height=4, fig.width=10}
bind_rows(
  bind_rows(lapply(contrasts_dte, function(ct) per_platform_mo(mat_list_dte[[ct]], ct))) %>%
    mutate(Filter = "Shared (AND)"),
  bind_rows(lapply(contrasts_dte, function(ct) per_platform_mo(mat_list_dte_indep[[ct]], ct))) %>%
    mutate(Filter = "Independent")
) %>%
  mutate(Platform = factor(Platform, levels = names(color_palette))) %>%
  ggplot(aes(x = Platform, y = MeanOverlap, fill = Platform, alpha = Filter)) +
  geom_col(position = position_dodge(0.7), width = 0.65) +
  facet_wrap(~ Contrast) +
  scale_fill_manual(values = color_palette) +
  scale_alpha_manual(values = c("Shared (AND)" = 1.0, "Independent" = 0.55)) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  labs(title = "Per-platform Mean Overlap: shared vs independent filtering — DTE",
       y = "Mean Overlap", x = NULL, alpha = "Filter") +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 35, hjust = 1))
```

### Length bias comparison: shared vs independent

```{r compare-length-bin, fig.height=10, fig.width=14}
compare_bin_df <- bind_rows(
  build_length_bin_df(mat_list_dte, all_tx_tested) %>% mutate(Filter = "Shared (AND)"),
  build_length_bin_df(mat_list_dte_indep, all_tx_tested_indep) %>% mutate(Filter = "Independent")
) %>%
  filter(!is.na(length_bin), Contrast == "AvsL") %>%
  group_by(Platform, length_bin, Filter) %>%
  summarise(n_DE = sum(is_DE), prop_DE = sum(is_DE) / n(), .groups = "drop") %>%
  mutate(Platform = factor(Platform, levels = names(color_palette)))

p_cmp_prop <- ggplot(compare_bin_df, aes(x = length_bin, y = prop_DE, fill = Platform, alpha = Filter)) +
  geom_col(position = position_dodge(0.7), width = 0.65) +
  facet_wrap(~ Platform, nrow = 1) +
  scale_fill_manual(values = color_palette) +
  scale_alpha_manual(values = c("Shared (AND)" = 1.0, "Independent" = 0.55)) +
  scale_y_continuous(labels = percent_format()) +
  labs(title = "DE rate by length bin: shared vs independent filtering (AvsL)",
       y = "Proportion called DE", x = "Length category", alpha = "Filter") +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))

p_cmp_count <- ggplot(compare_bin_df, aes(x = length_bin, y = n_DE, fill = Platform, alpha = Filter)) +
  geom_col(position = position_dodge(0.7), width = 0.65) +
  facet_wrap(~ Platform, nrow = 1) +
  scale_fill_manual(values = color_palette) +
  scale_alpha_manual(values = c("Shared (AND)" = 1.0, "Independent" = 0.55)) +
  labs(title = "DE count by length bin: shared vs independent filtering (AvsL)",
       y = "Number of DE transcripts", x = "Length category", alpha = "Filter") +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 30, hjust = 1))

p_cmp_prop / p_cmp_count

# logCPM by length bin (shared filter, for reference — same transcripts in both filters)
build_length_bin_df(mat_list_dte, all_tx_tested) %>%
  filter(!is.na(length_bin), !is.na(avg_logCPM), Contrast == "AvsL") %>%
  mutate(Platform = factor(Platform, levels = names(color_palette))) %>%
  ggplot(aes(x = length_bin, y = avg_logCPM, fill = Platform)) +
  geom_violin(scale = "width", alpha = 0.7, trim = TRUE, position = position_dodge(0.85)) +
  geom_boxplot(width = 0.08, outlier.shape = NA, fill = "white", alpha = 0.8,
               position = position_dodge(0.85)) +
  scale_fill_manual(values = color_palette) +
  labs(title = "Avg logCPM by length bin (AvsL) — expression confound check",
       subtitle = "If logCPM rises with length, DE-rate trend may be expression-driven",
       y = "Average logCPM", x = "Transcript length category") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```

### Platform-specific absence: shared vs independent

```{r compare-absence-counts, fig.height=5, fig.width=12}
bind_rows(
  shared_absence_df %>% filter(Category == "Platform-specific absence") %>%
    mutate(Filter = "Shared (AND)"),
  indep_absence_df  %>% filter(Category == "Platform-specific absence") %>%
    mutate(Filter = "Independent")
) %>%
  count(Platform, Contrast, Filter) %>%
  mutate(Platform = factor(Platform, levels = names(color_palette))) %>%
  ggplot(aes(x = Platform, y = n, fill = Platform, alpha = Filter)) +
  geom_col(position = position_dodge(0.7), width = 0.65) +
  geom_text(aes(label = n), position = position_dodge(0.7), vjust = -0.4, size = 2.5) +
  facet_wrap(~ Contrast) +
  scale_fill_manual(values = color_palette) +
  scale_alpha_manual(values = c("Shared (AND)" = 1.0, "Independent" = 0.55)) +
  labs(title = "Platform-specific absence: shared vs independent filtering",
       y = "Number of transcripts", x = NULL) +
  theme_minimal() +
  theme(legend.position = "none", axis.text.x = element_text(angle = 35, hjust = 1))
```

---

# Part 5: Summary evidence

The following figures directly support the key claims about cross-protocol DE consistency.

---

## 5.1 Gene vs transcript-level agreement

Mean Overlap quantifies what proportion of each platform's DE calls are independently
corroborated by ≥2 other platforms. Gene-level DE is substantially more reproducible than
transcript-level DE across all three contrasts.

```{r summary-mo, fig.height=4, fig.width=7}
ggplot(mo_long, aes(x = Contrast, y = MeanOverlap, fill = Level)) +
  geom_col(position = position_dodge(0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", MeanOverlap)),
            position = position_dodge(0.7), vjust = -0.4, size = 3) +
  scale_fill_manual(values = c(Gene = "#4e79a7", Transcript = "#f28e2b")) +
  scale_y_continuous(limits = c(0, 1), labels = percent_format()) +
  labs(title = "Mean Overlap: gene-level vs transcript-level DE",
       y = "Mean Overlap", x = "Contrast", fill = "Feature level") +
  theme_minimal()
```

---

## 5.2 Direction consistency of platform-specific DE (shared filter) {.tabset}

For DE features called significant by only one protocol, we extract the log2 fold change (logFC)
from **all four** protocols. Even when a feature fails the significance threshold in three other
protocols, its estimated effect should be in the same direction—confirming the biology is real
but below the detection threshold of those protocols.

```{r summary-logfc-fns}
# Extract logFC table from a named list of per-platform DE result lists
extract_logfc_df <- function(de_lists_named, contrast) {
  bind_rows(lapply(names(de_lists_named), function(p) {
    tbl <- de_lists_named[[p]][[paste0(contrast, ".DE")]]$table
    tibble(feature_id = rownames(tbl), logFC = tbl$logFC, Platform = p)
  })) %>%
  mutate(Platform = factor(Platform, levels = names(color_palette)))
}

# For platform-specific DE features in `mat`, join logFC from all platforms
build_specific_logfc_df <- function(mat, logfc_df, contrast_label) {
  platforms <- setdiff(names(mat), "Gene")
  mat_td    <- mat %>% mutate(TotalDetected = rowSums(select(., all_of(platforms))))
  bind_rows(lapply(platforms, function(p) {
    ids <- mat_td %>% filter(.data[[p]] == 1, TotalDetected == 1) %>% pull(Gene)
    logfc_df %>%
      filter(feature_id %in% ids) %>%
      mutate(Calling_platform = p, Contrast = contrast_label)
  })) %>%
  mutate(
    is_caller        = Platform == Calling_platform,
    Calling_platform = factor(Calling_platform, levels = names(color_palette))
  )
}

plot_direction_consistency <- function(df, title) {
  ggplot(df, aes(x = Platform, y = logFC, fill = is_caller)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray40", linewidth = 0.5) +
    geom_violin(scale = "width", alpha = 0.65, trim = TRUE) +
    geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", alpha = 0.9) +
    facet_wrap(~ Calling_platform, nrow = 1) +
    scale_fill_manual(values = c("TRUE" = "#E41A1C", "FALSE" = "#377EB8"),
                      labels = c("TRUE" = "Calling protocol", "FALSE" = "Other protocols")) +
    labs(title = title,
         subtitle = "Each panel: features DE only in that protocol; violins show logFC in all four protocols",
         y = "log2 Fold Change", x = NULL, fill = NULL) +
    theme_minimal() +
    theme(legend.position = "bottom", axis.text.x = element_text(angle = 35, hjust = 1))
}
```

### Gene level (AvsL)

```{r summary-logfc-gene, fig.height=5, fig.width=14}
logfc_gene_avsl <- extract_logfc_df(list(
  Illumina   = ill_bulk.dge_human.list,
  `ONT cDNA` = ont_bulk.dge_human.list,
  PacBio     = pb_bulk.dge_human.list,
  `ONT dRNA` = drna_bulk.deg_human.list
), "AvsL")

specific_gene_avsl <- build_specific_logfc_df(mat_list_dge[["AvsL"]], logfc_gene_avsl, "AvsL")
plot_direction_consistency(specific_gene_avsl,
  "logFC of platform-specific DE genes across all protocols — shared filter (AvsL)")
```

### Transcript level (AvsL)

```{r summary-logfc-tx, fig.height=5, fig.width=14}
logfc_tx_avsl <- extract_logfc_df(list(
  Illumina   = ill_bulk.scl_dte_human.list,
  `ONT cDNA` = ont_bulk.scl_dte_human.list,
  PacBio     = pb_bulk.scl_dte_human.list,
  `ONT dRNA` = drna_bulk.scl_dte_human.list
), "AvsL")

specific_tx_avsl <- build_specific_logfc_df(mat_list_dte[["AvsL"]], logfc_tx_avsl, "AvsL")
plot_direction_consistency(specific_tx_avsl,
  "logFC of platform-specific DE transcripts across all protocols — shared filter (AvsL)")
```

### Direction consistency summary

```{r summary-direction-table}
direction_summary <- bind_rows(
  specific_gene_avsl %>% mutate(Level = "Gene"),
  specific_tx_avsl   %>% mutate(Level = "Transcript")
) %>%
  group_by(Level, Calling_platform, feature_id) %>%
  summarise(
    calling_sign   = sign(logFC[is_caller]),
    all_consistent = all(sign(logFC) == calling_sign),
    .groups = "drop"
  ) %>%
  group_by(Level, Calling_platform) %>%
  summarise(
    n_specific   = n(),
    n_consistent = sum(all_consistent),
    pct          = round(100 * mean(all_consistent), 1),
    .groups = "drop"
  )

knitr::kable(direction_summary,
  caption = "Direction consistency: % of platform-specific DE with same logFC sign in all other protocols (AvsL, shared filter)",
  col.names = c("Level", "Calling protocol", "N platform-specific", "N all-consistent", "% consistent"),
  row.names = FALSE)
```

---

## 5.3 Transcript length capture bias (shared filter)

DE rate increases with transcript length for PacBio and ONT cDNA (full-length long-read
protocols), reflecting their superior capture of longer isoforms. This explains a substantial
fraction of protocol-specific transcript-level DE calls.

```{r summary-length-bin, fig.height=5, fig.width=18}
print(shared_bin_plots$avsl_prop)
```

---

## 5.4 Independent filtering: expression evidence in other protocols {.tabset}

Under platform-independent filtering, each protocol tests its own expressed feature set.
For features exclusively DE in one protocol, two scenarios are possible:

1. The feature is **undetectable** (near-zero count) in other protocols — it cannot be DE there.
2. The feature is **expressed but sub-threshold** in other protocols — the effect is real but weaker.

The cross-platform logCPM plot below distinguishes these: if the calling protocol shows high
logCPM while others show low/NA (not even tested), the feature is a genuine capture-bias effect.

```{r summary-indep-setup}
# logcpm_lookup in scope here is the INDEPENDENT-filter version (overwritten in indep-tx-filter)

# For each platform-specific DE transcript in the independent filter,
# pull its logCPM from ALL platforms (NA if not tested/expressed there)
build_indep_specific_cpm_df <- function(mat) {
  platforms <- setdiff(names(mat), "Gene")
  mat_td    <- mat %>% mutate(TotalDetected = rowSums(select(., all_of(platforms))))
  all_specific <- bind_rows(lapply(platforms, function(p) {
    mat_td %>% filter(.data[[p]] == 1, TotalDetected == 1) %>%
      select(Gene) %>% mutate(Calling_platform = p)
  }))
  # Left-join to logcpm_lookup: NA where not tested in a platform
  all_platform_ids <- tidyr::expand_grid(
    Gene     = all_specific$Gene,
    Platform = factor(names(color_palette), levels = names(color_palette))
  ) %>%
  left_join(all_specific, by = "Gene") %>%
  left_join(
    logcpm_lookup %>% rename(Gene = tx_id),
    by = c("Gene", "Platform")
  ) %>%
  mutate(
    is_caller        = Platform == Calling_platform,
    Calling_platform = factor(Calling_platform, levels = names(color_palette)),
    Tested           = !is.na(avg_logCPM)
  )
}

indep_specific_cpm_df <- build_indep_specific_cpm_df(mat_list_dte_indep[["AvsL"]])
```

### Cross-platform logCPM (AvsL)

```{r summary-indep-cpm, fig.height=5, fig.width=14}
ggplot(indep_specific_cpm_df, aes(x = Platform, y = avg_logCPM, fill = is_caller)) +
  geom_violin(scale = "width", alpha = 0.65, trim = TRUE, na.rm = TRUE) +
  geom_boxplot(width = 0.1, outlier.shape = NA, fill = "white", alpha = 0.9, na.rm = TRUE) +
  facet_wrap(~ Calling_platform, nrow = 1) +
  scale_fill_manual(values = c("TRUE" = "#E41A1C", "FALSE" = "#377EB8"),
                    labels = c("TRUE" = "Calling protocol", "FALSE" = "Other protocols")) +
  labs(title = "logCPM of platform-specific DE transcripts across all protocols — independent filter (AvsL)",
       subtitle = "NA (not tested) means the transcript was filtered out in that protocol; only tested features shown",
       y = "Average logCPM", x = NULL, fill = NULL) +
  theme_minimal() +
  theme(legend.position = "bottom", axis.text.x = element_text(angle = 35, hjust = 1))
```

### Testability summary (AvsL)

What fraction of a protocol's platform-specific DE transcripts were even tested in each other protocol?

```{r summary-indep-testability}
indep_specific_cpm_df %>%
  filter(!is_caller) %>%
  group_by(Calling_platform, Platform) %>%
  summarise(
    n_features = n_distinct(Gene),
    n_tested   = sum(Tested, na.rm = TRUE),
    pct_tested = round(100 * mean(Tested, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  knitr::kable(
    caption = "% of platform-specific DE transcripts (independent filter, AvsL) that were testable in each other protocol",
    col.names = c("Calling protocol", "Other protocol", "N features", "N tested in other", "% tested"),
    row.names = FALSE)
```

### logFC direction where testable (AvsL)

For the subset of features that WERE tested in other protocols, check direction consistency.

```{r summary-indep-logfc, fig.height=5, fig.width=14}
logfc_tx_indep_avsl <- extract_logfc_df(list(
  Illumina   = ill_dte_indep.list,
  `ONT cDNA` = ont_dte_indep.list,
  PacBio     = pb_dte_indep.list,
  `ONT dRNA` = drna_dte_indep.list
), "AvsL")

specific_tx_indep_avsl <- build_specific_logfc_df(
  mat_list_dte_indep[["AvsL"]], logfc_tx_indep_avsl, "AvsL")

plot_direction_consistency(specific_tx_indep_avsl,
  "logFC of platform-specific DE transcripts across tested protocols — independent filter (AvsL)") +
  labs(subtitle = "Only protocols that tested the feature are shown; NA = not in that protocol's tested set")
```

---

# Session summary

```{r save-rds}
output_list <- list(
  mo_summary          = mo_summary,
  mo_compare          = mo_compare,
  mat_list_dge        = mat_list_dge,
  mat_list_dte        = mat_list_dte,
  mat_list_dge_indep  = mat_list_dge_indep,
  mat_list_dte_indep  = mat_list_dte_indep
)
dir.create(dirname(params$output_rds), recursive = TRUE, showWarnings = FALSE)
saveRDS(output_list, params$output_rds)
cat("RDS saved to:", params$output_rds, "\n")
```

```{r session-info}
sessionInfo()
```
