---
title: "DE analysis continue"
author: "Yupei"
date: '`r Sys.Date()`'
version: 'v1.0_lfc0'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
  bulk_DE_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds"
  log2FC_cutoff: 0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 5,
	fig.width = 7,
	warning = FALSE,
	cache = FALSE
)
```

```{r Libraries, include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(ggplot2)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ggtree)
library(aplot)
library(patchwork)
output_list <- list() # used to store result for downstream analysis

# setup some conflict functions
rename <- dplyr::rename
select <- dplyr::select

knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),  # Enable caching only if cache_dir is provided
  fig.width = 15,                     # Set default figure width
  fig.height = 10,                    # Set default figure height
  echo = FALSE,                       # Hide code in the output
  message = FALSE,                    # Suppress messages in the output
  warning = FALSE,                    # Suppress warnings in the output
  fig.path = params$fig.path,         # Set figure path (NULL if not provided)
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL  # Set output devices if fig.path is provided
)

# Set cache.path only if params$cache_dir is not NULL
if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache.path = params$cache_dir)
}
```

# Setup:
```{r Load bulk DE result}
bulk_DE_rst <- readRDS(params$bulk_DE_rds)
dge_lists <- bulk_DE_rst$dge_lists
dte_lists <- bulk_DE_rst$dte_lists

# Create a mapping from old names to new platform names
name_map <- c(
  ill = "Illumina",
  ont = "ONT cDNA",
  pb = "PacBio",
  drna = "ONT dRNA"
)

# Update the names of the list using the mapping
names(dge_lists) <- name_map[names(dge_lists)]
names(dte_lists) <- name_map[names(dte_lists)]

platforms <- c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA")
comparisons <- c("AvsL", "PvsL", "AvsP")
```



Comparison of gene-level and transcript-level log fold changes (logFC) across platforms and conditions. Each scatter plot shows the relationship between gene-level logFC (x-axis) and the logFC of the most significantly differentially expressed transcript (y-axis) for genes with at least one significant transcript (P < 0.05). Points represent genes, and the dashed line indicates equality (logFC_gene = logFC_transcript). Plots are faceted by platform and pairwise comparison.
```{r,fig.height=10, fig.width=8}
scatter_plots <- list()
library(ggpubr)
for (platform in platforms) {
  for (comp in comparisons) {
    dge <- dge_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      rownames_to_column("gene_id")

    dte <- dte_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      rownames_to_column("transcript_id") %>%
      mutate(gene_id = str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))

    dte_sig <- dte %>% filter(PValue < 0.05)
    top_dte <- dte_sig %>% group_by(gene_id) %>% slice_min(PValue, n = 1) %>% ungroup()

    joined <- top_dte %>%
      inner_join(dge, by = "gene_id", suffix = c("_transcript", "_gene"))

    p <- ggplot(joined, aes(x = logFC_gene, y = logFC_transcript)) +
      geom_point(alpha = 0.2, size=0.3) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
      labs(title = paste(toupper(platform), comp), x = "Gene logFC", y = "Transcript logFC") +
      theme_minimal()

    scatter_plots[[paste(platform, comp)]] <- p
  }
}

ggarrange(plotlist = scatter_plots, ncol = 3, nrow = 4)
```


## Venn diagram showing overlap of DE genes and DE transcripts (overlap by gene name)
```{r Venn DTE vs DGE, fig.height=10, fig.width=8}
library(VennDiagram)
futile.logger::flog.threshold(futile.logger::ERROR, name = "VennDiagramLogger")
venn_grobs <- list()

for (platform in platforms) {
  for (comp in comparisons) {
    dge <- dge_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      rownames_to_column("gene_id") %>% filter(PValue < 0.05)

    dte <- dte_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      rownames_to_column("transcript_id") %>%
      mutate(gene_id = str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+")) %>%
      filter(PValue < 0.05)

  grob <- venn.diagram(
    list(Gene = dge$gene_id, Transcript = dte$gene_id),
    filename = NULL,
    output = FALSE,  # <- prevents log file
    fill = c("skyblue", "pink"),
    alpha = 0.5,
    cat.cex = 1.2,
    main = paste(toupper(platform), comp)
  )

    venn_grobs[[paste(platform, comp)]] <- grobTree(grob)
  }
}

grid.arrange(grobs = venn_grobs, ncol = 3)
```

# Log FC Correlation between platform pairs
```{r LFC correlation, fig.height=4, fig.width=8}
get_combined_logfc_plot <- function(level = "gene", p1, p2) {
  require(dplyr)
  require(ggplot2)
  require(stringr)
  require(purrr)

  dot_color <- if (level == "gene") "#B4CD55" else "#AE98EB"

  all_data <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- if (level == "gene") {
      dge_lists[[p1]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("gene_id")
    } else {
      dte_lists[[p1]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("transcript_id") %>%
        mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))
    }

    df2 <- if (level == "gene") {
      dge_lists[[p2]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("gene_id")
    } else {
      dte_lists[[p2]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column("transcript_id") %>%
        mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))
    }

    df1 <- df1 %>% filter(PValue < 0.05)
    df2 <- df2 %>% filter(PValue < 0.05)

    joined <- full_join(df1[, c("gene_id", "logFC")], 
                        df2[, c("gene_id", "logFC")],
                        by = "gene_id", suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2)

    joined$comparison <- comp
    joined
  })

  cors <- all_data %>%
    group_by(comparison) %>%
    summarise(
      r = cor(logFC_p1, logFC_p2, use = "complete.obs"),
      slope = coef(lm(logFC_p2 ~ logFC_p1))[2]
    ) %>%
    mutate(label = paste0("r = ", round(r, 2)))

  ggplot(all_data, aes(x = logFC_p1, y = logFC_p2)) +
    geom_point(color = dot_color, alpha = 0.4, size = 0.3) +
    geom_density_2d(color = "gray60", size = 0.3) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +  # identity line
    geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.3) +  # regression line
    facet_wrap(~comparison, ncol = 3) +
    geom_text(data = cors, aes(x = -Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = -0.1, vjust = 1.2, size = 3.5) +
    labs(
      title = paste0(level, ": ", p1, " vs ", p2),
      x = paste0("logFC (", p1, ")"),
      y = paste0("logFC (", p2, ")")
    ) +
    theme_minimal(base_size = 8)
}

# Arrange figures
platform_pairs <- combn(platforms, 2, simplify = FALSE)
make_6panel_plot <- function(level) {
  plotlist <- map(platform_pairs, ~ get_combined_logfc_plot(level, .x[1], .x[2]))
  ggarrange(plotlist = plotlist, ncol = 3, nrow = 2)
}

gene_logfc_combined <- make_6panel_plot("gene")
tx_logfc_combined <- make_6panel_plot("transcript")

gene_logfc_combined
tx_logfc_combined
```


# Top Expressed Features: logFC Correlation Analysis

One concern for the lower transcript-level correlations is insufficient sequencing depth across many transcript features. Here we restrict logFC correlation to the top-N most highly expressed genes or transcripts (ranked by mean logCPM across comparisons from the first platform in each pair), without any significance filter. If correlations improve markedly when restricting to highly expressed features, this implicates low-count noise rather than genuine isoform-level disagreement between platforms.

## Correlation vs expression rank cutoff
```{r cor vs topn, fig.height=6, fig.width=15}
# Helper: retrieve DE result table for a given platform / comparison / level
get_de_table <- function(platform, comp, level = "gene") {
  if (level == "gene") {
    dge_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("gene_id")
  } else {
    dte_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id")
  }
}

# Compute Pearson r of logFC for top-N expressed features across a platform pair
compute_cor_vs_topn <- function(level = "gene", p1, p2,
                                top_ns = c(100, 200, 500, 1000, 2000, 5000, NA)) {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"

  # Expression rank: mean logCPM across comparisons using p1 as reference
  expr_rank <- purrr::map_dfr(comparisons, ~ get_de_table(p1, .x, level) %>%
                                select(all_of(id_col), logCPM)) %>%
    group_by(across(all_of(id_col))) %>%
    summarise(mean_logCPM = mean(logCPM), .groups = "drop") %>%
    arrange(desc(mean_logCPM)) %>%
    mutate(rank = row_number())

  # All logFC data joined for p1 and p2, no significance filter
  all_joined <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- get_de_table(p1, comp, level) %>% select(all_of(id_col), logFC)
    df2 <- get_de_table(p2, comp, level) %>% select(all_of(id_col), logFC)
    full_join(df1, df2, by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      left_join(expr_rank %>% select(all_of(id_col), rank), by = id_col) %>%
      mutate(comparison = comp)
  })

  purrr::map_dfr(top_ns, function(n) {
    filtered <- if (is.na(n)) all_joined else dplyr::filter(all_joined, rank <= n)
    filtered %>%
      group_by(comparison) %>%
      summarise(
        r = cor(logFC_p1, logFC_p2, use = "complete.obs"),
        n_features = dplyr::n(),
        .groups = "drop"
      ) %>%
      mutate(
        top_n = if (is.na(n)) "All" else as.character(n),
        level = level,
        pair = paste(p1, "vs", p2)
      )
  })
}

top_ns <- c(100, 200, 500, 1000, 2000, 5000, 10000, NA)

cor_vs_topn <- bind_rows(
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn("gene",       .x[1], .x[2], top_ns)),
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn("transcript", .x[1], .x[2], top_ns))
) %>%
  mutate(top_n = factor(top_n, levels = c("100", "200", "500", "1000", "2000", "5000", "10000", "All")))

cor_vs_topn %>%
  group_by(pair, level, top_n) %>%
  summarise(mean_r = mean(r), .groups = "drop") %>%
  ggplot(aes(x = top_n, y = mean_r, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  labs(
    title = "logFC correlation vs expression rank cutoff (mean across comparisons)",
    x = "Top N most-expressed features",
    y = "Mean Pearson r",
    color = "Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Scatter plots: top 2000 expressed features
```{r top 2000 scatter, fig.height=4, fig.width=8}
make_top_n_scatter <- function(level, top_n = 2000) {
  id_col     <- if (level == "gene") "gene_id" else "transcript_id"
  dot_color  <- if (level == "gene") "#B4CD55" else "#AE98EB"

  plotlist <- map(platform_pairs, function(pair) {
    p1 <- pair[1]; p2 <- pair[2]

    top_ids <- purrr::map_dfr(comparisons, ~ get_de_table(p1, .x, level) %>%
                                select(all_of(id_col), logCPM)) %>%
      group_by(across(all_of(id_col))) %>%
      summarise(mean_logCPM = mean(logCPM), .groups = "drop") %>%
      slice_max(mean_logCPM, n = top_n) %>%
      pull(all_of(id_col))

    all_data <- purrr::map_dfr(comparisons, function(comp) {
      df1 <- get_de_table(p1, comp, level) %>% filter(.data[[id_col]] %in% top_ids)
      df2 <- get_de_table(p2, comp, level) %>% filter(.data[[id_col]] %in% top_ids)
      full_join(
        df1 %>% select(all_of(id_col), logFC),
        df2 %>% select(all_of(id_col), logFC),
        by = id_col, suffix = c("_p1", "_p2")
      ) %>% drop_na() %>% mutate(comparison = comp)
    })

    cors <- all_data %>%
      group_by(comparison) %>%
      summarise(r = cor(logFC_p1, logFC_p2, use = "complete.obs"), .groups = "drop") %>%
      mutate(label = paste0("r = ", round(r, 2)))

    ggplot(all_data, aes(x = logFC_p1, y = logFC_p2)) +
      geom_point(color = dot_color, alpha = 0.4, size = 0.3) +
      geom_density_2d(color = "gray60", size = 0.3) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
      geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.3) +
      facet_wrap(~comparison, ncol = 3) +
      geom_text(data = cors, aes(x = -Inf, y = Inf, label = label),
                inherit.aes = FALSE, hjust = -0.1, vjust = 1.2, size = 3.5) +
      labs(
        title = paste0(level, " top ", top_n, ": ", p1, " vs ", p2),
        x = paste0("logFC (", p1, ")"),
        y = paste0("logFC (", p2, ")")
      ) +
      theme_minimal(base_size = 8)
  })

  ggarrange(plotlist = plotlist, ncol = 3, nrow = 2)
}

make_top_n_scatter("gene",       top_n = 2000)
make_top_n_scatter("transcript", top_n = 2000)
```

## Version 2: Global mean logCPM ranking

Rather than ranking by a single platform's expression, here we compute the mean logCPM per feature averaged across **all 4 platforms × 3 comparisons** (12 values per feature). This gives a platform-neutral expression estimate for ranking, removing the winner's-curse bias of version 1.

```{r global mean logCPM rank, fig.height=6, fig.width=15}
# Mean logCPM per feature across all platforms and comparisons
get_global_expr <- function(level = "gene") {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  purrr::map_dfr(platforms, function(platform) {
    purrr::map_dfr(comparisons, function(comp) {
      get_de_table(platform, comp, level) %>%
        select(all_of(id_col), logCPM)
    })
  }) %>%
    group_by(across(all_of(id_col))) %>%
    summarise(mean_logCPM_global = mean(logCPM), .groups = "drop")
}

gene_global_expr <- get_global_expr("gene")
tx_global_expr   <- get_global_expr("transcript")

compute_cor_vs_topn_global <- function(level = "gene", p1, p2,
                                       top_ns = c(100, 200, 500, 1000, 2000, 5000, 10000, NA)) {
  id_col      <- if (level == "gene") "gene_id" else "transcript_id"
  global_expr <- if (level == "gene") gene_global_expr else tx_global_expr

  global_rank <- global_expr %>%
    arrange(desc(mean_logCPM_global)) %>%
    mutate(rank = row_number())

  all_joined <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- get_de_table(p1, comp, level) %>% select(all_of(id_col), logFC)
    df2 <- get_de_table(p2, comp, level) %>% select(all_of(id_col), logFC)
    full_join(df1, df2, by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      left_join(global_rank %>% select(all_of(id_col), rank), by = id_col) %>%
      mutate(comparison = comp)
  })

  purrr::map_dfr(top_ns, function(n) {
    filtered <- if (is.na(n)) all_joined else dplyr::filter(all_joined, rank <= n)
    filtered %>%
      group_by(comparison) %>%
      summarise(
        r          = cor(logFC_p1, logFC_p2, use = "complete.obs"),
        n_features = dplyr::n(),
        .groups    = "drop"
      ) %>%
      mutate(
        top_n = if (is.na(n)) "All" else as.character(n),
        level = level,
        pair  = paste(p1, "vs", p2)
      )
  })
}

cor_vs_topn_global <- bind_rows(
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn_global("gene",       .x[1], .x[2], top_ns)),
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn_global("transcript", .x[1], .x[2], top_ns))
) %>%
  mutate(top_n = factor(top_n, levels = c("100", "200", "500", "1000", "2000", "5000", "10000", "All")))

cor_vs_topn_global %>%
  group_by(pair, level, top_n) %>%
  summarise(mean_r = mean(r), .groups = "drop") %>%
  ggplot(aes(x = top_n, y = mean_r, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  labs(
    title = "logFC correlation vs top-N cutoff (global mean logCPM ranking, all platforms)",
    x     = "Top N features (ranked by global mean logCPM)",
    y     = "Mean Pearson r (across comparisons)",
    color = "Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Version 3: Global mean logCPM threshold filter

Instead of a rank cutoff, filter features by a minimum global mean logCPM threshold. This is more interpretable — each point on the x-axis corresponds to a biological expression level rather than an arbitrary rank.

```{r logCPM threshold filter, fig.height=6, fig.width=15}
thresholds <- seq(-2, 6, by = 1)

compute_cor_vs_threshold <- function(level = "gene", p1, p2, thresholds) {
  id_col      <- if (level == "gene") "gene_id" else "transcript_id"
  global_expr <- if (level == "gene") gene_global_expr else tx_global_expr

  all_joined <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- get_de_table(p1, comp, level) %>% select(all_of(id_col), logFC)
    df2 <- get_de_table(p2, comp, level) %>% select(all_of(id_col), logFC)
    full_join(df1, df2, by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      left_join(global_expr, by = id_col) %>%
      mutate(comparison = comp)
  })

  purrr::map_dfr(thresholds, function(thr) {
    filtered <- dplyr::filter(all_joined, mean_logCPM_global >= thr)
    filtered %>%
      group_by(comparison) %>%
      summarise(
        r          = cor(logFC_p1, logFC_p2, use = "complete.obs"),
        n_features = dplyr::n(),
        .groups    = "drop"
      ) %>%
      mutate(threshold = thr, level = level, pair = paste(p1, "vs", p2))
  })
}

cor_vs_threshold <- bind_rows(
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_threshold("gene",       .x[1], .x[2], thresholds)),
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_threshold("transcript", .x[1], .x[2], thresholds))
)

# Correlation vs threshold
cor_vs_threshold %>%
  group_by(pair, level, threshold) %>%
  summarise(mean_r = mean(r), .groups = "drop") %>%
  ggplot(aes(x = threshold, y = mean_r, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  labs(
    title = "logFC correlation vs global mean logCPM threshold",
    x     = "Minimum global mean logCPM",
    y     = "Mean Pearson r (across comparisons)",
    color = "Level"
  ) +
  theme_minimal()
```

```{r n features vs threshold, fig.height=6, fig.width=15}
# Number of features retained at each threshold
cor_vs_threshold %>%
  group_by(pair, level, threshold) %>%
  summarise(mean_n = mean(n_features), .groups = "drop") %>%
  ggplot(aes(x = threshold, y = mean_n, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  scale_y_log10() +
  labs(
    title = "Number of features retained vs global mean logCPM threshold",
    x     = "Minimum global mean logCPM",
    y     = "Mean N features (log scale)",
    color = "Level"
  ) +
  theme_minimal()
```

