---
title: "Bulk gene and transcript identification analysis (20M reads)"
author: "Yupei & Ash"
version: "v2.0"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  sirv_ercc_gtf: '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_ERCC_longSIRV_multi-fasta_20210507.gtf'
  sequins_gtf: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_annotation_2.4.gtf'
  sequins_tsv: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv'
  human_gtf: '/vast/projects/LongBench/reference_files/GRCh38/gencode.v44.annotation.gtf'
  sirv_csv: '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv'
  ont_bulk_oarfish_dir: '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/oarfish/ont_bulk/20M'
  pb_bulk_oarfish_dir: '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/oarfish/pb_bulk/20M'
  ont_drnd_oarfish_dir: '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/oarfish/dRNA_bulk/20M'
  ill_bulk_salmon_dir: '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/salmon/20M'
  Tx2Gene.map.rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Tx2Gene.map.rds"
  bulk_meta: '/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt'
  cache_dir: NULL
  fig.path: 'figures/'
  out.rds: '/vast/projects/LongBench/analysis/reports/RDS/bulk_de_human_summary.rds'
editor_options: 
  markdown: 
    wrap: 72
---

# Setup:

```{r Input filename setup, include=FALSE, eval=F}
# This chunk is used only when running the file chunk by chunk manually
params <- list(
  random_seed = 2024,
  cache_dir = NULL,
  sirv_ercc_gtf = '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_ERCC_longSIRV_multi-fasta_20210507.gtf',
  sequins_gtf = '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_annotation_2.4.gtf',
  sequins_tsv = '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv',
  human_gtf = '/vast/projects/LongBench/reference_files/GRCh38/gencode.v44.annotation.gtf',
  sirv_csv = '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv',
  ont_bulk_oarfish_dir = '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/oarfish/ont_bulk/20M',
  pb_bulk_oarfish_dir = '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/oarfish/pb_bulk/20M',
  ont_drnd_oarfish_dir = '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/oarfish/dRNA_bulk/20M',
  ill_bulk_salmon_dir = '/vast/projects/LongBench/analysis/main_workflow/result/rarefraction_analysis/salmon/20M',
  bulk_meta = "/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt",
  out.rds= '/vast/projects/LongBench/analysis/reports/RDS/bulk_de_human_summary.rds',
  Tx2Gene.map.rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Tx2Gene.map.rds"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 15, 
  fig.height = 10, 
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir), 
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"), 
                      params$cache_dir))
```

```{r Libraries,include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(ggplot2)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ggtree)
library(aplot)
library(patchwork)
source("/vast/projects/LongBench/analysis/workflow/scripts/Rfunctions.R")

# setup some conflict functions
rename <- dplyr::rename
select <- dplyr::select

# Set color palette
color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b",
  ONT_1 = "#04476c",
  ONT_2 = "#00789b",
  Illumina = "#e88b20"
)
```

```{r function for splitting the transcript origin}
is.human <- function(dge) {grepl("^ENS", rownames(dge))}
```

```{r, ge transcripts to gene map}
list2env(readRDS(params$Tx2Gene.map.rds), .GlobalEnv)
# This will load 
# "SIRV.G.Tx.map"  "sequins.G.Tx.map" "human.G.Tx.map"  
```

# Load quantification

```{r Create DGE from input, echo=F}
bulk.meta <- read.csv(params$bulk_meta)
rownames(bulk.meta) <- bulk.meta$sample

# SR salmon
ill_bulk.dge <- get_dge_from_salmon(params$ill_bulk_salmon_dir, ".*/")
# LR oarfish
ont_bulk.dge.oarfish <- get_dge_from_oarfish(params$ont_bulk_oarfish_dir, ".*/")
pb_bulk.dge.oarfish <- get_dge_from_oarfish(params$pb_bulk_oarfish_dir, ".*/")
ont_drna.dge.oarfish <- get_dge_from_oarfish(params$ont_drnd_oarfish_dir, ".*/")

# Extract human transcript counts
ill_bulk.dge <- ill_bulk.dge[is.human(ill_bulk.dge), ]
ont_bulk.dge.oarfish <- ont_bulk.dge.oarfish[is.human(ont_bulk.dge.oarfish), ]
pb_bulk.dge.oarfish <- pb_bulk.dge.oarfish[is.human(pb_bulk.dge.oarfish), ]
ont_drna.dge.oarfish <- ont_drna.dge.oarfish[is.human(ont_drna.dge.oarfish), ]
```

### 1. Aggregate transcript counts to gene level using `tximport` with unfiltered transcript counts

```{r, echo=F}
library(tximport) # install latest version from github

# Extract tx_name and gene_id columns from human and sequins
human_tx2gene <- human.G.Tx.map[, c("tx_name", "gene_id")]
sequins_tx2gene <- sequins.G.Tx.map[, c("tx_name", "gene_id")]
SIRV_tx2gene <- SIRV.G.Tx.map[, c("tx_name", "gene_id")]
combined_tx2gene <- rbind(human_tx2gene, sequins_tx2gene,SIRV_tx2gene)

# Get quant.sf file paths
get_quant_paths <- function(dirs, file_name) {
  # Generate file paths for quantification files
  quant_paths <- file.path(dirs, file_name)
  return(quant_paths)
}

# Define directories with their corresponding expected file names
oarfish_dirs <- list(
  ill_bulk = list(dir = params$ill_bulk_salmon_dir, file_name = "quant.sf"),
  ont_bulk = list(dir = params$ont_bulk_oarfish_dir, file_name = ".quant"),
  pb_bulk = list(dir = params$pb_bulk_oarfish_dir, file_name = ".quant"),
  drna_bulk = list(dir = params$ont_drnd_oarfish_dir, file_name = ".quant")
)

# Apply function to each directory to get all quant file paths
quant_paths <- lapply(oarfish_dirs, function(dir_info) {
  # Get the directory and file name from dir_info
  dir <- dir_info$dir
  file_name <- dir_info$file_name
  
  # List all subdirectories and get the quant file paths
  sub_dirs <- list.dirs(dir, full.names = TRUE, recursive = FALSE)
  get_quant_paths(sub_dirs, file_name)
})

# Flatten the list of quant_paths into a named list
quant_paths <- lapply(quant_paths, unlist)

# Function to import DGE from tximport
get_dge_from_txi <- function(quant_dir, sample_prefix_regex, type) {
  # Import data using tximport
  txi <- tximport(quant_dir,
                  type = type,
                  tx2gene = combined_tx2gene,
                  ignoreAfterBar = TRUE,
                  countsFromAbundance = "no") # raw counts
  counts <- as.data.frame(txi$counts, stringAsFactors = FALSE)
  sample_names <- gsub(sample_prefix_regex, '', basename(dirname(quant_dir)))
  colnames(counts) <- sample_names
  rst.dge <- DGEList(counts = counts)
  rst.dge$genes <- data.frame(Length=txi$length %>% rowMeans())
  rownames(rst.dge$samples) <- sub(sample_prefix_regex, '', rownames(rst.dge$samples))
  colnames(rst.dge$counts) <- sub(sample_prefix_regex, "", colnames(rst.dge$counts))
  # colnames(rst.dge$gene$Length) <- sub(sample_prefix_regex, "", colnames(rst.dge$gene$Length))
  
  rst.dge$samples <- merge(rst.dge$samples %>% select(-group), bulk.meta, by = "row.names", all.x = TRUE) %>% select(-Row.names)
  rst.dge$samples <- rst.dge$samples[match(colnames(rst.dge$counts), rst.dge$samples$sample),]
  return(rst.dge)
}

# Process each quant path for different bulk types
ill_bulk.gene.dge <- get_dge_from_txi(quant_paths$ill_bulk, ".*/", "salmon")
ont_bulk.gene.dge <- get_dge_from_txi(quant_paths$ont_bulk, ".*/", "oarfish")
pb_bulk.gene.dge <- get_dge_from_txi(quant_paths$pb_bulk, ".*/", "oarfish")
drna_bulk.gene.dge <- get_dge_from_txi(quant_paths$drna_bulk, ".*/", "oarfish")

# Extract human gene counts
ill_bulk.gene.dge <- ill_bulk.gene.dge[is.human(ill_bulk.gene.dge), ]
ont_bulk.gene.dge <- ont_bulk.gene.dge[is.human(ont_bulk.gene.dge), ]
pb_bulk.gene.dge <- pb_bulk.gene.dge[is.human(pb_bulk.gene.dge), ]
drna_bulk.gene.dge <- drna_bulk.gene.dge[is.human(drna_bulk.gene.dge), ]

# remove .* 
rownames(ill_bulk.gene.dge) <- ill_bulk.gene.dge %>% rownames %>% sub("\\..*", "", .)
rownames(ont_bulk.gene.dge) <- ont_bulk.gene.dge %>% rownames %>% sub("\\..*", "", .)
rownames(pb_bulk.gene.dge) <- pb_bulk.gene.dge %>% rownames %>% sub("\\..*", "", .)
rownames(drna_bulk.gene.dge) <- drna_bulk.gene.dge %>% rownames %>% sub("\\..*", "", .)
```

# Gene discovery analysis

## Get `gene_discovery`

Based on `filterByExpr` for each dataset, using default parameters

```{r gene identification}
# 1. Get union (superset) of genes
all_genes <- Reduce(union, lapply( # TODO: intersect for more stringent analysis?
  list(ill_bulk.gene.dge, ont_bulk.gene.dge, pb_bulk.gene.dge, drna_bulk.gene.dge),
  rownames
))

# Pad each DGE with missing genes (set to 0)
pad_dge <- function(dge, all_genes) {
  missing <- setdiff(all_genes, rownames(dge))
  if (length(missing) > 0) {
    pad <- matrix(0, nrow = length(missing), ncol = ncol(dge$counts),
                  dimnames = list(missing, colnames(dge$counts)))
    dge$counts <- rbind(dge$counts, pad)
  }
  dge[all_genes, ]
}

ill_bulk.gene.dge <- pad_dge(ill_bulk.gene.dge, all_genes)
ont_bulk.gene.dge <- pad_dge(ont_bulk.gene.dge, all_genes)
pb_bulk.gene.dge  <- pad_dge(pb_bulk.gene.dge, all_genes)
drna_bulk.gene.dge <- pad_dge(drna_bulk.gene.dge, all_genes)

# 2. Gene discovery plots
gene_discovery <- tibble(
  gene_id   = all_genes,
  Illumina  = filterByExpr(ill_bulk.gene.dge, group = ill_bulk.gene.dge$samples$group),
  `ONT cDNA`= filterByExpr(ont_bulk.gene.dge, group = ont_bulk.gene.dge$samples$group),
  PacBio    = filterByExpr(pb_bulk.gene.dge, group = pb_bulk.gene.dge$samples$group),
  `ONT dRNA`= filterByExpr(drna_bulk.gene.dge, group = drna_bulk.gene.dge$samples$group)
) %>%
  filter(Illumina | `ONT cDNA` | PacBio | `ONT dRNA`)
```

### Add GC content

#### GC% calculation at gene-level (reduced exons)

```{r Add GC% from GTF exons (hg38), message=FALSE, warning=FALSE}
library (GenomicRanges)
library(GenomicFeatures)
library(Biostrings)
#Choose the genome of interest
suppressPackageStartupMessages(library(BSgenome.Hsapiens.UCSC.hg38))
hg <- BSgenome.Hsapiens.UCSC.hg38

# Creation of TxDb from Genome
txdb <- GenomicFeatures::makeTxDbFromGFF(params$human_gtf, format = "gtf")

# Exon extraction and reduction
ex_by_gene <- GenomicFeatures::exonsBy(txdb, by = "gene")
ex_by_gene_red <- GenomicRanges::reduce(ex_by_gene)  # GRangesList

# Flatten to a single GRanges so getSeq is called once for all exons
flat_gr     <- unlist(ex_by_gene_red)
gene_labels <- names(flat_gr)

# Single vectorised getSeq + letterFrequency call, then aggregate per gene
all_seqs  <- BSgenome::getSeq(hg, flat_gr)
gc_per_exon <- letterFrequency(all_seqs, "GC", as.prob = FALSE)

tot_bases <- tapply(width(flat_gr), gene_labels, sum)
gc_counts <- tapply(gc_per_exon,    gene_labels, sum)

# Final table with GC%
gc_tbl <- tibble::tibble(
  gene_id    = names(ex_by_gene_red),
  GC_percent = as.numeric(100 * gc_counts / tot_bases)
) %>%
  dplyr::mutate(
    gene_id = sub("\\..*", "", gene_id)  
  )

#add to gene_discovery table the GC% for each gene
gene_discovery <- gene_discovery %>%
  dplyr::left_join(gc_tbl, by = "gene_id")

#assign each gene to a specific bin based on the GC%
gene_discovery <- transform(
  gene_discovery,
  gc_bin = cut(GC_percent, breaks = seq(20, 80, by = 5),
               include.lowest = TRUE, right = FALSE)
)



cat("GC% NA after count from GTF: ", sum(is.na(gene_discovery$GC_percent)), "\n")
colnames(gene_discovery)
summary(gene_discovery$GC_percent)
```

```{r GENE: reshape gene_discovery table to long format, message=FALSE}
platform_cols <- c("Illumina","ONT cDNA","PacBio","ONT dRNA")

gd_long <- gene_discovery %>%
  tidyr::pivot_longer(cols = dplyr::all_of(platform_cols),
                      names_to = "platform", values_to = "keep") %>%
  dplyr::mutate(keep = (keep==TRUE) | (keep==1),
                platform = factor(platform, levels=c("Illumina","ONT cDNA","ONT dRNA","PacBio"))) %>%
  dplyr::group_by(gene_id) %>%
  dplyr::mutate(n_detected = sum(keep, na.rm=TRUE)) %>%
  dplyr::ungroup()

```

```{r GC violin plot- platform-specific detected, fig.height=3.8, fig.width=8}
#VIOLIN PLOT: Platform-specific detected genes

plat_levels <- c("Illumina", "ONT cDNA", "ONT dRNA", "PacBio")

# platform specific genes
ps_detected_df <- gd_long %>%
  dplyr::filter(keep, n_detected == 1, !is.na(GC_percent)) %>%
  dplyr::mutate(platform = factor(platform, levels = plat_levels))


n_ps_detected <- ps_detected_df %>% dplyr::count(platform)

# violin + box plot
p_violin_ps_detected <- ggplot(ps_detected_df, aes(x = platform, y = GC_percent, fill = platform)) +
  geom_violin(trim = FALSE, alpha = 0.25, width = 0.9) +       # violin distribution
  geom_boxplot(width = 0.18, outlier.shape = NA, linewidth = 0.4) +  # embedded box
  stat_summary(fun = median, geom = "point", shape = 21, size = 2.2, 
               fill = "white", stroke = 0.4) +                 # median point
  theme_minimal() +
  labs(
    title = "GC% distribution (platform-specific detected genes)",
    subtitle = paste("N specific detected:",
                     paste(n_ps_detected$platform, n_ps_detected$n, sep = "=", collapse = "  ")),
    x = NULL,
    y = "GC%"
  ) +
  scale_fill_manual(values = color_palette[c("Illumina","ONT_1","ONT_2","PacBio")] %>% unname()) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10)
  )

# show the plot
p_violin_ps_detected

ps_detected_df %>%
  select(platform, GC_percent, platform) %>%
  group_by(platform) %>%
  summarise(
    mean = mean(GC_percent),
    median = median(GC_percent), 
    Q1 = quantile(GC_percent, 0.25),
    Q3 = quantile(GC_percent, 0.75),
    min = min(GC_percent),
    max = max(GC_percent)
  )
```


```{r r GC violin plot- platform-specific absent, fig.height=3.8, fig.width=8}
# VIOLIN PLOT: Platform-specific ABSENT genes 

# platform specific absent genes
ps_absent_df <- gd_long %>%
  dplyr::filter(!keep, n_detected == 3, !is.na(GC_percent)) %>%
  dplyr::mutate(platform = factor(platform, levels = plat_levels))


n_ps_absent <- ps_absent_df %>% dplyr::count(platform)

# Violin + Boxplot (same visual style)
p_violin_ps_absent <- ggplot(ps_absent_df, aes(x = platform, y = GC_percent, fill = platform)) +
  geom_violin(trim = FALSE, alpha = 0.25, width = 0.9) +                  # violin distribution
  geom_boxplot(width = 0.18, outlier.shape = NA, linewidth = 0.4) +       # embedded box
  stat_summary(fun = median, geom = "point", shape = 21, size = 2.2,
               fill = "white", stroke = 0.4) +                            # median point
  theme_minimal() +
  labs(
    title = "GC% distribution (platform-specific absent genes)",
    subtitle = paste("N platform-specific absent:",
                     paste(n_ps_absent$platform, n_ps_absent$n, sep = "=", collapse = "  ")),
    x = NULL,
    y = "GC%"
  ) +
  scale_fill_manual(values = color_palette[c("Illumina","ONT_1","ONT_2","PacBio")] %>% unname()) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10)
  )

# show the plot
p_violin_ps_absent

# Figure data
ps_absent_df %>%
  select(platform, GC_percent, platform) %>%
  group_by(platform) %>%
  summarise(
    mean = mean(GC_percent),
    median = median(GC_percent), 
    Q1 = quantile(GC_percent, 0.25),
    Q3 = quantile(GC_percent, 0.75),
    min = min(GC_percent),
    max = max(GC_percent)
  )
```

# Tx discovery analysis

## Get `tx_discovery`

```{r tx filtering}
all_tx <- Reduce(union, lapply(list(ill_bulk.dge, ont_bulk.dge.oarfish, pb_bulk.dge.oarfish, ont_drna.dge.oarfish), rownames))


ill_bulk.dge <- pad_dge(ill_bulk.dge, all_tx)
ont_bulk.dge.oarfish <- pad_dge(ont_bulk.dge.oarfish, all_tx)
pb_bulk.dge.oarfish <- pad_dge(pb_bulk.dge.oarfish, all_tx)
ont_drna.dge.oarfish <- pad_dge(ont_drna.dge.oarfish, all_tx)

# Filter by expression
tx_discovery <- tibble(
  tx_id = all_tx,
  "Illumina" = filterByExpr(ill_bulk.dge, group = ill_bulk.dge$samples$group, min.count=5),
  "ONT cDNA" = filterByExpr(ont_bulk.dge.oarfish, group = ont_bulk.dge.oarfish$samples$group, min.count=5),
  "PacBio" = filterByExpr(pb_bulk.dge.oarfish, group = pb_bulk.dge.oarfish$samples$group, min.count=5),
  "ONT dRNA" = filterByExpr(ont_drna.dge.oarfish, group = ont_drna.dge.oarfish$samples$group, min.count=5)
) %>% filter(
  Illumina | `ONT cDNA` | PacBio | `ONT dRNA`
)
# split and get biotype (last element of tx_id)
tx_discovery <- tx_discovery %>%  mutate(
    biotype = strsplit(tx_id, "\\|") %>% sapply(function(x) x[8]),
    length = strsplit(tx_id, "\\|") %>% sapply(function(x) x[7]) %>% as.integer(),
    gene_id = strsplit(tx_id, "\\|") %>% sapply(function(x) x[2]),
    tx_id = strsplit(tx_id, "\\|") %>% sapply(function(x) x[1])
  )

tx_discovery <- tx_discovery %>% mutate(biotype = ifelse(biotype %in% 
            names(sort(table(biotype), decreasing = TRUE)[1:6]), 
            biotype, "Others"))
```

### Add overdispersion per platform

```{r}
get_overdispersion <- function(dge) {
  dge$genes %>%
    rownames_to_column("tx_id") %>%
    select(tx_id, Overdispersion) %>%
    mutate(tx_id = sapply(strsplit(tx_id, "\\|"), `[`, 1))
}

overdispersion.ill  <- get_overdispersion(ill_bulk.dge)
overdispersion.ont  <- get_overdispersion(ont_bulk.dge.oarfish)
overdispersion.pb   <- get_overdispersion(pb_bulk.dge.oarfish)
overdispersion.drna <- get_overdispersion(ont_drna.dge.oarfish)

# Merge overdispersion into tx_discovery
tx_discovery <- tx_discovery %>%
  left_join(overdispersion.ill, by = "tx_id") %>%
  rename(Overdispersion_Illumina = Overdispersion) %>%
  left_join(overdispersion.ont, by = "tx_id") %>%
  rename(Overdispersion_ONT_cDNA = Overdispersion) %>%
  left_join(overdispersion.pb, by = "tx_id") %>%
  rename(Overdispersion_PacBio = Overdispersion) %>%
  left_join(overdispersion.drna, by = "tx_id") %>%
  rename(Overdispersion_ONT_dRNA = Overdispersion)
```

## Tx discovery plots

#### GC% calculation from transcript-level averages
```{r}
# helper for cleaning the transcript ID 
tx_clean <- function(x) sub("\\..*$", "", sub("\\|.*$", "", x))  

# each element of the list is a transcript
ex_by_tx <- GenomicFeatures::exonsBy(txdb, by = "tx", use.names = TRUE)

# clean transcript names from GTF
names(ex_by_tx) <- tx_clean(names(ex_by_tx))

# clean transcript from DGE (contains experimental data)
tx_discovery$tx_id <- tx_clean(tx_discovery$tx_id)
wanted_tx <- unique(tx_discovery$tx_id)

# controls
cat("N transcript in GTF(use.names=TRUE): ", length(ex_by_tx), "\n")
cat("Examples ENST from GTF: ", paste(utils::head(names(ex_by_tx), 5), collapse=", "), "\n")
cat("Examples ENST from DGE: ", paste(utils::head(wanted_tx, 5), collapse=", "), "\n")
cat("Intersection ENST (expected > 0): ", length(intersect(names(ex_by_tx), wanted_tx)), "\n")


common_tx <- intersect(names(ex_by_tx), wanted_tx)
ex_by_tx  <- ex_by_tx[common_tx]
cat("Transcript in GTF after matching: ", length(ex_by_tx), "\n") #present in GTF as well as in DGE

```

```{r}

ex_by_tx_red <- GenomicRanges::reduce(ex_by_tx)

flat_tx     <- unlist(ex_by_tx_red)
tx_labels   <- names(flat_tx)
all_tx_seqs <- BSgenome::getSeq(hg, flat_tx)
gc_per_exon_tx <- letterFrequency(all_tx_seqs, "GC", as.prob = FALSE)

tx_len <- tapply(width(flat_tx),      tx_labels, sum)
tx_gc  <- tapply(gc_per_exon_tx,      tx_labels, sum)

# table with GC% and length of transcript
tx_gc_tbl <- tibble::tibble(
  tx_id   = names(tx_gc),
  GC_tx   = as.numeric(100 * tx_gc / tx_len),
  L_tx_bp = as.numeric(tx_len)
)

# the transcripts ID need to be coherent between the two dataset
tx_clean <- function(x) sub("\\..*$", "", sub("\\|.*$", "", x))
tx_discovery <- tx_discovery %>% dplyr::mutate(tx_id = tx_clean(tx_id))
tx_gc_tbl    <- tx_gc_tbl    %>% dplyr::mutate(tx_id = tx_clean(tx_id))


if ("GC_percent_tx" %in% names(tx_gc_tbl))
  tx_gc_tbl <- dplyr::rename(tx_gc_tbl, GC_tx = GC_percent_tx)
if ("GC_percent" %in% names(tx_gc_tbl))
  tx_gc_tbl <- dplyr::rename(tx_gc_tbl, GC_tx = GC_percent)

cat("N tx in discovery:", nrow(tx_discovery), "\n")
cat("N tx in gc_tbl:",    nrow(tx_gc_tbl),    "\n")
cat("Possible match:",   sum(tx_discovery$tx_id %in% tx_gc_tbl$tx_id), "\n")

tx_discovery <- tx_discovery %>% dplyr::left_join(tx_gc_tbl, by = "tx_id")
stopifnot("GC_tx" %in% names(tx_discovery)) #each transcript has its own content of GC and lenght

# Join + bin
tx_discovery <- tx_discovery %>%
  dplyr::left_join(tx_gc_tbl, by = "tx_id") %>%
  dplyr::mutate(
    GC_tx     = dplyr::na_if(GC_tx.x, Inf),
    gc_bin_tx = cut(GC_tx, breaks = seq(20, 80, by = 5),
                    include.lowest = TRUE, right = FALSE)
  )

cat("With GC_tx no-NA: ", sum(!is.na(tx_discovery$GC_tx)), "\n")

```

```{r reshape tx_discovery to long format}
# Transcript-level: reshape table to long format
platform_cols <- c("Illumina","ONT cDNA","PacBio","ONT dRNA")

td_long <- tx_discovery %>%
  tidyr::pivot_longer(
    cols = dplyr::all_of(platform_cols),
    names_to = "platform",
    values_to = "keep"
  ) %>%
  dplyr::mutate(
    keep = (keep == TRUE) | (keep == 1),
    platform = factor(platform, levels = c("Illumina","ONT cDNA","ONT dRNA","PacBio"))
  ) %>%
  dplyr::group_by(tx_id) %>%
  dplyr::mutate(n_detected = sum(keep, na.rm = TRUE)) %>%
  dplyr::ungroup()

```

##### Detected and undetected GC content distribution

```{r GC transcript violin plot - platform specific detected, fig.width=8, fig.height=8}
#  VIOLIN PLOT: Transcript-level, platform-specific detected 

tx_detected_df <- td_long %>%
  dplyr::filter(keep, n_detected == 1, !is.na(GC_tx)) %>%
  dplyr::mutate(platform = factor(platform, levels = plat_levels))


n_tx_detected <- tx_detected_df %>% dplyr::count(platform)

# Violin + boxplot 
p_violin_tx_detected <- ggplot(tx_detected_df, aes(x = platform, y = GC_tx, fill = platform)) +
  geom_violin(trim = FALSE, alpha = 0.25, width = 0.9) +                  # distribution
  geom_boxplot(width = 0.18, outlier.shape = NA, linewidth = 0.4) +       # box 
  stat_summary(fun = median, geom = "point", shape = 21, size = 2.2, 
               fill = "white", stroke = 0.4) +                            
  theme_minimal() +
  labs(
    title = "GC% distribution (transcript-level, platform-specific detected)",
    subtitle = paste("N specific transcripts:",
                     paste(n_tx_detected$platform, n_tx_detected$n, sep = "=", collapse = "  ")),
    x = NULL,
    y = "GC% (transcript)"
  ) +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname()) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10)
  )

# show the plot
p_violin_tx_detected


tx_detected_df %>% select(platform, GC_tx) %>% group_by(platform) %>% summarise(
  mean = mean(GC_tx, na.rm = TRUE),
  median = median(GC_tx, na.rm = TRUE),
  Q1 = quantile(GC_tx, 0.25, na.rm = TRUE),
  Q3 = quantile(GC_tx, 0.75, na.rm = TRUE),
  min = min(GC_tx, na.rm = TRUE),
  max = max(GC_tx, na.rm = TRUE)
)
```

```{r GC transcript violin plot - platform specific absent, fig.width=8, fig.height=8}

# VIOLIN PLOT: Transcript-level, platform-specific absent 

tx_absent_df <- td_long %>%
  dplyr::filter(!keep, n_detected == 3, !is.na(GC_tx)) %>%
  dplyr::mutate(platform = factor(platform, levels = plat_levels))


n_tx_absent <- tx_absent_df %>% dplyr::count(platform)

# Violin + boxplot
p_violin_tx_absent <- ggplot(tx_absent_df, aes(x = platform, y = GC_tx, fill = platform)) +
  geom_violin(trim = FALSE, alpha = 0.25, width = 0.9) +                  
  geom_boxplot(width = 0.18, outlier.shape = NA, linewidth = 0.4) +       
  stat_summary(fun = median, geom = "point", shape = 21, size = 2.2, 
               fill = "white", stroke = 0.4) +                           
  theme_minimal() +
  labs(
    title = "GC% distribution (transcript-level, platform-specific absent)",
    subtitle = paste("N platform-specific absent:",
                     paste(n_tx_absent$platform, n_tx_absent$n, sep = "=", collapse = "  ")),
    x = NULL,
    y = "GC% (transcript)"
  ) +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname()) +
  theme(
    legend.position = "none",
    axis.text.x = element_text(size = 10)
  )


p_violin_tx_absent
tx_absent_df %>% select(platform, GC_tx) %>% group_by(platform) %>% summarise(
  mean = mean(GC_tx, na.rm = TRUE),
  median = median(GC_tx, na.rm = TRUE),
  Q1 = quantile(GC_tx, 0.25, na.rm = TRUE),
  Q3 = quantile(GC_tx, 0.75, na.rm = TRUE),
  min = min(GC_tx, na.rm = TRUE),
  max = max(GC_tx, na.rm = TRUE)
)
```

