---
title: "LongBench Quantification analysis - MiniQuant vs Oarfish"
author: "Yupei"
version: "v2.0"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL

  # Illumina + Oarfish bundle
  input_DGEobject_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"

  # IsoQuant
  isoquant_ont_bulk_rds:  "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds"
  isoquant_pb_bulk_rds:   "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds"
  isoquant_drna_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds"

  # MiniQuant
  hybrid_ont_bulk_rds:  "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.ont_bulk.rds"
  hybrid_pb_bulk_rds:   "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.pb_bulk.rds"
  hybrid_drna_bulk_rds: "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.dRNA_bulk.rds"

  # Resources
  Tx2Gene.map.rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Tx2Gene.map.rds"
  sequins_tsv: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv'
  sirv_csv: '/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv'
---

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
if (interactive()) {
  params <- list(
    input_DGEobject_rds    = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
    isoquant_ont_bulk_rds  = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds",
    isoquant_pb_bulk_rds   = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds",
    isoquant_drna_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds",
    hybrid_ont_bulk_rds    = "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.ont_bulk.rds",
    hybrid_pb_bulk_rds     = "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.pb_bulk.rds",
    hybrid_drna_bulk_rds   = "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.dRNA_bulk.rds",
    Tx2Gene.map.rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/Tx2Gene.map.rds",
    sequins_tsv = '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv',
    sirv_csv = '/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv'
  )
}

knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),
  fig.width = 15, fig.height = 10,
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.path = params$fig.path,
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL
)
if (!is.null(params$cache_dir))
  knitr::opts_chunk$set(cache.path = params$cache_dir)
```

```{r libraries, include=FALSE}
library(edgeR); library(limma)
library(tidyverse); library(ggplot2); library(patchwork)
library(ggExtra); library(gridExtra); library(scales)
library(knitr); library(glue); library(ComplexHeatmap); library(circlize)
rename <- dplyr::rename; select <- dplyr::select
```

```{r helpers}
# --- Feature-type predicates ---
is.sequins  <- function(dge) grepl("^R",           rownames(dge))
is.human    <- function(dge) grepl("^ENS",          rownames(dge))
is.sirv_all <- function(dge) grepl("^SIRV",         rownames(dge))
is.sirv_E0  <- function(dge) grepl("^SIRV\\d{3}$",  rownames(dge))
is.long_sirv<- function(dge) grepl("^SIRV\\d{4}",   rownames(dge))
is.ercc      <- function(dge) grepl("^(ERCC|DQ|EF)", rownames(dge))

# --- Normalisation helpers ---

# Within-subset log-CPM: used for spike-in scatter observed values.
# For Oarfish/IsoQuant (raw counts) and for MiniQuant (TPM), this re-normalises
# to the subset totals so the scale matches the ground truth (which is also
# normalised within the spike-in set).
# NOTE: for MiniQuant, the input TPM was computed over the full library, so we
# must re-normalise within the subset here — log2(TPM + 1) would be wrong.
lcpm_within <- function(x) log2(x * 1e6 / sum(x) + 1)

# Illumina: length-adjusted within-subset CPM
ltcpm_within <- function(counts, eff_len) {
  rpk <- counts / eff_len
  log2(rpk * 1e6 / sum(rpk) + 1)
}

# Full-library log2(CPM+1) — for MDS / correlation
log_cpm <- function(dge)
  log2(t(t(dge$counts) / colSums(dge$counts)) * 1e6 + 1)

# Full-library log2(TPM+1) for Illumina (length-adjusted)
log_tpm_illumina <- function(dge) {
  len <- dge$genes$EffectiveLength
  tpm <- apply(dge$counts, 2, function(col) {
    rpk <- col / len; rpk / sum(rpk) * 1e6
  })
  log2(tpm + 1)
}
```

```{r colors, include=FALSE}
tool_colors <- c(
  Oarfish   = "#2166AC",
  IsoQuant  = "#4DAC26",
  MiniQuant = "#D6604D",
  Illumina  = "#e88b20"
)
platform_colors <- c(
  PacBio     = "#df1995",
  "ONT dRNA" = "#00789b",
  "ONT cDNA" = "#04476c",
  Illumina   = "#e88b20"
)
```

# Load data

```{r load-data, echo=TRUE, results='hide'}
safe_rds <- function(p) { if (!is.null(p) && file.exists(p)) readRDS(p) else NULL }

# Illumina (Salmon) + Oarfish
bundle       <- readRDS(params$input_DGEobject_rds)
ill_tx.dge   <- bundle$ill_bulk.tx.dge
ont_oar.dge  <- bundle$ont_bulk.tx.dge
pb_oar.dge   <- bundle$pb_bulk.tx.dge
drna_oar.dge <- bundle$drna_bulk.tx.dge

# IsoQuant
isq_ont  <- safe_rds(params$isoquant_ont_bulk_rds)
isq_pb   <- safe_rds(params$isoquant_pb_bulk_rds)
isq_drna <- safe_rds(params$isoquant_drna_bulk_rds)
ont_isq.dge  <- isq_ont$ont_bulk.tx.dge
pb_isq.dge   <- isq_pb$pb_bulk.tx.dge
drna_isq.dge <- isq_drna$dRNA_bulk.tx.dge

# MiniQuant — $counts holds full-library TPM; versioned transcript IDs
mq_ont  <- safe_rds(params$hybrid_ont_bulk_rds)
mq_pb   <- safe_rds(params$hybrid_pb_bulk_rds)
mq_drna <- safe_rds(params$hybrid_drna_bulk_rds)

strip_version <- function(dge) {
  rownames(dge) <- sub("\\.\\d+$", "", rownames(dge))
  dge
}

ill_tx.dge <- strip_version(ill_tx.dge)
ont_oar.dge <- strip_version(ont_oar.dge)
pb_oar.dge <- strip_version(pb_oar.dge)
drna_oar.dge <- strip_version(drna_oar.dge)
ont_isq.dge <- strip_version(ont_isq.dge)
pb_isq.dge <- strip_version(pb_isq.dge)
drna_isq.dge <- strip_version(drna_isq.dge)
ont_mq.dge  <- strip_version(mq_ont$ont_bulk.TPM.tx.dge)
pb_mq.dge   <- strip_version(mq_pb$pb_bulk.TPM.tx.dge)
drna_mq.dge <- strip_version(mq_drna$dRNA_bulk.TPM.tx.dge)

list2env(readRDS(params$Tx2Gene.map.rds), .GlobalEnv)
```

# Feature selection

Expressed transcripts are defined by `filterByExpr` with a minimum count of 5 in at least one sample, applied to the subset of transcripts common to all tools. This ensures that downstream analyses (MDS, correlation) are based on a consistent set of expressed features across all datasets. The same filtering is applied to all DGEs to maintain comparability.

```{r feature-selection, echo=TRUE}
common_tx <- Reduce(intersect, lapply(
  list(
    ill_tx.dge,
    ont_oar.dge,
     pb_oar.dge, 
     drna_oar.dge, 
     ont_isq.dge, 
     pb_isq.dge, 
     drna_isq.dge, 
     ont_mq.dge, 
     pb_mq.dge, 
     drna_mq.dge
  ), rownames
))

ill_tx.dge <- ill_tx.dge[common_tx, ]
ont_oar.dge <- ont_oar.dge[common_tx, ]
pb_oar.dge <- pb_oar.dge[common_tx, ]
drna_oar.dge <- drna_oar.dge[common_tx, ]
ont_isq.dge <- ont_isq.dge[common_tx, ]
pb_isq.dge <- pb_isq.dge[common_tx, ]
drna_isq.dge <- drna_isq.dge[common_tx, ]
ont_mq.dge <- ont_mq.dge[common_tx, ]
pb_mq.dge <- pb_mq.dge[common_tx, ]
drna_mq.dge <- drna_mq.dge[common_tx, ]



keep <- #
  filterByExpr(ill_tx.dge, group = ill_tx.dge$samples$group, min.count = 5) &
    # filterByExpr(ont_oar.dge, group = ont_oar.dge$samples$group, min.count = 5) &
    # filterByExpr(pb_oar.dge, group = pb_oar.dge$samples$group, min.count = 5) &
    # filterByExpr(drna_oar.dge, group = drna_oar.dge$samples$group, min.count = 5) &
    filterByExpr(ont_isq.dge, group = ont_isq.dge$samples$group, min.count = 5) &
    filterByExpr(pb_isq.dge, group = pb_isq.dge$samples$group, min.count = 5) &
    filterByExpr(drna_isq.dge, group = drna_isq.dge$samples$group, min.count = 5)
  # Can't apply filterByExpr to MiniQuant TPM DGE in the same way — it expects raw counts and the scale is different, so we skip it here. 
  # filterByExpr(ont_mq.dge, group = ont_mq.dge$samples$group, min.count = 5) &
  # filterByExpr(pb_mq.dge, group = pb_mq.dge$samples$group, min.count = 5) &
  # filterByExpr(drna_mq.dge, group = drna_mq.dge$samples$group, min.count = 5)

table(keep)

expressed_tx <- common_tx[keep]
message(glue("Expressed transcripts: {length(expressed_tx)}"))

ill_tx.dge   <- ill_tx.dge[keep, ]
ont_oar.dge  <- ont_oar.dge[keep, ]
pb_oar.dge   <- pb_oar.dge[keep, ]
drna_oar.dge <- drna_oar.dge[keep, ]
ont_isq.dge  <- ont_isq.dge[keep, ]
pb_isq.dge <- pb_isq.dge[keep, ]
drna_isq.dge <- drna_isq.dge[keep, ]
ont_mq.dge <- ont_mq.dge[keep, ]
pb_mq.dge <- pb_mq.dge[keep, ]
drna_mq.dge <- drna_mq.dge[keep, ]

```

# Part 1: Spike-in quantification
For MiniQuant: although `$counts` are full-library TPM, working on a spike-in subset
requires re-normalising within that subset (same as for the other tools and consistent
with the ground truth normalisation).

```{r spikein-setup}
# Length lookup from Oarfish reference (which carries full genes metadata)
get_ref_length <- function(ids) {
  ont_oar.dge$genes$Length[match(ids, rownames(ont_oar.dge))]
}
```

## Sequins: expected vs observed
```{r sequins-scatter, fig.width=25, fig.height=9}
sequins.gt <- read_tsv(params$sequins_tsv, show_col_types = FALSE)
# Normalise a subset DGE to within-subset log-CPM
# is_illumina = TRUE: length-adjusted (EffectiveLength from genes slot)
# is_illumina = FALSE: plain log-CPM within subset (works for both raw counts and TPM)
norm_subset <- function(dge, filter_fn, is_illumina = FALSE) {
  sub <- dge[filter_fn(dge), ]
  if (nrow(sub) == 0) return(NULL)
  if (is_illumina) {
    eff <- sub$genes$EffectiveLength
    apply(sub$counts, 2, function(col) ltcpm_within(col, eff))
  } else {
    apply(sub$counts, 2, lcpm_within)
  }
}

make_sequins_scatter <- function(dge, label, is_illumina = FALSE) {
  norm_mat <- norm_subset(dge, is.sequins, is_illumina)
  if (is.null(norm_mat)) return(NULL)

  get_lcpm <- function(x) log2(x * 1e6 / sum(x) + 1)
  gt <- read_tsv(params$sequins_tsv, show_col_types = FALSE) %>%
    mutate(across(starts_with("MIX"), get_lcpm)) %>%
    select(NAME, MIX_A, MIX_B, LENGTH) %>%
    mutate(N_tx_per_gene = as.numeric(gsub("^R.*_", "", NAME)))

  sp <- dge[is.sequins(dge), ]$samples
  norm_df <- norm_mat %>% as_tibble(rownames = "NAME")

  combined <- bind_rows(
    norm_df %>% select(-sp$sample[sp$sequins == "B"]) %>%
      left_join(gt, by = "NAME") %>% mutate(GT = MIX_A) %>% select(-MIX_A, -MIX_B) %>%
      pivot_longer(sp$sample[sp$sequins == "A"], names_to = "Sample", values_to = "Obs"),
    norm_df %>% select(-sp$sample[sp$sequins == "A"]) %>%
      left_join(gt, by = "NAME") %>% mutate(GT = MIX_B) %>% select(-MIX_A, -MIX_B) %>%
      pivot_longer(sp$sample[sp$sequins == "B"], names_to = "Sample", values_to = "Obs")
  ) %>% mutate(`Tx length` = cut(LENGTH,
      breaks = quantile(LENGTH, probs = seq(0,1,.25)), include.lowest = TRUE,
      labels = c("Q1","Q2","Q3","Q4")))

  r   <- cor(combined$GT, combined$Obs, use = "complete")
  mse <- mean((combined$GT - combined$Obs)^2, na.rm = TRUE)

  ggplot(combined[sample(nrow(combined)), ], aes(x = Obs, y = GT, color = `Tx length`)) +
    geom_point(size = 0.8, alpha = 0.7) + geom_abline(linetype = "dotted") +
    theme_classic() +
    labs(title = label, x = "Observed log2(TPM+1)", y = "Expected log2(TPM+1)") +
    annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5,
             label = glue("r={round(r,2)}\nMSE={round(mse,2)}"), size = 3) +
    scale_color_viridis_d() + xlim(0, 20) + ylim(0, 20)
}

plots <- Filter(Negate(is.null), list(
  make_sequins_scatter(ill_tx.dge,      "Illumina",         is_illumina = TRUE),
  patchwork::plot_spacer(),
  # make_sequins_scatter(ont_oar.dge,  "ONT cDNA (Oarfish)"),
  make_sequins_scatter(ont_isq.dge,  "ONT cDNA (IsoQuant)"),
  make_sequins_scatter(ont_mq.dge,   "ONT cDNA (MiniQuant)"),
  # make_sequins_scatter(drna_oar.dge,  "ONT dRNA (Oarfish)"),
  make_sequins_scatter(drna_isq.dge,  "ONT dRNA (IsoQuant)"),
  make_sequins_scatter(drna_mq.dge,   "ONT dRNA (MiniQuant)"),
  # make_sequins_scatter(pb_oar.dge,   "PacBio (Oarfish)"),
  make_sequins_scatter(pb_isq.dge,   "PacBio (IsoQuant)"),
  make_sequins_scatter(pb_mq.dge,    "PacBio (MiniQuant)")

))
patchwork::wrap_plots(plots, ncol = 4, guides = "collect", byrow = FALSE)
```

```{r sequins-raw-data, echo=TRUE}
get_sequins_data <- function(dge, label, is_illumina = FALSE) {
  norm_mat <- norm_subset(dge, is.sequins, is_illumina)
  if (is.null(norm_mat)) return(NULL)

  get_lcpm <- function(x) log2(x * 1e6 / sum(x) + 1)
  gt <- read_tsv(params$sequins_tsv, show_col_types = FALSE) %>%
    mutate(across(starts_with("MIX"), get_lcpm)) %>%
    select(NAME, MIX_A, MIX_B, LENGTH) %>%
    mutate(N_tx_per_gene = as.numeric(gsub("^R.*_", "", NAME)))

  sp <- dge[is.sequins(dge), ]$samples
  norm_df <- norm_mat %>% as_tibble(rownames = "NAME")

  bind_rows(
    norm_df %>% select(-sp$sample[sp$sequins == "B"]) %>%
      left_join(gt, by = "NAME") %>% mutate(GT = MIX_A) %>% select(-MIX_A, -MIX_B) %>%
      pivot_longer(sp$sample[sp$sequins == "A"], names_to = "Sample", values_to = "Obs"),
    norm_df %>% select(-sp$sample[sp$sequins == "A"]) %>%
      left_join(gt, by = "NAME") %>% mutate(GT = MIX_B) %>% select(-MIX_A, -MIX_B) %>%
      pivot_longer(sp$sample[sp$sequins == "B"], names_to = "Sample", values_to = "Obs")
  ) %>% mutate(
    label = label,
    `Tx length` = cut(LENGTH, breaks = quantile(LENGTH, probs = seq(0, 1, .25)),
                      include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4"))
  )
}

sequins_raw <- bind_rows(
  get_sequins_data(ill_tx.dge,   "Illumina",              is_illumina = TRUE),
  get_sequins_data(ont_isq.dge,  "ONT cDNA (IsoQuant)"),
  get_sequins_data(ont_mq.dge,   "ONT cDNA (MiniQuant)"),
  get_sequins_data(drna_isq.dge, "ONT dRNA (IsoQuant)"),
  get_sequins_data(drna_mq.dge,  "ONT dRNA (MiniQuant)"),
  get_sequins_data(pb_isq.dge,   "PacBio (IsoQuant)"),
  get_sequins_data(pb_mq.dge,    "PacBio (MiniQuant)")
)

glimpse(sequins_raw)
```

## ERCC: expected vs observed
```{r ercc-scatter, fig.width=25, fig.height=9}
make_ercc_scatter <- function(dge, label, is_illumina = FALSE) {
  norm_mat <- norm_subset(dge, is.ercc, is_illumina)
  if (is.null(norm_mat)) return(NULL)

  get_lcpm <- function(x) log2(x * 1e6 / sum(x) + 1)
  ercc_gt <- read_csv(params$sirv_csv, show_col_types = FALSE) %>%
    filter(grepl("ERCC", txid)) %>%
    mutate(GT = get_lcpm(conc), NAME = GenBank,
           Length = get_ref_length(GenBank)) %>%
    select(NAME, GT, Length)

  combined <- norm_mat %>% as_tibble(rownames = "NAME") %>%
    left_join(ercc_gt, by = "NAME") %>%
    pivot_longer(colnames(dge[is.ercc(dge), ]$counts),
                 names_to = "Sample", values_to = "Obs") %>%
    mutate(`Tx length` = cut(Length,
      breaks = quantile(Length, probs = seq(0,1,.25), na.rm = TRUE),
      include.lowest = TRUE, labels = c("Q1","Q2","Q3","Q4")))

  r   <- cor(combined$GT, combined$Obs, use = "complete")
  mse <- mean((combined$GT - combined$Obs)^2, na.rm = TRUE)
  ggplot(combined[sample(nrow(combined)), ], aes(x = Obs, y = GT, color = `Tx length`)) +
    geom_point(size = 0.8, alpha = 0.7) + geom_abline(linetype = "dotted") +
    theme_classic() +
    labs(title = label, x = "Observed log2(TPM+1)", y = "Expected log2(TPM+1)") +
    annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5,
             label = glue("r={round(r,2)}\nMSE={round(mse,2)}"), size = 3) +
    scale_color_viridis_d() + xlim(10, 20) + ylim(10, 20)
}


plots <- Filter(Negate(is.null), list(
  make_ercc_scatter(ill_tx.dge,      "Illumina",         is_illumina = TRUE),
  patchwork::plot_spacer(),
  # make_ercc_scatter(ont_oar.dge,  "ONT cDNA (Oarfish)"),
  make_ercc_scatter(ont_isq.dge,  "ONT cDNA (IsoQuant)"),
  make_ercc_scatter(ont_mq.dge,   "ONT cDNA (MiniQuant)"),
  # make_ercc_scatter(drna_oar.dge,  "ONT dRNA (Oarfish)"),
  make_ercc_scatter(drna_isq.dge,  "ONT dRNA (IsoQuant)"),
  make_ercc_scatter(drna_mq.dge,   "ONT dRNA (MiniQuant)"),
  # make_ercc_scatter(pb_oar.dge,   "PacBio (Oarfish)"),
  make_ercc_scatter(pb_isq.dge,   "PacBio (IsoQuant)"),
  make_ercc_scatter(pb_mq.dge,    "PacBio (MiniQuant)")

))
patchwork::wrap_plots(plots, ncol =4, guides = "collect", byrow = FALSE)
```

```{r ercc-raw-data, echo=TRUE}
get_ercc_data <- function(dge, label, is_illumina = FALSE) {
  norm_mat <- norm_subset(dge, is.ercc, is_illumina)
  if (is.null(norm_mat)) return(NULL)

  get_lcpm <- function(x) log2(x * 1e6 / sum(x) + 1)
  ercc_gt <- read_csv(params$sirv_csv, show_col_types = FALSE) %>%
    filter(grepl("ERCC", txid)) %>%
    mutate(GT = get_lcpm(conc), NAME = GenBank,
           Length = get_ref_length(GenBank)) %>%
    select(NAME, GT, Length)

  norm_mat %>% as_tibble(rownames = "NAME") %>%
    left_join(ercc_gt, by = "NAME") %>%
    pivot_longer(colnames(dge[is.ercc(dge), ]$counts),
                 names_to = "Sample", values_to = "Obs") %>%
    mutate(
      label = label,
      `Tx length` = cut(Length, breaks = quantile(Length, probs = seq(0, 1, .25), na.rm = TRUE),
                        include.lowest = TRUE, labels = c("Q1", "Q2", "Q3", "Q4"))
    )
}

ercc_raw <- bind_rows(
  get_ercc_data(ill_tx.dge,   "Illumina",              is_illumina = TRUE),
  get_ercc_data(ont_isq.dge,  "ONT cDNA (IsoQuant)"),
  get_ercc_data(ont_mq.dge,   "ONT cDNA (MiniQuant)"),
  get_ercc_data(drna_isq.dge, "ONT dRNA (IsoQuant)"),
  get_ercc_data(drna_mq.dge,  "ONT dRNA (MiniQuant)"),
  get_ercc_data(pb_isq.dge,   "PacBio (IsoQuant)"),
  get_ercc_data(pb_mq.dge,    "PacBio (MiniQuant)")
)

if !is.null(params$fig.path) {
  write_csv(ercc_raw, file.path(params$fig.path, "ercc_raw_data.csv"))
}
```

## SIRV: transcript length vs observed abundance
```{r sirv-scatter, fig.width=25, fig.height=9}
make_sirv_scatter <- function(dge, label, is_illumina = FALSE) {
  norm_mat <- norm_subset(dge, is.sirv_all, is_illumina)
  if (is.null(norm_mat)) return(NULL)

  sub_sirv <- dge[is.sirv_all(dge), ]
  get_lcpm <- function(x) log2(x * 1e6 / sum(x) + 1)
  sirv_gt <- tibble(
    NAME    = rownames(sub_sirv),
    LENGTH  = get_ref_length(rownames(sub_sirv)),
    is_long = is.long_sirv(sub_sirv),
    GT      = get_lcpm(rep(1, nrow(sub_sirv)))
  ) %>% mutate(N_tx = SIRV.G.Tx.map$tx_count[match(NAME, SIRV.G.Tx.map$tx_name)])

  combined <- norm_mat %>% as_tibble(rownames = "NAME") %>%
    left_join(sirv_gt, by = "NAME") %>%
    pivot_longer(colnames(sub_sirv), names_to = "Sample", values_to = "Obs")

  SD  <- sd(combined$Obs, na.rm = TRUE)
  MAD <- mad(combined$Obs, na.rm = TRUE)
  MSE <- mean((combined$GT - combined$Obs)^2, na.rm = TRUE)

  p <- ggplot(combined[sample(nrow(combined)), ], aes(y = LENGTH, x = Obs, color = is_long)) +
    geom_point(size = 0.8, alpha = 0.7) +
    geom_vline(xintercept = combined$GT[1], linetype = "dashed", color = "red") +
    theme_classic() +
    labs(title = label, y = "Transcript length", x = "Observed log2(CPM+1)") +
    annotate("text", x = 0, y = 12000, hjust = 0, vjust = 1, size = 3,
             label = glue("SD={round(SD,2)}\nMAD={round(MAD,2)}\nMSE={round(MSE,2)}")) +
    scale_color_viridis_d() + xlim(0, 20)
  ggMarginal(p, type = "density", margins = "x", fill = NA, size = 2,
             groupColour = TRUE, groupFill = TRUE)
}

plots <- Filter(Negate(is.null), list(
  make_sirv_scatter(ill_tx.dge,      "Illumina",         is_illumina = TRUE),
  patchwork::plot_spacer(),
  # make_sirv_scatter(ont_oar.dge,  "ONT cDNA (Oarfish)"),
  make_sirv_scatter(ont_isq.dge,  "ONT cDNA (IsoQuant)"),
  make_sirv_scatter(ont_mq.dge,   "ONT cDNA (MiniQuant)"),
  # make_sirv_scatter(drna_oar.dge,  "ONT dRNA (Oarfish)"),
  make_sirv_scatter(drna_isq.dge,  "ONT dRNA (IsoQuant)"),
  make_sirv_scatter(drna_mq.dge,   "ONT dRNA (MiniQuant)"),
  # make_sirv_scatter(pb_oar.dge,   "PacBio (Oarfish)"),
  make_sirv_scatter(pb_isq.dge,   "PacBio (IsoQuant)"),
  make_sirv_scatter(pb_mq.dge,    "PacBio (MiniQuant)")
))
patchwork::wrap_plots(plots, ncol = 4, guides = "collect", byrow = FALSE)
```

```{r sirv-raw-data, echo=TRUE}
get_sirv_data <- function(dge, label, is_illumina = FALSE) {
  norm_mat <- norm_subset(dge, is.sirv_all, is_illumina)
  if (is.null(norm_mat)) return(NULL)

  sub_sirv <- dge[is.sirv_all(dge), ]
  get_lcpm <- function(x) log2(x * 1e6 / sum(x) + 1)
  sirv_gt <- tibble(
    NAME    = rownames(sub_sirv),
    LENGTH  = get_ref_length(rownames(sub_sirv)),
    is_long = is.long_sirv(sub_sirv),
    GT      = get_lcpm(rep(1, nrow(sub_sirv)))
  ) %>% mutate(N_tx = SIRV.G.Tx.map$tx_count[match(NAME, SIRV.G.Tx.map$tx_name)])

  norm_mat %>% as_tibble(rownames = "NAME") %>%
    left_join(sirv_gt, by = "NAME") %>%
    pivot_longer(colnames(sub_sirv), names_to = "Sample", values_to = "Obs") %>%
    mutate(label = label)
}

sirv_raw <- bind_rows(
  get_sirv_data(ill_tx.dge,   "Illumina",              is_illumina = TRUE),
  get_sirv_data(ont_isq.dge,  "ONT cDNA (IsoQuant)"),
  get_sirv_data(ont_mq.dge,   "ONT cDNA (MiniQuant)"),
  get_sirv_data(drna_isq.dge, "ONT dRNA (IsoQuant)"),
  get_sirv_data(drna_mq.dge,  "ONT dRNA (MiniQuant)"),
  get_sirv_data(pb_isq.dge,   "PacBio (IsoQuant)"),
  get_sirv_data(pb_mq.dge,    "PacBio (MiniQuant)")
)

if (!is.null(params$fig.path)) {
  write_csv(sirv_raw, file.path(params$fig.path, "sirv_raw_data.csv"))
}
```

### Per-sample correlation table
```{r cor-table}
get_cor_table <- function(oar_dge, mq_dge, platform) {
  common <- intersect(rownames(oar_dge), rownames(mq_dge))
  common <- common[grepl("^ENS", common)]
  if (length(common) == 0) return(NULL)
  oar_log <- log_cpm(oar_dge[common, ])
  mq_log  <- log2(mq_dge[common, ]$counts + 1)
  tibble(platform = platform,
         sample = colnames(oar_log),
         r = sapply(colnames(oar_log), function(s) cor(oar_log[, s], mq_log[, s])))
}
bind_rows(
  get_cor_table(ont_oar.dge,  ont_mq.dge,  "ONT cDNA"),
  get_cor_table(pb_oar.dge,   pb_mq.dge,   "PacBio"),
  get_cor_table(drna_oar.dge, drna_mq.dge, "ONT dRNA")
) %>% knitr::kable(digits = 3)
```

# Part 3: MDS — Illumina + Oarfish + MiniQuant

MDS computed on human transcripts common to all tools. Illumina: log2(TPM+1);
long-read tools: log2(CPM+1).

```{r mds-prep}
build_log_mat <- function(dge, is_illumina = FALSE, tool = "Oarfish", platform = "Unknown") {
  mat <- if (is_illumina) log_tpm_illumina(dge) else log_cpm(dge)
  colnames(mat) <- paste0(tool, "_", colnames(mat))
  list(mat = mat, samples = dge$samples %>% mutate(tool = tool, platform = platform))
}

parts <- list(
  build_log_mat(ill_tx.dge,    is_illumina = TRUE, tool = "Illumina", platform = "Illumina"),
  build_log_mat(ont_oar.dge,   tool = "Oarfish", platform = "ONT cDNA"),
  build_log_mat(ont_mq.dge,    tool = "MiniQuant", platform = "ONT cDNA"),
  build_log_mat(pb_oar.dge,    tool = "Oarfish", platform = "PacBio"),
  build_log_mat(pb_mq.dge,     tool = "MiniQuant", platform = "PacBio"),
  build_log_mat(drna_oar.dge,  tool = "Oarfish", platform = "ONT dRNA"),
  build_log_mat(drna_mq.dge,   tool = "MiniQuant", platform = "ONT dRNA")
)

log_mat_all <- do.call(cbind, lapply(parts, `[[`, "mat"))
samples_all <- bind_rows(lapply(parts, `[[`, "samples"))
```

```{r mds-plot, fig.height=4, fig.width=10}
mds <- plotMDS(log_mat_all, plot = FALSE)
mds_df <- data.frame(
  Dim1 = mds$x, 
  Dim2 = mds$y,
  group    = samples_all$group,
  platform = samples_all$platform,
  tool     = samples_all$tool,
  sample   = samples_all$sample
)

p1 <- ggplot(mds_df, aes(Dim1, Dim2, color = tool, shape = platform)) +
  geom_point(size = 3, alpha = 0.8) +
  scale_color_manual(values = tool_colors) +
  labs(title = "MDS: coloured by tool",
       x = glue("Dim1 ({round(mds$var.explained[1]*100,1)}%)"),
       y = glue("Dim2 ({round(mds$var.explained[2]*100,1)}%)")) +
  theme_minimal()

p2 <- ggplot(mds_df, aes(Dim1, Dim2, color = group, shape = tool)) +
  geom_point(size = 3, alpha = 0.8) + scale_color_viridis_d() +
  labs(title = "MDS: coloured by cell line",
       x = glue("Dim1 ({round(mds$var.explained[1]*100,1)}%)"),
       y = glue("Dim2 ({round(mds$var.explained[2]*100,1)}%)")) +
  theme_minimal()

p1 | p2
```

# Part 4: Sample-wise correlation heatmap
```{r heatmap, fig.width=14, fig.height=5}
cor_mat <- cor(log_mat_all, method = "pearson")
col_fun <- colorRamp2(c(0.5, 0.75, 1), c("#2166AC", "white", "#B2182B"))

ht_list <- lapply(unique(samples_all$sample), function(sid) {
  cols <- which(samples_all$sample == sid)
  mat  <- cor_mat[cols, cols]
  lbls <- paste0(samples_all$tool[cols], "\n", samples_all$platform[cols])
  rownames(mat) <- colnames(mat) <- lbls
  Heatmap(mat, name = as.character(sid), col = col_fun,
          cluster_rows = TRUE, cluster_columns = TRUE,
          show_row_names = TRUE, show_column_names = TRUE,
          row_dend_width = unit(4,"mm"), column_dend_height = unit(4,"mm"),
          row_names_gp = gpar(fontsize = 7), column_names_gp = gpar(fontsize = 7))
})

ht_grobs <- lapply(ht_list, function(ht) grid.grabExpr(draw(ht, merge_legend = TRUE)))
grid.arrange(grobs = ht_grobs, nrow = 2, ncol = 4,
             top = grid::textGrob("Sample-wise Pearson correlation (log2 CPM/TPM+1)",
                                   gp = grid::gpar(fontsize = 14, fontface = "bold")))
```
