---
title: "Longbench Cross-tool Long-read Quantification Comparison"
author: "Yupei"
version: "v1.0"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL

  # input oarfish/salmon
  input_DGEobject_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"

  # input isoquant
  isoquant_ont_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds"
  isoquant_pb_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds"
  isoquant_drna_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds"

  # input miniquant
  hybrid_ont_bulk_rds: "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.ont_bulk.rds"
  hybrid_pb_bulk_rds: "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.pb_bulk.rds"
  hybrid_drna_bulk_rds: "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.dRNA_bulk.rds"

  sequins_tsv: "/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv"
  sirv_csv: "/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv"
---

# Setup

```{r setup, include=FALSE, message=FALSE, warning=FALSE}
if (interactive()) {
  params <- list(
    random_seed = 2024,
    cache_dir = NULL,
    fig.path = NULL,
    input_DGEobject_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
    isoquant_ont_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds",
    isoquant_pb_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds",
    isoquant_drna_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds",
    hybrid_ont_bulk_rds = "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.ont_bulk.rds",
    hybrid_pb_bulk_rds = "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.pb_bulk.rds",
    hybrid_drna_bulk_rds = "/vast/projects/LongBench/analysis/hybrid_bulk/result/miniquant_output/miniquant_bulk_dge.dRNA_bulk.rds",
    sequins_tsv = "/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv",
    sirv_csv = "/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv"
  )
}

if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache = TRUE, fig.width = 7, fig.height = 5, warning = FALSE, cache.path = params$cache_dir)
} else {
  knitr::opts_chunk$set(cache = FALSE, fig.width = 7, fig.height = 5, warning = FALSE)
}

knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),
  fig.width = 12,
  fig.height = 6,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL
)

set.seed(params$random_seed)
```

```{r libraries, include=FALSE}
library(edgeR)
library(tidyverse)
library(knitr)
library(patchwork)

rename <- dplyr::rename
select <- dplyr::select
```

```{r helpers, include=FALSE}
safe_read_rds <- function(path) {
  if (is.null(path) || !file.exists(path)) return(NULL)
  readRDS(path)
}

is.sequins <- function(dge) grepl("^R", rownames(dge))
is.human <- function(dge) grepl("^ENS", rownames(dge))
is.sirv_E0 <- function(dge) grepl("^SIRV\\d{3}$", rownames(dge))
is.sirv_all <- is.sirv_E0 # temporarily only consider the E0 SIRVs as long SIRV is not captures by isoquant
is.ercc      <- function(dge) grepl("^(ERCC|DQ|EF)", rownames(dge))
color_palette <- c(
  PacBio = "#df1995",
  "ONT dRNA" = "#00789b",
  "ONT cDNA" = "#04476c"
)

normalise_feature_ids <- function(dge) {
  rownames(dge) <- rownames(dge) %>%
    sub("\\|.*", "", .) %>%
    sub("\\..*", "", .)
  dge
}

normalise_observed_matrix <- function(dge, tx_len_norm = FALSE, log2_1p = FALSE) {
  counts <- as.matrix(dge$counts)

  if (tx_len_norm) {
    feature_lengths <- dge$genes$EffectiveLength
    if (is.null(feature_lengths)) {
      feature_lengths <- dge$genes$Length
    }
    if (is.null(feature_lengths)) {
      feature_lengths <- rep(1, nrow(counts))
    }
    counts <- sweep(counts, 1, feature_lengths, "/")
  }

  lib_sizes <- colSums(counts, na.rm = TRUE)
  lib_sizes[lib_sizes == 0] <- NA_real_
  cpm <- sweep(counts, 2, lib_sizes, "/") * 1e6

  if (log2_1p) {
    log2(cpm + 1)
  } else {
    cpm
  }
}

log2_1p_CPM <- function(x) {
  log2(x * 1e6 / sum(x) + 1)
}

safe_cor <- function(x, y, method = "pearson") {
  ok <- is.finite(x) & is.finite(y)
  x <- x[ok]
  y <- y[ok]

  if (length(x) < 2 || length(unique(x)) < 2 || length(unique(y)) < 2) {
    return(NA_real_)
  }

  cor(x, y, method = method)
}

safe_length_bin <- function(x) {
  x <- as.numeric(x)
  breaks <- unique(quantile(x, probs = seq(0, 1, 0.25), na.rm = TRUE, names = FALSE))

  if (length(breaks) < 2) {
    return(factor(rep("All lengths", length(x)), levels = "All lengths"))
  }

  labels <- c("1st Quartile", "2nd Quartile", "3rd Quartile", paste0(seq(4, length(breaks) - 1), "th Quartile"))

  cut(
    x,
    breaks = breaks,
    include.lowest = TRUE,
    labels = labels[seq_len(length(breaks) - 1)]
  )
}

summarise_metric_frame <- function(metric_df, sample_name) {
  per_bin <- metric_df %>%
    group_by(Sample, `Transcript length`) %>%
    summarise(
      correlation = safe_cor(GT, Observed, method = "pearson"),
      `mean square error` = mean((GT - Observed)^2, na.rm = TRUE),
      `mean absolute error` = mean(abs(GT - Observed), na.rm = TRUE),
      nrmse = {
        rmse <- sqrt(mean((GT - Observed)^2, na.rm = TRUE))
        gt_mean <- mean(GT, na.rm = TRUE)
        if (is.na(gt_mean) || gt_mean == 0) NA_real_ else rmse / gt_mean
      },
      .groups = "drop"
    ) %>%
    rename(cell_line = Sample)

  overall <- metric_df %>%
    group_by(Sample) %>%
    summarise(
      correlation = safe_cor(GT, Observed, method = "pearson"),
      `mean square error` = mean((GT - Observed)^2, na.rm = TRUE),
      `mean absolute error` = mean(abs(GT - Observed), na.rm = TRUE),
      nrmse = {
        rmse <- sqrt(mean((GT - Observed)^2, na.rm = TRUE))
        gt_mean <- mean(GT, na.rm = TRUE)
        if (is.na(gt_mean) || gt_mean == 0) NA_real_ else rmse / gt_mean
      },
      .groups = "drop"
    ) %>%
    rename(cell_line = Sample) %>%
    mutate(`Transcript length` = "All transcript")

  bind_rows(per_bin, overall) %>%
    mutate(sample = sample_name)
}

build_sequin_metric_df <- function(dge, tx_len_norm = FALSE) {
  dge.sequins <- dge[is.sequins(dge), ]
  observed <- normalise_observed_matrix(dge.sequins, tx_len_norm = tx_len_norm, log2_1p = TRUE) %>%
    as_tibble(rownames = "NAME")

  sequins.gt <- readr::read_tsv(params$sequins_tsv, show_col_types = FALSE) %>%
    transmute(
      NAME,
      LENGTH,
      MIX_A = log2_1p_CPM(MIX_A),
      MIX_B = log2_1p_CPM(MIX_B)
    ) %>%
    pivot_longer(
      cols = c(MIX_A, MIX_B),
      names_to = "MIX",
      values_to = "GT"
    )

  sample_meta <- tibble(
    Sample = dge.sequins$samples$sample,
    MIX = dge.sequins$samples$sequins
  ) %>%
    mutate(
      MIX = toupper(trimws(as.character(MIX))),
      MIX = ifelse(MIX %in% c("A", "MIXA", "MIX_A"), "MIX_A", "MIX_B")
  )

  purrr::map_dfr(seq_len(nrow(sample_meta)), function(i) {
    sample_id <- sample_meta$Sample[[i]]
    mix_id <- sample_meta$MIX[[i]]

    observed %>%
      select(NAME, all_of(sample_id)) %>%
      rename(Observed = all_of(sample_id)) %>%
      mutate(MIX = mix_id) %>%
      left_join(sequins.gt, by = c("NAME", "MIX")) %>%
      mutate(
        Sample = sample_id,
        `Transcript length` = safe_length_bin(LENGTH)
      ) %>%
      select(NAME, Sample, GT, Observed, LENGTH, `Transcript length`)
  })
}

get_sequin_quartiled_states <- function(dge, sample_name, tx_len_norm = FALSE) {
  combined_df <- build_sequin_metric_df(dge, tx_len_norm = tx_len_norm)
  summarise_metric_frame(combined_df, sample_name)
}

build_ercc_metric_df <- function(dge, tx_len_norm = FALSE) {
  dge.ercc <- dge[is.ercc(dge), ]
  observed <- normalise_observed_matrix(dge.ercc, tx_len_norm = tx_len_norm, log2_1p = TRUE) %>%
    as_tibble(rownames = "NAME")

  ercc.gt <- readr::read_csv(params$sirv_csv, show_col_types = FALSE) %>%
    filter(grepl("ERCC", txid)) %>%
    transmute(NAME = GenBank, GT = log2_1p_CPM(conc))

  observed %>%
    left_join(ercc.gt, by = "NAME") %>%
    pivot_longer(cols = -c(NAME, GT), names_to = "Sample", values_to = "Observed") %>%
    mutate(
      LENGTH = dge.ercc$genes$Length[match(NAME, rownames(dge.ercc))],
      `Transcript length` = safe_length_bin(LENGTH)
    ) %>%
    select(NAME, Sample, GT, Observed, LENGTH, `Transcript length`)
}

get_ercc_quartiled_states <- function(dge, sample_name, tx_len_norm = FALSE) {
  combined_df <- build_ercc_metric_df(dge, tx_len_norm = tx_len_norm)
  summarise_metric_frame(combined_df, sample_name)
}

build_sirv_metric_df <- function(dge, tx_len_norm = FALSE) {
  dge.sirv <- dge[is.sirv_all(dge), ]
  observed <- normalise_observed_matrix(dge.sirv, tx_len_norm = tx_len_norm, log2_1p = TRUE) %>%
    as_tibble(rownames = "NAME")

  sirv_gt_value <- log2_1p_CPM(rep(1, nrow(dge.sirv)))[1]

  observed %>%
    pivot_longer(cols = -NAME, names_to = "Sample", values_to = "Observed") %>%
    mutate(
      GT = sirv_gt_value,
      LENGTH = dge.sirv$genes$Length[match(NAME, rownames(dge.sirv))],
      `Transcript length` = safe_length_bin(LENGTH)
    ) %>%
    select(NAME, Sample, GT, Observed, LENGTH, `Transcript length`)
}

get_sirv_quartiled_states <- function(dge, sample_name, tx_len_norm = FALSE) {
  combined_df <- build_sirv_metric_df(dge, tx_len_norm = tx_len_norm)
  summarise_metric_frame(combined_df, sample_name)
}

build_tool_registry <- function() {
  oarfish_bundle <- safe_read_rds(params$input_DGEobject_rds)
  stopifnot(!is.null(oarfish_bundle))

  isoquant_ont_obj <- safe_read_rds(params$isoquant_ont_bulk_rds)
  isoquant_pb_obj <- safe_read_rds(params$isoquant_pb_bulk_rds)
  isoquant_drna_obj <- safe_read_rds(params$isoquant_drna_bulk_rds)

  hybrid_ont_obj <- safe_read_rds(params$hybrid_ont_bulk_rds)
  hybrid_pb_obj <- safe_read_rds(params$hybrid_pb_bulk_rds)
  hybrid_drna_obj <- safe_read_rds(params$hybrid_drna_bulk_rds)

  quant_registry <- list(
    Oarfish = list(
      ont_bulk.tx.dge = oarfish_bundle$ont_bulk.tx.dge,
      pb_bulk.tx.dge = oarfish_bundle$pb_bulk.tx.dge,
      dRNA_bulk.tx.dge = oarfish_bundle$drna_bulk.tx.dge
    ),
    IsoQuant = list(
      ont_bulk.tx.dge = isoquant_ont_obj$ont_bulk.tx.dge,
      pb_bulk.tx.dge = isoquant_pb_obj$pb_bulk.tx.dge,
      dRNA_bulk.tx.dge = isoquant_drna_obj$dRNA_bulk.tx.dge
    ),
    MiniQuant = list(
      ont_bulk.tx.dge = hybrid_ont_obj$ont_bulk.TPM.tx.dge,
      pb_bulk.tx.dge = hybrid_pb_obj$pb_bulk.TPM.tx.dge,
      dRNA_bulk.tx.dge = hybrid_drna_obj$dRNA_bulk.TPM.tx.dge
    )
  )

  purrr::imap(quant_registry, function(tool_entry, tool_name) {
    list(
      `ONT cDNA` = if (!is.null(tool_entry$ont_bulk.tx.dge)) normalise_feature_ids(tool_entry$ont_bulk.tx.dge) else NULL,
      PacBio = if (!is.null(tool_entry$pb_bulk.tx.dge)) normalise_feature_ids(tool_entry$pb_bulk.tx.dge) else NULL,
      `ONT dRNA` = if (!is.null(tool_entry$dRNA_bulk.tx.dge)) normalise_feature_ids(tool_entry$dRNA_bulk.tx.dge) else NULL
    )
  })
}

average_log_abundance <- function(dge) {
  rowMeans(normalise_observed_matrix(dge, tx_len_norm = FALSE, log2_1p = TRUE), na.rm = TRUE)
}

compare_tool_pair <- function(left_tool, right_tool, platform_name, left_dge, right_dge) {
  if (is.null(left_dge) || is.null(right_dge)) return(NULL)

  common_tx <- intersect(rownames(left_dge), rownames(right_dge))
  common_samples <- intersect(colnames(left_dge$counts), colnames(right_dge$counts))

  if (length(common_tx) == 0 || length(common_samples) == 0) return(NULL)

  left_sub <- left_dge[common_tx, common_samples]
  right_sub <- right_dge[common_tx, common_samples]

  left_detected_ids <- rownames(left_dge)[rowSums(left_dge[, common_samples]$counts > 0, na.rm = TRUE) > 0]
  right_detected_ids <- rownames(right_dge)[rowSums(right_dge[, common_samples]$counts > 0, na.rm = TRUE) > 0]
  common_detected_ids <- intersect(left_detected_ids, right_detected_ids)
  common_detected_tx <- length(common_detected_ids)
  union_detected_tx <- length(union(left_detected_ids, right_detected_ids))

  if (common_detected_tx > 0) {
    left_signal <- average_log_abundance(left_dge[common_detected_ids, common_samples])
    right_signal <- average_log_abundance(right_dge[common_detected_ids, common_samples])
    rmse_value <- sqrt(mean((left_signal - right_signal)^2, na.rm = TRUE))
  } else {
    left_signal <- numeric(0)
    right_signal <- numeric(0)
    rmse_value <- NA_real_
  }

  tibble(
    platform = platform_name,
    tool_left = left_tool,
    tool_right = right_tool,
    common_tx = length(common_tx),
    common_detected_tx = common_detected_tx,
    union_detected_tx = union_detected_tx,
    common_samples = length(common_samples),
    jaccard_tx = ifelse(union_detected_tx == 0, NA_real_, common_detected_tx / union_detected_tx),
    pearson = safe_cor(left_signal, right_signal, method = "pearson"),
    spearman = safe_cor(left_signal, right_signal, method = "spearman"),
    rmse = rmse_value
  )
}

collect_pairwise_tool_metrics <- function(tool_dges, platform_name) {
  available_tools <- names(tool_dges)[sapply(tool_dges, function(x) !is.null(x[[platform_name]]))]

  if (length(available_tools) < 2) return(NULL)

  pair_index <- combn(available_tools, 2, simplify = FALSE)

  purrr::map_dfr(pair_index, function(tool_pair) {
    compare_tool_pair(
      left_tool = tool_pair[[1]],
      right_tool = tool_pair[[2]],
      platform_name = platform_name,
      left_dge = tool_dges[[tool_pair[[1]]]][[platform_name]],
      right_dge = tool_dges[[tool_pair[[2]]]][[platform_name]]
    )
  })
}

collect_spikein_metric <- function(tool_name, platform_name, dge) {
  if (is.null(dge)) return(NULL)

  bind_rows(
    get_sequin_quartiled_states(dge, sample_name = platform_name, tx_len_norm = FALSE) %>% mutate(spikein = "Sequins"),
    get_ercc_quartiled_states(dge, sample_name = platform_name, tx_len_norm = FALSE) %>% mutate(spikein = "ERCC"),
    get_sirv_quartiled_states(dge, sample_name = platform_name, tx_len_norm = FALSE) %>% mutate(spikein = "SIRV")
  ) %>%
    mutate(tool = tool_name, platform = platform_name)
}

collect_spikein_raw <- function(tool_name, platform_name, dge) {
  if (is.null(dge)) return(NULL)

  bind_rows(
    build_sequin_metric_df(dge, tx_len_norm = FALSE) %>% mutate(spikein = "Sequins"),
    build_ercc_metric_df(dge,   tx_len_norm = FALSE) %>% mutate(spikein = "ERCC"),
    build_sirv_metric_df(dge,   tx_len_norm = FALSE) %>% mutate(spikein = "SIRV")
  ) %>%
    mutate(tool = tool_name, platform = platform_name)
}
```

# Load data

```{r load-data}
tool_dges <- build_tool_registry()
tool_dges <- tool_dges[sapply(tool_dges, function(x) any(!sapply(x, is.null)))]

tool_inventory <- purrr::imap_dfr(tool_dges, function(platform_list, tool_name) {
  purrr::imap_dfr(platform_list, function(dge, platform_name) {
    if (is.null(dge)) return(NULL)
    tibble(
      tool = tool_name,
      platform = platform_name,
      tx_features = nrow(dge),
      samples = ncol(dge$counts),
      total_signal = sum(dge$counts, na.rm = TRUE),
      detected_tx = sum(rowSums(dge$counts > 0, na.rm = TRUE) > 0)
    )
  })
})

tool_pairwise <- purrr::map_dfr(c("ONT cDNA", "PacBio", "ONT dRNA"), function(platform_name) {
  collect_pairwise_tool_metrics(tool_dges, platform_name)
})

# The isoquant dge does not have LENGTH col in the gene, backfill it with the length of the corresponding feature in the oarfish dge if available, otherwise NA
tool_dges$IsoQuant$PacBio$genes <- tool_dges$Oarfish$PacBio$genes[rownames(tool_dges$IsoQuant$PacBio), ]
tool_dges$IsoQuant$`ONT cDNA`$genes <- tool_dges$Oarfish$`ONT cDNA`$genes[rownames(tool_dges$IsoQuant$`ONT cDNA`), ]
tool_dges$IsoQuant$`ONT dRNA`$genes <- tool_dges$Oarfish$`ONT dRNA`$genes[rownames(tool_dges$IsoQuant$`ONT dRNA`), ]
tool_dges$MiniQuant$PacBio$genes <- tool_dges$Oarfish$PacBio$genes[rownames(tool_dges$MiniQuant$PacBio), ]
tool_dges$MiniQuant$`ONT cDNA`$genes <- tool_dges$Oarfish$`ONT cDNA`$genes[rownames(tool_dges$MiniQuant$`ONT cDNA`), ]
tool_dges$MiniQuant$`ONT dRNA`$genes <- tool_dges$Oarfish$`ONT dRNA`$genes[rownames(tool_dges$MiniQuant$`ONT dRNA`), ]


# Compute per-platform common feature set across all tools for a fair spike-in comparison.
# A feature is kept if it (1) exists in all tools' DGEs and (2) passes filterByExpr in every tool.
platform_common_features <- purrr::map(c("ONT cDNA", "PacBio", "ONT dRNA"), function(platform_name) {
  dge_list <- purrr::map(tool_dges, ~ .x[[platform_name]]) %>% purrr::compact()

  # Rowname intersection first
  common_rows <- Reduce(intersect, purrr::map(dge_list, rownames))

  # filterByExpr on each tool's DGE subsetted to common rows; keep features that pass in all tools
  keep <- Reduce(`&`, purrr::map(dge_list, function(dge) {
    dge_sub <- dge[common_rows, ]
    filterByExpr(dge_sub, group = dge_sub$samples$group, min.count = 5)
  }))

  common_rows[keep]
}) %>% setNames(c("ONT cDNA", "PacBio", "ONT dRNA"))

tool_dges_common <- purrr::imap(tool_dges, function(platform_list, tool_name) {
  purrr::imap(platform_list, function(dge, platform_name) {
    if (is.null(dge)) return(NULL)
    common_features <- platform_common_features[[platform_name]]
    dge[intersect(rownames(dge), common_features), ]
  })
})

tool_spikein <- purrr::imap_dfr(tool_dges_common, function(platform_list, tool_name) {
  purrr::imap_dfr(platform_list, function(dge, platform_name) {
    collect_spikein_metric(tool_name, platform_name, dge)
  })
})

tool_spikein_raw <- purrr::imap_dfr(tool_dges_common, function(platform_list, tool_name) {
  purrr::imap_dfr(platform_list, function(dge, platform_name) {
    collect_spikein_raw(tool_name, platform_name, dge)
  })
})

tool_spikein_summary <- tool_spikein_raw %>%
  group_by(tool, platform, spikein) %>%
  summarise(
    mean_correlation = safe_cor(GT, Observed, method = "pearson"),
    mean_nrmse = {
      rmse <- sqrt(mean((GT - Observed)^2, na.rm = TRUE))
      gt_mean <- mean(GT, na.rm = TRUE)
      if (is.na(gt_mean) || gt_mean == 0) NA_real_ else rmse / gt_mean
    },
    mean_mse = mean((GT - Observed)^2, na.rm = TRUE),
    mean_mae = mean(abs(GT - Observed), na.rm = TRUE),
    .groups = "drop"
  )
```

# Inventory

This report compares the long-read transcript quantification outputs from Oarfish, IsoQuant, and MiniQuant. The comparisons are done per platform and focus on three questions:

1. Do the tools detect the same transcript set?
2. Do the tools assign similar relative abundances to the shared transcripts?
3. Which tool best matches spike-in truth sets?

```{r tool_inventory_table}
kable(tool_inventory)
```

```{r common_feature_summary}
common_feature_summary <- tibble(
  platform = names(platform_common_features),
  common_features = sapply(platform_common_features, length)
)
kable(common_feature_summary, caption = "Common features used for spike-in analysis (intersection across all tools)")
```

```{r tool_inventory_plot, fig.width=9, fig.height=4}
ggplot(tool_inventory, aes(x = platform, y = detected_tx, fill = tool)) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Detected transcripts by long-read quantification tool",
    x = NULL,
    y = "Detected transcripts"
  )
```

# Pairwise overlap

```{r pairwise_overlap_table}
kable(tool_pairwise %>% select(platform, tool_left, tool_right, common_tx, common_detected_tx, union_detected_tx, common_samples, jaccard_tx))
```

```{r pairwise_overlap_plot, fig.width=9, fig.height=4}
ggplot(tool_pairwise, aes(x = platform, y = jaccard_tx, fill = interaction(tool_left, tool_right, sep = " vs "))) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Detected Transcript-set overlap between long-read quantification tools",
    x = NULL,
    y = "Jaccard index",
    fill = "Tool pair"
  )
```

# Pairwise abundance concordance

The abundance concordance analysis is calculated only on the transcripts shared between each tool pair for the same platform. All matrices are put onto a common per-sample compositional scale before averaging.

```{r pairwise_concordance_table}
kable(tool_pairwise %>% select(platform, tool_left, tool_right, pearson, spearman, rmse))
```

```{r pairwise_concordance_plot, fig.width=10, fig.height=4}

p1 <- ggplot(tool_pairwise, aes(x = platform, y = pearson, fill = interaction(tool_left, tool_right, sep = " vs "))) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Shared-transcript abundance concordance",
    x = NULL,
    y = "Pearson correlation",
    fill = "Tool pair"
  )
p2 <- ggplot(tool_pairwise, aes(x = platform, y = spearman, fill = interaction(tool_left, tool_right, sep = " vs "))) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Shared-transcript rank concordance",
    x = NULL,
    y = "Spearman correlation",
    fill = "Tool pair"
  )

p3 <- ggplot(tool_pairwise, aes(x = platform, y = rmse, fill = interaction(tool_left, tool_right, sep = " vs "))) +
  geom_col(position = "dodge") +
  theme_minimal() +
  labs(
    title = "Shared-transcript abundance deviation",
    x = NULL,
    y = "RMSE",
    fill = "Tool pair"
  )

# Combine the three plots into a single layout and share the legend
(p1 | p2 | p3) + plot_layout(guides = "collect")
```

# Spike-in accuracy

Spike-ins provide an external reference so the tools are not only being compared to one another, but also against known truth.

```{r spikein_summary_table}
kable(tool_spikein_summary)
```

```{r spikein_per_quartile_comparison, fig.width=14, fig.height=10}
ggplot(tool_spikein,
       aes(x = platform, y = nrmse, fill = tool)) +
  geom_col(position = "dodge") +
  facet_grid(spikein ~ `Transcript length`, scales = "free_y") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Spike-in NRMSE by transcript length quartile and quantification tool",
    x = NULL,
    y = "NRMSE"
  )
```

```{r spikein_summary_plot, fig.width=10, fig.height=5}
p1 <- ggplot(tool_spikein_summary, aes(x = platform, y = mean_nrmse, fill = tool)) +
  geom_col(position = "dodge") +
  facet_wrap(~ spikein, scales = "free_y") +
  theme_minimal() +
  labs(
    title = "Spike-in accuracy across long-read quantification tools",
    x = NULL,
    y = "Mean NRMSE"
  )

p1
```

