---
title: "DE analysis"
author: "Yupei & Ash"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  sirv_ercc_gtf: '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_ERCC_longSIRV_multi-fasta_20210507.gtf'
  sequins_gtf: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_annotation_2.4.gtf'
  sequins_tsv: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv'
  sirv_csv: '/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv'
  human_gtf: '/vast/projects/LongBench/reference_files/GRCh38/gencode.v44.annotation.gtf'
  ont_bulk_oarfish_dir: "/vast/projects/LongBench/analysis/lr_bulk/result/oarfish_cov_output/ont_bulk"
  pb_bulk_oarfish_dir: "/vast/projects/LongBench/analysis/lr_bulk/result/oarfish_cov_output/pb_bulk"
  drna_bulk_oarfish_dir: "/vast/projects/LongBench/analysis/lr_bulk/result/oarfish_cov_output/dRNA_bulk"
  ont_sc_salmon_dir: "/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkSalmon/ont_sc"
  ont_sn_salmon_dir: "/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkSalmon/ont_sn"
  ill_bulk_salmon_dir: "/vast/projects/LongBench/analysis/illumina_bulk/tx_de_analysis/salmon_bs"
  bulk_meta: '/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt'
---

# Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.width = 20, fig.height=15, warning = F)
```

```{r Libraries,include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(ggplot2)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ggtree)
library(aplot)
source("/vast/projects/LongBench/analysis/workflow/scripts/Rfunctions.R")
```

```{r color code}
color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b",
  ONT_1 = "#04476c",
  ONT_2 = "#24cdcd",
  Illumina = "#e88b20"
)
```

```{r function for splitting the transcript origin}
is.sequins <- function(dge) {grepl("^R", rownames(dge))}
is.human <- function(dge) {grepl("^ENS", rownames(dge))}
is.sirv_or_ercc <- function(dge) {!is.human(dge) & !is.sequins(dge)} # thie is the whole SIRV set4 (ERCC + SIRV E0 + Long SIRV)
is.sirv_E0 <- function(dge) {grepl("^SIRV\\d{3}$", rownames(dge))} 
is.sirv_all <- function(dge) {grepl("^SIRV", rownames(dge))} # including all SIRV E0 and Long SIRV
is.long_sirv <- function(dge) {grepl("^SIRV\\d{4}", rownames(dge))}
is.ercc <- function(dge) {!is.human(dge) & !is.sequins(dge) & !is.sirv_all(dge)}
```

```{r Input filename setup, include=FALSE, eval=F}
params <- list(
  random_seed = 2024,
  cache_dir = NULL,
  sirv_ercc_gtf = '/vast/projects/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_ERCC_longSIRV_multi-fasta_20210507.gtf',
  sequins_gtf = '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_annotation_2.4.gtf',
  sequins_tsv = '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv',
  human_gtf = '/vast/projects/LongBench/reference_files/GRCh38/gencode.v44.annotation.gtf',
  ont_bulk_oarfish_dir = "/vast/projects/LongBench/analysis/lr_bulk/result/oarfish_cov_output/ont_bulk",
  pb_bulk_oarfish_dir = "/vast/projects/LongBench/analysis/lr_bulk/result/oarfish_cov_output/pb_bulk",
  drna_bulk_oarfish_dir = "/vast/projects/LongBench/analysis/lr_bulk/result/oarfish_cov_output/dRNA_bulk",
  sirv_csv= '/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv',
  # ont_sc_salmon_dir = "/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkSalmon/ont_sc",
  # ont_sn_salmon_dir = "/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkSalmon/ont_sn",
  # ont_sc_salmon_dir = "/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkSalmon/ont_sc",
  # ont_sn_salmon_dir = "/vast/projects/LongBench/analysis/lr_sc_sn/result/PseudoBulkSalmon/ont_sn",
  ill_bulk_salmon_dir = "/vast/projects/LongBench/analysis/illumina_bulk/tx_de_analysis/salmon_bs",
  bulk_meta = '/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt'
)
```

```{r Create DGE from input}
bulk.meta <- read.csv(params$bulk_meta)
rownames(bulk.meta) <- bulk.meta$sample

get_dge_from_salmon <- function(dir, sample_prefix_regex) {
  salmon.dirs <- file.path(dir, 
                               list.dirs(dir, full.names = FALSE, recursive = FALSE))
  catched.salmon <- edgeR::catchSalmon(salmon.dirs, verbose = TRUE)
  rst.dge <- DGEList(counts=catched.salmon$counts, gene=catched.salmon$annotation)
  rownames(rst.dge$samples) <- rownames(rst.dge$samples) %>% sub(sample_prefix_regex, '', .)
  colnames(rst.dge$counts) <- colnames(rst.dge$counts) %>% sub(sample_prefix_regex, '', .)
  rst.dge$samples <- merge(rst.dge$samples %>% select(-group), bulk.meta, by = "row.names", all.x = TRUE) %>% select(-Row.names)
  rst.dge$samples <- rst.dge$samples[match(colnames(rst.dge$counts), rst.dge$samples$sample),]
  return(rst.dge)
}

get_dge_from_oarfish <- function(dir, sample_prefix_regex) {
  salmon.dirs <- file.path(dir, 
                               list.dirs(dir, full.names = FALSE, recursive = FALSE))
  catched.oarfish <- catchOarfish(salmon.dirs, verbose = TRUE)
  rst.dge <- DGEList(counts=catched.oarfish$counts, gene=catched.oarfish$annotation)
  rownames(rst.dge$samples) <- rownames(rst.dge$samples) %>% sub(sample_prefix_regex, '', .)
  colnames(rst.dge$counts) <- colnames(rst.dge$counts) %>% sub(sample_prefix_regex, '', .)
  rst.dge$samples <- merge(rst.dge$samples %>% select(-group), bulk.meta, by = "row.names", all.x = TRUE) %>% select(-Row.names)
  rst.dge$samples <- rst.dge$samples[match(colnames(rst.dge$counts), rst.dge$samples$sample),]
  return(rst.dge)
}

ill_bulk.dge <- get_dge_from_salmon(params$ill_bulk_salmon_dir, ".*_S")
ont_bulk.dge <- get_dge_from_oarfish(params$ont_bulk_oarfish_dir, ".*/")
pb_bulk.dge <- get_dge_from_oarfish(params$pb_bulk_oarfish_dir, ".*/")
ont_drna.deg <- get_dge_from_oarfish(params$drna_bulk_oarfish_dir, ".*/")
# ont_sc.dge <- get_dge_from_salmon(params$ont_sc_salmon_dir, ".*/")
# ont_sn.dge <- get_dge_from_salmon(params$ont_sn_salmon_dir, ".*/")
# ont_sc.dge <- get_dge_from_salmon(params$ont_sc_salmon_dir, ".*/")
# ont_sn.dge <- get_dge_from_salmon(params$ont_sn_salmon_dir, ".*/")

# Bulk ill
salmon.ill_bulk.dirs <- file.path(params$ill_bulk_salmon_dir,
                               list.dirs(params$ill_bulk_salmon_dir, full.names = FALSE, recursive = FALSE))
ill_bulk.salmon <- edgeR::catchSalmon(salmon.ill_bulk.dirs, verbose = TRUE)
ill_bulk.dge <- DGEList(counts=ill_bulk.salmon$counts, gene=ill_bulk.salmon$annotation)
rownames(ill_bulk.dge$samples) <- rownames(ill_bulk.dge$samples) %>% sub(".*_S", '', .)
rownames(ill_bulk.dge$samples) <- bulk.meta$sample[match(bulk.meta$index, rownames(ill_bulk.dge$samples))]
colnames(ill_bulk.dge$counts) <- colnames(ill_bulk.dge$counts) %>% sub(".*_S", '', .)
colnames(ill_bulk.dge$counts) <- bulk.meta$sample[match(bulk.meta$index, colnames(ill_bulk.dge$counts))]
ill_bulk.dge$samples <- merge(ill_bulk.dge$samples %>% select(-group), bulk.meta, by = "row.names", all.x = TRUE) %>% select(-Row.names)
ill_bulk.dge$samples <- ill_bulk.dge$samples[match(colnames(ill_bulk.dge$counts), ill_bulk.dge$samples$sample),]
```

# Raw count distribution
```{r, fig.width=10, fig.height=5}
plot_raw_count <- function(filter_func, title, legend = TRUE) {
  plot_df <- data.frame(
    "Illumina" = ill_bulk.dge[filter_func(ill_bulk.dge), ]$counts %>% sum(),
    "ONT cDNA" = ont_bulk.dge[filter_func(ont_bulk.dge), ]$counts %>% sum(),
    "ONT dRNA" = ont_drna.deg[filter_func(ont_drna.deg), ]$counts %>% sum(),
    "PacBio" = pb_bulk.dge[filter_func(pb_bulk.dge), ]$counts %>% sum(),
    check.names = FALSE
  ) %>% pivot_longer(everything(), names_to = "Dataset", values_to = "Total Count")

  p <- ggplot(plot_df, aes(x = Dataset, y = `Total Count`/1000000, fill = Dataset)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    theme_minimal() +
    labs(title = title, x = "Datasets", y = "Total Count (Million)") +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    # change fill color scheme
    scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname())

    # # proportion of sequins counts
    # plot_df <- data.frame(
    #   ill_bulk = ill_bulk.dge[is.sequins(ill_bulk.dge), ]$counts %>% sum / ill_bulk.dge$counts %>% sum,
    #   ont_bulk = ont_bulk.dge[is.sequins(ont_bulk.dge), ]$counts %>% sum / ont_bulk.dge$counts %>% sum,
    #   pb_bulk = pb_bulk.dge[is.sequins(pb_bulk.dge), ]$counts %>% sum / pb_bulk.dge$counts %>% sum,
    #   ont_drna = ont_drna.deg[is.sequins(ont_drna.deg), ]$counts %>% sum / ont_drna.deg$counts %>% sum
    #   # ont_sc = ont_sc.dge[is.sequins(ont_sc.dge), ]$counts %>% sum / ont_sc.dge$counts %>% sum,
    #   # ont_sn = ont_sn.dge[is.sequins(ont_sn.dge), ]$counts %>% sum / ont_sn.dge$counts %>% sum,
    #   # ont_sn = ont_sn.dge[is.sequins(ont_sn.dge), ]$counts %>% sum / ont_sn.dge$counts %>% sum,
    #   # ont_sc = ont_sc.dge[is.sequins(ont_sc.dge), ]$counts %>% sum / ont_sc.dge$counts %>% sum
    # ) %>% pivot_longer(everything(), names_to = "Dataset", values_to = "Proportion of Sequins Count")
    # p3 <- ggplot(plot_df, aes(x = Dataset, y = `Proportion of Sequins Count`, fill = Dataset)) +
    #   geom_bar(stat = "identity", position = position_dodge()) +
    #   theme_minimal() +
    #   labs(title = "Proportion of Sequins counts", x = "Sample", y = "Proportion of Sequins Count") +
    #   theme(axis.text.x = element_text(angle = 45, hjust = 1))

  if (legend) {
    p <- p + theme(legend.position = "bottom")
  } else {
    p <- p + theme(legend.position = "none")
  }
  p
}

p1 <- plot_raw_count(is.human, "Human transcript quantification", legend=FALSE)
p2 <- plot_raw_count(is.sequins, "Sequins transcript quantification", legend=TRUE)
p3 <- plot_raw_count(is.sirv_or_ercc, "SIRV transcript quantification", legend=FALSE)

p1 | p2 | p3

```



# Comparison of Overdispesion
```{r, get number of transcripts gen from gtf}
calcTxNum <- function(G.Tx.map){
  geneid = rep(NA, nrow(G.Tx.map))
  txcount = table(G.Tx.map$gene_id)
  txnum = as.numeric(txcount[match(G.Tx.map$gene_id, names(txcount))])
}
# SIRV
SIRV_tx <- GenomicFeatures::makeTxDbFromGFF(params$sirv_ercc_gtf, format = "gtf") %>% GenomicFeatures::transcripts(columns = c("tx_id", "tx_name", "gene_id"))
SIRV.G.Tx.map <- tibble(
  gene_id = SIRV_tx$gene_id %>% unlist,
  tx_name = SIRV_tx$tx_name
)
rm(SIRV_tx)
SIRV.G.Tx.map$tx_count <- SIRV.G.Tx.map %>% calcTxNum

# Sequins
sequins_tx  <- GenomicFeatures::makeTxDbFromGFF(params$sequins_gtf, format = "gtf") %>% GenomicFeatures::transcripts(columns = c("tx_id", "tx_name", "gene_id"))
sequins.G.Tx.map <- tibble(
  gene_id = sequins_tx$gene_id %>% unlist,
  tx_name = sequins_tx$tx_name
)
rm(sequins_tx)
sequins.G.Tx.map$tx_count <- sequins.G.Tx.map %>% calcTxNum

# Human
human_tx  <- GenomicFeatures::makeTxDbFromGFF(params$human_gtf, format = "gtf") %>% GenomicFeatures::transcripts(columns = c("tx_id", "tx_name", "gene_id"))
human.G.Tx.map <- tibble(
  gene_id = human_tx$gene_id %>% unlist,
  tx_name = human_tx$tx_name
)
rm(human_tx)
human.G.Tx.map$tx_count <- human.G.Tx.map %>% calcTxNum
```

## Make boxplot of the overdispersion, stratified by transcript number, split figure by human, sequins and sirv
```{r, Overdispersion}

overdisp_boxplot <- function(filter_func, title, breaks= c(0,1,5,10,20,40, Inf), legend=TRUE) {
  ## human
  ill_bulk.dge.filtered <-  ill_bulk.dge[filter_func(ill_bulk.dge),]
  ont_bulk.dge.filtered <- ont_bulk.dge[filter_func(ont_bulk.dge),]
  pb_bulk.dge.filtered <- pb_bulk.dge[filter_func(pb_bulk.dge), ]
  ont_drna.deg.filtered <- ont_drna.deg[filter_func(ont_drna.deg),]
  # ont_sc.dge.filtered <- ont_sc.dge[filter_func(ont_sc.dge),]
  # ont_sn.dge.filtered <- ont_sn.dge[filter_func(ont_sn.dge),]

  tx_map_d <- rbind(human.G.Tx.map, sequins.G.Tx.map, SIRV.G.Tx.map)

  overdisp_d <- data.frame(
    "ONT cDNA" = ont_bulk.dge.filtered$genes$Overdispersion,
    Illumina = ill_bulk.dge.filtered$genes$Overdispersion[match(rownames(ont_bulk.dge.filtered), rownames(ill_bulk.dge.filtered))],
    PacBio = pb_bulk.dge.filtered$genes$Overdispersion[match(rownames(ont_bulk.dge.filtered), rownames(pb_bulk.dge.filtered))],
    "ONT dRNA" = ont_drna.deg.filtered$genes$Overdispersion[match(rownames(ont_bulk.dge.filtered), rownames(ont_drna.deg.filtered$genes))],
    tx_count = tx_map_d$tx_count[match(rownames(ont_bulk.dge.filtered) %>% sub("\\|.*", "", .), tx_map_d$tx_name)],
    check.names = FALSE
  )
  overdisp_d$tx_count_bin <- cut(overdisp_d$tx_count, breaks = breaks)
  overdisp_d$tx_count_bin <- factor(overdisp_d$tx_count_bin, ordered = TRUE)
  overdisp_d_pivot <- pivot_longer(overdisp_d, 
                              cols = c("ONT cDNA", PacBio, "ONT dRNA", Illumina), 
                              names_to = "Method", values_to = "Overdispersion")
  p <- ggplot(overdisp_d_pivot, aes(x = tx_count_bin, y = Overdispersion, fill = Method)) +
    geom_boxplot(position = position_dodge(width = 0.75), outliers = FALSE) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = title, x = "Transcript number", y = "Overdispersion") +
    theme_minimal() +
    scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname())
  if (legend) {
    p <- p + theme(legend.position = "right")
  } else {
    p <- p + theme(legend.position = "none")
  }
  rst <- list(
    all_data = p
  )

  overdisp_d <- pivot_longer(overdisp_d, 
                            cols = c("ONT cDNA", PacBio, "ONT dRNA"), 
                            names_to = "Method", values_to = "Overdispersion")

  p <- ggplot(overdisp_d, aes(x = tx_count_bin, y = Overdispersion, fill = Method)) +
    geom_boxplot(position = position_dodge(width = 0.75), outliers = FALSE) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    labs(title = paste(title, "(Long read datasets)"), x = "Transcript number", y = "Overdispersion") +
    scale_fill_manual(values = color_palette[c("ONT_1", "ONT_2", "PacBio")] %>% unname()) +
    theme_minimal()
  if (legend) {
    p <- p + theme(legend.position = "right")
  } else {
    p <- p + theme(legend.position = "none")
  }
  rst$long_only <- p

  rst
}

p1 <- overdisp_boxplot(is.human, "Human transcript assignment ambiguity", legend = TRUE)
p2 <- overdisp_boxplot(function(x) {is.sequins(x) | is.sirv_or_ercc(x)}, title = "Spike-ins assignment ambiguity", legend = TRUE)



{
  (p1$all_data + theme(legend.position = "none") | p2$all_data) /
    (p1$long_only + theme(legend.position = "none") | p2$long_only)
}

```

## Long read only overdispersion plot
```{r, fig.width=20, echo=FALSE}
# Make boxplot of the overdispersion, stratified by transcript number, split figure by human, sequins and sirv
## human
overdisp.human_pivot <- pivot_longer(overdisp.human, 
                            cols = c(ont_bulk, pb_bulk, ont_drna), 
                            names_to = "Method", values_to = "Overdispersion")

p1 <- ggplot(overdisp.human_pivot, aes(x = tx_count_bin, y = Overdispersion, fill = Method)) +
  geom_boxplot(position = position_dodge(width = 0.75), outliers=FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Human transcript assignment ambiguity", x = "Transcript number", y = "Overdispersion")
  

# ## sequins
overdisp.sequins_pivot <- pivot_longer(overdisp.sequins, 
                                        cols = c(ont_bulk, pb_bulk, ont_drna), 
                                        names_to = "Method", values_to = "Overdispersion")

p2 <- ggplot(overdisp.sequins_pivot, aes(x = tx_count_bin, y = Overdispersion, fill = Method)) +
  geom_boxplot(position = position_dodge(width = 0.75), outliers=FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Sequins assignment ambiguity", x = "Transcript number", y = "Overdispersion")

## sirv
overdisp.sirv_pivot <- pivot_longer(overdisp.sirv, 
                                    cols = c(ont_bulk, pb_bulk, ont_drna), 
                                    names_to = "Method", values_to = "Overdispersion")

p3 <- ggplot(overdisp.sirv_pivot, aes(x = tx_count_bin, y = Overdispersion, fill = Method)) +
  geom_boxplot(position = position_dodge(width = 0.75), outliers=FALSE) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "SIRV assignment ambiguity", x = "Transcript number", y = "Overdispersion") +   ylim(0,20)


grid.arrange(p1, p2, p3+ylim(1,2),  ncol=2)
```


# Tx length distribution of Spikeins
```{r}
# sequins transcript length
sequins.gt <- read_tsv(params$sequins_tsv)

# plot the distribution of the sequins transcript length, mark each quartile with lines
p1 <- ggplot(sequins.gt, aes(x = LENGTH)) +
  geom_histogram(binwidth = 100, fill="grey", color = "black", size=0.1) +
  geom_vline(xintercept = quantile(sequins.gt$LENGTH, c(0.25, 0.5, 0.75)), linetype = "dotted", color="red", size=0.4) +
  labs(title = "Sequins transcript length distribution", x = "Transcript length", y = "Count") +
  theme_minimal()
p1

ercc.len <- ont_bulk.dge[is.ercc(ont_bulk.dge),]$genes$Length
plot_d <- data.frame('LENGTH'=ercc.len, row.names = NULL)
# plot the distribution of the sequins transcript length, mark each quartile with lines
p2 <- ggplot(plot_d, aes(x = LENGTH)) +
  geom_histogram(binwidth = 100, fill="grey", color = "black", size=0.1) +
  geom_vline(xintercept = quantile(sequins.gt$LENGTH, c(0.25, 0.5, 0.75)), linetype = "dotted", color="red", size=0.4) +
  labs(title = "ERCC transcript length distribution", x = "Transcript length", y = "Count") +
  theme_minimal()


sirv.len <- ont_bulk.dge[is.sirv_all(ont_bulk.dge),]$genes$Length
plot_d <- data.frame('LENGTH'=sirv.len, row.names = NULL)
# plot the distribution of the sequins transcript length, mark each quartile with lines
p3 <- ggplot(plot_d, aes(x = LENGTH)) +
  geom_histogram(binwidth = 100, fill="grey", color = "black", size=0.1) +
  geom_vline(xintercept = quantile(sequins.gt$LENGTH, c(0.25, 0.5, 0.75)), linetype = "dotted", color="red", size=0.4) +
  labs(title = "SIRV transcript length distribution", x = "Transcript length", y = "Count") +
  theme_minimal()

p1 <- p1 + xlim(0, 10000)
p2 <- p2 + xlim(0, 10000)
p3 <- p3 + xlim(0, 10000)

p1 / p2 / p3
```

# Sequins (Scatter)

```{r, fig.width=20}
get_sequins_scatter_plot <- function(dge, figure_name, tx_len_norm = TRUE) {

    dge <- dge[is.sequins(dge),]
    dge.sequins <- dge[is.sequins(dge),]

    # Normalisation functions
    get_ltpm <- function(counts,len) {
        x <- counts/len
        get_lcpm(x)
    }

    get_lcpm <- function(x) { 
        return(log2(x*1e6/sum(x) + 1))
    }
    
    # Prepare Sequins ground truth
    sequins.gt <- read_tsv(params$sequins_tsv)
    sequins.gt.norm <- sequins.gt %>% mutate(across(starts_with("MIX"), get_lcpm)) %>% select(NAME, MIX_A, MIX_B, LENGTH) %>% mutate(N_transcript_per_gene = gsub("^R.*_", "", NAME) %>% as.numeric)
    
    # Normalise observed data
    if (tx_len_norm) {
        dge.sequins.norm <- dge.sequins$counts %>% as_tibble(rownames="NAME")  %>% mutate(across(where(is.numeric), ~ get_ltpm(.x, dge.sequins$gene$Length)))
    } else {
        dge.sequins.norm <- dge.sequins$counts %>% as_tibble(rownames="NAME")  %>% mutate(across(where(is.numeric), get_lcpm))
    }
    dge.sequins.norm  <- dge.sequins.norm %>% left_join(dge.sequins$gene %>% as_tibble(rownames="NAME") %>% select(NAME))

    # Merge the ground truth and observed data
    mixA_df <-dge.sequins.norm %>% select(-dge.sequins$samples$sample[dge.sequins$samples$sequins=='B']) %>% left_join(sequins.gt.norm)  %>% 
          mutate(GT=MIX_A, MIX='A')  %>% 
          select(-MIX_B, -MIX_A)
    mixB_df <- dge.sequins.norm %>% select(-dge.sequins$samples$sample[dge.sequins$samples$sequins=='A']) %>% left_join(sequins.gt.norm) %>% 
      mutate(GT=MIX_B, MIX='B')  %>% 
      select(-MIX_B, -MIX_A)
    
    mixA_df.long <- mixA_df %>% pivot_longer(cols = mixA_df %>% select(dge.sequins$samples$sample[dge.sequins$samples$sequins=='A']) %>% colnames, names_to = "Sample", values_to = "Observed log-CPM")
    mixB_df.long <- mixB_df %>% pivot_longer(cols = mixB_df %>% select(dge.sequins$samples$sample[dge.sequins$samples$sequins=='B']) %>% colnames, names_to = "Sample", values_to = "Observed log-CPM")
    combined_df <- rbind(mixA_df.long, mixB_df.long)
    
    per_sample_cor <- sapply(dge$samples$sample, function(x) {
      cor(combined_df[combined_df$Sample == x,]$GT, combined_df[combined_df$Sample == x,]$`Observed log-CPM`) %>% unlist
    })
    
    # Calculate the correlation
    combined_df$Sample <- factor(combined_df$Sample)
    combined_df$MIX <- factor(combined_df$MIX)
    correlation <- cor(combined_df$GT, combined_df$`Observed log-CPM`)
    mean_sq_error <- mean((combined_df$GT - combined_df$`Observed log-CPM`)^2)
    mean_abs_error <- mean(abs(combined_df$GT - combined_df$`Observed log-CPM`))
    
    combined_df <- combined_df %>% mutate(N_transcript_per_gene = N_transcript_per_gene %>% as.factor(), `Transcript length` = LENGTH)
    # color by the quantile of the transcript length
    combined_df$`Transcript length` <- cut(combined_df$`Transcript length`, 
                                                      breaks = quantile(combined_df$`Transcript length`, 
                                                      probs = seq(0, 1, 0.25)), 
                                                      include.lowest = TRUE,
                                                      labels = c("1st Quartile", "2nd Quartile", "3rd Quartile", "4th Quartile"))

     p2 <- ggplot(combined_df[rownames(combined_df) %>% sample(nrow(combined_df), replace = F),], aes(y = GT, x = `Observed log-CPM`, color = `Transcript length`)) +
      geom_point() +
      geom_abline(slope = 1, intercept = 0, linetype = "dotted") +  # Add a dotted line of x=y
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      theme_classic() + 
      # scale_y_log10(labels = label_number()) + 
      labs(title = figure_name, y = "Expected abundance log2(TPM+1)", x = "Observed log2(TPM+1)") +
      annotate("text", x = 10, y = 5, 
               label = paste("Correlation: ", round(correlation, 2),'\n'
                             ,"Mean Squared Error: ", round(mean_sq_error, 2),'\n'
                             ,"Mean Absolute Error: ", round(mean_abs_error, 2)),
               hjust = 0, vjust = 1, 
               color = "black", size =3, 
               fontface = "italic") +
      scale_color_viridis_d() +
      scale_size_continuous(limits = c(0, 5))
    
    t <- tibble(
      sample = dge$samples$sample,
      sample_correlation = per_sample_cor,
      merged_correlation = correlation
    )
    return(list(plot1 = p1,  plot2 = p2, table = t))
}


cor.ont_bulk <- get_sequins_scatter_plot(ont_bulk.dge, figure_name = "ONT BULK Sequins", tx_len_norm = FALSE)
cor.pb_bulk <- get_sequins_scatter_plot(pb_bulk.dge, figure_name = "Pacbio BULK Sequins", tx_len_norm = FALSE)
cor.ont_drna <- get_sequins_scatter_plot(ont_drna.deg, figure_name = "ONT dRNA Sequins", tx_len_norm = FALSE)
cor.ill_bulk <- get_sequins_scatter_plot(ill_bulk.dge, figure_name = "Illumina BULK Sequins", tx_len_norm = TRUE)

grid.arrange(
            #cor.ont_bulk$plot1, 
            cor.ont_bulk$plot2, 
            cor.ont_drna$plot2,
            #cor.pb_bulk$plot1,  
            cor.pb_bulk$plot2,
            #cor.ill_bulk$plot1, 
            cor.ill_bulk$plot2,
            ncol=2)


# plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT cDNA`=sample_correlation) %>% 
#             left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina`=sample_correlation), by = 'sample') %>% 
#             left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio`=sample_correlation), by = 'sample') %>% 
#             left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_correlation), by = 'sample') %>%
#             select(`ONT Bulk`, `ONT dRNA`, `PacBio Bulk`, `Illumina Bulk`) %>% 
#             pivot_longer(col=c(`Illumina`, `ONT cDNA`, `ONT dRNA`, `PacBio`, ), values_to = "correlation", names_to = "Sample")
# 
# bx_plot <- ggplot(plot_df, aes(x = Sample, y = correlation, fill = Sample)) +
#   geom_boxplot(position = position_dodge(width = 0.75)) +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
#   labs(title = "Sequins transcript: per sample correlation with true abundance") +
#   theme_minimal() +
#   scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio", )] %>% unname())
# 
# bx_plot
```


```{r}
get_sequin_quartiled_states <- function(dge, sample_name, tx_len_norm = TRUE) {
  # tx_len_norm <- TRUE
  # dge <- ont_bulk.dge.oarfish
  # sample_name <- "ONT BULK Sequins"

  dge <- dge[is.sequins(dge), ]
  dge.sequins <- dge[is.sequins(dge), ]

  # Normalisation functions
  get_ltpm <- function(counts, len) {
    x <- counts / len
    get_lcpm(x)
  }

  get_lcpm <- function(x) {
    return(log2(x * 1e6 / sum(x) + 1))
  }

  # Prepare Sequins ground truth
  sequins.gt <- read_tsv(params$sequins_tsv)
  sequins.gt.norm <- sequins.gt %>%
    mutate(across(starts_with("MIX"), get_lcpm)) %>%
    select(NAME, MIX_A, MIX_B, LENGTH) %>%
    mutate(N_transcript_per_gene = gsub("^R.*_", "", NAME) %>% as.numeric())

  # Normalise observed data
  if (tx_len_norm) {
    dge.sequins.norm <- dge.sequins$counts %>%
      as_tibble(rownames = "NAME") %>%
      mutate(across(where(is.numeric), ~ get_ltpm(.x, dge.sequins$gene$Length)))
  } else {
    dge.sequins.norm <- dge.sequins$counts %>%
      as_tibble(rownames = "NAME") %>%
      mutate(across(where(is.numeric), get_lcpm))
  }
  dge.sequins.norm <- dge.sequins.norm %>% left_join(dge.sequins$gene %>% as_tibble(rownames = "NAME") %>% select(NAME))

  # Merge the ground truth and observed data
  mixA_df <- dge.sequins.norm %>%
    select(-dge.sequins$samples$sample[dge.sequins$samples$sequins == "B"]) %>%
    left_join(sequins.gt.norm) %>%
    mutate(GT = MIX_A, MIX = "A") %>%
    select(-MIX_B, -MIX_A)
  mixB_df <- dge.sequins.norm %>%
    select(-dge.sequins$samples$sample[dge.sequins$samples$sequins == "A"]) %>%
    left_join(sequins.gt.norm) %>%
    mutate(GT = MIX_B, MIX = "B") %>%
    select(-MIX_B, -MIX_A)

  mixA_df.long <- mixA_df %>% pivot_longer(cols = mixA_df %>% select(dge.sequins$samples$sample[dge.sequins$samples$sequins == "A"]) %>% colnames(), names_to = "Sample", values_to = "Observed log-CPM")
  mixB_df.long <- mixB_df %>% pivot_longer(cols = mixB_df %>% select(dge.sequins$samples$sample[dge.sequins$samples$sequins == "B"]) %>% colnames(), names_to = "Sample", values_to = "Observed log-CPM")
  combined_df <- rbind(mixA_df.long, mixB_df.long)

  per_sample_cor <- sapply(dge$samples$sample, function(x) {
    cor(combined_df[combined_df$Sample == x, ]$GT, combined_df[combined_df$Sample == x, ]$`Observed log-CPM`) %>% unlist()
  })
  per_sample_mse <- sapply(dge$samples$sample, function(x) {
    mean((combined_df[combined_df$Sample == x, ]$GT - combined_df[combined_df$Sample == x, ]$`Observed log-CPM`)^2) %>% unlist()
  })

  # Calculate the correlation
  combined_df$Sample <- factor(combined_df$Sample)
  combined_df$MIX <- factor(combined_df$MIX)
  # correlation <- cor(combined_df$GT, combined_df$`Observed log-CPM`)
  # mean_sq_error <- mean((combined_df$GT - combined_df$`Observed log-CPM`)^2)
  # mean_abs_error <- mean(abs(combined_df$GT - combined_df$`Observed log-CPM`))

  combined_df <- combined_df %>% mutate(N_transcript_per_gene = N_transcript_per_gene %>% as.factor(), `Transcript length` = LENGTH)
  # color by the quantile of the transcript length
  combined_df$`Transcript length` <- cut(combined_df$`Transcript length`, 
                                                    breaks = quantile(combined_df$`Transcript length`, 
                                                    probs = seq(0, 1, 0.25)), 
                                                    include.lowest = TRUE,
                                                    labels = c("1st Quartile", "2nd Quartile", "3rd Quartile", "4th Quartile"))

                                              

  # calculate the correlation per quartile and per sample, and also give the correlation overall all four quartiles
  per_quartile_cor <- combined_df %>%
    group_by(Sample, `Transcript length`) %>%
    summarise(correlation = cor(GT, `Observed log-CPM`)) %>%
    rename(
      Sample = "cell_line"
    ) %>%
    ungroup()
  
  per_quartile_mse <- combined_df %>%
    group_by(Sample, `Transcript length`) %>%
    summarise(mean_sq_error = mean((GT - `Observed log-CPM`)^2)) %>%
    rename(
      Sample = "cell_line"
    ) %>%
    ungroup()

  # merge the overall correlation with the per quartile correlation
  per_sample_cor <- tibble(
    cell_line = names(per_sample_cor),
    new_correlation = as.numeric(per_sample_cor)
  )
  per_sample_mse <- tibble(
    cell_line = names(per_sample_mse),
    new_mean_sq_error = as.numeric(per_sample_mse)
  )

  # Combine the two tibbles
  combined_data <- per_quartile_cor %>%
    left_join(per_quartile_mse, by = c("cell_line", "Transcript length")) %>%
    rename(mean_sq_error = "mean square error")
  # Add "All transcript" row with the new correlations
  all_transcripts <- per_sample_cor %>%
    left_join(per_sample_mse, by = "cell_line") %>%
    mutate(`Transcript length` = "All transcript") %>%
    rename(new_correlation = "correlation",   new_mean_sq_error = "mean square error")

  # Combine with the original data
  combined_data <- bind_rows(combined_data, all_transcripts)
  combined_data$sample <- sample_name
  combined_data
}



cor.ont_bulk <- get_sequin_quartiled_states(ont_bulk.dge, sample_name="ONT cDNA" ,tx_len_norm = FALSE)
cor.pb_bulk <- get_sequin_quartiled_states(pb_bulk.dge, sample_name = "Pacbio", tx_len_norm = FALSE)
cor.ont_drna <- get_sequin_quartiled_states(ont_drna.deg, sample_name = "ONT dRNA", tx_len_norm = FALSE)
cor.ill_bulk <- get_sequin_quartiled_states(ill_bulk.dge, sample_name = "Illumina", tx_len_norm = TRUE)

plot_df <- rbind(cor.ont_bulk, cor.pb_bulk, cor.ont_drna, cor.ill_bulk)

# get_stats_comparison
## start a one column tibble called total counts
dge_list <- list(
  'ONT cDNA' = ont_bulk.dge,
  Pacbio = pb_bulk.dge,
  "ONT dRNA" = ont_drna.deg,
  Illumina = ill_bulk.dge
) %>%  lapply(function(x) {
  x[is.sequins(x), ]
})
  
total_counts <- tibble(
  sample = lapply(names(dge_list), function(x) {rep(x, nrow(dge_list[[x]]$samples))}) %>% unlist,
  cell_line = lapply(dge_list, function(x) {x$samples$sample}) %>% unlist,
  total_counts = lapply(dge_list, function(x) {x$counts %>% colSums()}) %>% unlist,
  Transcript_discovery = lapply(dge_list, function(x) {
     x$counts %>% apply(2, function(x) {
       mean(x > 0)
     })
  }) %>% unlist
)



plot_df <- plot_df %>% left_join(total_counts, by = c("cell_line", "sample"))


# working
plot_df$cell_line_sample <- paste(plot_df$sample, plot_df$cell_line, sep = " - ")
# order the cell line sample level first by the sample then by the cell line
plot_df$cell_line_sample <- factor(plot_df$cell_line_sample, levels = plot_df$cell_line_sample %>% unique() %>% sort())



p1 <- plot_df %>%
  filter(`Transcript length` == "All transcript") %>%
  ggplot(aes(x = cell_line_sample, y = total_counts, fill = sample)) +
  # bar plot
  geom_bar(stat = "identity") +
  # better color scheme
  scale_fill_viridis_d() +
  # minimal theme
  theme_minimal() +
  # remove the x axis labels
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank()
#    panel.grid.minor = element_blank()
  ) +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname()) +
  # log the y axis
  #scale_y_log10() +
  labs(y="Total counts")

p2 <- plot_df %>%
  filter(`Transcript length` == "All transcript") %>%
  ggplot(aes(x = cell_line_sample, y = Transcript_discovery, fill = sample)) +
  # bar plot
  geom_bar(stat = "identity") +
  # minimal theme
  theme_minimal() +
  # remove the x axis labels
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank()
#    panel.grid.minor = element_blank()
  ) +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname()) +
  labs(y="Transcripts discovery rate")
p3 <- plot_df %>%
  ggplot(aes(x = cell_line_sample, y = `Transcript length`, color = correlation, size = `mean square error` )) +
  geom_point() +
  # better color scheme
  scale_color_gradient(low = "#d2cccc", high = "black")  +
  # minimal theme
  theme_minimal() +
  # rotate the x axis labels
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Samples")
  
p1 / (p2 + theme(legend.position = "none")) / p3 +
  patchwork::plot_layout(
    heights = c(2, 2, 1), 
    guides = "collect"
  ) & theme(plot.margin = unit(c(0, 1, 0, 1), "cm"))
```

# ERCC


```{r, fig.width=20}
get_ercc_scatter_plot <- function(dge, figure_name, tx_len_norm = TRUE) {
    dge.ercc <- dge[is.sirv_or_ercc(dge) & !is.sirv_all(dge),]

    # Normalisation functions
    get_ltpm <- function(counts,len) {
        x <- counts/len
        get_lcpm(x)
    }

    get_lcpm <- function(x) { 
        return(log2(x*1e6/sum(x) + 1))
    }
    
    # Prepare Sequins ground truth
    ercc.gt <- read_csv(params$sirv_csv) %>% filter(grepl("ERCC", txid))
    ercc.gt.norm <- ercc.gt %>% mutate(GT=get_lcpm(conc), NAME=GenBank, Length=dge.ercc$genes$Length[match(GenBank, rownames(dge.ercc))]) %>% select(NAME, GT, Length)
    
    # Normalise observed data
    if (tx_len_norm) {
        dge.ercc.norm <- dge.ercc$counts %>% as_tibble(rownames="NAME")  %>% mutate(across(where(is.numeric), ~ get_ltpm(.x, dge.ercc$gene$Length)))
    } else {
        dge.ercc.norm <- dge.ercc$counts %>% as_tibble(rownames="NAME")  %>% mutate(across(where(is.numeric), get_lcpm))
    }
    dge.ercc.norm  <- dge.ercc.norm %>% left_join(dge.ercc$gene %>% as_tibble(rownames="NAME") %>% select(NAME))

    # Merge the ground truth and observed data
    df <-dge.ercc.norm %>% left_join(ercc.gt.norm)
    
    combined_df <- df %>% pivot_longer(cols = dge.ercc$samples$sample, names_to = "Sample", values_to = "Observed log-CPM")
    
    per_sample_cor <- sapply(dge$samples$sample, function(x) {
      cor(combined_df[combined_df$Sample == x, ]$GT, combined_df[combined_df$Sample == x, ]$`Observed log-CPM`) %>% unlist()
    })


    # Calculate the correlation
    combined_df$Sample <- factor(combined_df$Sample)
    correlation <- cor(combined_df$GT, combined_df$`Observed log-CPM`)
    mean_sq_error <- mean((combined_df$GT - combined_df$`Observed log-CPM`)^2)
    mean_abs_error <- mean(abs(combined_df$GT - combined_df$`Observed log-CPM`))
    
    combined_df <- combined_df %>% rename(Length = "Transcript length")

    # color by the quantile of the transcript length
    combined_df$`Transcript length` <- cut(combined_df$`Transcript length`, 
                                                      breaks = quantile(combined_df$`Transcript length`, 
                                                      probs = seq(0, 1, 0.25)), 
                                                      include.lowest = TRUE,
                                                      labels = c("1st Quartile", "2nd Quartile", "3rd Quartile", "4th Quartile"))

    #p2 <- ggplot(combined_df[rownames(combined_df) %>% sample(nrow(combined_df), replace = F),], aes(y = GT, x = `Observed log-CPM`, color = `Transcript length quartiles`, shape = N_transcript_per_gene, size = Overdispersion)) +
     p2 <- ggplot(combined_df[rownames(combined_df) %>% sample(nrow(combined_df), replace = F),], aes(y = GT, x = `Observed log-CPM`, color = `Transcript length`)) +
      geom_point() +
      geom_abline(slope = 1, intercept = 0, linetype = "dotted") +  # Add a dotted line of x=y
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      theme_classic() + 
      # scale_y_log10(labels = label_number()) + 
      labs(title = figure_name, y = "Expected abundance log2(TPM+1)", x = "Observed log2(TPM+1)") +
      annotate("text", x = 12, y = 5, 
               label = paste("Correlation: ", round(correlation, 2), '\n'
                             ,"Mean Squared Error: ", round(mean_sq_error, 2), '\n'
                             ,"Mean Absolute Error: ", round(mean_abs_error, 2)),
               hjust = 0, vjust = 1, 
               color = "black", size = 3, 
               fontface = "italic") +
      scale_color_viridis_d() +
      scale_size_continuous(limits = c(0, 5))
    
    t <- tibble(
      sample = dge$samples$sample,
      sample_correlation = per_sample_cor,
      merged_correlation = correlation
    )
    return(list(plot1 = p1,  plot2 = p2, table = t))
}
cor.ont_bulk <- get_ercc_scatter_plot(ont_bulk.dge, figure_name = "ONT BULK ERCC", tx_len_norm = FALSE)
cor.pb_bulk <- get_ercc_scatter_plot(pb_bulk.dge, figure_name = "Pacbio BULK ERCC", tx_len_norm = FALSE)
cor.ont_drna <- get_ercc_scatter_plot(ont_drna.deg, figure_name = "ONT dRNA ERCC", tx_len_norm = FALSE)
cor.ill_bulk <- get_ercc_scatter_plot(ill_bulk.dge, figure_name = "Illumina BULK ERCC", tx_len_norm = TRUE)



plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_correlation) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_correlation), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_correlation), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_correlation), by = 'sample') %>%
            select(`ONT Bulk`, `ONT dRNA`, `PacBio Bulk`, `Illumina Bulk`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `ONT dRNA`, `PacBio Bulk`, `Illumina Bulk`), values_to = "correlation", names_to = "Sample")

bx_plot <- ggplot(plot_df, aes(x = Sample, y = correlation, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "ERCC transcript: per sample correlation with true abundance")

grid.arrange(
            #cor.ont_bulk$plot1, 
            cor.ont_bulk$plot2, 
            cor.ont_drna$plot2,
            #cor.pb_bulk$plot1,  
            cor.pb_bulk$plot2,
            #cor.ill_bulk$plot1, 
            cor.ill_bulk$plot2, 
            bx_plot, ncol=2)

```

```{r}
get_ercc_quartiled_states <- function(dge, sample_name, tx_len_norm = TRUE) {
  dge.ercc <- dge[is.sirv_or_ercc(dge) & !is.sirv_all(dge), ]

  # Normalisation functions
  get_ltpm <- function(counts, len) {
    x <- counts / len
    get_lcpm(x)
  }

  get_lcpm <- function(x) {
    return(log2(x * 1e6 / sum(x) + 1))
  }

  # Prepare Sequins ground truth
  ercc.gt <- read_csv(params$sirv_csv) %>% filter(grepl("ERCC", txid))
  ercc.gt.norm <- ercc.gt %>%
    mutate(GT = get_lcpm(conc), NAME = GenBank, Length = dge.ercc$genes$Length[match(GenBank, rownames(dge.ercc))]) %>%
    select(NAME, GT, Length)

  # Normalise observed data
  if (tx_len_norm) {
    dge.ercc.norm <- dge.ercc$counts %>%
      as_tibble(rownames = "NAME") %>%
      mutate(across(where(is.numeric), ~ get_ltpm(.x, dge.ercc$gene$Length)))
  } else {
    dge.ercc.norm <- dge.ercc$counts %>%
      as_tibble(rownames = "NAME") %>%
      mutate(across(where(is.numeric), get_lcpm))
  }
  dge.ercc.norm <- dge.ercc.norm %>% left_join(dge.ercc$gene %>% as_tibble(rownames = "NAME") %>% select(NAME))

  # Merge the ground truth and observed data
  df <- dge.ercc.norm %>% left_join(ercc.gt.norm)

  combined_df <- df %>% pivot_longer(cols = dge.ercc$samples$sample, names_to = "Sample", values_to = "Observed log-CPM")

  per_sample_cor <- sapply(dge$samples$sample, function(x) {
    cor(combined_df[combined_df$Sample == x, ]$GT, combined_df[combined_df$Sample == x, ]$`Observed log-CPM`) %>% unlist()
  })

  # Calculate the correlation
  combined_df$Sample <- factor(combined_df$Sample)
  correlation <- cor(combined_df$GT, combined_df$`Observed log-CPM`)

  combined_df <- combined_df %>% rename(Length = "Transcript length")

  # color by the quantile of the transcript length
  combined_df$`Transcript length` <- cut(combined_df$`Transcript length`,
    breaks = quantile(combined_df$`Transcript length`,
      probs = seq(0, 1, 0.25)
    ),
    include.lowest = TRUE,
    labels = c("1st Quartile", "2nd Quartile", "3rd Quartile", "4th Quartile")
  )

  # calculate the correlation per quartile and per sample, and also give the correlation overall all four quartiles
  per_quartile_cor <- combined_df %>%
    group_by(Sample, `Transcript length`) %>%
    summarise(correlation = cor(GT, `Observed log-CPM`)) %>%
    rename(
      Sample = "cell_line"
    ) %>%
    ungroup()
  

  # merge the overall correlation with the per quartile correlation
  per_sample_cor <- tibble(
    cell_line = names(per_sample_cor),
    new_correlation = as.numeric(per_sample_cor)
  )
  # Combine the two tibbles
  combined_data <- per_quartile_cor %>%
    left_join(per_sample_cor, by = "cell_line") %>%
    mutate(correlation = ifelse(`Transcript length` == "All transcript", new_correlation, correlation))

  # Add "All transcript" row with the new correlations
  all_transcripts <- per_sample_cor %>%
    mutate(`Transcript length` = "All transcript", correlation = new_correlation) %>%
    select(cell_line, `Transcript length`, correlation)

  # Combine with the original data
  combined_data <- bind_rows(combined_data, all_transcripts) %>% select(-new_correlation)
  combined_data$sample <- sample_name
  combined_data
}

cor.ont_bulk <- get_ercc_quartiled_states(ont_bulk.dge, sample_name="ONT cDNA" ,tx_len_norm = FALSE)
cor.pb_bulk <- get_ercc_quartiled_states(pb_bulk.dge, sample_name = "Pacbio", tx_len_norm = FALSE)
cor.ont_drna <- get_ercc_quartiled_states(ont_drna.deg, sample_name = "ONT dRNA", tx_len_norm = FALSE)
cor.ill_bulk <- get_ercc_quartiled_states(ill_bulk.dge, sample_name = "Illumina", tx_len_norm = TRUE)

plot_df <- rbind(cor.ont_bulk, cor.pb_bulk, cor.ont_drna, cor.ill_bulk)

# get_stats_comparison
## start a one column tibble called total counts
dge_list <- list(
  'ONT cDNA' = ont_bulk.dge,
  Pacbio = pb_bulk.dge,
  "ONT dRNA" = ont_drna.deg,
  Illumina = ill_bulk.dge
) %>%  lapply(function(x) {
  x[is.sirv_or_ercc(x) & !is.sirv_all(x), ]
})
  
total_counts <- tibble(
  sample = lapply(names(dge_list), function(x) {rep(x, nrow(dge_list[[x]]$samples))}) %>% unlist,
  cell_line = lapply(dge_list, function(x) {x$samples$sample}) %>% unlist,
  total_counts = lapply(dge_list, function(x) {x$counts %>% colSums()}) %>% unlist,
  Transcript_discovery = lapply(dge_list, function(x) {
     x$counts %>% apply(2, function(x) {
       mean(x > 0)
     })
  }) %>% unlist
)


plot_df <- plot_df %>% left_join(total_counts, by = c("cell_line", "sample"))


# working
plot_df$cell_line_sample <- paste(plot_df$sample, plot_df$cell_line, sep = " - ")
# order the cell line sample level first by the sample then by the cell line
plot_df$cell_line_sample <- factor(plot_df$cell_line_sample, levels = plot_df$cell_line_sample %>% unique() %>% sort())




p1 <- plot_df %>%
  filter(`Transcript length` == "All transcript") %>%
  ggplot(aes(x = cell_line_sample, y = total_counts, fill = sample)) +
  # bar plot
  geom_bar(stat = "identity") +
  # better color scheme
  scale_fill_viridis_d() +
  # minimal theme
  theme_minimal() +
  # remove the x axis labels
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank()
#    panel.grid.minor = element_blank()
  ) +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname()) +
  # log the y axis
  #scale_y_log10() +
  labs(y="Total counts")

p2 <- plot_df %>%
  filter(`Transcript length` == "All transcript") %>%
  ggplot(aes(x = cell_line_sample, y = Transcript_discovery, fill = sample)) +
  # bar plot
  geom_bar(stat = "identity") +
  # minimal theme
  theme_minimal() +
  # remove the x axis labels
  theme(
    axis.text.x = element_blank(),
    axis.title.x = element_blank(),
    panel.grid.major = element_blank()
#    panel.grid.minor = element_blank()
  ) +
  scale_fill_manual(values = color_palette[c("Illumina", "ONT_1", "ONT_2", "PacBio")] %>% unname()) +
  labs(y="Transcripts discovery rate")

p3 <- plot_df %>%
  ggplot(aes(x = cell_line_sample, y = `Transcript length`, color = correlation, size = `mean square error` )) +
  geom_point() +
  # better color scheme
  scale_color_gradient(low = "#d2cccc", high = "black")  +
  # minimal theme
  theme_minimal() +
  # rotate the x axis labels
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  labs(x = "Samples")
  
p1 / (p2 + theme(legend.position = "none")) / p3 +
  patchwork::plot_layout(
    guides = "collect"
  ) & theme(plot.margin = unit(c(0, 1, 0, 1), "cm"))




```

# SIRV (Scatter)

## All sirv
```{r, fig.width=20, echo=FALSE}
get_sirv_scatter_plot <- function(dge, figure_name, tx_len_norm = TRUE, filter_func = function(x) {is.sirv_E0(dge) | is.long_sirv(dge)}) {

    dge.sirv <- dge[filter_func(dge),]

    # Normalisation functions
    get_ltpm <- function(counts,len) {
        x <- counts/len
        get_lcpm(x)
    }

    get_lcpm <- function(x) { 
        return(log2(x*1e6/sum(x) + 1))
    }

    # Prepare sirv_e0 ground truth
    sirv.gt <- tibble(
        NAME = dge.sirv %>% row.names(),
        LENGTH = dge.sirv$gene$Length,
        GT = rep(1, nrow(dge.sirv$gene)),
        is.long.sirv = is.long_sirv(dge.sirv)
    )
    sirv.gt.norm <- sirv.gt %>% mutate(GT = get_lcpm(GT)) %>%
                                    mutate(N_transcript_per_gene = SIRV.G.Tx.map$tx_count[match(NAME, SIRV.G.Tx.map$tx_name)])
    
    # Normalise observed data
    
    if (tx_len_norm) {
        dge.sirv.norm <- dge.sirv$counts %>% as_tibble(rownames="NAME")  %>% mutate(across(where(is.numeric), ~ get_ltpm(.x, dge.sirv$gene$Length)))
    } else {
        dge.sirv.norm <- dge.sirv$counts %>% as_tibble(rownames="NAME")  %>% mutate(across(where(is.numeric), get_lcpm))
    }
    dge.sirv.norm  <- dge.sirv.norm %>% left_join(dge.sirv$gene %>% as_tibble(rownames="NAME") %>% select(NAME, Overdispersion))


    # Merge the ground truth and observed data
    plot_df <- dge.sirv.norm %>% left_join(sirv.gt.norm) %>% pivot_longer(cols=dge.sirv %>% colnames(), names_to = "Sample", values_to = "Observed log-CPM")

    per_sample_sd <- sapply(dge$samples$sample, function(x) {
      sd(plot_df[plot_df$Sample == x,]$`Observed log-CPM`) %>% unlist
    })
    
    per_sample_mad <- sapply(dge$samples$sample, function(x) {
      mad(plot_df[plot_df$Sample == x,]$`Observed log-CPM`) %>% unlist
    })
    
    # Calculate the standard deviation
    plot_df$Sample <- factor(plot_df$Sample)
    SD <- sd(plot_df$`Observed log-CPM`)
    MAD <- mad(plot_df$`Observed log-CPM`)

    # Scatter plot
    plot_df <- plot_df %>% mutate(N_transcript_per_gene = N_transcript_per_gene %>% as.factor(), `Transcript length` = LENGTH, `is long sirv` = is.long.sirv)
    # p1 <- ggplot(plot_df[rownames(plot_df) %>% sample(nrow(plot_df), replace = F),], aes(y = `Transcript length`, x = `Observed log-CPM`, color = Sample)) +
    #   geom_point() +
    #   geom_vline(xintercept = plot_df$GT[1], linetype = "dashed", color = "red") +
    #   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
    #   theme_classic() + 
    #   labs(title = figure_name, y = "Transcript length", x = "Observed log2(TPM+1)") +
    #   annotate("text", x = min(plot_df$`Observed log-CPM`), y = max(plot_df$`Transcript length`), 
    #            label = paste("sd: ", round(SD, 2)), 
    #            hjust = 0, vjust = 1, 
    #            color = "black", size = 4, 
    #            fontface = "italic")
    
    p2 <- ggplot(plot_df[rownames(plot_df) %>% sample(nrow(plot_df), replace = F),], aes(y =`Transcript length`, x = `Observed log-CPM`, color = `is long sirv`)) +
      geom_point() +
      geom_vline(xintercept = plot_df$GT[1], linetype = "dashed", color = "red") + 
      theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
      theme_classic() + 
      labs(title = figure_name, y = "Transcript length", x = "Observed log2(TPM+1)") +
      annotate("text", x = min(plot_df$`Observed log-CPM`), y = max(plot_df$`Transcript length`), 
               label = paste("SD: ", round(SD, 2), "\nMAD: ", round(MAD, 2)),
               hjust = 0, vjust = 1, 
               color = "black", size = 4, 
               fontface = "italic") +
      scale_color_viridis_d()
    p2 <- ggMarginal(p2, type = "histogram", margins = "x", fill = NA)
    
    t <- tibble(
      sample = dge$samples$sample,
      sample_sd = per_sample_sd,
      sample_mad = per_sample_mad,
      merged_sd = SD,
      merged_mad = MAD
    )
    return(list(plot1 = p1,  plot2 = p2, table = t))
}

cor.ont_bulk <- get_sirv_scatter_plot(ont_bulk.dge, figure_name = "ONT BULK sirv", tx_len_norm = FALSE)
cor.pb_bulk <- get_sirv_scatter_plot(pb_bulk.dge, figure_name = "Pacbio BULK sirv", tx_len_norm = FALSE)
cor.ill_bulk <- get_sirv_scatter_plot(ill_bulk.dge, figure_name = "Illumina BULK sirv", tx_len_norm = TRUE)
cor.ont_drna <- get_sirv_scatter_plot(ont_drna.deg, figure_name = "ONT dRNA sirv", tx_len_norm = FALSE)


plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_sd) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_sd), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_sd), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_sd), by = 'sample') %>%
            select(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`), values_to = "Standard Deviation", names_to = "Sample")

bx_plot1 <- ggplot(plot_df, aes(x = Sample, y = `Standard Deviation`, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal() + 
  labs(title = "Per sample SD")

plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_mad) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_mad), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_mad), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_mad), by = 'sample') %>%
            select(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`), values_to = "Median Absolute Deviation", names_to = "Sample")

bx_plot2 <- ggplot(plot_df, aes(x = Sample, y = `Median Absolute Deviation`, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal() + 
  labs(title = "Per sample MAD")

grid.arrange(
            #cor.ont_bulk$plot1, 
            cor.ont_bulk$plot2, 
            cor.ont_drna$plot2,
            #cor.pb_bulk$plot1,  
            cor.pb_bulk$plot2,
            #cor.ill_bulk$plot1, 
            cor.ill_bulk$plot2, 
            bx_plot1, bx_plot2, ncol=2)

```

## E0 sirv or long sirv
```{r, fig.width=20, echo=FALSE}
cor.ont_bulk <- get_sirv_scatter_plot(ont_bulk.dge, figure_name = "ONT BULK sirv", tx_len_norm = FALSE, filter_func = is.sirv_E0)
cor.pb_bulk <- get_sirv_scatter_plot(pb_bulk.dge, figure_name = "Pacbio BULK sirv", tx_len_norm = FALSE, filter_func = is.sirv_E0)
cor.ill_bulk <- get_sirv_scatter_plot(ill_bulk.dge, figure_name = "Illumina BULK sirv", tx_len_norm = TRUE, filter_func = is.sirv_E0)
cor.ont_drna <- get_sirv_scatter_plot(ont_drna.deg, figure_name = "ONT dRNA sirv", tx_len_norm = FALSE, filter_func = is.sirv_E0)


plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_sd) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_sd), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_sd), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_sd), by = 'sample') %>%
            select(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`), values_to = "Standard Deviation", names_to = "Sample")

bx_plot1 <- ggplot(plot_df, aes(x = Sample, y = `Standard Deviation`, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal() + 
  labs(title = "Per sample SD")

plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_mad) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_mad), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_mad), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_mad), by = 'sample') %>%
            select(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`), values_to = "Median Absolute Deviation", names_to = "Sample")

bx_plot2 <- ggplot(plot_df, aes(x = Sample, y = `Median Absolute Deviation`, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal() + 
  labs(title = "Per sample MAD")

grid.arrange(
            #cor.ont_bulk$plot1, 
            cor.ont_bulk$plot2, 
            cor.ont_drna$plot2,
            #cor.pb_bulk$plot1,  
            cor.pb_bulk$plot2,
            #cor.ill_bulk$plot1, 
            cor.ill_bulk$plot2, 
            bx_plot1, bx_plot2, ncol=2)


cor.ont_bulk <- get_sirv_scatter_plot(ont_bulk.dge, figure_name = "ONT BULK sirv", tx_len_norm = FALSE, filter_func = is.long_sirv)
cor.pb_bulk <- get_sirv_scatter_plot(pb_bulk.dge, figure_name = "Pacbio BULK sirv", tx_len_norm = FALSE, filter_func = is.long_sirv)
cor.ill_bulk <- get_sirv_scatter_plot(ill_bulk.dge, figure_name = "Illumina BULK sirv", tx_len_norm = TRUE, filter_func = is.long_sirv)
cor.ont_drna <- get_sirv_scatter_plot(ont_drna.deg, figure_name = "ONT dRNA sirv", tx_len_norm = FALSE, filter_func = is.long_sirv)


plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_sd) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_sd), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_sd), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_sd), by = 'sample') %>%
            select(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`), values_to = "Standard Deviation", names_to = "Sample")

bx_plot1 <- ggplot(plot_df, aes(x = Sample, y = `Standard Deviation`, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal() + 
  labs(title = "Per sample SD")

plot_df <- cor.ont_bulk$table %>% dplyr::rename(`ONT Bulk`=sample_mad) %>% 
            left_join(cor.ill_bulk$table %>% dplyr::rename(`Illumina Bulk`=sample_mad), by = 'sample') %>% 
            left_join(cor.pb_bulk$table %>% dplyr::rename(`PacBio Bulk`=sample_mad), by = 'sample') %>% 
            left_join(cor.ont_drna$table %>% dplyr::rename(`ONT dRNA`=sample_mad), by = 'sample') %>%
            select(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`) %>% 
            pivot_longer(col=c(`ONT Bulk`, `PacBio Bulk`, `Illumina Bulk`, `ONT dRNA`), values_to = "Median Absolute Deviation", names_to = "Sample")

bx_plot2 <- ggplot(plot_df, aes(x = Sample, y = `Median Absolute Deviation`, fill = Sample)) +
  geom_boxplot(position = position_dodge(width = 0.75)) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal() + 
  labs(title = "Per sample MAD")

grid.arrange(
            #cor.ont_bulk$plot1, 
            cor.ont_bulk$plot2, 
            cor.ont_drna$plot2,
            #cor.pb_bulk$plot1,  
            cor.pb_bulk$plot2,
            #cor.ill_bulk$plot1, 
            cor.ill_bulk$plot2, 
            bx_plot1, bx_plot2, ncol=2)


```


## Transcript discovery rate (transcript with non-zero count)
```{r, fig.width=20}
get_tx_discovery_rate <- function(dge_list, dataset_list, figure_name="Transcript discovery rate"){
  plot_df <- tibble()
  for (i in 1:length(dge_list)) {
    dge <- dge_list[[i]]
    dataset <- dataset_list[[i]]
    non_zero_proportions <- dge$counts %>% apply(2, function(x) mean(x != 0)) %>% tibble()
    colnames(non_zero_proportions) <- "Transcript discovery rate"
    non_zero_proportions$sample <- dge$counts %>% colnames
    non_zero_proportions$Dataset <- dataset
    plot_df <- rbind(plot_df, non_zero_proportions)
  }
  ggplot(plot_df, aes(x = sample, y = `Transcript discovery rate`, fill = Dataset)) +
    geom_bar(stat = "identity", position = position_dodge()) +
    theme_minimal() +
    labs(title = figure_name, x = "Sample", y = "Transcript Discovery Rate") +
    geom_text(aes(label = scales::percent(`Transcript discovery rate`)), 
              position = position_dodge(width = 0.9), 
              vjust = -0.5, size = 2) 
}

p1 <- get_tx_discovery_rate(
  list(ont_bulk.dge[is.sequins(ont_bulk.dge),], 
      ont_drna.deg[is.sequins(ont_drna.deg),],
      ill_bulk.dge[is.sequins(ill_bulk.dge),], 
      pb_bulk.dge[is.sequins(pb_bulk.dge),]),
  list("ONT BULK", "ONT dRNA", "Illumina BULK", "PacBio BULK"),
  figure_name = "Sequins transcript discovery rate"
)

p2 <- get_tx_discovery_rate(
  list(ont_bulk.dge[is.sirv_or_ercc(ont_bulk.dge),], 
        ont_drna.deg[is.sirv_or_ercc(ont_drna.deg),],
        ill_bulk.dge[is.sirv_or_ercc(ill_bulk.dge),], 
        pb_bulk.dge[is.sirv_or_ercc(pb_bulk.dge),]),
  list("ONT BULK", "ONT dRNA", "Illumina BULK", "PacBio BULK"),
  figure_name = "SIRV transcript discovery rate"
)

p3 <- get_tx_discovery_rate(
  list(ont_bulk.dge[is.human(ont_bulk.dge),], 
      ont_drna.deg[is.human(ont_drna.deg),],
       ill_bulk.dge[is.human(ill_bulk.dge),], 
      #  ont_sc.dge[is.human(ont_sc.dge),], 
      #  ont_sn.dge[is.human(ont_sn.dge),], 
       pb_bulk.dge[is.human(pb_bulk.dge),]),
  list("ONT BULK", "ONT dRNA", "Illumina BULK", "PacBio BULK"),
  figure_name = "Human transcript discovery rate"
)

# p4 <- get_tx_discovery_rate(
#   list(ont_sc.dge[is.human(ont_sc.dge),], ont_sn.dge[is.human(ont_sn.dge),], ont_sc.dge[is.human(ont_sc.dge),], ont_sn.dge[is.human(ont_sn.dge),]),
#   list("ONT SC", "ONT SN", "ONT SC (cleaned)", "ONT SN (cleaned)"),
#   figure_name = "Human transcript discovery rate (SC/SN data)"
# )

# Combine all plots
grid.arrange(p1, p2, p3, ncol=2)


```


# Normalisation and MDS
```{r, fig.height=15, fig.width=10}
norm_and_mds <- function(dge, dataset='ont_bulk', col = NULL, text = NULL, norm.method = "TMM") {
  # filter 
  filter <- filterByExpr(dge, group=dge$samples$sample)
  dge <- dge[filter, , keep.lib.sizes = FALSE]
  
  
   # normalisation
  if (!is.null(norm.method)) {
     dge <- calcNormFactors(dge, method = norm.method)
  }
 
  cpm <- cpm(dge, log = TRUE)

  # MDS
  if (is.null(col)) {
    col <- dge$samples$group
  }
  
  if (is.null(text)) {
    text <- dge$samples$group
  }

  factor_vector <- as.factor(col)
  col <- brewer.pal(n = length(levels(factor_vector)), name = "Set1")
  color_mapping <- setNames(col, levels(factor_vector))
  boxplot(cpm, las=2, col=color_mapping[as.character(factor_vector)], main="")
  plotMDS(cpm, main = dataset,
          text = text,
          col = color_mapping[as.character(factor_vector)])
}


par(mfrow=c(3,2))

norm_and_mds(ont_bulk.dge[is.human(ont_bulk.dge),], 
            dataset='ont bulk human (colored by sample groups)',
            col = ont_bulk.dge$samples$group)

norm_and_mds(ont_bulk.dge[is.sequins(ont_bulk.dge),], 
            dataset='ont bulk sequins (colored by sequins mix A or B)',
            col = ont_bulk.dge$samples$sequins,
            norm.method = NULL)
norm_and_mds(ont_bulk.dge[is.sirv_or_ercc(ont_bulk.dge),], 
            dataset='ont bulk SIRV (colored by sample groups)',
            col = ont_bulk.dge$samples$group,
            norm.method = NULL)


norm_and_mds(ont_drna.deg[is.human(ont_drna.deg),], 
            dataset='ont dRNA human (colored by sample groups)',
            col = ont_drna.deg$samples$group)
norm_and_mds(ont_drna.deg[is.sequins(ont_drna.deg),], 
            dataset='ont dRNA sequins (colored by sequins mix A or B)',
            col = ont_drna.deg$samples$sequins,
            norm.method = NULL)
norm_and_mds(ont_drna.deg[is.sirv_or_ercc(ont_drna.deg),],
            dataset='ont dRNA SIRV (colored by sample groups)',
            col = ont_drna.deg$samples$group,
            norm.method = NULL)


norm_and_mds(pb_bulk.dge[is.human(pb_bulk.dge),], 
            dataset='PacBio bulk human (colored by sample groups)',
            col = pb_bulk.dge$samples$group)

norm_and_mds(pb_bulk.dge[is.sequins(pb_bulk.dge),], 
            dataset='PacBio bulk sequins (colored by sequins mix A or B)',
            col = pb_bulk.dge$samples$sequins,
            norm.method = NULL)
norm_and_mds(pb_bulk.dge[is.sirv_or_ercc(pb_bulk.dge),], 
            dataset='PacBio bulk SIRV (colored by sample groups)',
            col = pb_bulk.dge$samples$group,
            norm.method = NULL)


norm_and_mds(ill_bulk.dge[is.human(ill_bulk.dge),], 
            dataset='ILL bulk human (colored by sample groups)',
            col = ill_bulk.dge$samples$group)

norm_and_mds(ill_bulk.dge[is.sequins(ill_bulk.dge),], 
            dataset='ILL bulk sequins (colored by sequins mix A or B)',
            col = ill_bulk.dge$samples$sequins,
            norm.method = NULL)

norm_and_mds(ill_bulk.dge[is.sirv_or_ercc(ill_bulk.dge),], 
            dataset='ILL bulk SIRV (colored by sample groups)',
            col = ill_bulk.dge$samples$group,
            norm.method = NULL)

# par(mfrow=c(3,2))
# norm_and_mds(ont_sc.dge[is.human(ont_sc.dge),], 
#             dataset='ONT SC pseudobulk human (colored by sample groups)',
#             col = ont_sc.dge$samples$group,
#             norm.method = NULL)
# 
# norm_and_mds(ont_sn.dge[is.human(ont_sn.dge),], 
#             dataset='ONT SN pseudobulk human (colored by sample groups)',
#             col = ont_sn.dge$samples$group,
#             norm.method = NULL)
# 
# par(mfrow=c(3,2))
# norm_and_mds(ont_sc.dge[is.human(ont_sc.dge),], 
#             dataset='ONT SC pseudobulk human (colored by sample groups)',
#             col = ont_sc.dge$samples$group,
#             norm.method = NULL)
# 
# norm_and_mds(ont_sn.dge[is.human(ont_sn.dge),], 
#             dataset='ONT SN pseudobulk human (colored by sample groups)',
#             col = ont_sn.dge$samples$group,
#             norm.method = NULL)


```

# Differential gene expression
## Aggregate counts (tximport)

```{r}
library(tximport) # install latest version from github

# Extract tx_name and gene_id columns from human and sequins
human_tx2gene <- human.G.Tx.map[, c("tx_name", "gene_id")]
sequins_tx2gene <- sequins.G.Tx.map[, c("tx_name", "gene_id")]
SIRV_tx2gene <- SIRV.G.Tx.map[, c("tx_name", "gene_id")]
combined_tx2gene <- rbind(human_tx2gene, sequins_tx2gene,SIRV_tx2gene)

# Get quant.sf file paths
get_quant_paths <- function(dirs, file_name) {
  # Generate file paths for quantification files
  quant_paths <- file.path(dirs, file_name)
  return(quant_paths)
}

# Define directories with their corresponding expected file names
oarfish_dirs <- list(
  ill_bulk = list(dir = params$ill_bulk_salmon_dir, file_name = "quant.sf"),
  ont_bulk = list(dir = params$ont_bulk_oarfish_dir, file_name = ".quant"),
  pb_bulk = list(dir = params$pb_bulk_oarfish_dir, file_name = ".quant"),
  drna_bulk = list(dir = params$drna_bulk_oarfish_dir, file_name = ".quant")
)

# Apply function to each directory to get all quant file paths
quant_paths <- lapply(oarfish_dirs, function(dir_info) {
  # Get the directory and file name from dir_info
  dir <- dir_info$dir
  file_name <- dir_info$file_name
  
  # List all subdirectories and get the quant file paths
  sub_dirs <- list.dirs(dir, full.names = TRUE, recursive = FALSE)
  get_quant_paths(sub_dirs, file_name)
})

# Flatten the list of quant_paths into a named list
quant_paths <- lapply(quant_paths, unlist)

# Function to import DGE from tximport
get_dge_from_txi <- function(quant_dir, sample_prefix_regex, type) {
  # Import data using tximport
  txi <- tximport(quant_dir,
                  type = type,
                  tx2gene = combined_tx2gene,
                  ignoreAfterBar = TRUE,
                  countsFromAbundance = "no") # raw counts
  counts <- as.data.frame(txi$counts, stringAsFactors = FALSE)
  sample_names <- gsub(sample_prefix_regex, '', basename(dirname(quant_dir)))
  colnames(counts) <- sample_names
  rst.dge <- DGEList(counts = counts)
  rownames(rst.dge$samples) <- sub(sample_prefix_regex, '', rownames(rst.dge$samples))
  colnames(rst.dge$counts) <- sub(sample_prefix_regex, '', colnames(rst.dge$counts))
  rst.dge$samples <- merge(rst.dge$samples %>% select(-group), bulk.meta, by = "row.names", all.x = TRUE) %>% select(-Row.names)
  rst.dge$samples <- rst.dge$samples[match(colnames(rst.dge$counts), rst.dge$samples$sample),]
  return(rst.dge)
}

# Process each quant path for different bulk types
#txi_ill_bulk.dge <- get_dge_from_txi(quant_paths$ill_bulk, ".*_S", "salmon")
txi_ont_bulk.dge <- get_dge_from_txi(quant_paths$ont_bulk, ".*/", "oarfish")
txi_pb_bulk.dge <- get_dge_from_txi(quant_paths$pb_bulk, ".*/", "oarfish")
txi_drna_bulk.dge <- get_dge_from_txi(quant_paths$drna_bulk, ".*/", "oarfish")

# Separate bulk illumina 
salmon_ill_bulk_quant <- quant_paths$ill_bulk

txi <- tximport(salmon_ill_bulk_quant,
                type = "salmon",
                tx2gene = combined_tx2gene,
                ignoreAfterBar = TRUE,
                countsFromAbundance = "no") # raw counts
counts <- as.data.frame(txi$counts, stringAsFactors = FALSE)
sample_names <- basename(dirname(salmon_ill_bulk_quant))
colnames(counts) <- sample_names

# Create DGEList for ill bulk
txi_ill_bulk.dge <- DGEList(counts = counts)
rownames(txi_ill_bulk.dge$samples) <- sub(".*_S", '', rownames(txi_ill_bulk.dge$samples))
rownames(txi_ill_bulk.dge$samples) <- bulk.meta$sample[match(bulk.meta$index, rownames(txi_ill_bulk.dge$samples))]
colnames(txi_ill_bulk.dge$counts) <- sub(".*_S", '', colnames(txi_ill_bulk.dge$counts))
colnames(txi_ill_bulk.dge$counts) <- bulk.meta$sample[match(bulk.meta$index, colnames(txi_ill_bulk.dge$counts))]
txi_ill_bulk.dge$samples <- merge(txi_ill_bulk.dge$samples %>% select(-group), bulk.meta, by = "row.names", all.x = TRUE) %>% select(-Row.names)
correct_order <- match(colnames(txi_ill_bulk.dge$counts), txi_ill_bulk.dge$samples$sample)
txi_ill_bulk.dge$samples <- txi_ill_bulk.dge$samples[correct_order, ]

```


## DGE edgeR
```{r}

DGE.edgeR_human <- function(dge, norm.method = "TMM") {
  # QC and normalization
  filter <- filterByExpr(dge, group = dge$samples$sample)
  dge <- dge[filter, , keep.lib.sizes = FALSE]
  
  if (!is.null(norm.method)) {
    dge <- calcNormFactors(dge, method = norm.method)
  }
  
  design <- model.matrix(~0 + group, data = dge$samples)
  dge <- estimateDisp(dge, design)
  qlfit <- glmQLFit(dge, design)
  
  contrasts <- makeContrasts(
    AvsL = groupsclc_a - groupluad,
    PvsL = groupsclc_p - groupluad,
    AvsP = groupsclc_a - groupsclc_p,
    levels = design
  )
  
  AvsL.QLF.test <- glmQLFTest(qlfit, contrast = contrasts[, "AvsL"]) %>% decideTests
  PvsL.QLF.test <- glmQLFTest(qlfit, contrast = contrasts[, "PvsL"]) %>% decideTests
  AvsP.QLF.test <- glmQLFTest(qlfit, contrast = contrasts[, "AvsP"]) %>% decideTests
  
  list(
    AvsL.DE = rownames(AvsL.QLF.test)[AvsL.QLF.test != 0],
    PvsL.DE = rownames(PvsL.QLF.test)[PvsL.QLF.test != 0],
    AvsP.DE = rownames(AvsP.QLF.test)[AvsP.QLF.test != 0]
  )
}

DGE.edgeR_sequins <- function(dge, norm.method = "TMM") {
  # QC and normalization
  filter <- filterByExpr(dge, group = dge$samples$sample)
  dge <- dge[filter, , keep.lib.sizes = FALSE]
  
  if (!is.null(norm.method)) {
    dge <- calcNormFactors(dge, method = norm.method)
  }
  
  design <- model.matrix(~0 + sequins, data = dge$samples)
  dge <- estimateDisp(dge, design)
  qlfit <- glmQLFit(dge, design)
  
  contrasts <- makeContrasts(
    AvsB = sequinsA - sequinsB,
    levels = design
  )
  
  AvsB.QLF.test <- glmQLFTest(qlfit, contrast = contrasts[, "AvsB"]) %>% decideTests
  
  list(
    AvsB.DE = rownames(AvsB.QLF.test)[AvsB.QLF.test != 0]
  )
}


ill_bulk.dge_human.list <- DGE.edgeR_human(txi_ill_bulk.dge[is.human(txi_ill_bulk.dge),])
ont_bulk.dge_human.list <- DGE.edgeR_human(txi_ont_bulk.dge[is.human(txi_ont_bulk.dge),])
drna_bulk.deg_human.list <- DGE.edgeR_human(txi_drna_bulk.dge[is.human(txi_drna_bulk.dge),])
pb_bulk.dge_human.list <- DGE.edgeR_human(txi_pb_bulk.dge[is.human(txi_pb_bulk.dge),])
# ont_sc.dge_human.list <- DGE.edgeR_human(txi_ont_sc.dge[is.human(txi_ont_sc.dge),])
# ont_sn.dge_human.list <- DGE.edgeR_human(txi_ont_sn.dge[is.human(txi_ont_sn.dge),])
# ont_sn.dge_human.list <- DGE.edgeR_human(txi_ont_sc.dge[is.human(txi_ont_sc.dge),])
# ont_sc.dge_human.list <- DGE.edgeR_human(txi_ont_sn.dge[is.human(txi_ont_sn.dge),])

ill_bulk.dge_sequins.list <- DGE.edgeR_sequins(txi_ill_bulk.dge[is.sequins(txi_ill_bulk.dge),])
ont_bulk.dge_sequins.list <- DGE.edgeR_sequins(txi_ont_bulk.dge[is.sequins(txi_ont_bulk.dge),])
drna_bulk.dge_sequins.list <- DGE.edgeR_sequins(txi_drna_bulk.dge[is.sequins(txi_drna_bulk.dge),])
pb_bulk.dge_sequins.list <- DGE.edgeR_sequins(txi_pb_bulk.dge[is.sequins(txi_pb_bulk.dge),])

#This is still using the seuqins function to do A vs B DE.
ill_bulk.dge_sirv.list <- DGE.edgeR_sequins(txi_ill_bulk.dge[is.sirv_or_ercc(txi_ill_bulk.dge),])
ont_bulk.dge_sirv.list <- DGE.edgeR_sequins(txi_ont_bulk.dge[is.sirv_or_ercc(txi_ont_bulk.dge),])
drna_bulk.dge_sirv.list <- DGE.edgeR_sequins(txi_drna_bulk.dge[is.sirv_or_ercc(txi_drna_bulk.dge),])
pb_bulk.dge_sirv.list <- DGE.edgeR_sequins(txi_pb_bulk.dge[is.sirv_or_ercc(txi_pb_bulk.dge),])


```


# Differential transcript expression
## DTE edgeR 
```{r get_dte, message=FALSE}

DTE.edgeR_human <- function(dge, norm.method="TMM") {
  # QC and normalisation:
  filter <- filterByExpr(dge, group=dge$samples$sample)
  dge <- dge[filter, , keep.lib.sizes = FALSE]

  if (!is.null(norm.method)) {
     dge <- calcNormFactors(dge, method = norm.method)
  }

  design <- model.matrix(~0+group, data = dge$samples)
  dge <- estimateDisp(dge, design)
  qlfit <- glmQLFit(dge, design)
  contrasts <- makeContrasts(AvsL=groupsclc_a-groupluad, PvsL=groupsclc_p-groupluad, AvsP = groupsclc_a-groupsclc_p, levels=design)
  # SCLC_A vs LUCA
  AvsL.QLF.test <- glmQLFTest(qlfit, contrast=contrasts[,"AvsL"]) %>% decideTests
  # SCLC_P vs LUCA
  PvsL.QLF.test <- glmQLFTest(qlfit, contrast=contrasts[,"PvsL"]) %>% decideTests
  # SCLC_A vs SCLC_P
  AvsP.QLF.test <- glmQLFTest(qlfit, contrast=contrasts[,"AvsP"]) %>% decideTests

 list(
    AvsL.test_rst = AvsL.QLF.test,
    PvsL.test_rst = PvsL.QLF.test,
    AvsP.test_rst = AvsP.QLF.test,
    AvsL.DE = rownames(AvsL.QLF.test)[AvsL.QLF.test!=0],
    PvsL.DE = rownames(PvsL.QLF.test)[PvsL.QLF.test!=0],
    AvsP.DE = rownames(AvsP.QLF.test)[AvsP.QLF.test!=0]
  )
}

DTE.edgeR_sequins <- function(dge, norm.method = "TMM") {
  # QC and normalization
  filter <- filterByExpr(dge, group = dge$samples$sample)
  dge <- dge[filter, , keep.lib.sizes = FALSE]
  
  if (!is.null(norm.method)) {
    dge <- calcNormFactors(dge, method = norm.method)
  }
  
  design <- model.matrix(~0 + sequins, data = dge$samples)
  dge <- estimateDisp(dge, design)
  qlfit <- glmQLFit(dge, design)
  
  contrasts <- makeContrasts(
    AvsB = sequinsA - sequinsB,
    levels = design
  )
  
  AvsB.QLF.test <- glmQLFTest(qlfit, contrast = contrasts[, "AvsB"]) 
  d <-  AvsB.QLF.test %>% decideTests
  
  list(
    test_rst = AvsB.QLF.test,
    AvsB.DE = rownames(d)[d != 0]
  )
}



ill_bulk.dte_human.list <- DTE.edgeR_human(ill_bulk.dge[is.human(ill_bulk.dge),])
ont_bulk.dte_human.list <- DTE.edgeR_human(ont_bulk.dge[is.human(ont_bulk.dge),])
drna_bulk.dte_human.list <- DTE.edgeR_human(ont_drna.deg[is.human(ont_drna.deg),])
pb_bulk.dte_human.list <- DTE.edgeR_human(pb_bulk.dge[is.human(pb_bulk.dge),])

# ont_sc.dte_human.list <- DTE.edgeR_human(ont_sc.dge[is.human(ont_sc.dge),])
# ont_sn.dte_human.list <- DTE.edgeR_human(ont_sn.dge[is.human(ont_sn.dge),])
# ont_sn.dte_human.list <- DTE.edgeR_human(ont_sn.dge[is.human(ont_sn.dge),])
# ont_sc.dte_human.list <- DTE.edgeR_human(ont_sc.dge[is.human(ont_sc.dge),])

ill_bulk.dte_sequins.list <- DTE.edgeR_sequins(ill_bulk.dge[is.sequins(ill_bulk.dge),])
ont_bulk.dte_sequins.list <- DTE.edgeR_sequins(ont_bulk.dge[is.sequins(ont_bulk.dge),])
drna_bulk.dte_sequins.list <- DTE.edgeR_sequins(ont_drna.deg[is.sequins(ont_drna.deg),])
pb_bulk.dte_sequins.list <- DTE.edgeR_sequins(pb_bulk.dge[is.sequins(pb_bulk.dge),])


# This is still using the sequins function to do A vs B DE.
ill_bulk.dte_sirv.list <- DTE.edgeR_sequins(ill_bulk.dge[is.sirv_or_ercc(ill_bulk.dge),])
ont_bulk.dte_sirv.list <- DTE.edgeR_sequins(ont_bulk.dge[is.sirv_or_ercc(ont_bulk.dge),])
drna_bulk.dte_sirv.list <- DTE.edgeR_sequins(ont_drna.deg[is.sirv_or_ercc(ont_drna.deg),])
pb_bulk.dte_sirv.list <- DTE.edgeR_sequins(pb_bulk.dge[is.sirv_or_ercc(pb_bulk.dge),])
```

## DTE edgeR+scaling

```{r}
# scale
ill_bulk.scale_dge <- ill_bulk.dge
ont_bulk.scale_dge <- ont_bulk.dge
pb_bulk.scale_dge <- pb_bulk.dge
ont_drna.scale_dge <- ont_drna.deg

ill_bulk.scale_dge$counts  <- ill_bulk.scale_dge$counts / ill_bulk.scale_dge$genes$Overdispersion
ont_bulk.scale_dge$counts  <- ont_bulk.scale_dge$counts / ont_bulk.scale_dge$genes$Overdispersion
pb_bulk.scale_dge$counts  <- pb_bulk.scale_dge$counts / pb_bulk.scale_dge$genes$Overdispersion
ont_drna.scale_dge$counts  <- ont_drna.scale_dge$counts / ont_drna.scale_dge$genes$Overdispersion

```


```{r}
ill_bulk.scl_dte_human.list <- DTE.edgeR_human(ill_bulk.scale_dge[is.human(ill_bulk.scale_dge),])
ont_bulk.scl_dte_human.list <- DTE.edgeR_human(ont_bulk.scale_dge[is.human(ont_bulk.scale_dge),])
pb_bulk.scl_dte_human.list <- DTE.edgeR_human(pb_bulk.scale_dge[is.human(pb_bulk.scale_dge),])
drna_bulk.scl_dte_human.list <- DTE.edgeR_human(ont_drna.scale_dge[is.human(ont_drna.scale_dge),])
# ont_sc.scl_dte_human.list <- DTE.edgeR_human(ont_sc.scale_dge[is.human(ont_sc.scale_dge),])
# ont_sn.scl_dte_human.list <- DTE.edgeR_human(ont_sn.scale_dge[is.human(ont_sn.scale_dge),])
# ont_sn.scl_dte_human.list <- DTE.edgeR_human(ont_sn.scale_dge[is.human(ont_sn.scale_dge),])
# ont_sc.scl_dte_human.list <- DTE.edgeR_human(ont_sc.scale_dge[is.human(ont_sc.scale_dge),])

ill_bulk.scl_dte_sequins.list <- DTE.edgeR_sequins(ill_bulk.scale_dge[is.sequins(ill_bulk.scale_dge),])
ont_bulk.scl_dte_sequins.list <- DTE.edgeR_sequins(ont_bulk.scale_dge[is.sequins(ont_bulk.scale_dge),])
pb_bulk.scl_dte_sequins.list <- DTE.edgeR_sequins(pb_bulk.scale_dge[is.sequins(pb_bulk.scale_dge),])
drna_bulk.scl_dte_sequins.list <- DTE.edgeR_sequins(ont_drna.scale_dge[is.sequins(ont_drna.scale_dge),])

# This is still using the seuqins function to do A vs B DE.
ill_bulk.scl_dte_sirv.list <- DTE.edgeR_sequins(ill_bulk.scale_dge[is.sirv_or_ercc(ill_bulk.scale_dge),])
ont_bulk.scl_dte_sirv.list <- DTE.edgeR_sequins(ont_bulk.scale_dge[is.sirv_or_ercc(ont_bulk.scale_dge),])
pb_bulk.scl_dte_sirv.list <- DTE.edgeR_sequins(pb_bulk.scale_dge[is.sirv_or_ercc(pb_bulk.scale_dge),])
drna_bulk.scl_dte_sirv.list <- DTE.edgeR_sequins(ont_drna.scale_dge[is.sirv_or_ercc(ont_drna.scale_dge),])

```


# Summary DE
## DGE human
```{r upset plot dge human, fig.width=8, fig.height=6}

# Upset plot
## SCLC_A vs LUCA (AvsL)
all_elements <- unique(c(
  ILL_BULK = ill_bulk.dge_human.list$AvsL.DE,
  ONT_BULK = ont_bulk.dge_human.list$AvsL.DE,
  PB_BULK = pb_bulk.dge_human.list$AvsL.DE,
  dRNA_BULK = drna_bulk.deg_human.list$AvsL.DE
  #ONT_SC = ont_sc.dge_human.list$AvsL.DE,
  #ONT_SN = ont_sn.dge_human.list$AvsL.DE,
  #ONT_SN_CLEAN = ont_sn.dge_human.list$AvsL.DE,
  #ONT_SC_CLEAN = ont_sc.dge_human.list$AvsL.DE
))

plot_d_AvsL.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dge_human.list$AvsL.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dge_human.list$AvsL.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dge_human.list$AvsL.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.deg_human.list$AvsL.DE)
  #ont_sc= as.integer(all_elements %in% ont_sc.dge_human.list$AvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.dge_human.list$AvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.dge_human.list$AvsL.DE),
  #ont_sc= as.integer(all_elements %in% ont_sc.dge_human.list$AvsL.DE)
)
plt1 <- upset(plot_d_AvsL.DE,  order.by = "freq", nsets = 8) 


## SCLC_P vs LUCA (PvsL)
all_elements <- unique(c(
 ILL_BULK = ill_bulk.dge_human.list$PvsL.DE,
 ONT_BULK = ont_bulk.dge_human.list$PvsL.DE,
 PB_BULK = pb_bulk.dge_human.list$PvsL.DE,
 dRNA_BULK = drna_bulk.deg_human.list$PvsL.DE
 #ONT_SC = ont_sc.dge_human.list$PvsL.DE,
 #ONT_SN = ont_sn.dge_human.list$PvsL.DE,
  #ONT_SN_CLEAN = ont_sn.dge_human.list$PvsL.DE,
  #ONT_SC_CLEAN = ont_sc.dge_human.list$PvsL.DE
))

plot_d_PvsL.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dge_human.list$PvsL.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dge_human.list$PvsL.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dge_human.list$PvsL.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.deg_human.list$PvsL.DE)
  #ont_sc= as.integer(all_elements %in% ont_sc.dge_human.list$PvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.dge_human.list$PvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.dge_human.list$PvsL.DE),
  #ont_sc= as.integer(all_elements %in% ont_sc.dge_human.list$PvsL.DE)
)
plt2 <- upset(plot_d_PvsL.DE,  order.by = "freq", nsets = 8) 


## SCLC_A vs SCLC_P (AvsP)
all_elements <- unique(c(
 ILL_BULK = ill_bulk.dge_human.list$AvsP.DE,
 ONT_BULK = ont_bulk.dge_human.list$AvsP.DE,
 PB_BULK = pb_bulk.dge_human.list$AvsP.DE,
 dRNA_BULK = drna_bulk.deg_human.list$AvsP.DE
 #ONT_SC = ont_sc.dge_human.list$AvsP.DE,
 #ONT_SN = ont_sn.dge_human.list$AvsP.DE,
  #ONT_SN_CLEAN = ont_sn.dge_human.list$AvsP.DE,
  #ONT_SC_CLEAN = ont_sc.dge_human.list$AvsP.DE
))

plot_d_AvsP.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dge_human.list$AvsP.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dge_human.list$AvsP.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dge_human.list$AvsP.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.deg_human.list$AvsP.DE)
  #ont_sc= as.integer(all_elements %in% ont_sc.dge_human.list$AvsP.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.dge_human.list$AvsP.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.dge_human.list$AvsP.DE),
  #ont_sc= as.integer(all_elements %in% ont_sc.dge_human.list$AvsP.DE)
)
plt3 <- upset(plot_d_AvsP.DE,  order.by = "freq", nsets = 8) 

plt1
grid::grid.text("DGE SCLC_A vs LUAD (EdgeR)", x = 0.5, y = 0.95)

plt2
grid::grid.text("DGE SCLC_P vs LUAD (EdgeR)", x = 0.5, y = 0.95)

plt3
grid::grid.text("DGE SCLC_P vs SCLC_A (EdgeR)", x = 0.5, y = 0.95)

```


## DGE sequins 

```{r upset plot dge sequins, fig.width=8, fig.height=6}

# Upset plot
## A vs B (AvsB)
all_elements <- unique(c(
  ILL_BULK = ill_bulk.dge_sequins.list$AvsB.DE,
  ONT_BULK = ont_bulk.dge_sequins.list$AvsB.DE,
  PB_BULK = pb_bulk.dge_sequins.list$AvsB.DE,
  dRNA_BULK = drna_bulk.dge_sequins.list$AvsB.DE
))

plot_d_AvsB.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dge_sequins.list$AvsB.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dge_sequins.list$AvsB.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dge_sequins.list$AvsB.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.dge_sequins.list$AvsB.DE)
)
plt4 <- upset(plot_d_AvsB.DE,  order.by = "freq", nsets = 8) 


plt4
grid::grid.text("DGE SequinsA vs SequinsB (EdgeR)", x = 0.5, y = 0.95)


```



## DTE human(edgeR)
```{r upset plot dte human edgeR, fig.width=8, fig.height=6}

# Upset plot
## SCLC_A vs LUCA (AvsL)
all_elements <- unique(c(
  ILL_BULK = ill_bulk.dte_human.list$AvsL.DE,
  ONT_BULK = ont_bulk.dte_human.list$AvsL.DE,
  PB_BULK = pb_bulk.dte_human.list$AvsL.DE,
  dRNA_BULK = drna_bulk.dte_human.list$AvsL.DE
  # ONT_SC = ont_sc.dte_human.list$AvsL.DE,
  # ONT_SN = ont_sn.dte_human.list$AvsL.DE,
  # ONT_SN_CLEAN = ont_sn.dte_human.list$AvsL.DE,
  # ONT_SC_CLEAN = ont_sc.dte_human.list$AvsL.DE
))

plot_d_AvsL.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dte_human.list$AvsL.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dte_human.list$AvsL.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dte_human.list$AvsL.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.dte_human.list$AvsL.DE)
  # ont_sc= as.integer(all_elements %in% ont_sc.dte_human.list$AvsL.DE),
  # ont_sn= as.integer(all_elements %in% ont_sn.dte_human.list$AvsL.DE),
  # ont_sn= as.integer(all_elements %in% ont_sn.dte_human.list$AvsL.DE),
  # ont_sc= as.integer(all_elements %in% ont_sc.dte_human.list$AvsL.DE)
)
p1 <- upset(plot_d_AvsL.DE,  order.by = "freq", nsets = 8) 

## SCLC_P vs LUCA (PvsL)
all_elements <- unique(c(
 ILL_BULK = ill_bulk.dte_human.list$PvsL.DE,
 ONT_BULK = ont_bulk.dte_human.list$PvsL.DE,
 PB_BULK = pb_bulk.dte_human.list$PvsL.DE,
 dRNA_BULK = drna_bulk.dte_human.list$PvsL.DE
 #  ONT_SC = ont_sc.dte_human.list$PvsL.DE,
#  ONT_SN = ont_sn.dte_human.list$PvsL.DE,
#   ONT_SN_CLEAN = ont_sn.dte_human.list$PvsL.DE,
#   ONT_SC_CLEAN = ont_sc.dte_human.list$PvsL.DE
))

plot_d_PvsL.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dte_human.list$PvsL.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dte_human.list$PvsL.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dte_human.list$PvsL.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.dte_human.list$PvsL.DE)
  # ont_sc= as.integer(all_elements %in% ont_sc.dte_human.list$PvsL.DE),
  # ont_sn= as.integer(all_elements %in% ont_sn.dte_human.list$PvsL.DE),
  # ont_sn= as.integer(all_elements %in% ont_sn.dte_human.list$PvsL.DE),
  # ont_sc= as.integer(all_elements %in% ont_sc.dte_human.list$PvsL.DE)
)
p2 <- upset(plot_d_PvsL.DE,  order.by = "freq", nsets = 8) 


## SCLC_A vs SCLC_P (AvsP)
all_elements <- unique(c(
 ILL_BULK = ill_bulk.dte_human.list$AvsP.DE,
 ONT_BULK = ont_bulk.dte_human.list$AvsP.DE,
 PB_BULK = pb_bulk.dte_human.list$AvsP.DE,
 dRNA_BULK = drna_bulk.dte_human.list$AvsP.DE
#  ONT_SC = ont_sc.dte_human.list$AvsP.DE,
#  ONT_SN = ont_sn.dte_human.list$AvsP.DE,
#   ONT_SN_CLEAN = ont_sn.dte_human.list$AvsP.DE,
#   ONT_SC_CLEAN = ont_sc.dte_human.list$AvsP.DE
))

plot_d_AvsP.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dte_human.list$AvsP.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dte_human.list$AvsP.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dte_human.list$AvsP.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.dte_human.list$AvsP.DE)
  # ont_sc= as.integer(all_elements %in% ont_sc.dte_human.list$AvsP.DE),
  # ont_sn= as.integer(all_elements %in% ont_sn.dte_human.list$AvsP.DE),
  # ont_sn= as.integer(all_elements %in% ont_sn.dte_human.list$AvsP.DE),
  # ont_sc= as.integer(all_elements %in% ont_sc.dte_human.list$AvsP.DE)
)
p3 <- upset(plot_d_AvsP.DE,  order.by = "freq", nsets = 8) 

p1
grid::grid.text("DTE SCLC_A vs LUAD (EdgeR)", x = 0.5, y = 0.95)

p2
grid::grid.text("DTE SCLC_P vs LUAD (EdgeR)", x = 0.5, y = 0.95)

p3
grid::grid.text("DTE SCLC_P vs SCLC_A (EdgeR)", x = 0.5, y = 0.95)
```

## DTE sequins(edgeR) 
```{r upset plot dte sequins edgeR, fig.width=8, fig.height=6}

# Upset plot
## A vs B (AvsB)
all_elements <- unique(c(
  ILL_BULK = ill_bulk.dte_sequins.list$AvsB.DE,
  ONT_BULK = ont_bulk.dte_sequins.list$AvsB.DE,
  PB_BULK = pb_bulk.dte_sequins.list$AvsB.DE,
  dRNA_bulk = drna_bulk.dte_sequins.list$AvsB.DE
))

plot_d_AvsB.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.dte_sequins.list$AvsB.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.dte_sequins.list$AvsB.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.dte_sequins.list$AvsB.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.dte_sequins.list$AvsB.DE)
)
p4 <- upset(plot_d_AvsB.DE,  order.by = "freq", nsets = 8) 


p4
grid::grid.text("DTE SequinsA vs SequinsB", x = 0.5, y = 0.95)


```

## DTE human(edgeR + scaling)
```{r upset plot dte human edgeR + scaling, fig.width=8, fig.height=6}

# Upset plot
## SCLC_A vs LUCA (AvsL)
all_elements <- unique(c(
  ILL_BULK = ill_bulk.scl_dte_human.list$AvsL.DE,
  ONT_BULK = ont_bulk.scl_dte_human.list$AvsL.DE,
  PB_BULK = pb_bulk.scl_dte_human.list$AvsL.DE,
  dRNA_BULK = drna_bulk.scl_dte_human.list$AvsL.DE
  #ONT_SC = ont_sc.scl_dte_human.list$AvsL.DE,
  #ONT_SN = ont_sn.scl_dte_human.list$AvsL.DE,
  #ONT_SN_CLEAN = ont_sn.scl_dte_human.list$AvsL.DE,
  #ONT_SC_CLEAN = ont_sc.scl_dte_human.list$AvsL.DE
))

plot_d_AvsL.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.scl_dte_human.list$AvsL.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.scl_dte_human.list$AvsL.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.scl_dte_human.list$AvsL.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.scl_dte_human.list$AvsL.DE)
  #ont_sc= as.integer(all_elements %in% ont_sc.scl_dte_human.list$AvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.scl_dte_human.list$AvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.scl_dte_human.list$AvsL.DE),
  #ont_sc= as.integer(all_elements %in% ont_sc.scl_dte_human.list$AvsL.DE)
)
pp1 <- upset(plot_d_AvsL.DE,  order.by = "freq", nsets = 8) 


## SCLC_P vs LUCA (PvsL)
all_elements <- unique(c(
 ILL_BULK = ill_bulk.scl_dte_human.list$PvsL.DE,
 ONT_BULK = ont_bulk.scl_dte_human.list$PvsL.DE,
 PB_BULK = pb_bulk.scl_dte_human.list$PvsL.DE,
 dRNA_BULK = drna_bulk.scl_dte_human.list$PvsL.DE
 #ONT_SC = ont_sc.scl_dte_human.list$PvsL.DE,
 #ONT_SN = ont_sn.scl_dte_human.list$PvsL.DE,
  #ONT_SN_CLEAN = ont_sn.scl_dte_human.list$PvsL.DE,
  #ONT_SC_CLEAN = ont_sc.scl_dte_human.list$PvsL.DE
))

plot_d_PvsL.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.scl_dte_human.list$PvsL.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.scl_dte_human.list$PvsL.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.scl_dte_human.list$PvsL.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.scl_dte_human.list$PvsL.DE)
  #ont_sc= as.integer(all_elements %in% ont_sc.scl_dte_human.list$PvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.scl_dte_human.list$PvsL.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.scl_dte_human.list$PvsL.DE),
  #ont_sc= as.integer(all_elements %in% ont_sc.scl_dte_human.list$PvsL.DE)
)
pp2 <- upset(plot_d_PvsL.DE,  order.by = "freq", nsets = 8) 


## SCLC_A vs SCLC_P (AvsP)
all_elements <- unique(c(
 ILL_BULK = ill_bulk.scl_dte_human.list$AvsP.DE,
 ONT_BULK = ont_bulk.scl_dte_human.list$AvsP.DE,
 PB_BULK = pb_bulk.scl_dte_human.list$AvsP.DE,
 dRNA_BULK = drna_bulk.scl_dte_human.list$AvsP.DE
 #ONT_SC = ont_sc.scl_dte_human.list$AvsP.DE,
 #ONT_SN = ont_sn.scl_dte_human.list$AvsP.DE,
  #ONT_SN_CLEAN = ont_sn.scl_dte_human.list$AvsP.DE,
  #ONT_SC_CLEAN = ont_sc.scl_dte_human.list$AvsP.DE
))

plot_d_AvsP.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.scl_dte_human.list$AvsP.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.scl_dte_human.list$AvsP.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.scl_dte_human.list$AvsP.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.scl_dte_human.list$AvsP.DE)
  #ont_sc= as.integer(all_elements %in% ont_sc.scl_dte_human.list$AvsP.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.scl_dte_human.list$AvsP.DE),
  #ont_sn= as.integer(all_elements %in% ont_sn.scl_dte_human.list$AvsP.DE),
  #ont_sc= as.integer(all_elements %in% ont_sc.scl_dte_human.list$AvsP.DE)
)
pp3 <- upset(plot_d_AvsP.DE,  order.by = "freq", nsets = 8) 

pp1
grid::grid.text("DTE SCLC_A vs LUAD (EdgeR + scale)", x = 0.5, y = 0.95)

pp2
grid::grid.text("DTE SCLC_P vs LUAD (EdgeR + scale)", x = 0.5, y = 0.95)

pp3
grid::grid.text("DTE SCLC_P vs SCLC_A (EdgeR + scale)", x = 0.5, y = 0.95)
```

## DTE sequins(edgeR + scaling)
```{r upset plot dte sequins edgeR+scaling, fig.width=8, fig.height=6}

# Upset plot
## A vs B (AvsB)
all_elements <- unique(c(
  ILL_BULK = ill_bulk.scl_dte_sequins.list$AvsB.DE,
  ONT_BULK = ont_bulk.scl_dte_sequins.list$AvsB.DE,
  PB_BULK = pb_bulk.scl_dte_sequins.list$AvsB.DE,
  dRNA_BULK = drna_bulk.scl_dte_sequins.list$AvsB.DE
))

plot_d_AvsB.DE <- data.frame(
  Element = all_elements,
  ill_bulk= as.integer(all_elements %in% ill_bulk.scl_dte_sequins.list$AvsB.DE),
  ont_bulk= as.integer(all_elements %in% ont_bulk.scl_dte_sequins.list$AvsB.DE),
  pb_bulk = as.integer(all_elements %in% pb_bulk.scl_dte_sequins.list$AvsB.DE),
  drna_bulk = as.integer(all_elements %in% drna_bulk.scl_dte_sequins.list$AvsB.DE)
)
pp4 <- upset(plot_d_AvsB.DE,  order.by = "freq", nsets = 8) 


pp4
grid::grid.text("DTE SequinsA vs SequinsB (EdgeR + scale)", x = 0.5, y = 0.95)


```





## Spice in DE 
```{r spice in de, fig.width=8, fig.height=6}

compare_sequins_logFC <- function(glm.test.rst, fig.title){

  sequins.gt <- read_tsv(params$sequins_tsv)
  sequins.gt$gt.logFC <- log(sequins.gt$MIX_A/sequins.gt$MIX_B)


  # pb_bulk.dte_sequins.list
  df <- glm.test.rst %>% 
    topTags(adjust.method = 'BH', n = nrow(glm.test.rst))  %>% 
    data.frame() %>% tibble::rownames_to_column(var="NAME") %>% 
    select(NAME, logFC, FDR) 

  # join with sequins ground truth
  df <- df %>% 
    left_join(sequins.gt %>% select(NAME, gt.logFC), by = c( "NAME"))

  # add a column for significant DE, "down" for downregulated, "up" for upregulated
  df$DE <- ifelse(df$logFC > 0 & df$FDR < 0.05, "up", ifelse(df$logFC < 0 & df$FDR < 0.05, "down", "not_sig"))

  # plot gt.logFC vs logFC, add correlation
  p <- ggplot(df, aes(x = logFC, y = gt.logFC, color = DE)) +
    geom_point() +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    scale_color_viridis_d() +
    theme_minimal() +
    labs(x = "Estimated logFC", y = " Ground truth logFC", color = "DE") +
    annotate("text", x = min(df$logFC), y = max(df$gt.logFC), 
          label = paste("Correlation: ", round(cor(df$logFC, df$gt.logFC), 2),
                        "\nMean Absolute Error: ", round(mean(abs(df$logFC - df$gt.logFC)), 2)),
          hjust = 0, vjust = 1, 
          color = "black", size = 4, 
          fontface = "italic") +
    ggtitle(fig.title)
  return(p)
}


compare_sirv_logFC <- function(glm.test.rst, fig.title){
  # pb_bulk.dte_sequins.list
  df <- glm.test.rst %>% 
    topTags(adjust.method = 'BH', n = nrow(glm.test.rst))  %>% 
    data.frame() %>% tibble::rownames_to_column(var="NAME") %>% 
    select(NAME, logFC, FDR)  %>% mutate(gt.logFC = 0)

  # add a column for significant DE, "down" for downregulated, "up" for upregulated
  df$DE <- ifelse(df$logFC > 0 & df$FDR < 0.05, "up", ifelse(df$logFC < 0 & df$FDR < 0.05, "down", "not_sig"))

  # plot gt.logFC vs logFC, add correlation
  p <- ggplot(df, aes(x = logFC, color = DE)) +
    # bine dot plot
    geom_dotplot(binwidth = (max(df$logFC) - min(df$logFC))/200, stackdir = "center") + 
    # remove y axis
    scale_color_viridis_d() +
    theme_minimal() +
    theme(axis.title.y=element_blank(), axis.text.y=element_blank(), axis.ticks.y=element_blank()) +
    # add a vertical line at 0
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "Estimated logFC", color = "DE") +
    annotate("text", x = min(df$logFC), y = 2, 
          label = paste("Mean Absolute Error: ", round(mean(abs(df$logFC - df$gt.logFC)), 2)),
          hjust = 0, vjust = 1, 
          color = "black", size = 4, 
          fontface = "italic") +
    ggtitle(fig.title)
  return(p)
}
```

```{r spice in de edgeR + scaling, fig.width=8, fig.height=6}
p1 <- compare_sequins_logFC(ill_bulk.dte_sequins.list$test_rst, "Illumina Bulk")
p2 <- compare_sequins_logFC(ont_bulk.dte_sequins.list$test_rst, "ONT Bulk")
p3 <- compare_sequins_logFC(pb_bulk.dte_sequins.list$test_rst, "PacBio Bulk")
p4 <- compare_sequins_logFC(ont_drna.dte_sequins.list$test_rst, "ONT dRNA")
grid.arrange(p1, p2, p3, p4, ncol = 2)

p1 <- compare_sequins_logFC(ill_bulk.scl_dte_sequins.list$test_rst, "Illumina Bulk (scaled by Overdispersion)")
p2 <- compare_sequins_logFC(ont_bulk.scl_dte_sequins.list$test_rst, "ONT Bulk (scaled by Overdispersion)")
p3 <- compare_sequins_logFC(pb_bulk.scl_dte_sequins.list$test_rst, "PacBio Bulk (scaled by Overdispersion)")
p4 <- compare_sequins_logFC(ont_drna.scl_dte_sequins.list$test_rst, "ONT dRNA (scaled by Overdispersion)")
grid.arrange(p1, p2, p3, p4, ncol = 2)


p1 <- compare_sirv_logFC(ill_bulk.dte_sirv.list$test_rst, "Illumina Bulk")
p2 <- compare_sirv_logFC(ont_bulk.dte_sirv.list$test_rst, "ONT Bulk")
p3 <- compare_sirv_logFC(pb_bulk.dte_sirv.list$test_rst, "PacBio Bulk")
p4 <- compare_sirv_logFC(ont_drna.dte_sirv.list$test_rst, "ONT dRNA")
grid.arrange(p1, p2, p3, p4, ncol = 1)


p1 <- compare_sirv_logFC(ill_bulk.scl_dte_sirv.list$test_rst, "Illumina Bulk (scaled by Overdispersion)")
p2 <- compare_sirv_logFC(ont_bulk.scl_dte_sirv.list$test_rst, "ONT Bulk (scaled by Overdispersion)")
p3 <- compare_sirv_logFC(pb_bulk.scl_dte_sirv.list$test_rst, "PacBio Bulk (scaled by Overdispersion)")
p4 <- compare_sirv_logFC(ont_drna.scl_dte_sirv.list$test_rst, "ONT dRNA (scaled by Overdispersion)")
grid.arrange(p1, p2, p3, p4, ncol = 1)


# save Rdata
# save.image("~/vast/DE_with_salmon_oarfish.RData")
# load("~/vast/DE_with_salmon_oarfish.RData")
```