---
title: "Mutation/Variant Calling (Downsampled Long Reads)"
author: "Yupei & Margaux"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
version: "v1.0"
params:
  lr_bulk_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation"
  sr_bulk_dir: "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation"
  lr_whatshap_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation/whatshap"
  sr_whatshap_dir: "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation/whatshap"
  variant_analysis_dir: "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds"
  variant_analysis_phased_dir: "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds_phased"
  summary_dir: "/vast/projects/LongBench/analysis/variant_allele_specific_analysis/"
  longcallR_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/LongcallR"
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 8,
  fig.height = 6,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir),
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"),
                      params$cache_dir))


```

```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(patchwork)
library(GenomicRanges)
library(rtracklayer)


# # Temporarily add params as a runable list
# params <- list(
#   lr_bulk_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation",
#   sr_bulk_dir = "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation",
#   lr_whatshap_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation/whatshap",
#   sr_whatshap_dir = "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation/whatshap",
#   variant_analysis_dir = "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds",
#   variant_analysis_phased_dir = "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds_phased",
#   summary_dir = "/vast/projects/LongBench/analysis/variant_allele_specific_analysis/",
#   longcallR_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/LongcallR",
#   random_seed = 2024,
#   cache_dir = NULL,
#   fig.path = NULL
# )

color_map <- c(
  "Illumina"               = "#E87D24",
  "Illumina_DV"   = "#914f19",
  "PacBio"                 = "#DA1884",
  "ONT dRNA"               = "#007EA7",
  "ONT cDNA"               = "#003F5C"
)
```

# Part 1: Genomic region covered by >=30 read_tsv
```{r Genomic region coverage ge 30, fig.height=6, fig.width=8}
files <- list.files(params$lr_bulk_dir, pattern = "\\.genomic_coverage\\.txt$", full.names = TRUE)
files <- c(files, list.files(params$sr_bulk_dir, pattern = "\\.genomic_coverage\\.txt$", full.names = TRUE))

extract_coverage <- function(file) {
    table <- read_tsv(file, col_names = TRUE)
    table$file <- basename(file)
    return(table)
}

# Apply the function to all files and combine results
coverage_df <- lapply(files, extract_coverage) %>% bind_rows()
# Separate the file name to extract sample ID and Cell line
coverage_df %>%
    mutate(file = sub("\\.genomic_coverage\\.txt$", "", file)) %>%
    mutate(pct_covered = Covered / Total * 100) %>%
    mutate(
        Protocols = case_when(
            # match substring in file name
            grepl("dRNA", file) ~ "ONT dRNA",
            grepl("ont", file) ~ "ONT cDNA",
            grepl("pb", file) ~ "PacBio",
            TRUE ~ "Illumina"
        ),
        # remove any the before _
        Cell_lines = sub(".*_", "", file)
    ) %>%
    select(-file) -> coverage_df

# Assuming your data frame is called coverage_df

coverage.p <- coverage_df %>%
  ggplot(aes(x = Protocols, y = Covered/1000000, color = Protocols)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(aes(color = Protocols), width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~ Region, nrow = 1, scales = "free_y") +
  labs(title = "Genome Coverage (>=30 reads)",
       x = "Protocol",
       y = "Genome (Mbp) Covered by ≥30 Reads ",
       fill = "Protocol") +
  scale_fill_manual(values = color_map) +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))



# Assuming your data frame is called coverage_df
coverage.p.pct <- coverage_df %>%
  ggplot(aes(x = Protocols, y = pct_covered, color = Protocols)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(aes(color = Protocols), width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~ Region, nrow = 1, scales = "free_y") +
  labs(title = "Genome Coverage (>=30 reads)",
       x = "Protocol",
       y = "Percent Covered by ≥30 Reads (%)",
       fill = "Protocol") +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

coverage.p / coverage.p.pct
```


# Part 2: Variant calling and phasing summary
## Whatshap stat:
```{r parse the whatshap stat files, echo=FALSE, fig.height=3, fig.width=3}
# List all .stat files
files <- list.files(params$lr_whatshap_dir, pattern = "\\.vcf\\.stat$", full.names = TRUE)
files <- c(files, list.files(params$sr_whatshap_dir, pattern = "\\.vcf\\.stat$", full.names = TRUE))

# Function to extract summary stats per file
valid_chr <- c(as.character(1:22), "X", "Y")

extract_and_sum <- function(file) {
  lines <- readLines(file)
  chr <- NA
  stats <- tibble()

  for (i in seq_along(lines)) {
    line <- lines[i]
    if (grepl("^---------------- Chromosome ", line)) {
      chr <- sub("^---------------- Chromosome chr(.+) ----------------$", "\\1", line)
    }

    if (grepl("Variants in VCF:", line)) {
        vcf_variants <- as.integer(trimws(strsplit(line, ":")[[1]][2]))
    }
    if (grepl("^\\s*Heterozygous:", line)) {
    matches <- regmatches(line, regexec("Heterozygous:\\s+([0-9]+).*?\\((\\s*[0-9]+)\\s+SNVs\\)", line))[[1]]
    heterozygous <- as.integer(matches[2])
    heterozygous_snv <- as.integer(matches[3])
    }
    if (grepl("^\\s*Phased:", line)) {
    matches <- regmatches(line, regexec("Phased:\\s+([0-9]+).*?\\((\\s*[0-9]+)\\s+SNVs\\)", line))[[1]]
    phased <- as.integer(matches[2])
    phased_snv <- as.integer(matches[3])
    }
    if (grepl("Longest block:", line)) {
      longest_block <- as.numeric(gsub("bp", "", trimws(strsplit(line, ":")[[1]][2])))
    }
    # end of the chunk
    if (grepl("Block NG50:", line)) {
      if (!is.na(chr) && chr %in% valid_chr) {
        stats <- bind_rows(stats, tibble(
          file = basename(file),
          chromosome = chr,
          vcf_variants = vcf_variants,
          heterozygous = heterozygous,
            heterozygous_snv = heterozygous_snv,
          phased = phased,
            phased_snv = phased_snv,
          longest_block = longest_block
        ))
      }
    }
  }

stats %>%
  group_by(file) %>%
  summarise(
    across(c(vcf_variants, heterozygous, heterozygous_snv, phased, phased_snv), sum, na.rm = TRUE),
    longest_block = max(longest_block, na.rm = TRUE),
    .groups = "drop"
  )
}

# Apply the function to all files and combine results
stats_df <- lapply(files, extract_and_sum) %>% bind_rows()
# Separate the file name to extract sample ID and Cell line
stats_df %>%
    mutate(file = sub("\\.phased\\.vcf\\.stat$", "", file)) %>%
    mutate(
        Protocols = case_when(
            # match substring in file name
            grepl("dRNA", file) ~ "ONT dRNA",
            grepl("ont", file) ~ "ONT cDNA",
            grepl("pb", file) ~ "PacBio",
            TRUE ~ "Illumina"
        ),
        # remove any the before _
        Cell_lines = sub(".*_", "", file)
    ) %>%
    select(-file) -> stats_df
```

```{r phased variance summary, fig.height=3, fig.width=8}
# Dot plot: Total variant calls per cell line, colored by protocol
stats_df %>%
    ggplot(aes(x = Cell_lines, y = vcf_variants, color = Protocols)) +
    geom_point(size = 3) +
    labs(title = "Total Variant Calls per Cell Line",
         x = "Cell Lines", y = "Total Variants") +
    scale_color_manual(values = color_map) +
    theme_minimal() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Bar plot: Total variants per protocol
stats_df %>%
    ggplot() +
    geom_bar(aes(x = Cell_lines, y = vcf_variants, fill = Protocols), stat = "identity", position = "dodge", alpha = 1) +
    facet_wrap(~Protocols, nrow = 1) +
    labs(title = "Total Variants per Protocol",
         x = "Cell Lines", y = "Total Variants") +
    theme_minimal() +
    scale_fill_manual(values = color_map) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Add Phasing status
stats_df %>%
    mutate(
        "Unphased heterozygous" = heterozygous - phased,
        Homosygous = vcf_variants - heterozygous
    ) %>%
    dplyr::rename(`Phased heterozygous` = phased) %>%
    select(Cell_lines, Protocols, `Phased heterozygous`, `Unphased heterozygous`, Homosygous) %>%
    pivot_longer(cols = c(`Unphased heterozygous`, `Phased heterozygous`, Homosygous), names_to = "variant_type", values_to = "count") %>%
    mutate(variant_type = factor(variant_type, levels = c("Homosygous", "Unphased heterozygous", "Phased heterozygous"))) %>%
    group_by(Cell_lines, Protocols) %>%
    mutate(total = sum(count)) %>%
    ungroup() %>%
    ggplot(aes(x = Cell_lines)) +
    # Stacked bars
    geom_bar(aes(y = count, fill = Protocols, alpha = variant_type), stat = "identity", position = "stack") +
    geom_col(aes(y = total, color = Protocols, group = Protocols), fill = NA, position = position_dodge(width = 0.9), linewidth = 0.1) +
    facet_wrap(~Protocols, nrow = 1) +
    scale_alpha_manual(values = c(`Phased heterozygous` = 1, `Unphased heterozygous` = 0.5, Homosygous = 0.0)) +
    scale_fill_manual(values = color_map) +
    scale_color_manual(values = color_map) +
  labs(title = "Phased vs Unphased Heterozygous Variants per Cell Line",
       x = "Cell Lines", y = "Variant Count", alpha = "Variant Type") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Variance calls summary

```{r load gtf and var summary, cache=TRUE}
gtf   <- import('/vast/projects/LongBench/reference_files/GRCh38.add.spikein/gencode.v44.transcripts.add.spikein.gtf')
exons <- gtf[gtf$type == "exon"]
genes <- gtf[gtf$type == "gene"]
cell_lines <- c("H146", "H1975", "H211", "H2228", "H526", "H69", "HCC827", "SHP77")
var_summary <- bind_rows(
  expand_grid(sample = c("ont_bulk", "dRNA_bulk", "pb_bulk"), cell_line = cell_lines) %>%
    mutate(
      vcf_file = file.path(params$lr_whatshap_dir, paste0(sample, "_", cell_line, ".phased.vcf")),
      protocol = recode(sample, "ont_bulk" = "ont", "dRNA_bulk" = "drna", "pb_bulk" = "pacbio")
    ),
  tibble(cell_line = cell_lines) %>%
    mutate(
      vcf_file = file.path(params$sr_whatshap_dir, paste0(cell_line, ".phased.vcf")),
      protocol = "illumina",
      sample   = "illumina"
    )
) %>%
  pmap_dfr(function(sample, cell_line, vcf_file, protocol) {
    read_tsv(vcf_file, comment = "##", show_col_types = FALSE) %>%
      dplyr::rename(CHROM = `#CHROM`) %>%
      mutate(protocol = protocol, cell_line = cell_line)
  }) %>%
  filter(FILTER == "PASS", QUAL >= 10) %>%
  {
    gr <- GRanges(seqnames = .$CHROM, ranges = IRanges(as.integer(.$POS), width = 1))
    mutate(., intergenic = countOverlaps(gr, genes) == 0, exonic = countOverlaps(gr, exons) > 0)
  }
```

```{r Variance calls summary, fig.height=5, fig.width=6}

var_summary <- var_summary %>%
  mutate(intronic = !exonic & !intergenic) %>%
  mutate(Protocols = recode(protocol,
    "ont"      = "ONT cDNA",
    "drna"     = "ONT dRNA",
    "pacbio"   = "PacBio",
    "illumina" = "Illumina"
  ))

# Plot1 Total number of variants per protocol, stacked by intergenic, exonic, intronic
p.box <- var_summary %>%
  group_by(Protocols, cell_line) %>%
  summarise(
    total_variants = n(),
    intergenic = sum(intergenic),
    exonic = sum(exonic),
    intronic = sum(intronic),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(intergenic, exonic, intronic), names_to = "region_type", values_to = "count") %>%
  ggplot(aes(x = Protocols, y = count, color = Protocols)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(aes(color = Protocols), width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~region_type, nrow = 1, scales = "free_y") +
  labs(title = "Total Variants per Protocol by Region Type",
       x = "Cell Lines", y = "Total Variants") +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p.bar <- var_summary %>%
  group_by(Protocols) %>%
  summarise(
    total_variants = n(),
    intergenic = sum(intergenic),
    exonic = sum(exonic),
    intronic = sum(intronic),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(intergenic, exonic, intronic), names_to = "region_type", values_to = "count") %>%
  ggplot(aes(x = Protocols, y = count, fill = Protocols)) +
  geom_bar(stat = "identity", position = "stack") +
  facet_wrap(~region_type, nrow = 1, scales = "free_y") +
  labs(title = "Total Variants per Protocol by Region Type",
       x = "Cell Lines", y = "Total Variants") +
  scale_fill_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

coverage.p / p.box
```

```{r, upset plot overlap site and genotype, fig.height=3, fig.width=10}
variants.upset.mat <- var_summary %>%
  mutate(var_pos = paste(CHROM, POS, sep = "_")) %>%
  mutate(variant_id = paste(CHROM, POS, REF, ALT, cell_line, sep = "_")) %>%
  distinct(variant_id, Protocols, .keep_all = TRUE) %>%
  mutate(present = TRUE) %>%
  pivot_wider(id_cols = variant_id, names_from = Protocols, values_from = present, values_fill = FALSE)


variants.upset.mat <- variants.upset.mat %>% left_join(
  var_summary %>%
    mutate(variant_id = paste(CHROM, POS, REF, ALT, cell_line, sep = "_")) %>%
    mutate(genomic_region = case_when(
      intergenic ~ "intergenic",
      exonic ~ "exonic",
      TRUE ~ "intronic"
    )) %>%
    distinct(variant_id, genomic_region),
  by = "variant_id"
)


ComplexUpset::upset(
  variants.upset.mat %>%
    filter((`ONT cDNA` | PacBio | `ONT dRNA` | Illumina) & genomic_region == "exonic"),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio", "Illumina"),
  name = "Variant Count",
  width_ratio = 0.25
)
```


# Part 3: Accuracy of mutation calling using Spike-in  and CCLE
```{r}
# Load the data
FP.matrix <- read_tsv(file.path(params$variant_analysis_dir, "SNP_FP_QUAL_detail.tsv"))
TP.matrix <- list.files(file.path(params$variant_analysis_dir, "genotype_comparisons_qual"),
                        pattern = "genotype_comparison_.*\\.tsv$", full.names = TRUE) %>%
  lapply(function(f) {
    read_tsv(f) %>% mutate(sample = sub("genotype_comparison_(.*)\\.tsv", "\\1", basename(f)))
  }) %>%
  bind_rows()
```

### Plotting the number of FP  per method
```{r FP}
# Organize the data
FP.matrix <- FP.matrix %>%
  # Split the SNP_ID column into two new columns: sample and locus
  separate(SNP_ID, into = c("sample", "locus"), sep = ":", extra = "merge") %>%
  dplyr::rename(
    PacBio = pacbio,
    `ONT dRNA` = drna,
    `ONT cDNA` = ont,
    Illumina = illumina,
    Illumina_DV = deepvariant
  )

```


```{r Spikein_FP, fig.height=3, fig.width=4}

# Correct color map
color_map <- c(
  "Illumina"               = "#E87D24",
  "Illumina_DV"   = "#914f19",
  "PacBio"                 = "#DA1884",
  "ONT dRNA"               = "#007EA7",
  "ONT cDNA"               = "#003F5C"
)

FP.summary <- FP.matrix %>%
  select(-locus) %>%
  group_by(sample) %>%
  summarise(across(everything(), sum, na.rm = TRUE), .groups = "drop")

# Prepare data
FP.summary %>%
  select(-Illumina_DV) %>%
  pivot_longer(-sample, names_to = "platform", values_to = "false_positives") %>%
  ggplot(aes(x = platform, y = false_positives, color = platform)) +
    geom_jitter(width = 0.3, size = 1.5) +  # smaller dots
    scale_color_manual(values = color_map) +
    labs(title = "False Positives by Platform",
        x = "Platform", y = "False Positives", color = "Platform") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 20, hjust = 1),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )


# Sup  with DV
FP.summary %>%
  pivot_longer(-sample, names_to = "platform", values_to = "false_positives") %>%
  ggplot(aes(x = platform, y = false_positives, color = platform)) +
    geom_jitter(width = 0.3, size = 1.5) +
    scale_color_manual(values = color_map) +
    labs(title = "False Positives by Platform",
        x = "Platform", y = "False Positives", color = "Platform") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 20, hjust = 1),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )
```

### Explore illumina
```{r, include=F, eval=T}
FP.matrix.split <- FP.matrix %>%
  separate(locus, into = c("chr", "pos", "REF", "ALT"), sep = ":", extra = "merge")

# Convert to GRanges
pos_gr <- GRanges(
  seqnames = FP.matrix.split$chr,
  ranges = IRanges(as.integer(FP.matrix.split$pos), width = 1)
)

FP.matrix.split$is.exonic <- countOverlaps(pos_gr, exons) > 0

color_map <- c(
  "Illumina"     = "#E87D24",
  "PacBio"       = "#DA1884",
  "ONT dRNA"     = "#007EA7",
  "ONT cDNA"     = "#003F5C"
)

# Plotting the proportion of exonic and non-exonic variants called by each platform
FP.matrix.split %>%
  pivot_longer(cols = names(color_map),
               names_to = "platform", values_to = "called") %>%
  filter(called == TRUE) %>%
  group_by(platform) %>%
  summarise(proportion_exonic = mean(is.exonic == TRUE)) %>%
  ggplot(aes(x = platform, y = proportion_exonic, fill = platform)) +
  geom_col() +
  scale_fill_manual(values = color_map) +
  labs(y = "Proportion exonic", x = "Platform") +
  theme_minimal()

# Plotting the proportion of non-exonic variants called by each platform
FP.matrix.split %>%
  select(-Illumina_DV) %>%
  pivot_longer(cols = names(color_map),
               names_to = "platform", values_to = "called") %>%
  filter(called == TRUE) %>%
  group_by(platform) %>%
  summarise(proportion_non_exonic = mean(is.exonic == FALSE)) %>%
  ggplot(aes(x = platform, y = proportion_non_exonic, fill = platform)) +
  geom_col() +
  scale_fill_manual(values = color_map) +
  labs(y = "Proportion non-exonic variants", x = "Platform") +
  theme_minimal()

# FP on exonic only
FP.summary <- FP.matrix.split %>%
  filter(is.exonic == TRUE) %>%
  select( -c(chr, pos, REF, ALT, is.exonic)) %>%
  group_by(sample) %>%
  summarise(across(everything(), sum, na.rm = TRUE), .groups = "drop")

# Prepare data
FP.summary %>%
  select(-Illumina_DV) %>%
  pivot_longer(-sample, names_to = "platform", values_to = "false_positives") %>%
  ggplot(aes(x = platform, y = false_positives, color = platform)) +
    geom_jitter(width = 0.3, size = 1.5) +
    scale_color_manual(values = color_map) +
    labs(title = "False Positives by Platform (on exonic variants only)",
        x = "Platform", y = "False Positives", color = "Platform") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 20, hjust = 1),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )

# FP on SIRV only
FP.matrix.split %>%
  mutate(panel = case_when(
    str_starts(chr, "SIRV") ~ "SIRV",
    str_starts(chr, "ERCC") ~ "ERCC",
    str_starts(chr, "chrIS") ~ "chrIS",
    TRUE ~ NA_character_
  )) %>%
  pivot_longer(
    cols = c(PacBio, `ONT dRNA`, `ONT cDNA`, Illumina),
    names_to = "platform", values_to = "called"
  ) %>%
  group_by(panel, platform, sample) %>%
  summarise(
    called = sum(called),
    .groups = "drop"
  ) %>%
  ggplot(aes(x = platform, y = called, fill = platform)) +
  geom_boxplot() +
  facet_wrap(~panel) +
  scale_fill_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## TP

### Add genomic region type
```{r }
pos_gr <- GRanges(
  seqnames = TP.matrix$CHROM,
  ranges = IRanges(as.integer(TP.matrix$POS), width = 1)
)
is.exonic <- countOverlaps(pos_gr, exons) > 0
is.gene <- countOverlaps(pos_gr, genes) > 0
is.intronic <- is.gene & !is.exonic
is.intergenic <- !is.gene

TP.matrix$region <- ifelse(is.exonic, "exonic",
                     ifelse(is.intronic, "intronic", "intergenic"))
```

```{r CCLE TP v1}
# Main without DV
TP.matrix %>%
  select(variant_type, ends_with("_STATUS"), region) %>%
  select(-deepvariant_STATUS) %>%
  filter(variant_type %in% c("SNP", "indel")) %>%
  pivot_longer(
    cols = ends_with("_STATUS"),
    names_to = "platform",
    values_to = "status"
  ) %>%
  mutate(
    status = recode(
      status,
      "match" = "Match",
      "discordant" = "Discordant",
      "lowqual" = "Low Quality",
      "not_called" = "Not Called"
    )
  )  %>%
  mutate(
    platform = sub("_STATUS", "", platform),
    platform = recode(platform,
                      "illumina" = "Illumina",
                      "deepvariant" = "Illumina deepvariant",
                      "pacbio" = "PacBio",
                      "ont" = "ONT cDNA",
                      "drna" = "ONT dRNA")
  )  %>%
  mutate(status = recode(status, "Low Quality" = "Not Called")) %>%
  mutate(status = factor(status, levels = c("Not Called", "Discordant",  "Match")))-> TP.summary.mtx

# Plotting
TP.summary.mtx %>%
  group_by(variant_type, platform, status, region) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = platform, y = count, fill = status)) +
  geom_col(position = "stack") +
  facet_wrap(~  region + variant_type, scales = "free_y") +
  scale_fill_manual(values = c("Not Called" = "#C9C9C9", "Match" = "#1b9e77", "Discordant" = "#CF9741")) +
  labs(x = "Status", y = "Count", fill = "Platform") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# Facet by region
TP.summary.mtx %>%
  group_by(platform, status, region) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = platform, y = count, fill = status)) +
  geom_col(position = "stack") +
  facet_wrap(~  region, scales = "free_y") +
  scale_fill_manual(values = c("Not Called" = "#C9C9C9", "Match" = "#1b9e77", "Discordant" = "#CF9741")) +
  labs(x = "Status", y = "Count", fill = "Platform") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
# Facet by variant type
TP.summary.mtx %>%
  group_by(platform, status, variant_type) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = platform, y = count, fill = status)) +
  geom_col(position = "stack") +
  facet_wrap(~  variant_type, scales = "free_y") +
  scale_fill_manual(values = c("Not Called" = "#C9C9C9", "Match" = "#1b9e77", "Discordant" = "#CF9741")) +
  labs(x = "Status", y = "Count", fill = "Platform") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Part 3.5 (v2, enable phasing): Accuracy of mutation calling using Spike-in  and CCLE
```{r}
# Load the data

FP.matrix.phase <- read_tsv(file.path(params$variant_analysis_phased_dir, "SNP_FP_QUAL_detail.tsv"))
TP.matrix.phase <- list.files(file.path(params$variant_analysis_phased_dir, "genotype_comparisons_qual"),
                               pattern = "genotype_comparison_.*\\.tsv$", full.names = TRUE) %>%
  lapply(function(f) {
    read_tsv(f) %>% mutate(sample = sub("genotype_comparison_(.*)\\.tsv", "\\1", basename(f)))
  }) %>%
  bind_rows()
```

### Plotting the number of FP  per method
```{r FP phasing model}
# Organize the data
FP.matrix.phase <- FP.matrix.phase %>%
  # Split the SNP_ID column into two new columns: sample and locus
  separate(SNP_ID, into = c("sample", "locus"), sep = ":", extra = "merge") %>%
  dplyr::rename(
    `PacBio phased` = pacbio,
    `ONT dRNA phased` = drna,
    `ONT cDNA phased` = ont
  ) %>% full_join(FP.matrix, by=c('sample', 'locus')) %>%
  mutate(across(everything(), ~replace_na(.x, FALSE)))

```


```{r Spikein_FP phasing model, fig.height=3, fig.width=6}

# Correct color map
color_map <- c(
  "Illumina"               = "#E87D24",
  "Illumina_DV"   = "#914f19",
  "PacBio"                 = "#DA1884",
  "ONT dRNA"               = "#007EA7",
  "ONT cDNA"               = "#003F5C",
  "PacBio phased" = "#DA1884",
  "ONT dRNA phased" = "#007EA7",
  "ONT cDNA phased" = "#003F5C"
)

FP.summary <- FP.matrix.phase %>%
  select(-locus) %>%
  group_by(sample) %>%
  summarise(across(everything(), sum, na.rm = TRUE), .groups = "drop")

# Prepare data
FP.summary %>%
  pivot_longer(-sample, names_to = "platform", values_to = "false_positives") %>%
  ggplot(aes(x = platform, y = false_positives, color = platform)) +
    geom_boxplot(width = 0.3, size = 1) +
    scale_color_manual(values = color_map) +
    labs(title = "False Positives by Platform",
        x = "Platform", y = "False Positives", color = "Platform") +
    theme_minimal(base_size = 14) +
    theme(
      axis.text.x = element_text(angle = 20, hjust = 1),
      legend.title = element_text(size = 13),
      legend.text = element_text(size = 12)
    )
```


## TP
```{r CCLE TP}
# Main without DV
TP.matrix.phase %>%
  pivot_longer(
    cols = ends_with("_STATUS"),
    names_to = "platform",
    values_to = "status"
  ) %>%
  mutate(
    status = recode(
      status,
      "match" = "Match",
      "discordant" = "Discordant",
      "lowqual" = "Low Quality",
      "not_called" = "Not Called"
    )
  )  %>%
  mutate(
    platform = sub("_STATUS", "", platform),
    platform = recode(platform,
                      "pacbio" = "PacBio phased",
                      "ont" = "ONT cDNA phased",
                      "drna" = "ONT dRNA phased")
  )  %>%
  mutate(status = recode(status, "Low Quality" = "Not Called")) %>%
  mutate(status = factor(status, levels = c("Not Called", "Discordant",  "Match")))-> TP.summary.mtx.phase
```


```{r CCLE TP plot, fig.height=3,fig.width=4}

TP.matrix %>%
  select(variant_type, ends_with("_STATUS"), region) %>%
  pivot_longer(
    cols = ends_with("_STATUS"),
    names_to = "platform",
    values_to = "status"
  ) %>%
  mutate(
    status = recode(
      status,
      "match" = "Match",
      "discordant" = "Discordant",
      "lowqual" = "Low Quality",
      "not_called" = "Not Called"
    )
  )  %>%
  mutate(
    platform = sub("_STATUS", "", platform),
    platform = recode(platform,
                      "illumina" = "Illumina",
                      "deepvariant" = "Illumina deepvariant",
                      "pacbio" = "PacBio",
                      "ont" = "ONT cDNA",
                      "drna" = "ONT dRNA")
  )  %>%
  mutate(status = recode(status, "Low Quality" = "Not Called")) %>%
  mutate(status = factor(status, levels = c("Not Called", "Discordant",  "Match")))-> TP.summary.mtx


# Plotting
TP.summary.mtx.phase %>%
  select(platform, status) %>%
  bind_rows(TP.summary.mtx %>% select(platform, status)) %>%
  group_by(platform, status) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = platform, y = count, fill = status)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Not Called" = "#C9C9C9", "Match" = "#1b9e77", "Discordant" = "#CF9741")) +
  labs(x = "Status", y = "Count", fill = "Platform") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


# Part 4:  Allelic imbalance
## 1. Allele frequency distribution by platform
### 1.1. Read and arrange the data
```{r AF distribution DP ge 30, fig.height=3, fig.width=4}

full_table <- read_tsv(file.path(params$summary_dir, "Allelic_imbalance/summary.tsv"))

colnames(full_table) <- c("CHROM","POS","ID","REF","ALT","QUAL","FILTER","INFO","FORMAT","SAMPLE","ont","pacbio","drna","illumina")


split_platform <- function(df, colname, prefix) {
  format_keys <- strsplit(df$FORMAT[1], ":")[[1]]
  col_names <- paste0(prefix, "_", format_keys)

  df %>%
    separate_wider_delim(
      !!sym(colname),
      delim = ":",
      names = col_names,
      too_few = "align_start"
    ) %>%
    mutate(
      across(all_of(paste0(prefix, "_AD")), ~ as.integer(.)),
      across(all_of(paste0(prefix, "_DP")), ~ as.integer(.))
    )
}

# Start processing
processed <- full_table %>%
  split_platform("ont", "ont") %>%
  split_platform("pacbio", "pb") %>%
  split_platform("drna", "drna") %>%
  split_platform("illumina", "ill") %>%
  select(
    -c(INFO, FILTER, FORMAT),
    -matches("(_PL|_ALL|_OTH)$")
  ) %>%
  filter(across(matches("_DP$"), ~ . >= 30))

af_long <- processed %>%
  mutate(
    `ONT cDNA` = ont_AD / ont_DP,
    PacBio     = pb_AD / pb_DP,
    `ONT dRNA` = drna_AD / drna_DP,
    Illumina   = ill_AD / ill_DP
  ) %>%
  pivot_longer(cols = c(`ONT cDNA`, PacBio, `ONT dRNA`, Illumina), names_to = "Platform", values_to = "AF") %>%
  filter(!is.na(AF), AF >= 0, AF <= 1) %>%
  mutate(GT = case_when(
    Platform == "ONT cDNA" ~ ont_GT,
    Platform == "ONT dRNA" ~ drna_GT,
    Platform == "PacBio" ~ pb_GT,
    Platform == "Illumina" ~ ill_GT,
    TRUE ~ NA_character_
  ), DP = case_when(
    Platform == "ONT cDNA" ~ ont_DP,
    Platform == "ONT dRNA" ~ drna_DP,
    Platform == "PacBio" ~ pb_DP,
    Platform == "Illumina" ~ ill_DP,
    TRUE ~ NA_integer_
  )) %>%
  filter(GT %in% c("0/1", "1/0"))

ggplot(af_long, aes(x = Platform, y = AF, color = Platform)) +
  geom_violin(trim = FALSE, fill = NA) +
  geom_boxplot(width = 0.1, outlier.size = 0.5) +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  labs(title = "Allele Frequency Distribution by Platform",
       x = "Platform",
       y = "ALT Allele Frequency") +
  theme(legend.position = "none") +
  coord_flip()


# Depth distribution
ggplot(af_long, aes(x = Platform, y = DP, color = Platform)) +
  geom_boxplot(outliers = F) +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  labs(
    title = "Depth Distribution by Platform (Filtered for DP >= 30 already)",
    x = "Platform",
    y = "Read Depth"
  ) +
  scale_y_log10() +
  theme(legend.position = "none") +
  coord_flip()
```

# 3. Binomial Test: Detection of allelic imbalance in full data with dp>=30

```{r Binomial test p value distribution}
binom_test_p <- function(alt, dp) {
  if (is.na(alt) || is.na(dp) || dp == 0) return(NA)
  binom.test(alt, dp, p = 0.5)$p.value
}

processed_with_binom_p <- processed %>%
  filter(across(matches("_GT$"), ~ !is.na(.) & . != "0/0")) %>%
  mutate(
    p_ont  = mapply(binom_test_p, ont_AD, ont_DP),
    p_pb   = mapply(binom_test_p, pb_AD, pb_DP),
    p_drna = mapply(binom_test_p, drna_AD, drna_DP),
    p_ill  = mapply(binom_test_p, ill_AD, ill_DP)
  ) %>%
  mutate(
    adj_p_ont  = p.adjust(p_ont, method = "BH"),
    adj_p_pb   = p.adjust(p_pb, method = "BH"),
    adj_p_drna = p.adjust(p_drna, method = "BH"),
    adj_p_ill  = p.adjust(p_ill, method = "BH")
  )

# plotting the  p-values distribution
p <- ggplot(processed_with_binom_p, aes(x = p_ont)) +
  stat_bin(bins = 20, color = color_map["ONT cDNA"], alpha = 0.7,geom="step") +
  stat_bin(aes(x = p_pb), bins = 20, color = color_map["PacBio"], alpha = 0.7,geom="step") +
  stat_bin(aes(x = p_drna), bins = 20, color = color_map["ONT dRNA"], alpha = 0.7,geom="step") +
  stat_bin(aes(x = p_ill), bins = 20, color = color_map["Illumina"], alpha = 0.7,geom="step") +
  labs(
    title = "Distribution of p-values for Allelic Imbalance",
    x = "p-value",
    y = "Count"
  ) +
  theme_minimal()

p
```

## Upset plot for binomial test
```{r Upset binomial test, fig.height=3, fig.width=8}
library(ComplexUpset)
  processed_with_binom_p %>% mutate(
    `ONT cDNA` = adj_p_ont < 0.05,
    PacBio = adj_p_pb < 0.05,
    `ONT dRNA` = adj_p_drna < 0.05,
    Illumina = adj_p_ill < 0.05
  ) -> processed_with_binom_p

ComplexUpset::upset(
  processed_with_binom_p %>% filter(
    `ONT cDNA` | PacBio | `ONT dRNA` | Illumina
  ),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio", "Illumina"),
  name = "Allelic Imbalance",
  width_ratio = 0.25
)

df_long <- processed_with_binom_p %>%
  filter (
    `ONT dRNA` | `ONT cDNA` | `PacBio` | Illumina
  ) %>%
  pivot_longer(c(`ONT dRNA`, `ONT cDNA`, `PacBio`, Illumina), names_to = "Platform", values_to = "Detected") %>%
  group_by(CHROM, POS, SAMPLE) %>%
  mutate(TotalDetected = sum(Detected)) %>%
  ungroup() %>%
  mutate(
    Category = case_when(
      Detected == 1 & TotalDetected == 1 ~ "Platform-Specific Detection",
      Detected == 0 & TotalDetected == 3 ~ "Platform-Specific Absence",
      Detected == 1 & TotalDetected > 1 ~ "Common Detection (also detected in >0 other platform)",
      Detected == 0 & TotalDetected < 3 ~ "Common Absence (also absent in >0 other platform)",
      TRUE ~ "Other"
    )
  ) %>%
  count(Platform, Category)

df_long$Category <- factor(df_long$Category,
                           levels = c(
                                      "Platform-Specific Absence",
                                      "Common Absence (also absent in >0 other platform)",
                                      "Platform-Specific Detection",
                                      "Common Detection (also detected in >0 other platform)"
                                      ))
df_long %>%
  filter(Category %in% c("Platform-Specific Absence", "Common Detection (also detected in >0 other platform)", "Platform-Specific Detection")) %>%
  mutate(n = if_else(Category == "Platform-Specific Absence", -n, n)) %>%
  ggplot(aes(x = Platform, y = n, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(
    y = "Number of Loci",
    x = "Platform",
    fill = "Category"
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = c(
      "Common Detection (also detected in >0 other platform)" = "#8fbbd4",
      "Platform-Specific Detection" = "#0072B2",
      "Platform-Specific Absence" = "#E69F00"
    )
  ) +
  coord_flip() +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.title.x = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )


```

## Binomial test with standardised power (Subsample reads)
```{r}
dp_target <- 30
set.seed(params$random_seed)
downsample_ad <- function(AD, DP, target = dp_target) {
  prob <- AD / DP
  rbinom(1, size = target, prob = prob)
}

downsample_platform <- function(df, ad_col, dp_col, prefix) {
  df %>%
    mutate(
      !!ad_col := map2_dbl(.data[[ad_col]], .data[[dp_col]],
                                             ~downsample_ad(.x, .y)),
      !!dp_col := dp_target
    )
}

# Apply to all platforms
processed_ds <- processed %>%
  downsample_platform("ont_AD",  "ont_DP",  "ont") %>%
  downsample_platform("pb_AD",   "pb_DP",   "pb") %>%
  downsample_platform("drna_AD", "drna_DP", "drna") %>%
  downsample_platform("ill_AD",  "ill_DP",  "ill")

  # Compute downsampled AF
af_long <- processed_ds %>%
  mutate(
    `ONT cDNA` = ont_AD / ont_DP,
    PacBio     = pb_AD / pb_DP,
    `ONT dRNA` = drna_AD / drna_DP,
    Illumina   = ill_AD / ill_DP
  ) %>%
  pivot_longer(cols = c(`ONT cDNA`, PacBio, `ONT dRNA`, Illumina), names_to = "Platform", values_to = "AF") %>%
  filter(!is.na(AF), AF >= 0, AF <= 1) %>%
  mutate(GT = case_when(
    Platform == "ONT cDNA" ~ ont_GT,
    Platform == "ONT dRNA" ~ drna_GT,
    Platform == "PacBio" ~ pb_GT,
    Platform == "Illumina" ~ ill_GT,
    TRUE ~ NA_character_
  ), DP = case_when(
    Platform == "ONT cDNA" ~ ont_DP,
    Platform == "ONT dRNA" ~ drna_DP,
    Platform == "PacBio" ~ pb_DP,
    Platform == "Illumina" ~ ill_DP,
    TRUE ~ NA_integer_
  )) %>%
  filter(GT %in% c("0/1", "1/0"))

ggplot(af_long, aes(x = Platform, y = AF, color = Platform)) +
  geom_violin(trim = FALSE, fill = NA) +
  geom_boxplot(width = 0.1, outlier.size = 0.5) +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  labs(title = "Allele Frequency Distribution by Platform",
       x = "Platform",
       y = "ALT Allele Frequency") +
  theme(legend.position = "none") +
  coord_flip()
```



# 3. Binomial Test: Detection of allelic imbalance in full data with dp>=30

```{r}

processed_with_binom_p <- processed_ds %>%
  filter(across(matches("_GT$"), ~ !is.na(.) & . != "0/0")) %>%
  mutate(
    p_ont  = mapply(binom_test_p, ont_AD, ont_DP),
    p_pb   = mapply(binom_test_p, pb_AD, pb_DP),
    p_drna = mapply(binom_test_p, drna_AD, drna_DP),
    p_ill  = mapply(binom_test_p, ill_AD, ill_DP)
  ) %>%
  mutate(
    adj_p_ont  = p.adjust(p_ont, method = "BH"),
    adj_p_pb   = p.adjust(p_pb, method = "BH"),
    adj_p_drna = p.adjust(p_drna, method = "BH"),
    adj_p_ill  = p.adjust(p_ill, method = "BH")
  )

```

## Upset plot for binomial test
```{r, fig.width=6, fig.height=3}
library(ComplexUpset)
  processed_with_binom_p %>% mutate(
    `ONT cDNA` = adj_p_ont < 0.05,
    PacBio = adj_p_pb < 0.05,
    `ONT dRNA` = adj_p_drna < 0.05,
    Illumina = adj_p_ill < 0.05
  ) -> processed_with_binom_p

ComplexUpset::upset(
  processed_with_binom_p %>% filter(
    `ONT cDNA` | PacBio | `ONT dRNA` | Illumina
  ),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio", "Illumina"),
  name = "Allelic Imbalance",
  width_ratio = 0.25
)

df_long <- processed_with_binom_p %>%
  filter (
    `ONT dRNA` | `ONT cDNA` | `PacBio` | Illumina
  ) %>%
  pivot_longer(c(`ONT dRNA`, `ONT cDNA`, `PacBio`, Illumina), names_to = "Platform", values_to = "Detected") %>%
  group_by(CHROM, POS, SAMPLE) %>%
  mutate(TotalDetected = sum(Detected)) %>%
  ungroup() %>%
  mutate(
    Category = case_when(
      Detected == 1 & TotalDetected == 1 ~ "Platform-Specific Detection",
      Detected == 0 & TotalDetected == 3 ~ "Platform-Specific Absence",
      Detected == 1 & TotalDetected > 1 ~ "Common Detection (also detected in >0 other platform)",
      Detected == 0 & TotalDetected < 3 ~ "Common Absence (also absent in >0 other platform)",
      TRUE ~ "Other"
    )
  ) %>%
  count(Platform, Category)

df_long$Category <- factor(df_long$Category,
                           levels = c(
                                      "Platform-Specific Absence",
                                      "Common Absence (also absent in >0 other platform)",
                                      "Platform-Specific Detection",
                                      "Common Detection (also detected in >0 other platform)"
                                      ))
df_long %>%
  filter(Category %in% c("Platform-Specific Absence", "Common Detection (also detected in >0 other platform)", "Platform-Specific Detection")) %>%
  mutate(n = if_else(Category == "Platform-Specific Absence", -n, n)) %>%
  ggplot(aes(x = Platform, y = n, fill = Category)) +
  geom_bar(stat = "identity") +
  labs(
    y = "Number of Loci",
    x = "Platform",
    fill = "Category"
  ) +
  theme_minimal() +
  scale_fill_manual(
    values = c(
      "Common Detection (also detected in >0 other platform)" = "#8fbbd4",
      "Platform-Specific Detection" = "#0072B2",
      "Platform-Specific Absence" = "#E69F00"
    )
  ) +
  coord_flip() +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.title.x = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )


df_long %>%
  filter(Category %in% c("Platform-Specific Absence", "Common Detection (also detected in >0 other platform)", "Platform-Specific Detection")) %>%
  mutate(n = if_else(Category == "Platform-Specific Absence", 0, n)) %>%
  ggplot(aes(x = Platform, y = n, fill=Platform)) +
  geom_bar(stat = "identity") +
  labs(
    y = "Number of Loci",
    x = "Platform",
    fill = "Sequencing Protocols"
  ) +
  theme_minimal() +
  scale_fill_manual(values = color_map) +
  theme(
    axis.text.x = element_text(angle = 0),
    axis.title.x = element_text(size = 12),
    axis.text.y = element_text(size = 10)
  )

```

# 4. Beta-binomial test with loci dp>=30 convervative overdispersion (0.001)
```{r}
library(VGAM)

beta_binom_pval <- function(alt, dp, p = 0.5, rho = 0.01) {
  2 * min(
    pbetabinom(alt, size = dp, prob=p, rho = rho),
    1 - pbetabinom(alt - 1, size = dp, prob=p, rho = rho)
  )
}

processed_with_beta_binom_p <- processed %>%
  mutate(
    p_ont  = mapply(beta_binom_pval, ont_AD, ont_DP, rho = 0.001),
    p_pb   = mapply(beta_binom_pval, pb_AD, pb_DP, rho = 0.001),
    p_drna = mapply(beta_binom_pval, drna_AD, drna_DP, rho = 0.001),
    p_ill  = mapply(beta_binom_pval, ill_AD, ill_DP, rho = 0.001)
  ) %>%
  mutate(
    adj_p_ont  = p.adjust(p_ont, method = "BH"),
    adj_p_pb   = p.adjust(p_pb, method = "BH"),
    adj_p_drna = p.adjust(p_drna, method = "BH"),
    adj_p_ill  = p.adjust(p_ill, method = "BH")
  )

# plotting the  p-values distribution
p <- ggplot(processed_with_beta_binom_p, aes(x = p_ont)) +
  stat_bin(bins = 20, color = color_map["ONT cDNA"], alpha = 0.7,geom="step") +
  stat_bin(aes(x = p_pb), bins = 20, color = color_map["PacBio"], alpha = 0.7,geom="step") +
  stat_bin(aes(x = p_drna), bins = 20, color = color_map["ONT dRNA"], alpha = 0.7,geom="step") +
  stat_bin(aes(x = p_ill), bins = 20, color = color_map["Illumina"], alpha = 0.7,geom="step") +
  labs(
    title = "Distribution of p-values for Allelic Imbalance",
    x = "p-value",
    y = "Count"
  ) +
  theme_minimal()

p


```


## Upset plot for beta-binomial test
```{r Upset plot for beta-binomial test}
library(ComplexUpset)
ComplexUpset::upset(
  processed_with_beta_binom_p %>% mutate(
    `ONT cDNA` = adj_p_ont < 0.05,
    PacBio = adj_p_pb < 0.05,
    `ONT dRNA` = adj_p_drna < 0.05,
    Illumina = adj_p_ill < 0.05
  ) %>% filter(
    `ONT cDNA` | PacBio | `ONT dRNA` | Illumina
  ),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio", "Illumina"),
  name = "Allelic Imbalance",
  width_ratio = 0.25
)
```


```{r}
  processed_with_beta_binom_p %>% mutate(
    `ONT cDNA` = adj_p_ont < 0.05,
    PacBio = adj_p_pb < 0.05,
    `ONT dRNA` = adj_p_drna < 0.05,
    Illumina = adj_p_ill < 0.05
  ) %>% filter(
    `ONT cDNA` & PacBio & `ONT dRNA` & Illumina
  )
```


# Part 5: Allele specific expression (1-R^2)


```{r, allele specific splicing 1, fig.height=7, fig.width=4}
ASE <- read_tsv(file.path(params$summary_dir,"Allelic_imbalance/CoverageDiv_summary.tsv")) %>% dplyr::rename(
    "ONT dRNA" = div_drna,
    "ONT cDNA" =  div_ont,
     "PacBio" = div_pacbio,
     "Illumina" = div_illumina
) %>%  mutate(across(everything(), ~ replace_na(., 0)))


library(pheatmap)

# Step 1: Filter rows with max > 0.5
ASE_filtered <- ASE %>%
  filter(pmax(PacBio, `ONT cDNA`, `ONT dRNA`, Illumina) > 0.2)

# Step 2: Convert to matrix and order by max value
mat <- ASE_filtered %>%
  column_to_rownames("locus") %>%
  as.matrix()

row_order <- order(apply(mat, 1, max), decreasing = TRUE)
mat_ordered <- mat[row_order, ]

# Step 3: Plot heatmap
pheatmap(mat_ordered,
         cluster_rows = FALSE,
         cluster_cols = FALSE,
         scale = "none",
         color = viridis::viridis(100),
         fontsize_row = 6)

mat_ordered
```



```{r, allele specific splicing, fig.height=4, fig.width=8}
threshold <- 0.0
ComplexUpset::upset(
  ASE %>%
  mutate(across(c(`PacBio`, `ONT cDNA`, `ONT dRNA`, Illumina), ~ .x > threshold)) %>%
  filter (
    `ONT dRNA` | `ONT cDNA` | `PacBio` | Illumina
  ),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio", "Illumina"),
  name = "Allele Specific Splicing Events",
  width_ratio = 0.25
)
```


# Part 6: LongcallR
```{r ase upsetplot}
longcallR_files <- list.files(
  file.path(params$longcallR_dir),
  pattern = "\\.ase\\.tsv$",
  full.names = TRUE
)

# Read and process the LongcallR results
longcallR_results.ase <- lapply(longcallR_files, function(x) {
  read_tsv(x) %>%
    mutate(file = sub("\\.longcallR\\.ase\\.tsv$", "", basename(x)))%>%
    mutate(
        Protocols = case_when(
            grepl("dRNA", file) ~ "ONT dRNA",
            grepl("ont", file) ~ "ONT cDNA",
            grepl("pb", file) ~ "PacBio",
            TRUE ~ "Illumina"
        ),
        Cell_lines = sub(".*_", "", file)
    )
})  %>%
    bind_rows() %>%
    select(-file) %>%
    dplyr::rename(
        Gene_name = `#Gene_name`
    )
```

## Plotting the LongcallR results (upset plot)
```{r ase upsetplot plot}
library(ComplexUpset)
LongcallR.ase.upset.mat <-
  longcallR_results.ase %>%
  filter(P_value < 0.05) %>%
  mutate(depth_pass = H1 + H2 >= 30) %>%
  group_by(Gene_name, Cell_lines, Protocols) %>%
  filter(all(depth_pass)) %>%
  distinct(Gene_name, Cell_lines, Protocols, depth_pass) %>%
  mutate(found = TRUE) %>%
  pivot_wider(names_from = Protocols, values_from = found, values_fill = FALSE)

ComplexUpset::upset(
  LongcallR.ase.upset.mat %>% filter(
    `ONT dRNA` | `ONT cDNA` | `PacBio`
  ),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio"),
  name = "LongcallR ASE Results",
  width_ratio = 0.25
)
```


```{r ase example}
LongcallR.ase.upset.example <-
  longcallR_results.ase %>%
  filter(P_value < 0.05) %>%
  mutate(depth_pass = H1 + H2 >= 300,
         minor_af = min(H1,H2)/(H1 + H2)) %>%
  group_by(Gene_name, Cell_lines, Protocols) %>%
  filter(all(depth_pass) & all(minor_af < 0.1)) %>%
  distinct(Gene_name, Cell_lines, Protocols, depth_pass) %>%
  mutate(found = TRUE) %>%
  pivot_wider(names_from = Protocols, values_from = found, values_fill = FALSE)

LongcallR.ase.upset.example %>% filter(
  Cell_lines == "SHP77" & `ONT dRNA` & `ONT cDNA` & `PacBio`
) %>% pull(Gene_name)


example <- longcallR_results.ase %>%
  filter(Gene_name == "ZCCHC17" & Cell_lines == "SHP77") %>%
  select(Gene_name, Cell_lines, Protocols, H1, H2, P_value) %>%
  arrange(Protocols)

example <- example %>%
  mutate(
    H1.plot = pmin(H1, H2),
    H2.plot = pmax(H1, H2)
  )

example_long <- example %>%
  pivot_longer(cols = c(H1.plot, H2.plot), names_to = "Allele", values_to = "Read_Count")

ggplot(example_long, aes(x = Protocols, y = Read_Count, fill = Allele)) +
  geom_bar(stat = "identity") +
  labs(
    title = paste("LongcallR Results for", example$Gene_name[1], "in", example$Cell_lines[1]),
    x = "Protocols",
    y = "Read Count",
    fill = "Allele"
  ) +
  scale_fill_manual(values = c("H1.plot" = "#CE9478", "H2.plot" = "#77A3D1")) +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))
```

```{r, MORE ASE example}
example <- longcallR_results.ase %>%
  filter(Gene_name == "SERINC2" & Cell_lines == "SHP77") %>%
  select(Gene_name, Cell_lines, Protocols, H1, H2, P_value) %>%
  arrange(Protocols)

example <- example %>%
  mutate(
    H1.plot = pmin(H1, H2),
    H2.plot = pmax(H1, H2)
  )

example_long <- example %>%
  pivot_longer(cols = c(H1.plot, H2.plot), names_to = "Allele", values_to = "Read_Count")

ggplot(example_long, aes(x = Protocols, y = Read_Count, fill = Allele)) +
  geom_bar(stat = "identity") +
  labs(
    title = paste("LongcallR Results for", example$Gene_name[1], "in", example$Cell_lines[1]),
    x = "Protocols",
    y = "Read Count",
    fill = "Allele"
  ) +
  scale_fill_manual(values = c("H1.plot" = "#CE9478", "H2.plot" = "#77A3D1")) +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))


example <- longcallR_results.ase %>%
  filter(Gene_name == "TM2D1" & Cell_lines == "SHP77") %>%
  select(Gene_name, Cell_lines, Protocols, H1, H2, P_value) %>%
  arrange(Protocols)

example <- example %>%
  mutate(
    H1.plot = pmin(H1, H2),
    H2.plot = pmax(H1, H2)
  )

example_long <- example %>%
  pivot_longer(cols = c(H1.plot, H2.plot), names_to = "Allele", values_to = "Read_Count")

ggplot(example_long, aes(x = Protocols, y = Read_Count, fill = Allele)) +
  geom_bar(stat = "identity") +
  labs(
    title = paste("LongcallR Results for", example$Gene_name[1], "in", example$Cell_lines[1]),
    x = "Protocols",
    y = "Read Count",
    fill = "Allele"
  ) +
  scale_fill_manual(values = c("H1.plot" = "#CE9478", "H2.plot" = "#77A3D1")) +
  theme_minimal() +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1)))

```

## ASJ gene
```{r}
longcallR_files <- list.files(
  file.path(params$longcallR_dir),
  pattern = "merged_genes_exons\\.asj_gene\\.tsv$",
  full.names = TRUE
)

longcallR_results.asj <- lapply(longcallR_files, function(x) {
  read_tsv(x) %>%
    mutate(file = sub("\\.merged_genes_exons\\.asj_gene\\.tsv$", "", basename(x)))%>%
    mutate(
        Protocols = case_when(
            grepl("dRNA", file) ~ "ONT dRNA",
            grepl("ont", file) ~ "ONT cDNA",
            grepl("pb", file) ~ "PacBio",
            TRUE ~ "Illumina"
        ),
        Cell_lines = sub(".*_", "", file)
    )
})  %>%
    bind_rows() %>%
    select(-file) %>%
    dplyr::rename(
        Gene_name = `#Gene_name`
    )

# gene coverage
longcallR_files <- list.files(
  file.path(params$longcallR_dir),
  pattern = "\\.merged_genes_exons\\.gene_coverage\\.tsv$",
  full.names = TRUE
)

longcallR_results.cov <- lapply(longcallR_files, function(x) {
  read_tsv(x) %>%
    mutate(file = sub("\\.merged_genes_exons\\.gene_coverage\\.tsv$", "", basename(x))) %>%
    mutate(
      Protocols = case_when(
        grepl("dRNA", file) ~ "ONT dRNA",
        grepl("ont", file) ~ "ONT cDNA",
        grepl("pb", file) ~ "PacBio",
        TRUE ~ "Illumina"
      ),
      Cell_lines = sub(".*_", "", file)
    )
}) %>%
  bind_rows() %>%
  select(-file) %>%
  dplyr::rename(
    Gene_name = `#Gene_name`
  )

longcallR_results.asj <- longcallR_results.asj %>% left_join(
  longcallR_results.cov,
  by = c("Chr", "Gene_name", "Protocols", "Cell_lines")
)

# Filter by Num_reads
longcallR_results.asj <- longcallR_results.asj %>%
  group_by(Cell_lines, Gene_name) %>%
  filter(n_distinct(Protocols) == 3) %>%
  filter(all(Num_reads > 30)) %>%
  ungroup()

```

```{r}
LongcallR.asj.upset.mat <-
  longcallR_results.asj %>%
  filter(P_value < 0.05) %>%
  group_by(Gene_name, Cell_lines, Protocols) %>%
  distinct(Gene_name, Cell_lines, Protocols) %>%
  mutate(found = TRUE) %>%
  pivot_wider(names_from = Protocols, values_from = found, values_fill = FALSE)

ComplexUpset::upset(
  LongcallR.asj.upset.mat %>% filter(
    `ONT dRNA` | `ONT cDNA` | `PacBio`
  ),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio"),
  name = "LongcallR ASJ Genes",
  width_ratio = 0.25
)

```
