---
title: "Mutation/Variant Calling (Downsampled Long Reads)"
author: "Yupei & Margaux"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
version: "v2.0"
params:
  lr_bulk_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation"
  sr_bulk_dir: "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation"
  lr_whatshap_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation/whatshap"
  sr_whatshap_dir: "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation/whatshap"
  variant_analysis_dir: "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds"
  variant_analysis_phased_dir: "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds_phased"
  summary_dir: "/vast/projects/LongBench/analysis/variant_allele_specific_analysis/"
  longcallR_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/LongcallR_with_spike_in"
  longcallR_ase_dir: "/vast/projects/LongBench/analysis/lr_bulk_ds/result/LongcallR"
  variant_analysis_longcallR_dir: "/vast/projects/LongBench/analysis/variant_analysis/bcftools_longcallR_ds"
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
---
```{r setup, include=FALSE}


if (!exists("params")) {
  params <- list(
  lr_bulk_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation",
  sr_bulk_dir = "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation",
  lr_whatshap_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/Mutation/whatshap",
  sr_whatshap_dir = "/vast/projects/LongBench/analysis/sr_bulk/result/Mutation/whatshap",
  variant_analysis_dir = "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds",
  variant_analysis_phased_dir = "/vast/projects/LongBench/analysis/variant_analysis/bcftools_ds_phased",
  summary_dir = "/vast/projects/LongBench/analysis/variant_allele_specific_analysis/",
  longcallR_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/LongcallR_with_spike_in",
  longcallR_ase_dir = "/vast/projects/LongBench/analysis/lr_bulk_ds/result/LongcallR",
  variant_analysis_longcallR_dir = "/vast/projects/LongBench/analysis/variant_analysis/bcftools_longcallR_ds",
  random_seed = 2024,
  cache_dir = NULL,
  fig.path = NULL
)
}

knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 8,
  fig.height = 6,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir),
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"),
                      params$cache_dir))

if (!is.null(params$fig.path)) {
  dir.create(params$fig.path, recursive = TRUE, showWarnings = FALSE)
}

```



```{r, include=FALSE}
library(tidyverse)
library(dplyr)
library(ggplot2)
library(patchwork)
library(GenomicRanges)
library(rtracklayer)

color_map <- c(
  "Illumina"               = "#E87D24",
  "Illumina_DV"   = "#914f19",
  "PacBio"                 = "#DA1884",
  "ONT dRNA"               = "#007EA7",
  "ONT cDNA"               = "#003F5C"
)
```

# Part 1: Genomic region covered by >=30 read_tsv
```{r Genomic region coverage ge 30, fig.height=6, fig.width=8}
files <- list.files(params$lr_bulk_dir, pattern = "\\.genomic_coverage\\.txt$", full.names = TRUE)
files <- c(files, list.files(params$sr_bulk_dir, pattern = "\\.genomic_coverage\\.txt$", full.names = TRUE))

extract_coverage <- function(file) {
    table <- read_tsv(file, col_names = TRUE)
    table$file <- basename(file)
    return(table)
}

# Apply the function to all files and combine results
coverage_df <- lapply(files, extract_coverage) %>% bind_rows()
# Separate the file name to extract sample ID and Cell line
coverage_df %>%
    mutate(file = sub("\\.genomic_coverage\\.txt$", "", file)) %>%
    mutate(pct_covered = Covered / Total * 100) %>%
    mutate(
        Protocols = case_when(
            # match substring in file name
            grepl("dRNA", file) ~ "ONT dRNA",
            grepl("ont", file) ~ "ONT cDNA",
            grepl("pb", file) ~ "PacBio",
            TRUE ~ "Illumina"
        ),
        # remove any the before _
        Cell_lines = sub(".*_", "", file)
    ) %>%
    select(-file) -> coverage_df

# Assuming your data frame is called coverage_df

coverage.p <- coverage_df %>%
  ggplot(aes(x = Protocols, y = Covered/1000000, color = Protocols)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(aes(color = Protocols), width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~ Region, nrow = 1, scales = "free_y") +
  labs(title = "Genome Coverage (>=30 reads)",
       x = "Protocol",
       y = "Genome (Mbp) Covered by ≥30 Reads ",
       fill = "Protocol") +
  scale_fill_manual(values = color_map) +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


# Save a csv with no of this chunk in fig.path if exists
if (!is.null(params$fig.path)) {
  write_csv(coverage_df, file.path(params$fig.path, "genome_coverage_summary.csv"))
}
```


# Part 2: Variant calling and phasing summary
## Whatshap stat:

Parse WhatsHap phasing statistics from `.vcf.stat` files for each sample. Summarises total variant counts, heterozygous and phased variant counts, and longest phasing block, summed across chromosomes 1–22, X and Y.

```{r parse the whatshap stat files, echo=FALSE, fig.height=3, fig.width=3}
# List all .stat files
files <- list.files(params$lr_whatshap_dir, pattern = "\\.vcf\\.stat$", full.names = TRUE)
files <- c(files, list.files(params$sr_whatshap_dir, pattern = "\\.vcf\\.stat$", full.names = TRUE))

# Function to extract summary stats per file
valid_chr <- c(as.character(1:22), "X", "Y")

extract_and_sum <- function(file) {
  lines <- readLines(file)
  chr <- NA
  stats <- tibble()

  for (i in seq_along(lines)) {
    line <- lines[i]
    if (grepl("^---------------- Chromosome ", line)) {
      chr <- sub("^---------------- Chromosome chr(.+) ----------------$", "\\1", line)
    }

    if (grepl("Variants in VCF:", line)) {
        vcf_variants <- as.integer(trimws(strsplit(line, ":")[[1]][2]))
    }
    if (grepl("^\\s*Heterozygous:", line)) {
    matches <- regmatches(line, regexec("Heterozygous:\\s+([0-9]+).*?\\((\\s*[0-9]+)\\s+SNVs\\)", line))[[1]]
    heterozygous <- as.integer(matches[2])
    heterozygous_snv <- as.integer(matches[3])
    }
    if (grepl("^\\s*Phased:", line)) {
    matches <- regmatches(line, regexec("Phased:\\s+([0-9]+).*?\\((\\s*[0-9]+)\\s+SNVs\\)", line))[[1]]
    phased <- as.integer(matches[2])
    phased_snv <- as.integer(matches[3])
    }
    if (grepl("Longest block:", line)) {
      longest_block <- as.numeric(gsub("bp", "", trimws(strsplit(line, ":")[[1]][2])))
    }
    # end of the chunk
    if (grepl("Block NG50:", line)) {
      if (!is.na(chr) && chr %in% valid_chr) {
        stats <- bind_rows(stats, tibble(
          file = basename(file),
          chromosome = chr,
          vcf_variants = vcf_variants,
          heterozygous = heterozygous,
            heterozygous_snv = heterozygous_snv,
          phased = phased,
            phased_snv = phased_snv,
          longest_block = longest_block
        ))
      }
    }
  }

stats %>%
  group_by(file) %>%
  summarise(
    across(c(vcf_variants, heterozygous, heterozygous_snv, phased, phased_snv), sum, na.rm = TRUE),
    longest_block = max(longest_block, na.rm = TRUE),
    .groups = "drop"
  )
}

# Apply the function to all files and combine results
stats_df <- lapply(files, extract_and_sum) %>% bind_rows()
# Separate the file name to extract sample ID and Cell line
stats_df %>%
    mutate(file = sub("\\.phased\\.vcf\\.stat$", "", file)) %>%
    mutate(
        Protocols = case_when(
            # match substring in file name
            grepl("dRNA", file) ~ "ONT dRNA",
            grepl("ont", file) ~ "ONT cDNA",
            grepl("pb", file) ~ "PacBio",
            TRUE ~ "Illumina"
        ),
        # remove any the before _
        Cell_lines = sub(".*_", "", file)
    ) %>%
    select(-file) -> stats_df

# Save a csv with no of this chunk in fig.path if exists
if (!is.null(params$fig.path)) {
  write_csv(stats_df, file.path(params$fig.path, "whatshap_stat_summary.csv"))
}
```

## Variance calls summary

Load phased VCF files for all protocols and cell lines (FILTER == PASS, QUAL ≥ 10), then annotate each variant as exonic, intronic, or intergenic using the GENCODEv44 reference GTF. This variant table (`var_summary`) is carried forward into Parts 4 and 5.

```{r load gtf and var summary, cache=TRUE}
gtf   <- import('/vast/projects/LongBench/reference_files/GRCh38.add.spikein/gencode.v44.transcripts.add.spikein.gtf')
exons <- gtf[gtf$type == "exon"]
genes <- gtf[gtf$type == "gene"]
cell_lines <- c("H146", "H1975", "H211", "H2228", "H526", "H69", "HCC827", "SHP77")
var_summary <- bind_rows(
  expand_grid(sample = c("ont_bulk", "dRNA_bulk", "pb_bulk"), cell_line = cell_lines) %>%
    mutate(
      vcf_file = file.path(params$lr_whatshap_dir, paste0(sample, "_", cell_line, ".phased.vcf")),
      protocol = recode(sample, "ont_bulk" = "ont", "dRNA_bulk" = "drna", "pb_bulk" = "pacbio")
    ),
  tibble(cell_line = cell_lines) %>%
    mutate(
      vcf_file = file.path(params$sr_whatshap_dir, paste0(cell_line, ".phased.vcf")),
      protocol = "illumina",
      sample   = "illumina"
    )
) %>%
  pmap_dfr(function(sample, cell_line, vcf_file, protocol) {
    read_tsv(vcf_file, comment = "##", show_col_types = FALSE) %>%
      dplyr::rename(CHROM = `#CHROM`) %>%
      mutate(protocol = protocol, cell_line = cell_line)
  }) %>%
  filter(FILTER == "PASS", QUAL >= 10) %>%
  {
    gr <- GRanges(seqnames = .$CHROM, ranges = IRanges(as.integer(.$POS), width = 1))
    mutate(., intergenic = countOverlaps(gr, genes) == 0, exonic = countOverlaps(gr, exons) > 0)
  }
```

```{r Variance calls summary, fig.height=5, fig.width=6}

var_summary <- var_summary %>%
  mutate(intronic = !exonic & !intergenic) %>%
  mutate(Protocols = recode(protocol,
    "ont"      = "ONT cDNA",
    "drna"     = "ONT dRNA",
    "pacbio"   = "PacBio",
    "illumina" = "Illumina"
  ))

# Save a csv with no of this chunk in fig.path if exists
if (!is.null(params$fig.path)) {
  write_csv(var_summary, file.path(params$fig.path, "variant_summary.csv"))
}

# Plot1 Total number of variants per protocol, stacked by intergenic, exonic, intronic
p.box.data <- var_summary %>%
  group_by(Protocols, cell_line) %>%
  summarise(
    total_variants = n(),
    intergenic = sum(intergenic),
    exonic = sum(exonic),
    intronic = sum(intronic),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(intergenic, exonic, intronic), names_to = "region_type", values_to = "count")

p.box <- p.box.data %>%
  ggplot(aes(x = Protocols, y = count, color = Protocols)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(aes(color = Protocols), width = 0.2, size = 1, alpha = 0.5) +
  facet_wrap(~region_type, nrow = 1, scales = "free_y") +
  labs(title = "Total Variants per Protocol by Region Type",
       x = "Cell Lines", y = "Total Variants") +
  scale_color_manual(values = color_map) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# p.bar <- var_summary %>%
#   group_by(Protocols) %>%
#   summarise(
#     total_variants = n(),
#     intergenic = sum(intergenic),
#     exonic = sum(exonic),
#     intronic = sum(intronic),
#     .groups = "drop"
#   ) %>%
#   pivot_longer(cols = c(intergenic, exonic, intronic), names_to = "region_type", values_to = "count") %>%
#   ggplot(aes(x = Protocols, y = count, fill = Protocols)) +
#   geom_bar(stat = "identity", position = "stack") +
#   facet_wrap(~region_type, nrow = 1, scales = "free_y") +
#   labs(title = "Total Variants per Protocol by Region Type",
#        x = "Cell Lines", y = "Total Variants") +
#   scale_fill_manual(values = color_map) +
#   theme_minimal() +
#   theme(axis.text.x = element_text(angle = 45, hjust = 1))

if (!is.null(params$fig.path)) {
  write_csv(p.box.data, file.path(params$fig.path, "variant_counts_by_region_per_cell_line.csv"))
}

coverage.p / p.box
```

```{r, upset plot overlap site and genotype (exonic only), fig.height=3, fig.width=10}
variants.upset.mat <- var_summary %>%
  mutate(var_pos = paste(CHROM, POS, sep = "_")) %>%
  mutate(variant_id = paste(CHROM, POS, REF, ALT, cell_line, sep = "_")) %>%
  distinct(variant_id, Protocols, .keep_all = TRUE) %>%
  mutate(present = TRUE) %>%
  pivot_wider(id_cols = variant_id, names_from = Protocols, values_from = present, values_fill = FALSE)


variants.upset.mat <- variants.upset.mat %>% left_join(
  var_summary %>%
    mutate(variant_id = paste(CHROM, POS, REF, ALT, cell_line, sep = "_")) %>%
    mutate(genomic_region = case_when(
      intergenic ~ "intergenic",
      exonic ~ "exonic",
      TRUE ~ "intronic"
    )) %>%
    distinct(variant_id, genomic_region),
  by = "variant_id"
)


ComplexUpset::upset(
  variants.upset.mat %>%
    filter((`ONT cDNA` | PacBio | `ONT dRNA` | Illumina) & genomic_region == "exonic"),
  intersect = c("ONT dRNA", "ONT cDNA", "PacBio", "Illumina"),
  name = "Variant Count (Exonic Only)",
  width_ratio = 0.25
)
```

# Part 3: Accuracy of mutation calling using CCLE

```{r load_accuracy_data}
# Unphased Illumina calls — no phased version available for short reads
FP.matrix <- read_tsv(file.path(params$variant_analysis_dir, "SNP_FP_QUAL_detail.tsv"),
                      show_col_types = FALSE) %>%
  separate(SNP_ID, into = c("sample", "locus"), sep = ":", extra = "merge") %>%
  dplyr::rename(Illumina = illumina, Illumina_DV = deepvariant,
                PacBio = pacbio, `ONT dRNA` = drna, `ONT cDNA` = ont)

TP.matrix <- list.files(file.path(params$variant_analysis_dir, "genotype_comparisons_qual"),
                        pattern = "genotype_comparison_.*\\.tsv$", full.names = TRUE) %>%
  lapply(function(f) read_tsv(f, show_col_types = FALSE) %>%
           mutate(sample = sub("genotype_comparison_(.*)\\.tsv", "\\1", basename(f)))) %>%
  bind_rows()

# Phased long-read calls (WhatsHap-phased VCFs — same source as var_summary)
FP.matrix.phase <- read_tsv(file.path(params$variant_analysis_phased_dir, "SNP_FP_QUAL_detail.tsv"),
                             show_col_types = FALSE) %>%
  separate(SNP_ID, into = c("sample", "locus"), sep = ":", extra = "merge") %>%
  dplyr::rename(`PacBio phased` = pacbio,
                `ONT dRNA phased` = drna,
                `ONT cDNA phased` = ont) %>%
  full_join(FP.matrix %>% select(sample, locus, Illumina, Illumina_DV),
            by = c("sample", "locus")) %>%
  mutate(across(where(is.logical), ~replace_na(.x, FALSE)))

TP.matrix.phase <- list.files(file.path(params$variant_analysis_phased_dir, "genotype_comparisons_qual"),
                               pattern = "genotype_comparison_.*\\.tsv$", full.names = TRUE) %>%
  lapply(function(f) read_tsv(f, show_col_types = FALSE) %>%
           mutate(sample = sub("genotype_comparison_(.*)\\.tsv", "\\1", basename(f)))) %>%
  bind_rows()

# Shared setup
qual_threshold <- 0

color_map_accuracy <- c(
  "Illumina"        = "#E87D24",
  "Illumina_DV"     = "#914f19",
  "PacBio phased"   = "#DA1884",
  "ONT dRNA phased" = "#007EA7",
  "ONT cDNA phased" = "#003F5C"
)

recode_lowqual <- function(df, thresh) {
  status_cols <- grep("_STATUS$", names(df), value = TRUE)
  qual_cols   <- sub("_STATUS$", "_QUAL", status_cols)
  for (i in seq_along(status_cols)) {
    qc <- qual_cols[i]; sc <- status_cols[i]
    if (qc %in% names(df))
      df[[sc]] <- ifelse(df[[sc]] != "not_called" & !is.na(df[[qc]]) &
                           as.numeric(df[[qc]]) < thresh, "lowqual", df[[sc]])
  }
  df
}

# Annotate TP.matrix region using GTF (needed for Illumina region breakdown)
pos_gr <- GRanges(seqnames = TP.matrix$CHROM,
                  ranges   = IRanges(as.integer(TP.matrix$POS), width = 1))
TP.matrix$region <- ifelse(countOverlaps(pos_gr, exons) > 0, "exonic",
                     ifelse(countOverlaps(pos_gr, genes) == 0, "intergenic", "intronic"))

TP.matrix.display       <- recode_lowqual(TP.matrix,       qual_threshold)
TP.matrix.phase.display <- recode_lowqual(TP.matrix.phase, qual_threshold)
```

## FP (spike-in contigs)

```{r accuracy_fp, fig.height=3, fig.width=6}
FP.summary <- FP.matrix.phase %>%
  select(-locus) %>%
  group_by(sample) %>%
  summarise(across(everything(), ~sum(!is.na(.) & . >= qual_threshold, na.rm = TRUE)),
            .groups = "drop")

if (!is.null(params$fig.path))
  write_csv(FP.summary, file.path(params$fig.path, "false_positive_summary.csv"))

FP.summary %>%
  pivot_longer(-sample, names_to = "platform", values_to = "false_positives") %>%
  ggplot(aes(x = platform, y = false_positives, color = platform)) +
  geom_boxplot(width = 0.4, outlier.shape = NA) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.7) +
  scale_color_manual(values = color_map_accuracy) +
  labs(title = "False Positives by Platform (spike-in contigs)",
       x = "Platform", y = paste0("FPs (QUAL ≥ ", qual_threshold, ")")) +
  theme_minimal(base_size = 14) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1), legend.position = "none")
```

## TP (CCLE genotype confirmation)

```{r accuracy_tp, fig.height=4, fig.width=6}
make_tp_summary <- function(display_df, platform_recode) {
  display_df %>%
    pivot_longer(cols = ends_with("_STATUS"), names_to = "platform", values_to = "status") %>%
    mutate(
      platform = recode(sub("_STATUS", "", platform), !!!platform_recode),
      status   = recode(status, match = "Match", discordant = "Discordant",
                        lowqual = "Not Called", not_called = "Not Called"),
      status   = factor(status, levels = c("Not Called", "Discordant", "Match"))
    )
}

tp_phased <- make_tp_summary(TP.matrix.phase.display,
  c(pacbio = "PacBio phased", ont = "ONT cDNA phased", drna = "ONT dRNA phased"))

tp_illumina <- make_tp_summary(TP.matrix.display,
  c(illumina = "Illumina")) %>%
  filter(platform == "Illumina")

bind_rows(tp_phased, tp_illumina) %>%
  group_by(platform, status) %>%
  summarise(count = n(), .groups = "drop") %>%
  ggplot(aes(x = platform, y = count, fill = status)) +
  geom_col(position = "stack") +
  scale_fill_manual(values = c("Not Called" = "#C9C9C9", "Match" = "#1b9e77",
                               "Discordant" = "#CF9741")) +
  labs(title = "CCLE genotype confirmation",
       subtitle = "Long reads: WhatsHap-phased clair3 calls; Illumina: unphased",
       x = "Platform", y = "Count", fill = "Status") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## ROC: TP vs total calls across QUAL thresholds

```{r roc_setup}
color_map_roc <- c(
  "Illumina" = "#E87D24",
  "PacBio"   = "#DA1884",
  "ONT cDNA" = "#003F5C",
  "ONT dRNA" = "#007EA7"
)

make_roc <- function(tp_qual, tp_status, method_name, min_q = 0) {
  tp_qual    <- as.numeric(tp_qual)
  thresholds <- seq(min_q, floor(max(tp_qual, na.rm = TRUE)),0.2)
  purrr::map_dfr(thresholds, function(q) {
    tibble(method    = method_name,
           threshold = q,
           tp    = sum(!is.na(tp_qual) & tp_qual >= q & tp_status == "match", na.rm = TRUE),
           total = sum(!is.na(tp_qual) & tp_qual >= q, na.rm = TRUE))
  })
}

# Plot 2 data: total calls AT CCLE positions
# Long reads use phased TP.matrix; Illumina uses unphased (only version available)
roc_data <- bind_rows(
  make_roc(TP.matrix$illumina_QUAL,          TP.matrix$illumina_STATUS,          "Illumina"),
  make_roc(TP.matrix.phase$pacbio_QUAL,      TP.matrix.phase$pacbio_STATUS,      "PacBio"),
  make_roc(TP.matrix.phase$ont_QUAL,         TP.matrix.phase$ont_STATUS,         "ONT cDNA"),
  make_roc(TP.matrix.phase$drna_QUAL,        TP.matrix.phase$drna_STATUS,        "ONT dRNA")
)

roc_q0 <- roc_data %>% filter(threshold == qual_threshold)

# # Plot 1 data: total genome-wide PASS calls
# # var_summary is from phased VCFs (lr_whatshap_dir) but pre-filtered QUAL >= 10 at load;
# # long-read curves therefore cannot extend below QUAL = 10
# make_roc_full <- function(all_qual, ccle_qual, ccle_status, method_name) {
#   all_qual  <- as.numeric(all_qual)
#   ccle_qual <- as.numeric(ccle_qual)
#   thresholds <- sort(unique(c(0, na.omit(all_qual))))
#   purrr::map_dfr(thresholds, function(q) {
#     tibble(method       = method_name,
#            threshold    = q,
#            total_genome = sum(!is.na(all_qual) & all_qual >= q),
#            tp           = sum(!is.na(ccle_qual) & ccle_qual >= q &
#                                 ccle_status == "match", na.rm = TRUE))
#   })
# }

# roc_full <- bind_rows(
#   make_roc_full(var_summary %>% filter(Protocols == "Illumina") %>% pull(QUAL),
#                 TP.matrix$illumina_QUAL,          TP.matrix$illumina_STATUS,          "Illumina"),
#   make_roc_full(var_summary %>% filter(Protocols == "PacBio")   %>% pull(QUAL),
#                 TP.matrix.phase$pacbio_QUAL,      TP.matrix.phase$pacbio_STATUS,      "PacBio"),
#   make_roc_full(var_summary %>% filter(Protocols == "ONT cDNA") %>% pull(QUAL),
#                 TP.matrix.phase$ont_QUAL,         TP.matrix.phase$ont_STATUS,         "ONT cDNA"),
#   make_roc_full(var_summary %>% filter(Protocols == "ONT dRNA") %>% pull(QUAL),
#                 TP.matrix.phase$drna_QUAL,        TP.matrix.phase$drna_STATUS,        "ONT dRNA")
# )
# 
# roc_full_ref <- roc_full %>% filter(threshold == qual_threshold)
```


```{r roc_clair3_illumina_4curves, fig.height=10, fig.width=7}
all_methods   <- c("Illumina", "PacBio", "ONT cDNA", "ONT dRNA")
label_points  <- function(df) df %>%
  filter(method %in% all_methods, threshold %in% c(1, 10))

p2c <- roc_data %>%
  filter(method %in% all_methods, threshold >= 1) %>%
  mutate(other_calls = total - tp) %>%
  ggplot(aes(x = other_calls, y = tp, color = method)) +
  geom_path(linewidth = 0.8) +
  geom_point(data = label_points(roc_data %>% mutate(other_calls = total - tp) %>% filter(threshold >= 1)),
             shape = 21, fill = "white", size = 2, stroke = 1) +
  ggrepel::geom_text_repel(data = label_points(roc_data %>% mutate(other_calls = total - tp) %>% filter(threshold >= 1)),
                           aes(label = threshold),
                           size = 2.5, show.legend = FALSE, max.overlaps = 20) +
  scale_color_manual(values = color_map_roc) +
  labs(title = "Plot 2: CCLE-position calls (QUAL threshold labelled, starting QUAL = 1)",
       x = "Inconsistent calls at CCLE positions passing QUAL threshold",
       y = "Calls consistent with CCLE reference", color = "Method") +
  theme_minimal(base_size = 13) + theme(legend.position = "right")

p2c
```
