---

title: "Bulk gene/transcript identification: Oarfish vs IP-filtered vs IsoQuant"
author: "Yupei"
date: '`r Sys.Date()`'
version: "v1.0"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  # Oarfish + Illumina DGE bundle (original)
  oarfish_dge_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"
  # IP-filtered Oarfish DGE bundle
  ip_filtered_dge_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.IP_filtered.obj.rds"
  # IsoQuant (no_novel) DGE bundles
  isoquant_ont_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds"
  isoquant_pb_rds:  "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds"
  isoquant_drna_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds"
  # Pre-computed biotype lookup (avoids biomaRt call)
  ip_filtered_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification_IP_filtered.rds"
  cache_dir: NULL
  fig.path:  NULL
---

# Setup
```{r Input filename setup, include=FALSE, eval=F}
params <- list(
  random_seed = 2024,
  cache_dir = NULL,
  fig.path = NULL,
  oarfish_dge_rds    = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
  ip_filtered_dge_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.IP_filtered.obj.rds",
  isoquant_ont_rds   = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds",
  isoquant_pb_rds    = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds",
  isoquant_drna_rds  = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds",
  ip_filtered_rds    = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_identification_IP_filtered.rds"
)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  fig.width = 15,
  fig.height = 10,
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = c("png", "svg"),
  cache.path = ifelse(is.null(params$cache_dir),
                      paste0(tools::file_path_sans_ext(knitr::current_input()), "_cache/"),
                      params$cache_dir))
```

```{r Libraries, include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(scales)
library(patchwork)
library(ComplexUpset)
library(ComplexHeatmap)
source("/home/users/allstaff/you.yu/LongBench/analysis/workflow/scripts/Rfunctions.R")

rename <- dplyr::rename
select <- dplyr::select

color_palette <- c(
  PacBio    = "#df1995",
  ONT       = "#00789b",
  ONT_1     = "#04476c",
  ONT_2     = "#00789b",
  Illumina  = "#e88b20"
)

method_colors <- c(
  Oarfish       = "#2166ac",
  `IP-filtered` = "#4dac26",
  IsoQuant      = "#d6604d"
)

output_list <- list()
```

```{r helpers}
is.human <- function(dge) { grepl("^ENS", rownames(dge)) }

normalize_rownames <- function(dge) {
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}

pad_dge <- function(dge, all_features) {
  missing <- setdiff(all_features, rownames(dge))
  if (length(missing) > 0) {
    pad <- matrix(0, nrow = length(missing), ncol = ncol(dge$counts),
                  dimnames = list(missing, colnames(dge$counts)))
    dge$counts <- rbind(dge$counts, pad)
  }
  dge[all_features, ]
}

strip_version <- function(dge) { rownames(dge) <- sub("\\..*", "", rownames(dge)); dge }
```


# Load DGE objects

```{r Load Oarfish DGE}
oarfish_obj <- readRDS(params$oarfish_dge_rds)
ill_bulk.tx.dge    <- oarfish_obj[["ill_bulk.tx.dge"]]
ill_bulk.gene.dge  <- oarfish_obj[["ill_bulk.gene.dge"]]

oar.ont_bulk.tx.dge   <- oarfish_obj[["ont_bulk.tx.dge"]]
oar.pb_bulk.tx.dge    <- oarfish_obj[["pb_bulk.tx.dge"]]
oar.drna_bulk.tx.dge  <- oarfish_obj[["drna_bulk.tx.dge"]]

oar.ont_bulk.gene.dge  <- oarfish_obj[["ont_bulk.gene.dge"]]
oar.pb_bulk.gene.dge   <- oarfish_obj[["pb_bulk.gene.dge"]]
oar.drna_bulk.gene.dge <- oarfish_obj[["drna_bulk.gene.dge"]]

rm(oarfish_obj)
```

```{r Load IP-filtered DGE}
ip_obj <- readRDS(params$ip_filtered_dge_rds)
ip.ont_bulk.tx.dge    <- ip_obj[["ont_bulk.tx.dge"]]
ip.pb_bulk.tx.dge     <- ip_obj[["pb_bulk.tx.dge"]]
ip.drna_bulk.tx.dge   <- ip_obj[["drna_bulk.tx.dge"]]

ip.ont_bulk.gene.dge  <- ip_obj[["ont_bulk.gene.dge"]]
ip.pb_bulk.gene.dge   <- ip_obj[["pb_bulk.gene.dge"]]
ip.drna_bulk.gene.dge <- ip_obj[["drna_bulk.gene.dge"]]

rm(ip_obj)
```

```{r Load IsoQuant DGE}
isoquant_ont_obj  <- readRDS(params$isoquant_ont_rds)
isoquant_pb_obj   <- readRDS(params$isoquant_pb_rds)
isoquant_drna_obj <- readRDS(params$isoquant_drna_rds)

iq.ont_bulk.tx.dge   <- isoquant_ont_obj[["ont_bulk.tx.dge"]]
iq.pb_bulk.tx.dge    <- isoquant_pb_obj[["pb_bulk.tx.dge"]]
iq.drna_bulk.tx.dge  <- isoquant_drna_obj[["dRNA_bulk.tx.dge"]]

iq.ont_bulk.gene.dge  <- isoquant_ont_obj[["ont_bulk.gene.dge"]]
iq.pb_bulk.gene.dge   <- isoquant_pb_obj[["pb_bulk.gene.dge"]]
iq.drna_bulk.gene.dge <- isoquant_drna_obj[["dRNA_bulk.gene.dge"]]

rm(isoquant_ont_obj, isoquant_pb_obj, isoquant_drna_obj)
```

```{r Normalize IDs}
# IsoQuant rownames: strip pipe annotations and version numbers
iq.ont_bulk.tx.dge   <- normalize_rownames(iq.ont_bulk.tx.dge)
iq.pb_bulk.tx.dge    <- normalize_rownames(iq.pb_bulk.tx.dge)
iq.drna_bulk.tx.dge  <- normalize_rownames(iq.drna_bulk.tx.dge)
iq.ont_bulk.gene.dge  <- normalize_rownames(iq.ont_bulk.gene.dge)
iq.pb_bulk.gene.dge   <- normalize_rownames(iq.pb_bulk.gene.dge)
iq.drna_bulk.gene.dge <- normalize_rownames(iq.drna_bulk.gene.dge)

# Strip version numbers from all gene-level DGEs
ill_bulk.gene.dge      <- strip_version(ill_bulk.gene.dge)
oar.ont_bulk.gene.dge  <- strip_version(oar.ont_bulk.gene.dge)
oar.pb_bulk.gene.dge   <- strip_version(oar.pb_bulk.gene.dge)
oar.drna_bulk.gene.dge <- strip_version(oar.drna_bulk.gene.dge)
ip.ont_bulk.gene.dge   <- strip_version(ip.ont_bulk.gene.dge)
ip.pb_bulk.gene.dge    <- strip_version(ip.pb_bulk.gene.dge)
ip.drna_bulk.gene.dge  <- strip_version(ip.drna_bulk.gene.dge)
```

```{r Build TX annotation reference}
get_full_id_col <- function(dge) {
  if ("full_id" %in% colnames(dge$genes)) {
    dge$genes$full_id
  } else {
    rownames(dge$genes)
  }
}

full_ids <- get_full_id_col(ill_bulk.tx.dge)

tx_annotation <- tibble(full_id = full_ids) %>%
  mutate(
    tx_id   = strsplit(full_id, "\\|") %>% sapply(function(x) x[1]),
    gene_id = strsplit(full_id, "\\|") %>% sapply(function(x) x[2]),
    length  = strsplit(full_id, "\\|") %>% sapply(function(x) x[7]) %>% as.integer(),
    biotype = strsplit(full_id, "\\|") %>% sapply(function(x) x[8])
  ) %>%
  mutate(
    tx_id   = sub("\\..*", "", tx_id),
    gene_id = sub("\\..*", "", gene_id)
  ) %>%
  select(tx_id, gene_id, length, biotype) %>%
  distinct(tx_id, .keep_all = TRUE)
```


# Gene discovery analysis

## Build `gene_discovery`

```{r Gene: Build gene_discovery}
ill_bulk.gene.dge      <- ill_bulk.gene.dge[is.human(ill_bulk.gene.dge), ]
oar.ont_bulk.gene.dge  <- oar.ont_bulk.gene.dge[is.human(oar.ont_bulk.gene.dge), ]
oar.pb_bulk.gene.dge   <- oar.pb_bulk.gene.dge[is.human(oar.pb_bulk.gene.dge), ]
oar.drna_bulk.gene.dge <- oar.drna_bulk.gene.dge[is.human(oar.drna_bulk.gene.dge), ]
ip.ont_bulk.gene.dge   <- ip.ont_bulk.gene.dge[is.human(ip.ont_bulk.gene.dge), ]
ip.pb_bulk.gene.dge    <- ip.pb_bulk.gene.dge[is.human(ip.pb_bulk.gene.dge), ]
ip.drna_bulk.gene.dge  <- ip.drna_bulk.gene.dge[is.human(ip.drna_bulk.gene.dge), ]
iq.ont_bulk.gene.dge   <- iq.ont_bulk.gene.dge[is.human(iq.ont_bulk.gene.dge), ]
iq.pb_bulk.gene.dge    <- iq.pb_bulk.gene.dge[is.human(iq.pb_bulk.gene.dge), ]
iq.drna_bulk.gene.dge  <- iq.drna_bulk.gene.dge[is.human(iq.drna_bulk.gene.dge), ]

all_genes <- Reduce(intersect, lapply(
  list(ill_bulk.gene.dge,
       oar.ont_bulk.gene.dge, oar.pb_bulk.gene.dge, oar.drna_bulk.gene.dge,
       ip.ont_bulk.gene.dge,  ip.pb_bulk.gene.dge,  ip.drna_bulk.gene.dge,
       iq.ont_bulk.gene.dge,  iq.pb_bulk.gene.dge,  iq.drna_bulk.gene.dge),
  rownames
))

ill_bulk.gene.dge      <- pad_dge(ill_bulk.gene.dge,      all_genes)
oar.ont_bulk.gene.dge  <- pad_dge(oar.ont_bulk.gene.dge,  all_genes)
oar.pb_bulk.gene.dge   <- pad_dge(oar.pb_bulk.gene.dge,   all_genes)
oar.drna_bulk.gene.dge <- pad_dge(oar.drna_bulk.gene.dge, all_genes)
ip.ont_bulk.gene.dge   <- pad_dge(ip.ont_bulk.gene.dge,   all_genes)
ip.pb_bulk.gene.dge    <- pad_dge(ip.pb_bulk.gene.dge,    all_genes)
ip.drna_bulk.gene.dge  <- pad_dge(ip.drna_bulk.gene.dge,  all_genes)
iq.ont_bulk.gene.dge   <- pad_dge(iq.ont_bulk.gene.dge,   all_genes)
iq.pb_bulk.gene.dge    <- pad_dge(iq.pb_bulk.gene.dge,    all_genes)
iq.drna_bulk.gene.dge  <- pad_dge(iq.drna_bulk.gene.dge,  all_genes)

gene_discovery <- tibble(
  gene_id                  = sub("\\..*", "", all_genes),
  Illumina                 = filterByExpr(ill_bulk.gene.dge,      group = ill_bulk.gene.dge$samples$group),
  `ONT cDNA (Oarfish)`     = filterByExpr(oar.ont_bulk.gene.dge,  group = oar.ont_bulk.gene.dge$samples$group),
  `PacBio (Oarfish)`       = filterByExpr(oar.pb_bulk.gene.dge,   group = oar.pb_bulk.gene.dge$samples$group),
  `ONT dRNA (Oarfish)`     = filterByExpr(oar.drna_bulk.gene.dge, group = oar.drna_bulk.gene.dge$samples$group),
  `ONT cDNA (IP-filtered)` = filterByExpr(ip.ont_bulk.gene.dge,   group = ip.ont_bulk.gene.dge$samples$group),
  `PacBio (IP-filtered)`   = filterByExpr(ip.pb_bulk.gene.dge,    group = ip.pb_bulk.gene.dge$samples$group),
  `ONT dRNA (IP-filtered)` = filterByExpr(ip.drna_bulk.gene.dge,  group = ip.drna_bulk.gene.dge$samples$group),
  `ONT cDNA (IsoQuant)`    = filterByExpr(iq.ont_bulk.gene.dge,   group = iq.ont_bulk.gene.dge$samples$group),
  `PacBio (IsoQuant)`      = filterByExpr(iq.pb_bulk.gene.dge,    group = iq.pb_bulk.gene.dge$samples$group),
  `ONT dRNA (IsoQuant)`    = filterByExpr(iq.drna_bulk.gene.dge,  group = iq.drna_bulk.gene.dge$samples$group)
) %>% filter(
  Illumina |
  `ONT cDNA (Oarfish)` | `PacBio (Oarfish)` | `ONT dRNA (Oarfish)` |
  `ONT cDNA (IP-filtered)` | `PacBio (IP-filtered)` | `ONT dRNA (IP-filtered)` |
  `ONT cDNA (IsoQuant)` | `PacBio (IsoQuant)` | `ONT dRNA (IsoQuant)`
)
```

## Add biotype and gene length

```{r Gene: Add biotype and gene length}
gene_length_lookup <- tx_annotation %>%
  group_by(gene_id) %>%
  summarise(mean_length = mean(length, na.rm = TRUE), .groups = "drop")

biotype_lookup <- readRDS(params$ip_filtered_rds)[["gene_discovery"]] %>%
  select(gene_id, biotype) %>%
  distinct(gene_id, .keep_all = TRUE)

gene_discovery <- gene_discovery %>%
  left_join(gene_length_lookup, by = "gene_id") %>%
  rename(`Gene length` = mean_length) %>%
  left_join(biotype_lookup, by = "gene_id")

gene_discovery$biotype <- factor(
  gene_discovery$biotype,
  levels = c(sort(unique(gene_discovery$biotype)) %>% setdiff("Others"), "Others")
)
```

## Add intronic gene info

```{r Gene: Add intronic gene info}
rds_list        <- readRDS("/home/users/allstaff/you.yu/LongBench/analysis/workflow/rmarkdown/RDS/intronic_gene_and_exon_count.rds")
overlaps_tibble <- rds_list$exon_intron_overlaps %>%
  mutate(exon_gene   = sub("\\..*", "", exon_gene),
         intron_gene = sub("\\..*", "", intron_gene))
exons_count <- rds_list$exons_count %>% mutate(Gene = sub("\\..*", "", Gene))
rm(rds_list)

gene_discovery <- gene_discovery %>%
  mutate(
    `Intronic Gene` = gene_id %in% overlaps_tibble[overlaps_tibble$intron_gene %in% gene_id, ]$exon_gene
  ) %>%
  left_join(exons_count, by = c("gene_id" = "Gene"))

output_list[["gene_discovery"]] <- gene_discovery
```

## Gene detection plots

### Method-grouped upset plots

### Combined upset (all methods and platforms)

```{r GENE: Upset combined, fig.height=6, fig.width=18}
lr_cols_gene <- c(
  "ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)",
  "ONT cDNA (IP-filtered)", "PacBio (IP-filtered)", "ONT dRNA (IP-filtered)",
  "ONT cDNA (IsoQuant)", "PacBio (IsoQuant)", "ONT dRNA (IsoQuant)"
)
all_cols_gene <- c("Illumina", lr_cols_gene)

ComplexUpset::upset(
  gene_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = all_cols_gene,
  n_intersections = 30,
  annotations = list(
    "Gene length" = list(
      aes  = aes(x = intersection, y = `Gene length`),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) + ggtitle("Gene detection — all methods and platforms (top 30 intersections)")
```

### Per-platform 3-way overlap bar chart

```{r GENE: Per-platform 3-way overlap, fig.height=4, fig.width=6}
platform_triples_gene <- list(
  `ONT cDNA` = c("ONT cDNA (Oarfish)",  "ONT cDNA (IP-filtered)",  "ONT cDNA (IsoQuant)"),
  `PacBio`   = c("PacBio (Oarfish)",    "PacBio (IP-filtered)",    "PacBio (IsoQuant)"),
  `ONT dRNA` = c("ONT dRNA (Oarfish)",  "ONT dRNA (IP-filtered)",  "ONT dRNA (IsoQuant)")
)

gene_3way <- map_dfr(names(platform_triples_gene), function(platform) {
  oar_col <- platform_triples_gene[[platform]][1]
  ip_col  <- platform_triples_gene[[platform]][2]
  iq_col  <- platform_triples_gene[[platform]][3]
  gene_discovery %>%
    mutate(
      category = case_when(
        .data[[oar_col]] & .data[[ip_col]] & .data[[iq_col]]   ~ "Oarfish & IP-filtered & IsoQuant",
        .data[[oar_col]] & .data[[ip_col]] & !.data[[iq_col]]  ~ "Oarfish & IP-filtered",
        .data[[oar_col]] & !.data[[ip_col]] & .data[[iq_col]]  ~ "Oarfish & IsoQuant",
        !.data[[oar_col]] & .data[[ip_col]] & .data[[iq_col]]  ~ "IP-filtered & IsoQuant",
        .data[[oar_col]] & !.data[[ip_col]] & !.data[[iq_col]] ~ "Oarfish only",
        !.data[[oar_col]] & .data[[ip_col]] & !.data[[iq_col]] ~ "IP-filtered only",
        !.data[[oar_col]] & !.data[[ip_col]] & .data[[iq_col]] ~ "IsoQuant only",
        TRUE ~ "None"
      )
    ) %>%
    filter(category != "None") %>%
    count(category) %>%
    mutate(platform = platform)
})

gene_3way$category <- factor(gene_3way$category, levels = c(
   "IP-filtered only", 
   "IsoQuant only",
   "Oarfish & IsoQuant",
   "IP-filtered & IsoQuant", 
   "Oarfish only",
   "Oarfish & IP-filtered", 
   "Oarfish & IP-filtered & IsoQuant"
))

ggplot(gene_3way, aes(x = platform, y = n, fill = category)) +
  geom_col(position = "stack") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Gene detection: 3-way overlap per platform",
       y = "Number of genes", x = NULL, fill = "Category") +
  theme(axis.text.x = element_text(size = 11))

# save gene_3way to TSV if fig.path is provided
if (!is.null(params$fig.path)) {
  write_tsv(gene_3way, file.path(params$fig.path, "gene_3way_overlap.tsv"))
}
```

### Retention of Oarfish platform-specific genes

Platform-specific genes are defined as genes detected by Oarfish on exactly one LR platform (and not the other two). The plot asks: of those platform-specific genes, how many are still detected by IP-filtered oarfish or IsoQuant?

```{r GENE: Platform-specific retention, fig.height=4, fig.width=4}
oar_lr_cols <- c("ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)")

platform_specific_gene_retention <- map_dfr(oar_lr_cols, function(this_col) {
  other_cols <- setdiff(oar_lr_cols, this_col)
  platform   <- sub(" \\(Oarfish\\)", "", this_col)
  ip_col     <- paste0(platform, " (IP-filtered)")
  iq_col     <- paste0(platform, " (IsoQuant)")

  # Platform-specific = detected by this Oarfish LR platform, not the other two Oarfish LR platforms, and not Illumina
  specific <- gene_discovery %>%
    filter(.data[[this_col]] & !.data[[other_cols[1]]] & !.data[[other_cols[2]]] & !Illumina)

  tibble(
    platform  = platform,
    method    = c("Oarfish", "IP-filtered", "IsoQuant"),
    n         = c(nrow(specific),
                  sum(specific[[ip_col]]),
                  sum(specific[[iq_col]]))
  )
})

platform_specific_gene_retention$method <- factor(
  platform_specific_gene_retention$method,
  levels = c("Oarfish", "IP-filtered", "IsoQuant")
)

ggplot(platform_specific_gene_retention, aes(x = method, y = n, fill = method)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  facet_wrap(~platform) +
  scale_fill_manual(values = method_colors) +
  theme_minimal() +
  labs(
    title = "Oarfish platform-specific genes: retained by each method",
    subtitle = "Platform-specific = detected by one Oarfish LR platform only (not the other two LR platforms or Illumina)",
    y = "Number of genes", x = NULL
  ) +
  theme(legend.position = "none")

if (!is.null(params$fig.path)) {
  write_tsv(platform_specific_gene_retention, file.path(params$fig.path, "platform_specific_gene_retention.tsv"))
}
```

### Per-platform biotype composition by method

```{r GENE: Per-platform method comparison by biotype, fig.height=4, fig.width=13}
gene_method_biotype <- map_dfr(names(platform_triples_gene), function(platform) {
  oar_col <- platform_triples_gene[[platform]][1]
  ip_col  <- platform_triples_gene[[platform]][2]
  iq_col  <- platform_triples_gene[[platform]][3]
  gene_discovery %>%
    select(gene_id, biotype, all_of(c(oar_col, ip_col, iq_col))) %>%
    pivot_longer(cols = c(oar_col, ip_col, iq_col), names_to = "method_col", values_to = "detected") %>%
    filter(detected) %>%
    mutate(
      method   = sub(paste0(" \\(|\\)"), "", sub(platform, "", method_col)) %>% trimws(),
      biotype2 = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding"),
      platform = platform
    ) %>%
    count(method, biotype2, platform)
})


gene_method_biotype$method <- factor(
  gene_method_biotype$method,
  levels = c("Oarfish)", "IP-filtered)", "IsoQuant)")
)

ggplot(gene_method_biotype, aes(x = method, y = n, fill = biotype2)) +
  geom_col(position = "fill") +
  facet_wrap(~platform) +
  scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
  theme_minimal() +
  labs(title = "Biotype composition of detected genes by method and platform",
       y = "Proportion", x = NULL, fill = "Biotype") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```


### Per-platform intronic gene number by method

```{r GENE: Per-platform method comparison by intronic, fig.height=4, fig.width=13}
gene_method_intronic <- map_dfr(names(platform_triples_gene), function(platform) {
  oar_col <- platform_triples_gene[[platform]][1]
  ip_col  <- platform_triples_gene[[platform]][2]
  iq_col  <- platform_triples_gene[[platform]][3]
  gene_discovery %>%
    select(gene_id, `Intronic Gene`, all_of(c(oar_col, ip_col, iq_col))) %>%
    pivot_longer(cols = c(oar_col, ip_col, iq_col), names_to = "method_col", values_to = "detected") %>%
    filter(detected) %>%
    mutate(
      method   = sub(paste0(" \\(|\\)"), "", sub(platform, "", method_col)) %>% trimws(),
      intronic = factor(ifelse(`Intronic Gene`, "Intronic", "Non-intronic"), levels = c("Non-intronic","Intronic")),
      platform = platform
    ) %>%
    count(method, intronic, platform)
})

gene_method_intronic$method <- factor(
  gene_method_intronic$method,
  levels = c("Oarfish)", "IP-filtered)", "IsoQuant)")
)

ggplot(gene_method_intronic, aes(x = method, y = n, fill = intronic)) +
  geom_col(position = "fill") +
  facet_wrap(~platform) +
  scale_fill_manual(values = c("Non-intronic" = "#bec4c7", "Intronic" = "#E07B54")) +
  theme_minimal() +
  labs(
    title = "Intronic gene proportion of detected genes by method and platform",
    y = "Proportion", x = NULL, fill = "Gene type"
  ) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

if (!is.null(params$fig.path)) {
  write_tsv(gene_method_intronic, file.path(params$fig.path, "gene_method_intronic.tsv"))
}
```


### Jaccard similarity heatmap (per platform)

One heatmap per platform comparing Illumina + Oarfish + IP-filtered + IsoQuant. Shared colour scale across all panels.

```{r GENE: Jaccard similarity per platform, fig.height=3.5, fig.width=12}
jaccard_sim <- function(x, y) { sum(x & y) / sum(x | y) }
jac_col <- circlize::colorRamp2(seq(0, 1, length.out = 50),
                                 colorRampPalette(c("grey90", "#06317F"))(50))

make_platform_jac_gene <- function(platform) {
  cols <- c(paste0(platform, " (Oarfish)"),
            paste0(platform, " (IP-filtered)"),
            paste0(platform, " (IsoQuant)"))
  labels <- c("Oarfish", "IP-filtered", "IsoQuant")
  bm <- gene_discovery %>%
    select(all_of(cols)) %>%
    mutate(across(everything(), as.integer))
  jm <- outer(cols, cols, Vectorize(function(i, j) jaccard_sim(bm[[i]], bm[[j]])))
  colnames(jm) <- rownames(jm) <- labels
  Heatmap(jm, name = "Jaccard\nSimilarity", col = jac_col,
          column_title = platform,
          cell_fun = function(j, i, x, y, w, h, fill) {
            grid.text(round(jm[i, j], 2), x, y,
                      gp = gpar(col = ifelse(jm[i, j] > 0.6, "white", "black"), fontsize = 9))
          },
          row_names_gp    = gpar(fontsize = 9),
          column_names_gp = gpar(fontsize = 9),
          show_heatmap_legend = FALSE)
}

ht_gene <- make_platform_jac_gene("ONT cDNA") +
           make_platform_jac_gene("PacBio") +
           make_platform_jac_gene("ONT dRNA")

lgd <- Legend(col_fun = jac_col, title = "Jaccard\nSimilarity", direction = "vertical")
draw(ht_gene, annotation_legend_list = list(lgd))
```

### Gene length distribution

```{r GENE: Length distribution, fig.height=4, fig.width=12}
gene_discovery$length_bin <- cut(
  gene_discovery$`Gene length`,
  breaks = c(0, 500, 1000, 1500, 2000, 2500, Inf),
  labels = c("0-0.5k", "0.5-1k", "1-1.5k", "1.5-2k", "2-2.5k", "2.5k+")
)

bg_dist_gene <- gene_discovery %>%
  filter(!is.na(length_bin)) %>%
  count(length_bin) %>%
  mutate(label = "All annotated genes")

gene_discovery %>%
  pivot_longer(cols = all_of(lr_cols_gene), names_to = "method_platform", values_to = "keep") %>%
  filter(keep, !is.na(length_bin)) %>%
  mutate(
    method   = case_when(
      grepl("IP-filtered", method_platform) ~ "IP-filtered",
      grepl("Oarfish",     method_platform) ~ "Oarfish",
      TRUE                                  ~ "IsoQuant"
    ),
    platform = sub(" \\(.*\\)", "", method_platform)
  ) %>%
  count(length_bin, method, platform) %>%
  ggplot(aes(x = length_bin, y = n, color = method, group = method)) +
  geom_line(data = bg_dist_gene, aes(x = length_bin, y = n, group = 1),
            color = "grey60", linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~platform, ncol = 3) +
  scale_color_manual(values = method_colors) +
  theme_minimal() +
  labs(title = "Detected genes by length bin",
       x = "Gene length (avg transcript length)", y = "# of genes", color = "Method",
       caption = "Dashed grey line: all annotated genes")
```


# Transcript discovery analysis

## Build `tx_discovery`

```{r TX: Filter to human}
ill_bulk.tx.dge       <- ill_bulk.tx.dge[is.human(ill_bulk.tx.dge), ]
oar.ont_bulk.tx.dge   <- oar.ont_bulk.tx.dge[is.human(oar.ont_bulk.tx.dge), ]
oar.pb_bulk.tx.dge    <- oar.pb_bulk.tx.dge[is.human(oar.pb_bulk.tx.dge), ]
oar.drna_bulk.tx.dge  <- oar.drna_bulk.tx.dge[is.human(oar.drna_bulk.tx.dge), ]
ip.ont_bulk.tx.dge    <- ip.ont_bulk.tx.dge[is.human(ip.ont_bulk.tx.dge), ]
ip.pb_bulk.tx.dge     <- ip.pb_bulk.tx.dge[is.human(ip.pb_bulk.tx.dge), ]
ip.drna_bulk.tx.dge   <- ip.drna_bulk.tx.dge[is.human(ip.drna_bulk.tx.dge), ]
iq.ont_bulk.tx.dge    <- iq.ont_bulk.tx.dge[is.human(iq.ont_bulk.tx.dge), ]
iq.pb_bulk.tx.dge     <- iq.pb_bulk.tx.dge[is.human(iq.pb_bulk.tx.dge), ]
iq.drna_bulk.tx.dge   <- iq.drna_bulk.tx.dge[is.human(iq.drna_bulk.tx.dge), ]
```

```{r TX: Normalize TX rownames}
save_full_id <- function(dge) {
  if (!"full_id" %in% colnames(dge$genes)) {
    dge$genes$full_id <- rownames(dge)
  }
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}

ill_bulk.tx.dge      <- save_full_id(ill_bulk.tx.dge)
oar.ont_bulk.tx.dge  <- save_full_id(oar.ont_bulk.tx.dge)
oar.pb_bulk.tx.dge   <- save_full_id(oar.pb_bulk.tx.dge)
oar.drna_bulk.tx.dge <- save_full_id(oar.drna_bulk.tx.dge)
ip.ont_bulk.tx.dge   <- save_full_id(ip.ont_bulk.tx.dge)
ip.pb_bulk.tx.dge    <- save_full_id(ip.pb_bulk.tx.dge)
ip.drna_bulk.tx.dge  <- save_full_id(ip.drna_bulk.tx.dge)
iq.ont_bulk.tx.dge   <- save_full_id(iq.ont_bulk.tx.dge)
iq.pb_bulk.tx.dge    <- save_full_id(iq.pb_bulk.tx.dge)
iq.drna_bulk.tx.dge  <- save_full_id(iq.drna_bulk.tx.dge)
```

```{r TX: Build tx_discovery}
all_tx <- Reduce(intersect, lapply(
  list(ill_bulk.tx.dge,
       oar.ont_bulk.tx.dge, oar.pb_bulk.tx.dge, oar.drna_bulk.tx.dge,
       ip.ont_bulk.tx.dge,  ip.pb_bulk.tx.dge,  ip.drna_bulk.tx.dge,
       iq.ont_bulk.tx.dge,  iq.pb_bulk.tx.dge,  iq.drna_bulk.tx.dge),
  rownames
))

ill_bulk.tx.dge      <- pad_dge(ill_bulk.tx.dge,      all_tx)
oar.ont_bulk.tx.dge  <- pad_dge(oar.ont_bulk.tx.dge,  all_tx)
oar.pb_bulk.tx.dge   <- pad_dge(oar.pb_bulk.tx.dge,   all_tx)
oar.drna_bulk.tx.dge <- pad_dge(oar.drna_bulk.tx.dge, all_tx)
ip.ont_bulk.tx.dge   <- pad_dge(ip.ont_bulk.tx.dge,   all_tx)
ip.pb_bulk.tx.dge    <- pad_dge(ip.pb_bulk.tx.dge,    all_tx)
ip.drna_bulk.tx.dge  <- pad_dge(ip.drna_bulk.tx.dge,  all_tx)
iq.ont_bulk.tx.dge   <- pad_dge(iq.ont_bulk.tx.dge,   all_tx)
iq.pb_bulk.tx.dge    <- pad_dge(iq.pb_bulk.tx.dge,    all_tx)
iq.drna_bulk.tx.dge  <- pad_dge(iq.drna_bulk.tx.dge,  all_tx)

tx_discovery <- tibble(
  tx_id                    = all_tx,
  Illumina                 = filterByExpr(ill_bulk.tx.dge,      group = ill_bulk.tx.dge$samples$group,      min.count = 5),
  `ONT cDNA (Oarfish)`     = filterByExpr(oar.ont_bulk.tx.dge,  group = oar.ont_bulk.tx.dge$samples$group,  min.count = 5),
  `PacBio (Oarfish)`       = filterByExpr(oar.pb_bulk.tx.dge,   group = oar.pb_bulk.tx.dge$samples$group,   min.count = 5),
  `ONT dRNA (Oarfish)`     = filterByExpr(oar.drna_bulk.tx.dge, group = oar.drna_bulk.tx.dge$samples$group, min.count = 5),
  `ONT cDNA (IP-filtered)` = filterByExpr(ip.ont_bulk.tx.dge,   group = ip.ont_bulk.tx.dge$samples$group,   min.count = 5),
  `PacBio (IP-filtered)`   = filterByExpr(ip.pb_bulk.tx.dge,    group = ip.pb_bulk.tx.dge$samples$group,    min.count = 5),
  `ONT dRNA (IP-filtered)` = filterByExpr(ip.drna_bulk.tx.dge,  group = ip.drna_bulk.tx.dge$samples$group,  min.count = 5),
  `ONT cDNA (IsoQuant)`    = filterByExpr(iq.ont_bulk.tx.dge,   group = iq.ont_bulk.tx.dge$samples$group,   min.count = 5),
  `PacBio (IsoQuant)`      = filterByExpr(iq.pb_bulk.tx.dge,    group = iq.pb_bulk.tx.dge$samples$group,    min.count = 5),
  `ONT dRNA (IsoQuant)`    = filterByExpr(iq.drna_bulk.tx.dge,  group = iq.drna_bulk.tx.dge$samples$group,  min.count = 5)
) %>% filter(
  Illumina |
  `ONT cDNA (Oarfish)` | `PacBio (Oarfish)` | `ONT dRNA (Oarfish)` |
  `ONT cDNA (IP-filtered)` | `PacBio (IP-filtered)` | `ONT dRNA (IP-filtered)` |
  `ONT cDNA (IsoQuant)` | `PacBio (IsoQuant)` | `ONT dRNA (IsoQuant)`
)
```

## Add biotype and length

```{r TX: Add annotation}
tx_biotype_lookup <- readRDS(params$ip_filtered_rds)[["tx_discovery"]] %>%
  select(tx_id, biotype, length) %>%
  distinct(tx_id, .keep_all = TRUE)

tx_discovery <- tx_discovery %>%
  left_join(tx_annotation %>% select(tx_id, gene_id, length, biotype), by = "tx_id") %>%
  mutate(biotype = ifelse(
    !is.na(biotype) & biotype %in% names(sort(table(biotype), decreasing = TRUE)[1:6]),
    biotype, "Others"
  ))

output_list[["tx_discovery"]] <- tx_discovery
# saveRDS(output_list, params$out.rds)
```

## Transcript detection plots


### Combined upset (all methods and platforms)

```{r TX: Upset combined, fig.height=6, fig.width=18}
lr_cols_tx <- c(
  "ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)",
  "ONT cDNA (IP-filtered)", "PacBio (IP-filtered)", "ONT dRNA (IP-filtered)",
  "ONT cDNA (IsoQuant)", "PacBio (IsoQuant)", "ONT dRNA (IsoQuant)"
)
all_cols_tx <- c("Illumina", lr_cols_tx)

ComplexUpset::upset(
  tx_discovery %>% mutate(biotype = case_when(biotype == "protein_coding" ~ "Protein-coding", TRUE ~ "Non-protein-coding")),
  intersect = all_cols_tx,
  n_intersections = 30,
  annotations = list(
    "Transcript length" = list(
      aes  = aes(x = intersection, y = length),
      geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
    )
  ),
  base_annotations = list(
    'Intersection size' = ComplexUpset::intersection_size(
      counts  = TRUE,
      mapping = aes(fill = biotype),
      text    = list(element_text(size = 5, vjust = -0.2))
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952'))
  ),
  set_sizes = (
    ComplexUpset::upset_set_size(
      geom = geom_bar(aes(fill = biotype), width = 0.8)
    ) + scale_fill_manual(values = c("Non-protein-coding" = "#BAC5CB", "Protein-coding" = '#B5C952')) +
      theme(legend.position = "none")
  ),
  guides = 'over', wrap = TRUE
) + ggtitle("Transcript detection — all methods and platforms (top 30 intersections)")
```

### Per-platform 3-way overlap bar chart

```{r TX: Per-platform 3-way overlap, fig.height=4, fig.width=12}
platform_triples_tx <- list(
  `ONT cDNA` = c("ONT cDNA (Oarfish)",  "ONT cDNA (IP-filtered)",  "ONT cDNA (IsoQuant)"),
  `PacBio`   = c("PacBio (Oarfish)",    "PacBio (IP-filtered)",    "PacBio (IsoQuant)"),
  `ONT dRNA` = c("ONT dRNA (Oarfish)",  "ONT dRNA (IP-filtered)",  "ONT dRNA (IsoQuant)")
)

tx_3way <- map_dfr(names(platform_triples_tx), function(platform) {
  oar_col <- platform_triples_tx[[platform]][1]
  ip_col  <- platform_triples_tx[[platform]][2]
  iq_col  <- platform_triples_tx[[platform]][3]
  tx_discovery %>%
    mutate(
      category = case_when(
        .data[[oar_col]] & .data[[ip_col]] & .data[[iq_col]]   ~ "Oarfish & IP-filtered & IsoQuant",
        .data[[oar_col]] & .data[[ip_col]] & !.data[[iq_col]]  ~ "Oarfish & IP-filtered",
        .data[[oar_col]] & !.data[[ip_col]] & .data[[iq_col]]  ~ "Oarfish & IsoQuant",
        !.data[[oar_col]] & .data[[ip_col]] & .data[[iq_col]]  ~ "IP-filtered & IsoQuant",
        .data[[oar_col]] & !.data[[ip_col]] & !.data[[iq_col]] ~ "Oarfish only",
        !.data[[oar_col]] & .data[[ip_col]] & !.data[[iq_col]] ~ "IP-filtered only",
        !.data[[oar_col]] & !.data[[ip_col]] & .data[[iq_col]] ~ "IsoQuant only",
        TRUE ~ "None"
      )
    ) %>%
    filter(category != "None") %>%
    count(category) %>%
    mutate(platform = platform)
})

tx_3way$category <- factor(tx_3way$category, levels = c(
   "IP-filtered only", 
   "IsoQuant only",
   "Oarfish & IsoQuant",
   "IP-filtered & IsoQuant", 
   "Oarfish only",
   "Oarfish & IP-filtered", 
   "Oarfish & IP-filtered & IsoQuant"
))

ggplot(tx_3way, aes(x = platform, y = n, fill = category)) +
  geom_col(position = "stack") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal() +
  labs(title = "Transcript detection: 3-way overlap per platform",
       y = "Number of transcripts", x = NULL, fill = "Category") +
  theme(axis.text.x = element_text(size = 11))

if (!is.null(params$fig.path)) {
  write_tsv(tx_3way, file.path(params$fig.path, "tx_3way_overlap.tsv"))
}
```

### Retention of Oarfish platform-specific transcripts

```{r TX: Platform-specific retention, fig.height=4, fig.width=9}
oar_lr_cols_tx <- c("ONT cDNA (Oarfish)", "PacBio (Oarfish)", "ONT dRNA (Oarfish)")

platform_specific_tx_retention <- map_dfr(oar_lr_cols_tx, function(this_col) {
  other_cols <- setdiff(oar_lr_cols_tx, this_col)
  platform   <- sub(" \\(Oarfish\\)", "", this_col)
  ip_col     <- paste0(platform, " (IP-filtered)")
  iq_col     <- paste0(platform, " (IsoQuant)")

  specific <- tx_discovery %>%
    filter(.data[[this_col]] & !.data[[other_cols[1]]] & !.data[[other_cols[2]]] & !Illumina)

  tibble(
    platform  = platform,
    method    = c("Oarfish", "IP-filtered", "IsoQuant"),
    n         = c(nrow(specific),
                  sum(specific[[ip_col]]),
                  sum(specific[[iq_col]]))
  )
})

platform_specific_tx_retention$method <- factor(
  platform_specific_tx_retention$method,
  levels = c("Oarfish", "IP-filtered", "IsoQuant")
)

ggplot(platform_specific_tx_retention, aes(x = method, y = n, fill = method)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = n), vjust = -0.4, size = 3.5) +
  facet_wrap(~platform) +
  scale_fill_manual(values = method_colors) +
  theme_minimal() +
  labs(
    title = "Oarfish platform-specific transcripts: retained by each method",
    subtitle = "Platform-specific = detected by one Oarfish LR platform only (not the other two LR platforms or Illumina)",
    y = "Number of transcripts", x = NULL
  ) +
  theme(legend.position = "none")

if (!is.null(params$fig.path)) {
  write_tsv(platform_specific_tx_retention, file.path(params$fig.path, "platform_specific_tx_retention.tsv"))
}
```

### Jaccard similarity heatmap (per platform)

```{r TX: Jaccard similarity per platform, fig.height=3.5, fig.width=12}
make_platform_jac_tx <- function(platform) {
  cols <- c(paste0(platform, " (Oarfish)"),
            paste0(platform, " (IP-filtered)"),
            paste0(platform, " (IsoQuant)"))
  labels <- c("Oarfish", "IP-filtered", "IsoQuant")
  bm <- tx_discovery %>%
    select(all_of(cols)) %>%
    mutate(across(everything(), as.integer))
  jm <- outer(cols, cols, Vectorize(function(i, j) jaccard_sim(bm[[i]], bm[[j]])))
  colnames(jm) <- rownames(jm) <- labels
  Heatmap(jm, name = "Jaccard\nSimilarity", col = jac_col,
          column_title = platform,
          cell_fun = function(j, i, x, y, w, h, fill) {
            grid.text(round(jm[i, j], 2), x, y,
                      gp = gpar(col = ifelse(jm[i, j] > 0.6, "white", "black"), fontsize = 9))
          },
          row_names_gp    = gpar(fontsize = 9),
          column_names_gp = gpar(fontsize = 9),
          show_heatmap_legend = FALSE)
}

ht_tx <- make_platform_jac_tx("ONT cDNA") +
         make_platform_jac_tx("PacBio") +
         make_platform_jac_tx("ONT dRNA")

draw(ht_tx, annotation_legend_list = list(lgd))
```

### Transcript length distribution

```{r TX: Length distribution, fig.height=4, fig.width=12}
tx_discovery$length_bin <- cut(
  tx_discovery$length,
  breaks = c(0, 500, 1000, 1500, 2000, 2500, Inf),
  labels = c("0-0.5k", "0.5-1k", "1-1.5k", "1.5-2k", "2-2.5k", "2.5k+")
)

bg_dist_tx <- tx_discovery %>%
  filter(!is.na(length_bin)) %>%
  count(length_bin) %>%
  mutate(label = "All annotated transcripts")

tx_length_dist <- tx_discovery %>%
  pivot_longer(cols = all_of(lr_cols_tx), names_to = "method_platform", values_to = "keep") %>%
  filter(keep, !is.na(length_bin)) %>%
  mutate(
    method   = case_when(
      grepl("IP-filtered", method_platform) ~ "IP-filtered",
      grepl("Oarfish",     method_platform) ~ "Oarfish",
      TRUE                                  ~ "IsoQuant"
    ),
    platform = sub(" \\(.*\\)", "", method_platform)
  ) %>%
  count(length_bin, method, platform)
  
tx_length_dist%>%
  ggplot(aes(x = length_bin, y = n, color = method, group = method)) +
  geom_line(data = bg_dist_tx, aes(x = length_bin, y = n, group = 1),
            color = "grey60", linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~platform, ncol = 3) +
  scale_color_manual(values = method_colors) +
  theme_minimal() +
  labs(title = "Detected transcripts by length bin",
       x = "Transcript length", y = "# of transcripts", color = "Method",
       caption = "Dashed grey line: all annotated transcripts")

# log scale y-axis
tx_length_dist%>%
  ggplot(aes(x = length_bin, y = n, color = method, group = method)) +
  geom_line(data = bg_dist_tx, aes(x = length_bin, y = n, group = 1),
            color = "grey60", linewidth = 1, linetype = "dashed", inherit.aes = FALSE) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  facet_wrap(~platform, ncol = 3) +
  scale_y_log10() +
  scale_color_manual(values = method_colors) +
  theme_minimal() +
  labs(title = "Detected transcripts by length bin",
       x = "Transcript length", y = "# of transcripts", color = "Method",
       caption = "Dashed grey line: all annotated transcripts")


if (!is.null(params$fig.path)) {
  write_tsv(tx_length_dist, file.path(params$fig.path, "tx_length_dist.tsv"))
}
```


# Library size summary

Total library size (sum of counts across all features) per sample, at gene and transcript level, for each method and platform.

```{r Library size, fig.height=5, fig.width=12}
extract_libsize <- function(dge, method, platform, level) {
  tibble(
    sample   = colnames(dge$counts),
    lib_size = colSums(dge$counts),
    method   = method,
    platform = platform,
    level    = level
  )
}

libsize_df <- bind_rows(
  # Gene level
  extract_libsize(ill_bulk.gene.dge,      "Illumina",     "Illumina", "Gene"),
  extract_libsize(oar.ont_bulk.gene.dge,  "Oarfish",      "ONT cDNA", "Gene"),
  extract_libsize(oar.pb_bulk.gene.dge,   "Oarfish",      "PacBio",   "Gene"),
  extract_libsize(oar.drna_bulk.gene.dge, "Oarfish",      "ONT dRNA", "Gene"),
  extract_libsize(ip.ont_bulk.gene.dge,   "IP-filtered",  "ONT cDNA", "Gene"),
  extract_libsize(ip.pb_bulk.gene.dge,    "IP-filtered",  "PacBio",   "Gene"),
  extract_libsize(ip.drna_bulk.gene.dge,  "IP-filtered",  "ONT dRNA", "Gene"),
  extract_libsize(iq.ont_bulk.gene.dge,   "IsoQuant",     "ONT cDNA", "Gene"),
  extract_libsize(iq.pb_bulk.gene.dge,    "IsoQuant",     "PacBio",   "Gene"),
  extract_libsize(iq.drna_bulk.gene.dge,  "IsoQuant",     "ONT dRNA", "Gene"),
  # Transcript level
  extract_libsize(ill_bulk.tx.dge,        "Illumina",     "Illumina", "Transcript"),
  extract_libsize(oar.ont_bulk.tx.dge,    "Oarfish",      "ONT cDNA", "Transcript"),
  extract_libsize(oar.pb_bulk.tx.dge,     "Oarfish",      "PacBio",   "Transcript"),
  extract_libsize(oar.drna_bulk.tx.dge,   "Oarfish",      "ONT dRNA", "Transcript"),
  extract_libsize(ip.ont_bulk.tx.dge,     "IP-filtered",  "ONT cDNA", "Transcript"),
  extract_libsize(ip.pb_bulk.tx.dge,      "IP-filtered",  "PacBio",   "Transcript"),
  extract_libsize(ip.drna_bulk.tx.dge,    "IP-filtered",  "ONT dRNA", "Transcript"),
  extract_libsize(iq.ont_bulk.tx.dge,     "IsoQuant",     "ONT cDNA", "Transcript"),
  extract_libsize(iq.pb_bulk.tx.dge,      "IsoQuant",     "PacBio",   "Transcript"),
  extract_libsize(iq.drna_bulk.tx.dge,    "IsoQuant",     "ONT dRNA", "Transcript")
)

libsize_df$method   <- factor(libsize_df$method,   levels = c("Illumina", "Oarfish", "IP-filtered", "IsoQuant"))
libsize_df$platform <- factor(libsize_df$platform, levels = c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA"))

ggplot(libsize_df,
       aes(x = method, y = lib_size / 1e6, fill = level)) +
  geom_boxplot(outlier.size = 1, width = 0.6) +
  facet_wrap(~platform, nrow = 1, scales = "free_x") +
  scale_fill_manual(values = c("Gene" = "#4C72B0", "Transcript" = "#DD8452")) +
  theme_minimal() +
  labs(title = "Total library size per sample by method and platform",
       y = "Library size (millions of counts)", x = NULL, fill = "Level") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

if (!is.null(params$fig.path)) {
  write_tsv(libsize_df, file.path(params$fig.path, "library_size_summary.tsv"))
}
```


