---
title: "DE analysis continue"
author: "Yupei"
date: '`r Sys.Date()`'
version: 'v1.1_lfc0'
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
  bulk_DE_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds"
  log2FC_cutoff: 0
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.height = 5,
	fig.width = 7,
	warning = FALSE,
	cache = FALSE
)

if(!exists("params")) {
  params <- list(
    random_seed = 2024,
    cache_dir = NULL,
    fig.path = NULL,
    bulk_DE_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DE.rds",
    log2FC_cutoff = 0
  )
}
```

```{r Libraries, include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(ggplot2)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ggtree)
library(aplot)
library(ggpubr)
library(patchwork)
output_list <- list() # used to store result for downstream analysis

# setup some conflict functions
rename <- dplyr::rename
select <- dplyr::select

knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),  # Enable caching only if cache_dir is provided
  fig.width = 15,                     # Set default figure width
  fig.height = 10,                    # Set default figure height
  echo = FALSE,                       # Hide code in the output
  message = FALSE,                    # Suppress messages in the output
  warning = FALSE,                    # Suppress warnings in the output
  fig.path = params$fig.path,         # Set figure path (NULL if not provided)
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL  # Set output devices if fig.path is provided
)

# Set cache.path only if params$cache_dir is not NULL
if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache.path = params$cache_dir)
}
```

# Setup:
```{r Load bulk DE result}
bulk_DE_rst <- readRDS(params$bulk_DE_rds)
dge_lists <- bulk_DE_rst$dge_lists
dte_lists <- bulk_DE_rst$dte_lists

# Create a mapping from old names to new platform names
name_map <- c(
  ill = "Illumina",
  ont = "ONT cDNA",
  pb = "PacBio",
  drna = "ONT dRNA"
)

# Update the names of the list using the mapping
names(dge_lists) <- name_map[names(dge_lists)]
names(dte_lists) <- name_map[names(dte_lists)]

platforms <- c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA")
comparisons <- c("AvsL", "PvsL", "AvsP")
```

# Log FC Correlation between platform pairs
```{r LFC correlation, fig.height=4, fig.width=8}
get_de_union <- function(level, comp) {
  lists <- if (level == "gene") dge_lists else dte_lists
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  purrr::map(platforms, function(plt) {
    tbl <- lists[[plt]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column(id_col) %>%
      mutate(FDR = p.adjust(PValue, method = "BH"))
    tbl[[id_col]][tbl$FDR < 0.05]
  }) %>% purrr::reduce(union)
}

get_combined_logfc_plot <- function(level = "gene", p1, p2, filter_fdr = TRUE, min_cpm = NULL) {
  require(dplyr)
  require(ggplot2)
  require(stringr)
  require(purrr)

  all_data <- purrr::map_dfr(comparisons, function(comp) {
    lists <- if (level == "gene") dge_lists else dte_lists
    id_col <- if (level == "gene") "gene_id" else "transcript_id"

    get_tbl <- function(plt) {
      tbl <- lists[[plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column(id_col)
      if (level == "transcript")
        tbl <- tbl %>% mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))
      tbl
    }

    df1 <- get_tbl(p1)
    df2 <- get_tbl(p2)

    if (filter_fdr) {
      de_union <- get_de_union(level, comp)
      df1 <- df1 %>% filter(.data[[id_col]] %in% de_union)
      df2 <- df2 %>% filter(.data[[id_col]] %in% de_union)
    }

    joined <- full_join(df1[, c(id_col, "logFC")],
                        df2[, c(id_col, "logFC")],
                        by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      mutate(dir_match = sign(logFC_p1) == sign(logFC_p2))

    joined$comparison <- comp
    joined
  })

  cors <- all_data %>%
    group_by(comparison) %>%
    summarise(
      r  = cor(logFC_p1, logFC_p2, use = "complete.obs"),
      dc = mean(sign(logFC_p1) == sign(logFC_p2), na.rm = TRUE)
    ) %>%
    mutate(label = paste0("r = ", round(r, 2), "\nDC = ", round(dc * 100, 1), "%"))

  ggplot(all_data, aes(x = logFC_p1, y = logFC_p2)) +
    geom_point(aes(color = dir_match), alpha = 0.4, size = 0.3) +
    scale_color_manual(name = "Direction", values = c("TRUE" = "#2196F3", "FALSE" = "#F44336"),
                       labels = c("TRUE" = "Consistent", "FALSE" = "Discordant")) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    facet_wrap(~comparison, ncol = 3) +
    geom_text(data = cors, aes(x = -Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = -0.1, vjust = 1.2, size = 3.5) +
    labs(
      title = paste0(level, ": ", p1, " vs ", p2),
      x = paste0("logFC (", p1, ")"),
      y = paste0("logFC (", p2, ")")
    ) +
    theme_minimal(base_size = 8)
}

# Arrange figures
platform_pairs <- combn(platforms, 2, simplify = FALSE)
make_6panel_plot <- function(level, filter_fdr = TRUE) {
  plotlist <- map(platform_pairs, ~ get_combined_logfc_plot(level, .x[1], .x[2], filter_fdr = filter_fdr))
  ggarrange(plotlist = plotlist, ncol = 3, nrow = 2)
}

gene_logfc_combined <- make_6panel_plot("gene")
gene_logfc_combined
rm(gene_logfc_combined); gc()

tx_logfc_combined <- make_6panel_plot("transcript")
tx_logfc_combined
rm(tx_logfc_combined); gc()

# Boxplot: logFC Pearson r across ALL features (no FDR filter)
# x = comparison, box outline colour = level, dots coloured by pair
compute_all_cors <- function() {
  pair_labels <- map_chr(platform_pairs, ~ paste(.x[1], "vs", .x[2]))
  purrr::map_dfr(c("gene", "transcript"), function(level) {
    lists <- if (level == "gene") dge_lists else dte_lists
    id_col <- if (level == "gene") "gene_id" else "transcript_id"
    purrr::map_dfr(seq_along(platform_pairs), function(i) {
      p1 <- platform_pairs[[i]][1]; p2 <- platform_pairs[[i]][2]
      purrr::map_dfr(comparisons, function(comp) {
        df1 <- lists[[p1]][[paste0(comp, ".DE")]]$table %>% tibble::rownames_to_column(id_col)
        df2 <- lists[[p2]][[paste0(comp, ".DE")]]$table %>% tibble::rownames_to_column(id_col)
        joined <- full_join(df1[, c(id_col, "logFC")], df2[, c(id_col, "logFC")],
                            by = id_col, suffix = c("_p1", "_p2")) %>%
          drop_na(logFC_p1, logFC_p2)
        tibble(level = level, pair = pair_labels[[i]], comparison = comp,
               r = cor(joined$logFC_p1, joined$logFC_p2, use = "complete.obs"))
      })
    })
  })
}

pair_labels  <- map_chr(platform_pairs, ~ paste(.x[1], "vs", .x[2]))
pair_colors  <- setNames(RColorBrewer::brewer.pal(6, "Set2"), pair_labels)
level_colors <- c(gene = "#B4CD55", transcript = "#AE98EB")

plot_cor_boxplot <- function(cors_df, title) {
  ggplot(cors_df, aes(x = comparison, y = r, group = interaction(comparison, level))) +
    geom_boxplot(aes(color = level), fill = NA, width = 0.6, outlier.shape = NA,
                 position = position_dodge(width = 0.7)) +
    geom_point(aes(fill = pair), shape = 21, color = "gray30",
               position = position_jitterdodge(jitter.width = 0.15, dodge.width = 0.7),
               size = 2.5, alpha = 0.9) +
    scale_color_manual(name = "Level", values = level_colors) +
    scale_fill_manual(name = "Pair", values = pair_colors) +
    coord_cartesian(ylim = c(0.7, 1)) +
    labs(x = "Comparison", y = "Pearson r (logFC)", title = title) +
    theme_minimal(base_size = 11) +
    theme(legend.position = "right")
}

all_cors <- compute_all_cors()
p.box.all <- plot_cor_boxplot(all_cors, "logFC correlation across all features (no FDR filter)")

# Boxplot: FDR < 0.05 in any platform (union)
compute_fdr_cors <- function() {
  purrr::map_dfr(c("gene", "transcript"), function(level) {
    lists <- if (level == "gene") dge_lists else dte_lists
    id_col <- if (level == "gene") "gene_id" else "transcript_id"
    purrr::map_dfr(seq_along(platform_pairs), function(i) {
      p1 <- platform_pairs[[i]][1]; p2 <- platform_pairs[[i]][2]
      purrr::map_dfr(comparisons, function(comp) {
        de_union <- get_de_union(level, comp)
        df1 <- lists[[p1]][[paste0(comp, ".DE")]]$table %>%
          tibble::rownames_to_column(id_col) %>%
          filter(.data[[id_col]] %in% de_union)
        df2 <- lists[[p2]][[paste0(comp, ".DE")]]$table %>%
          tibble::rownames_to_column(id_col) %>%
          filter(.data[[id_col]] %in% de_union)
        joined <- full_join(df1[, c(id_col, "logFC")], df2[, c(id_col, "logFC")],
                            by = id_col, suffix = c("_p1", "_p2")) %>%
          drop_na(logFC_p1, logFC_p2)
        tibble(level = level, pair = pair_labels[[i]], comparison = comp,
               r = cor(joined$logFC_p1, joined$logFC_p2, use = "complete.obs"))
      })
    })
  })
}

fdr_cors <- compute_fdr_cors()
p.box.fdr <- plot_cor_boxplot(fdr_cors, "logFC correlation: FDR < 0.05 in any platform (union)")

(p.box.all | p.box.fdr)
```


```{r raw data points}
# Reconstruct the scatter plot data for all platform pairs and export as TSV

collect_logfc_scatter_data <- function(level, filter_fdr = TRUE) {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  lists  <- if (level == "gene") dge_lists else dte_lists

  purrr::map_dfr(platform_pairs, function(pair) {
    p1 <- pair[1]; p2 <- pair[2]
    purrr::map_dfr(comparisons, function(comp) {
      get_tbl <- function(plt) {
        tbl <- lists[[plt]][[paste0(comp, ".DE")]]$table %>%
          tibble::rownames_to_column(id_col)
        if (level == "transcript")
          tbl <- tbl %>% mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))
        tbl %>% mutate(FDR = p.adjust(PValue, method = "BH"))
      }

      df1 <- get_tbl(p1)
      df2 <- get_tbl(p2)

      if (filter_fdr) {
        de_union <- get_de_union(level, comp)
        df1 <- df1 %>% filter(.data[[id_col]] %in% de_union)
        df2 <- df2 %>% filter(.data[[id_col]] %in% de_union)
      }

      full_join(
        df1 %>% select(all_of(id_col), logFC, logCPM, FDR),
        df2 %>% select(all_of(id_col), logFC, logCPM, FDR),
        by = id_col, suffix = c("_p1", "_p2")
      ) %>%
        drop_na(logFC_p1, logFC_p2) %>%
        mutate(
          dir_match  = sign(logFC_p1) == sign(logFC_p2),
          comparison = comp,
          platform1  = p1,
          platform2  = p2,
          level      = level
        )
    })
  })
}

scatter_gene <- collect_logfc_scatter_data("gene",       filter_fdr = TRUE)
scatter_tx   <- collect_logfc_scatter_data("transcript", filter_fdr = TRUE)


if (!is.null(params$fig.path)) {
  out_dir <- dirname(params$fig.path)
  readr::write_tsv(scatter_gene, file.path(out_dir, "logFC_scatter_raw_data.gene.tsv"))
  readr::write_tsv(scatter_tx, file.path(out_dir, "logFC_scatter_raw_data.transcript.tsv"))
}
```
## LogFC Correlation: platform-specific DE features

For each comparison, identify features DE (FDR < 0.05) in **exactly one platform** across all four platforms. Then plot the logFC correlation of those platform-specific features between each pair — showing how the "missed" platform captures those signals.

```{r platform-specific DE directional support, fig.height=5, fig.width=10}
get_excl_ids <- function(level, comp, specific_plt) {
  lists  <- if (level == "gene") dge_lists else dte_lists
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  de_sets <- purrr::map(platforms, function(plt) {
    tbl <- lists[[plt]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column(id_col) %>%
      mutate(FDR = p.adjust(PValue, method = "BH"))
    tbl[[id_col]][tbl$FDR < 0.05]
  })
  names(de_sets) <- platforms
  setdiff(de_sets[[specific_plt]],
          purrr::reduce(de_sets[setdiff(platforms, specific_plt)], union))
}

get_dir_support <- function(level) {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  lists  <- if (level == "gene") dge_lists else dte_lists
  purrr::map_dfr(platforms, function(specific_plt) {
    other_plts <- setdiff(platforms, specific_plt)
    purrr::map_dfr(comparisons, function(comp) {
      excl_ids <- get_excl_ids(level, comp, specific_plt)
      if (length(excl_ids) == 0)
        return(tibble(platform = specific_plt, comparison = comp, level = level,
                      n_specific = 0L, n_supported = 0L))
      df_ref <- lists[[specific_plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column(id_col) %>%
        filter(.data[[id_col]] %in% excl_ids) %>%
        select(all_of(id_col), logFC_ref = logFC)
      support_tbl <- purrr::map_dfr(other_plts, function(op) {
        lists[[op]][[paste0(comp, ".DE")]]$table %>%
          tibble::rownames_to_column(id_col) %>%
          filter(.data[[id_col]] %in% excl_ids) %>%
          select(all_of(id_col), logFC_other = logFC) %>%
          left_join(df_ref, by = id_col) %>%
          mutate(dir_agree = sign(logFC_other) == sign(logFC_ref))
      }) %>%
        group_by(.data[[id_col]]) %>%
        summarise(all_agree = all(dir_agree, na.rm = TRUE), .groups = "drop")
      tibble(platform = specific_plt, comparison = comp, level = level,
             n_specific  = length(excl_ids),
             n_supported = sum(support_tbl$all_agree, na.rm = TRUE))
    })
  })
}

dir_support_raw <- bind_rows(get_dir_support("gene"), get_dir_support("transcript")) %>%
  mutate(n_not_supported = n_specific - n_supported,
         pct_supported   = ifelse(n_specific > 0, round(n_supported / n_specific * 100, 1), NA))

dir_support_long <- dir_support_raw %>%
  pivot_longer(c(n_supported, n_not_supported),
               names_to = "support", values_to = "count") %>%
  mutate(support = factor(support,
                          levels = c("n_not_supported", "n_supported"),
                          labels = c("Discordant", "Directionally consistent")))

ggplot(dir_support_long, aes(x = platform, y = count, fill = support)) +
  geom_bar(stat = "identity") +
  geom_text(data = dir_support_raw,
            aes(x = platform, y = n_specific,
                label = paste0(pct_supported, "%")),
            inherit.aes = FALSE, vjust = -0.3, size = 2.5) +
  facet_grid(level ~ comparison) +
  scale_fill_manual(name = "Direction in other platforms",
                    values = c("Directionally consistent" = "#2196F3",
                               "Discordant" = "#F44336")) +
  labs(x = "Platform (exclusively DE here)",
       y = "# platform-specific DE features",
       title = "Platform-specific DE: directional support from all other 3 platforms") +
  theme_minimal(base_size = 9) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "bottom")
```

## Expression level vs number of platforms detecting DE

The directional consistency above suggests inter-platform discrepancy is primarily a **sensitivity** effect: the same signal is present but only crosses the FDR threshold in some platforms. Here we confirm this directly — features detected as DE by fewer platforms have lower expression (logCPM), consistent with being near the detection limit.

```{r detection count vs CPM, fig.height=5, fig.width=8}
get_detection_count_vs_cpm <- function(level) {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  lists  <- if (level == "gene") dge_lists else dte_lists

  # Global mean logCPM: single value per feature averaged across all platforms x comparisons
  global_cpm <- purrr::map_dfr(platforms, function(plt) {
    purrr::map_dfr(comparisons, function(comp) {
      lists[[plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column(id_col) %>%
        select(all_of(id_col), logCPM)
    })
  }) %>%
    group_by(.data[[id_col]]) %>%
    summarise(mean_logCPM = mean(logCPM, na.rm = TRUE), .groups = "drop")

  # n_platforms per feature per comparison, then pool comparisons
  purrr::map_dfr(comparisons, function(comp) {
    purrr::map_dfr(platforms, function(plt) {
      lists[[plt]][[paste0(comp, ".DE")]]$table %>%
        tibble::rownames_to_column(id_col) %>%
        mutate(FDR = p.adjust(PValue, method = "BH"), is_DE = FDR < 0.05)
    }) %>%
      group_by(.data[[id_col]]) %>%
      summarise(n_platforms = sum(is_DE), .groups = "drop")
  }) %>%
    left_join(global_cpm, by = id_col) %>%
    mutate(level = level)
}

det_cpm_raw <- bind_rows(
  get_detection_count_vs_cpm("gene"),
  get_detection_count_vs_cpm("transcript")
)

det_cpm <- bind_rows(
  det_cpm_raw %>% mutate(n_platforms = as.character(n_platforms)),
  det_cpm_raw %>% mutate(n_platforms = "All")
) %>%
  mutate(n_platforms = factor(n_platforms,
                              levels = c("All", "0", "1", "2", "3", "4"),
                              labels = c("All", "0\n(not DE)", "1", "2", "3", "4")))

ggplot(det_cpm, aes(x = n_platforms, y = mean_logCPM, fill = n_platforms)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.12, fill = "white", outlier.shape = NA) +
  facet_wrap(~level, ncol = 2) +
  scale_fill_manual(values = c("All" = "#E8E8E8", "0\n(not DE)" = "gray60",
                               "1" = "#C6DBEF", "2" = "#6BAED6",
                               "3" = "#2171B5", "4" = "#08306B")) +
  labs(x = "Number of platforms detecting DE (FDR < 0.05)",
       y = "Global mean logCPM (all platforms × comparisons)",
       title = "Expression level by number of platforms detecting DE") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "none")

# Save raw data as TSV
det_cpm %>% group_by(level, n_platforms) %>%
  summarise(median_logCPM = median(mean_logCPM, na.rm = TRUE),
            n_features = n(),
            Q1_logCPM = quantile(mean_logCPM, 0.25, na.rm = TRUE),
            Q3_logCPM = quantile(mean_logCPM, 0.75, na.rm = TRUE),
            min_logCPM = min(mean_logCPM, na.rm = TRUE),
            max_logCPM = max(mean_logCPM, na.rm = TRUE),
             .groups = "drop") %>%
  arrange(level, n_platforms)
```

## LogFC Correlation: multi-transcript genes only

Restrict transcript-level logFC correlation to transcripts belonging to genes with **≥ 2 significant transcripts (P < 0.05) in at least one platform** of each pair. This isolates genes where multiple isoforms are detected, where platform disagreement is most biologically informative.

```{r LFC correlation multi-tx, fig.height=4, fig.width=8}
get_multitx_logfc_plot <- function(p1, p2) {
  require(dplyr); require(ggplot2); require(purrr); require(stringr)

  all_data <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- dte_lists[[p1]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id") %>%
      mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+")) %>%
      mutate(FDR = p.adjust(PValue, method = "BH")) %>% filter(FDR < 0.05)

    df2 <- dte_lists[[p2]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id") %>%
      mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+")) %>%
      mutate(FDR = p.adjust(PValue, method = "BH")) %>% filter(FDR < 0.05)

    # Genes with >= 2 expressed transcripts in p1 OR p2 (no significance filter)
    all1 <- dte_lists[[p1]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id") %>%
      mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))
    all2 <- dte_lists[[p2]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id") %>%
      mutate(gene_id = stringr::str_extract(transcript_id, "ENSG[0-9]+\\.[0-9]+"))

    multi_genes <- union(
      all1 %>% count(gene_id) %>% filter(n >= 2) %>% pull(gene_id),
      all2 %>% count(gene_id) %>% filter(n >= 2) %>% pull(gene_id)
    )

    df1 <- df1 %>% filter(gene_id %in% multi_genes)
    df2 <- df2 %>% filter(gene_id %in% multi_genes)

    full_join(df1[, c("transcript_id", "logFC", "logCPM")],
              df2[, c("transcript_id", "logFC", "logCPM")],
              by = "transcript_id", suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      mutate(mean_logCPM = (logCPM_p1 + logCPM_p2) / 2,
             comparison  = comp)
  })

  cors <- all_data %>%
    group_by(comparison) %>%
    summarise(r = cor(logFC_p1, logFC_p2, use = "complete.obs"),
              n = dplyr::n(), .groups = "drop") %>%
    mutate(label = paste0("r = ", round(r, 2), "\nn = ", n))

  ggplot(all_data, aes(x = logFC_p1, y = logFC_p2)) +
    geom_point(aes(color = mean_logCPM), alpha = 0.4, size = 0.3) +
    scale_color_viridis_c(name = "mean\nlogCPM", option = "plasma") +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
    geom_smooth(method = "lm", se = FALSE, color = "black", linewidth = 0.3) +
    facet_wrap(~comparison, ncol = 3) +
    geom_text(data = cors, aes(x = -Inf, y = Inf, label = label),
              inherit.aes = FALSE, hjust = -0.1, vjust = 1.2, size = 3) +
    labs(
      title = paste0("transcript (multi-tx genes): ", p1, " vs ", p2),
      x = paste0("logFC (", p1, ")"),
      y = paste0("logFC (", p2, ")")
    ) +
    theme_minimal(base_size = 8)
}

multitx_combined <- map(platform_pairs, ~ get_multitx_logfc_plot(.x[1], .x[2]))
multitx_plot <- ggarrange(plotlist = multitx_combined, ncol = 3, nrow = 2)
multitx_plot
rm(multitx_combined, multitx_plot); gc()
```


# Top Expressed Features: logFC Correlation Analysis

One concern for the lower transcript-level correlations is insufficient sequencing depth across many transcript features. Here we restrict logFC correlation to the top-N most highly expressed genes or transcripts (ranked by mean logCPM across comparisons from the first platform in each pair), without any significance filter. If correlations improve markedly when restricting to highly expressed features, this implicates low-count noise rather than genuine isoform-level disagreement between platforms.

## Correlation vs expression rank cutoff
```{r cor vs topn, fig.height=6, fig.width=15}
# Helper: retrieve DE result table for a given platform / comparison / level
get_de_table <- function(platform, comp, level = "gene") {
  if (level == "gene") {
    dge_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("gene_id")
  } else {
    dte_lists[[platform]][[paste0(comp, ".DE")]]$table %>%
      tibble::rownames_to_column("transcript_id")
  }
}

# Compute Pearson r of logFC for top-N expressed features across a platform pair
compute_cor_vs_topn <- function(level = "gene", p1, p2,
                                top_ns = c(100, 200, 500, 1000, 2000, 5000, NA)) {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"

  # Expression rank: mean logCPM across comparisons using p1 as reference
  expr_rank <- purrr::map_dfr(comparisons, ~ get_de_table(p1, .x, level) %>%
                                select(all_of(id_col), logCPM)) %>%
    group_by(across(all_of(id_col))) %>%
    summarise(mean_logCPM = mean(logCPM), .groups = "drop") %>%
    arrange(desc(mean_logCPM)) %>%
    mutate(rank = row_number())

  # All logFC data joined for p1 and p2, no significance filter
  all_joined <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- get_de_table(p1, comp, level) %>% select(all_of(id_col), logFC)
    df2 <- get_de_table(p2, comp, level) %>% select(all_of(id_col), logFC)
    full_join(df1, df2, by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      left_join(expr_rank %>% select(all_of(id_col), rank), by = id_col) %>%
      mutate(comparison = comp)
  })

  purrr::map_dfr(top_ns, function(n) {
    filtered <- if (is.na(n)) all_joined else dplyr::filter(all_joined, rank <= n)
    filtered %>%
      group_by(comparison) %>%
      summarise(
        r = cor(logFC_p1, logFC_p2, use = "complete.obs"),
        n_features = dplyr::n(),
        .groups = "drop"
      ) %>%
      mutate(
        top_n = if (is.na(n)) "All" else as.character(n),
        level = level,
        pair = paste(p1, "vs", p2)
      )
  })
}

top_ns <- c(100, 200, 500, 1000, 2000, 5000, 10000, NA)

cor_vs_topn <- bind_rows(
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn("gene",       .x[1], .x[2], top_ns)),
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn("transcript", .x[1], .x[2], top_ns))
) %>%
  mutate(top_n = factor(top_n, levels = c("100", "200", "500", "1000", "2000", "5000", "10000", "All")))

cor_vs_topn %>%
  group_by(pair, level, top_n) %>%
  summarise(mean_r = mean(r), .groups = "drop") %>%
  ggplot(aes(x = top_n, y = mean_r, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  labs(
    title = "logFC correlation vs expression rank cutoff (mean across comparisons)",
    x = "Top N most-expressed features",
    y = "Mean Pearson r",
    color = "Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Scatter plots: top 2000 expressed features
```{r top 2000 scatter, fig.height=4, fig.width=8}
make_top_n_scatter <- function(level, top_n = 2000) {
  id_col     <- if (level == "gene") "gene_id" else "transcript_id"
  dot_color  <- if (level == "gene") "#B4CD55" else "#AE98EB"

  plotlist <- map(platform_pairs, function(pair) {
    p1 <- pair[1]; p2 <- pair[2]

    top_ids <- purrr::map_dfr(comparisons, ~ get_de_table(p1, .x, level) %>%
                                select(all_of(id_col), logCPM)) %>%
      group_by(across(all_of(id_col))) %>%
      summarise(mean_logCPM = mean(logCPM), .groups = "drop") %>%
      slice_max(mean_logCPM, n = top_n) %>%
      pull(all_of(id_col))

    all_data <- purrr::map_dfr(comparisons, function(comp) {
      df1 <- get_de_table(p1, comp, level) %>% filter(.data[[id_col]] %in% top_ids)
      df2 <- get_de_table(p2, comp, level) %>% filter(.data[[id_col]] %in% top_ids)
      full_join(
        df1 %>% select(all_of(id_col), logFC),
        df2 %>% select(all_of(id_col), logFC),
        by = id_col, suffix = c("_p1", "_p2")
      ) %>% drop_na() %>% mutate(comparison = comp)
    })

    cors <- all_data %>%
      group_by(comparison) %>%
      summarise(r = cor(logFC_p1, logFC_p2, use = "complete.obs"), .groups = "drop") %>%
      mutate(label = paste0("r = ", round(r, 2)))

    ggplot(all_data, aes(x = logFC_p1, y = logFC_p2)) +
      geom_point(color = dot_color, alpha = 0.4, size = 0.3) +
      geom_density_2d(color = "gray60", size = 0.3) +
      geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "gray50") +
      geom_smooth(method = "lm", se = FALSE, color = "black", size = 0.3) +
      facet_wrap(~comparison, ncol = 3) +
      geom_text(data = cors, aes(x = -Inf, y = Inf, label = label),
                inherit.aes = FALSE, hjust = -0.1, vjust = 1.2, size = 3.5) +
      labs(
        title = paste0(level, " top ", top_n, ": ", p1, " vs ", p2),
        x = paste0("logFC (", p1, ")"),
        y = paste0("logFC (", p2, ")")
      ) +
      theme_minimal(base_size = 8)
  })

  ggarrange(plotlist = plotlist, ncol = 3, nrow = 2)
}

make_top_n_scatter("gene",       top_n = 2000)
make_top_n_scatter("transcript", top_n = 2000)
```

## Version 2: Global mean logCPM ranking

Rather than ranking by a single platform's expression, here we compute the mean logCPM per feature averaged across **all 4 platforms × 3 comparisons** (12 values per feature). This gives a platform-neutral expression estimate for ranking, removing the winner's-curse bias of version 1.

```{r global mean logCPM rank, fig.height=6, fig.width=15}
# Mean logCPM per feature across all platforms and comparisons
get_global_expr <- function(level = "gene") {
  id_col <- if (level == "gene") "gene_id" else "transcript_id"
  purrr::map_dfr(platforms, function(platform) {
    purrr::map_dfr(comparisons, function(comp) {
      get_de_table(platform, comp, level) %>%
        select(all_of(id_col), logCPM)
    })
  }) %>%
    group_by(across(all_of(id_col))) %>%
    summarise(mean_logCPM_global = mean(logCPM), .groups = "drop")
}

gene_global_expr <- get_global_expr("gene")
tx_global_expr   <- get_global_expr("transcript")

compute_cor_vs_topn_global <- function(level = "gene", p1, p2,
                                       top_ns = c(100, 200, 500, 1000, 2000, 5000, 10000, NA)) {
  id_col      <- if (level == "gene") "gene_id" else "transcript_id"
  global_expr <- if (level == "gene") gene_global_expr else tx_global_expr

  global_rank <- global_expr %>%
    arrange(desc(mean_logCPM_global)) %>%
    mutate(rank = row_number())

  all_joined <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- get_de_table(p1, comp, level) %>% select(all_of(id_col), logFC)
    df2 <- get_de_table(p2, comp, level) %>% select(all_of(id_col), logFC)
    full_join(df1, df2, by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      left_join(global_rank %>% select(all_of(id_col), rank), by = id_col) %>%
      mutate(comparison = comp)
  })

  purrr::map_dfr(top_ns, function(n) {
    filtered <- if (is.na(n)) all_joined else dplyr::filter(all_joined, rank <= n)
    filtered %>%
      group_by(comparison) %>%
      summarise(
        r          = cor(logFC_p1, logFC_p2, use = "complete.obs"),
        n_features = dplyr::n(),
        .groups    = "drop"
      ) %>%
      mutate(
        top_n = if (is.na(n)) "All" else as.character(n),
        level = level,
        pair  = paste(p1, "vs", p2)
      )
  })
}

cor_vs_topn_global <- bind_rows(
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn_global("gene",       .x[1], .x[2], top_ns)),
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_topn_global("transcript", .x[1], .x[2], top_ns))
) %>%
  mutate(top_n = factor(top_n, levels = c("100", "200", "500", "1000", "2000", "5000", "10000", "All")))

cor_vs_topn_global %>%
  group_by(pair, level, top_n) %>%
  summarise(mean_r = mean(r), .groups = "drop") %>%
  ggplot(aes(x = top_n, y = mean_r, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  labs(
    title = "logFC correlation vs top-N cutoff (global mean logCPM ranking, all platforms)",
    x     = "Top N features (ranked by global mean logCPM)",
    y     = "Mean Pearson r (across comparisons)",
    color = "Level"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Version 3: Global mean logCPM threshold filter

Instead of a rank cutoff, filter features by a minimum global mean logCPM threshold. This is more interpretable — each point on the x-axis corresponds to a biological expression level rather than an arbitrary rank.

```{r logCPM threshold filter, fig.height=6, fig.width=15}
thresholds <- seq(-2, 6, by = 1)

compute_cor_vs_threshold <- function(level = "gene", p1, p2, thresholds) {
  id_col      <- if (level == "gene") "gene_id" else "transcript_id"
  global_expr <- if (level == "gene") gene_global_expr else tx_global_expr

  all_joined <- purrr::map_dfr(comparisons, function(comp) {
    df1 <- get_de_table(p1, comp, level) %>% select(all_of(id_col), logFC)
    df2 <- get_de_table(p2, comp, level) %>% select(all_of(id_col), logFC)
    full_join(df1, df2, by = id_col, suffix = c("_p1", "_p2")) %>%
      drop_na(logFC_p1, logFC_p2) %>%
      left_join(global_expr, by = id_col) %>%
      mutate(comparison = comp)
  })

  purrr::map_dfr(thresholds, function(thr) {
    filtered <- dplyr::filter(all_joined, mean_logCPM_global >= thr)
    filtered %>%
      group_by(comparison) %>%
      summarise(
        r          = cor(logFC_p1, logFC_p2, use = "complete.obs"),
        n_features = dplyr::n(),
        .groups    = "drop"
      ) %>%
      mutate(threshold = thr, level = level, pair = paste(p1, "vs", p2))
  })
}

cor_vs_threshold <- bind_rows(
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_threshold("gene",       .x[1], .x[2], thresholds)),
  purrr::map_dfr(platform_pairs, ~ compute_cor_vs_threshold("transcript", .x[1], .x[2], thresholds))
)

# Correlation vs threshold
cor_vs_threshold %>%
  group_by(pair, level, threshold) %>%
  summarise(mean_r = mean(r), .groups = "drop") %>%
  ggplot(aes(x = threshold, y = mean_r, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  labs(
    title = "logFC correlation vs global mean logCPM threshold",
    x     = "Minimum global mean logCPM",
    y     = "Mean Pearson r (across comparisons)",
    color = "Level"
  ) +
  theme_minimal()
```

```{r n features vs threshold, fig.height=6, fig.width=15}
# Number of features retained at each threshold
cor_vs_threshold %>%
  group_by(pair, level, threshold) %>%
  summarise(mean_n = mean(n_features), .groups = "drop") %>%
  ggplot(aes(x = threshold, y = mean_n, color = level, group = level)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~pair, ncol = 3) +
  scale_color_manual(values = c(gene = "#B4CD55", transcript = "#AE98EB")) +
  scale_y_log10() +
  labs(
    title = "Number of features retained vs global mean logCPM threshold",
    x     = "Minimum global mean logCPM",
    y     = "Mean N features (log scale)",
    color = "Level"
  ) +
  theme_minimal()
```

