---
title: "Spike in DE analysis (IsoQuant quantification)"
author: "Yupei & Ash"
version: "v1.0"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  sequins_tsv: '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv'
  sirv_csv: '/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv'
  bulk_meta: "/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt"
  # Illumina DGE (Salmon bundle)
  bulk_dge_rds: "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds"
  # IsoQuant DGE bundles (no_novel set)
  isoquant_ont_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds"
  isoquant_pb_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds"
  isoquant_drna_bulk_rds: "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds"
  fig.path: NULL
---

# Setup: Create DGE from input
```{r setup, include=FALSE}
knitr::opts_chunk$set(cache = TRUE, fig.width = 7, fig.height=5, warning = F)

if (!exists("params")) {
  params <- list(
    random_seed = 2024,
    cache_dir = NULL,
    sequins_tsv = '/vast/projects/LongBench/reference_files/Sequin_resources/v2.4/rnasequin_isoforms_2.4.tsv',
    sirv_csv = '/home/users/allstaff/you.yu/LongBench/reference_files/SIRV_Set4_Norm_Sequences_20210507/SIRV_set4_concentration.csv',
    bulk_meta = "/vast/projects/LongBench/sequencing_data/illumina_bulk/metadata.txt",
    bulk_dge_rds = "/vast/projects/LongBench/analysis/workflow/rmarkdown/RDS/bulk_DGE.obj.rds",
    isoquant_ont_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.ont_bulk.rds",
    isoquant_pb_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.pb_bulk.rds",
    isoquant_drna_bulk_rds = "/home/users/allstaff/you.yu/LongBench/analysis/lr_bulk/result/IsoQuant_output_no_novel/isoquant_bulk_dge.dRNA_bulk.rds",
    fig.path = NULL
  )
}
```

```{r Libraries,include=FALSE}
library(limma)
library(edgeR)
library(tidyverse)
library(RColorBrewer)
library(dplyr)
library(tidyr)
library(ggplot2)
library(ggrepel)
library(glue)
library(ggplot2)
library(gridExtra)
library(ggExtra)
library(knitr)
library(kableExtra)
library(rtracklayer)
library(scales)
library(UpSetR)
library(ggtree)
library(aplot)
library(patchwork)

# setup some conflict functions
rename <- dplyr::rename
select <- dplyr::select


knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),  # Enable caching only if cache_dir is provided
  fig.width = 7,                     # Set default figure width
  fig.height = 5,                    # Set default figure height
  echo = FALSE,                       # Hide code in the output
  message = FALSE,                    # Suppress messages in the output
  warning = FALSE,                    # Suppress warnings in the output
  fig.path = params$fig.path,         # Set figure path (NULL if not provided)
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL  # Set output devices if fig.path is provided
)

# Set cache.path only if params$cache_dir is not NULL
if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache.path = params$cache_dir)
}
```

```{r color code, include=FALSE, eval=T}
color_palette <- c(
  PacBio = "#df1995",
  ONT = "#00789b",
  ONT_1 = "#04476c", #cDNA
  ONT_2 = "#00789b", # dRNA
  Illumina = "#e88b20"
)
```

```{r function for splitting the transcript origin}
is.sequins <- function(dge) {grepl("^R", rownames(dge))}
is.human <- function(dge) {grepl("^ENS", rownames(dge))}
is.sirv_or_ercc <- function(dge) {!is.human(dge) & !is.sequins(dge)} # thie is the whole SIRV set4 (ERCC + SIRV E0 + Long SIRV)
is.sirv_E0 <- function(dge) {grepl("^SIRV\\d{3}$", rownames(dge))}
is.sirv_all <- function(dge) {grepl("^SIRV", rownames(dge))} # including all SIRV E0 and Long SIRV
is.long_sirv <- function(dge) {grepl("^SIRV\\d{4}", rownames(dge))}
is.ercc      <- function(dge) grepl("^(ERCC|DQ|EF)", rownames(dge))
```

# Load the count
## Get the count data from RDS

LR platforms (ONT cDNA / PacBio / ONT dRNA) use IsoQuant-derived DGE objects.
Illumina uses the Salmon-derived bundle (same as the main DE analysis).

```{r helper functions}
normalize_rownames <- function(dge) {
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}

extract_and_normalize_ids <- function(dge) {
  if (!"full_id" %in% colnames(dge$genes)) {
    dge$genes$full_id <- rownames(dge)
  }
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}
```

```{r load-dge, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
bulk_obj          <- readRDS(params$bulk_dge_rds)
isoquant_ont_obj  <- readRDS(params$isoquant_ont_bulk_rds)
isoquant_pb_obj   <- readRDS(params$isoquant_pb_bulk_rds)
isoquant_drna_obj <- readRDS(params$isoquant_drna_bulk_rds)

# Illumina — from Salmon bundle
ill_bulk.tx.dge <- extract_and_normalize_ids(bulk_obj[["ill_bulk.tx.dge"]])

# LR — IsoQuant (no_novel)
ont_bulk.tx.dge   <- normalize_rownames(isoquant_ont_obj[["ont_bulk.tx.dge"]])
pb_bulk.tx.dge    <- normalize_rownames(isoquant_pb_obj[["pb_bulk.tx.dge"]])
drna_bulk.tx.dge  <- normalize_rownames(isoquant_drna_obj[["dRNA_bulk.tx.dge"]])
```


# Separate spike in types
```{r}
# Get seqins and sirv transcript dge
ill_bulk.txi.dge_sequins <- ill_bulk.tx.dge[is.sequins(ill_bulk.tx.dge), ]
ont_bulk.txi.dge_sequins <- ont_bulk.tx.dge[is.sequins(ont_bulk.tx.dge), ]
pb_bulk.txi.dge_sequins <- pb_bulk.tx.dge[is.sequins(pb_bulk.tx.dge), ]
drna_bulk.txi.dge_sequins <- drna_bulk.tx.dge[is.sequins(drna_bulk.tx.dge), ]

## get sirv transcript dge
ill_bulk.txi.dge_sirv <- ill_bulk.tx.dge[is.sirv_all(ill_bulk.tx.dge), ]
ont_bulk.txi.dge_sirv <- ont_bulk.tx.dge[is.sirv_all(ont_bulk.tx.dge), ]
pb_bulk.txi.dge_sirv <- pb_bulk.tx.dge[is.sirv_all(pb_bulk.tx.dge), ]
drna_bulk.txi.dge_sirv <- drna_bulk.tx.dge[is.sirv_all(drna_bulk.tx.dge), ]

## get ercc transcript dge
ill_bulk.txi.dge_ercc <- ill_bulk.tx.dge[is.ercc(ill_bulk.tx.dge), ]
ont_bulk.txi.dge_ercc <- ont_bulk.tx.dge[is.ercc(ont_bulk.tx.dge), ]
pb_bulk.txi.dge_ercc <- pb_bulk.tx.dge[is.ercc(pb_bulk.tx.dge), ]
drna_bulk.txi.dge_ercc <- drna_bulk.tx.dge[is.ercc(drna_bulk.tx.dge), ]
```

# Gene Transcript discovery analysis
Using `filterByExpr` for each dataset, using min.count = 10 for gene and min.count = 5 for transcript.
```{r Gene Transcript discovery upset plot}
# Transcript
get_discovery_upset <- function(dge_lst, min.count = 10) {
  # align all platforms to a common transcript order before stacking
  common_tx <- Reduce(intersect, lapply(dge_lst, rownames))

  make_keep <- function(dge) {
    keep <- filterByExpr(dge[common_tx, ], group = dge$samples$sample, min.count = min.count)
    keep[common_tx]
  }

  keeps <- lapply(dge_lst, make_keep)

  tx_discovery <- tibble(
    "Illumina" = keeps[["Illumina"]],
    "ONT cDNA" = keeps[["ONT cDNA"]],
    "PacBio"   = keeps[["PacBio"]],
    "ONT dRNA" = keeps[["ONT dRNA"]],
    Length = dge_lst[["Illumina"]][common_tx, ]$genes$Length
  )

  p <- ComplexUpset::upset(
    tx_discovery,
    intersect = c('Illumina', "ONT cDNA", "PacBio", "ONT dRNA"),
    annotations = list(
      "Gene length" = list(
        aes = aes(x = intersection, y = Length),
        geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
      )
    ),
    wrap = TRUE
  )

  count_tbl <- tibble(
    Platform   = names(keeps),
    Identified = sapply(keeps, sum),
    Total      = length(common_tx)
  )

  list(plot = p, table = count_tbl)
}


# Transcript
res_seq <- list(
  "Illumina" = ill_bulk.txi.dge_sequins,
  "ONT cDNA" = ont_bulk.txi.dge_sequins,
  "PacBio" = pb_bulk.txi.dge_sequins,
  "ONT dRNA" = drna_bulk.txi.dge_sequins
) %>% get_discovery_upset(min.count = 5)
res_seq$plot + ggtitle("Sequins transcript discovery (IsoQuant)")
res_seq$table %>% kable(caption = "Sequins transcripts identified per platform", format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

res_sirv <- list(
  "Illumina" = ill_bulk.txi.dge_sirv,
  "ONT cDNA" = ont_bulk.txi.dge_sirv,
  "PacBio" = pb_bulk.txi.dge_sirv,
  "ONT dRNA" = drna_bulk.txi.dge_sirv
) %>% get_discovery_upset(min.count = 5)
res_sirv$plot + ggtitle("SIRV transcript discovery (IsoQuant)")
res_sirv$table %>% kable(caption = "SIRV transcripts identified per platform", format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)

## ERCC with true abundance
dge_lst <- list(
  "Illumina" = ill_bulk.txi.dge_ercc,
  "ONT cDNA" = ont_bulk.txi.dge_ercc,
  "PacBio" = pb_bulk.txi.dge_ercc,
  "ONT dRNA" = drna_bulk.txi.dge_ercc
)
min.count = 5
ercc_common_tx <- Reduce(intersect, lapply(dge_lst, rownames))
make_ercc_keep <- function(dge) {
  keep <- filterByExpr(dge[ercc_common_tx, ], group = dge$samples$sample, min.count = min.count)
  keep[ercc_common_tx]
}
tx_discovery <- tibble(
    "Illumina" = make_ercc_keep(dge_lst[["Illumina"]]),
    "ONT cDNA" = make_ercc_keep(dge_lst[["ONT cDNA"]]),
    "PacBio"   = make_ercc_keep(dge_lst[["PacBio"]]),
    "ONT dRNA" = make_ercc_keep(dge_lst[["ONT dRNA"]]),
    Length = dge_lst[["Illumina"]][ercc_common_tx, ]$genes$Length,
    Name = ercc_common_tx
  ) %>% left_join(read_csv(params$sirv_csv) %>% filter(grepl("ERCC", txid)), by = c("Name" = "GenBank"))

  ComplexUpset::upset(
    tx_discovery,
    intersect = c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA"),
    annotations = list(
      # 1st method - passing list:
      "Gene length" = list(
        aes = aes(x = intersection, y = Length),
        # provide a list if you wish to add several geoms
        geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
      ),
      "True abundance (log10 amol/ul)" = list(
        aes = aes(x = intersection, y = log10(conc)),
        geom = geom_boxplot(na.rm = TRUE, outliers = FALSE)
      )
    ),
    wrap = TRUE
  ) + ggtitle("ERCC transcript discovery (IsoQuant)")

tx_discovery %>%
  summarise(across(c(Illumina, `ONT cDNA`, PacBio, `ONT dRNA`), sum)) %>%
  pivot_longer(everything(), names_to = "Platform", values_to = "Identified") %>%
  mutate(Total = length(ercc_common_tx)) %>%
  kable(caption = "ERCC transcripts identified per platform", format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

## IsoQuant vs Oarfish discovery comparison
```{r isoquant-vs-oarfish-discovery, fig.show='asis'}
# Oarfish LR DGEs are in the same bundle as Illumina (bulk_obj), loaded in load-dge chunk
oarfish_ont  <- bulk_obj[["ont_bulk.tx.dge"]]
oarfish_pb   <- bulk_obj[["pb_bulk.tx.dge"]]
oarfish_drna <- bulk_obj[["drna_bulk.tx.dge"]]

count_detected <- function(dge, filter_fn, min.count = 5) {
  sub <- dge[filter_fn(dge), ]
  sum(filterByExpr(sub, group = sub$samples$sample, min.count = min.count))
}

spike_fns <- list(Sequins = is.sequins, SIRV = is.sirv_all, ERCC = is.ercc)

# Illumina uses Salmon in both analyses — same value for IsoQuant and Oarfish columns
comparison <- bind_rows(lapply(names(spike_fns), function(sp) {
  fn  <- spike_fns[[sp]]
  tot <- sum(fn(ill_bulk.tx.dge))
  bind_rows(
    tibble(`Spike-in` = sp, Platform = "Illumina",
           IsoQuant = count_detected(ill_bulk.tx.dge, fn),
           Oarfish  = count_detected(ill_bulk.tx.dge, fn),
           Total = tot),
    tibble(`Spike-in` = sp, Platform = "ONT cDNA",
           IsoQuant = count_detected(ont_bulk.tx.dge,  fn),
           Oarfish  = count_detected(oarfish_ont,       fn),
           Total = tot),
    tibble(`Spike-in` = sp, Platform = "PacBio",
           IsoQuant = count_detected(pb_bulk.tx.dge,   fn),
           Oarfish  = count_detected(oarfish_pb,        fn),
           Total = tot),
    tibble(`Spike-in` = sp, Platform = "ONT dRNA",
           IsoQuant = count_detected(drna_bulk.tx.dge, fn),
           Oarfish  = count_detected(oarfish_drna,      fn),
           Total = tot)
  )
}))

comparison %>%
  kable(caption = "Detected spike-in transcripts: IsoQuant vs Oarfish (LR platforms)",
        format = "html") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

# Differential Transcript expression analysis

```{r filter for common gene and transcript only}
# Transcript
# Sequin
common_tx <- Reduce(intersect, lapply(list(ill_bulk.txi.dge_sequins, ont_bulk.txi.dge_sequins, pb_bulk.txi.dge_sequins, drna_bulk.txi.dge_sequins), rownames))
ill_bulk.txi.dge_sequins <- ill_bulk.txi.dge_sequins[common_tx, ]
ont_bulk.txi.dge_sequins <- ont_bulk.txi.dge_sequins[common_tx, ]
pb_bulk.txi.dge_sequins <- pb_bulk.txi.dge_sequins[common_tx, ]
drna_bulk.txi.dge_sequins <- drna_bulk.txi.dge_sequins[common_tx, ]
keep <- filterByExpr(ill_bulk.txi.dge_sequins, group = ill_bulk.txi.dge_sequins$samples$group, min.count=5) &
          filterByExpr(ont_bulk.txi.dge_sequins, group = ont_bulk.txi.dge_sequins$samples$group, min.count=5) &
          filterByExpr(pb_bulk.txi.dge_sequins, group = pb_bulk.txi.dge_sequins$samples$group, min.count=5) &
          filterByExpr(drna_bulk.txi.dge_sequins, group = drna_bulk.txi.dge_sequins$samples$group, min.count=5)
ill_bulk.txi.dge_sequins <- ill_bulk.txi.dge_sequins[keep, ,keep.lib.sizes=FALSE]
ont_bulk.txi.dge_sequins <- ont_bulk.txi.dge_sequins[keep, ,keep.lib.sizes=FALSE]
pb_bulk.txi.dge_sequins <- pb_bulk.txi.dge_sequins[keep, ,keep.lib.sizes=FALSE]
drna_bulk.txi.dge_sequins <- drna_bulk.txi.dge_sequins[keep, ,keep.lib.sizes=FALSE]


# Sirv
common_tx <- Reduce(intersect, lapply(list(ill_bulk.txi.dge_sirv, ont_bulk.txi.dge_sirv, pb_bulk.txi.dge_sirv, drna_bulk.txi.dge_sirv), rownames))
ill_bulk.txi.dge_sirv <- ill_bulk.txi.dge_sirv[common_tx, ]
ont_bulk.txi.dge_sirv <- ont_bulk.txi.dge_sirv[common_tx, ]
pb_bulk.txi.dge_sirv <- pb_bulk.txi.dge_sirv[common_tx, ]
drna_bulk.txi.dge_sirv <- drna_bulk.txi.dge_sirv[common_tx, ]

keep <- filterByExpr(ill_bulk.txi.dge_sirv, group = ill_bulk.txi.dge_sirv$samples$group, min.count=5) &
          filterByExpr(ont_bulk.txi.dge_sirv, group = ont_bulk.txi.dge_sirv$samples$group, min.count=5) &
          filterByExpr(pb_bulk.txi.dge_sirv, group = pb_bulk.txi.dge_sirv$samples$group, min.count=5) &
          filterByExpr(drna_bulk.txi.dge_sirv, group = drna_bulk.txi.dge_sirv$samples$group, min.count=5)
ill_bulk.txi.dge_sirv <- ill_bulk.txi.dge_sirv[keep, ,keep.lib.sizes=FALSE]
ont_bulk.txi.dge_sirv <- ont_bulk.txi.dge_sirv[keep, ,keep.lib.sizes=FALSE]
pb_bulk.txi.dge_sirv <- pb_bulk.txi.dge_sirv[keep, ,keep.lib.sizes=FALSE]
drna_bulk.txi.dge_sirv <- drna_bulk.txi.dge_sirv[keep, ,keep.lib.sizes=FALSE]



# ERCC
common_tx <- Reduce(intersect, lapply(list(ill_bulk.txi.dge_ercc, ont_bulk.txi.dge_ercc, pb_bulk.txi.dge_ercc, drna_bulk.txi.dge_ercc), rownames))
ill_bulk.txi.dge_ercc <- ill_bulk.txi.dge_ercc[common_tx, ]
ont_bulk.txi.dge_ercc <- ont_bulk.txi.dge_ercc[common_tx, ]
pb_bulk.txi.dge_ercc <- pb_bulk.txi.dge_ercc[common_tx, ]
drna_bulk.txi.dge_ercc <- drna_bulk.txi.dge_ercc[common_tx, ]

keep <- filterByExpr(ill_bulk.txi.dge_ercc, group = ill_bulk.txi.dge_ercc$samples$group, min.count=5) &
          filterByExpr(ont_bulk.txi.dge_ercc, group = ont_bulk.txi.dge_ercc$samples$group, min.count=5) &
          filterByExpr(pb_bulk.txi.dge_ercc, group = pb_bulk.txi.dge_ercc$samples$group, min.count=5) &
          filterByExpr(drna_bulk.txi.dge_ercc, group = drna_bulk.txi.dge_ercc$samples$group, min.count=5)
ill_bulk.txi.dge_ercc <- ill_bulk.txi.dge_ercc[keep, ,keep.lib.sizes=FALSE]
ont_bulk.txi.dge_ercc <- ont_bulk.txi.dge_ercc[keep, ,keep.lib.sizes=FALSE]
pb_bulk.txi.dge_ercc <- pb_bulk.txi.dge_ercc[keep, ,keep.lib.sizes=FALSE]
drna_bulk.txi.dge_ercc <- drna_bulk.txi.dge_ercc[keep, ,keep.lib.sizes=FALSE]

```


## Transcript level
### Run DTE edgeR+scaling (IsoQuant LR / Salmon Illumina)

> **Note:** Illumina counts are scaled by overdispersion estimates from Salmon bootstrap samples. IsoQuant LR platforms (ONT cDNA, PacBio, ONT dRNA) do not provide bootstrap overdispersion, so their counts are used unscaled.

```{r}
DTE.edgeR_sequins <- function(dge, norm.method = "TMM") {

  if (!is.null(norm.method)) {
    dge <- calcNormFactors(dge, method = norm.method)
  }

  design <- model.matrix(~0 + sequins, data = dge$samples)
  dge <- estimateDisp(dge, design)
  qlfit <- glmQLFit(dge, design)

  contrasts <- makeContrasts(
    AvsB = sequinsA - sequinsB,
    levels = design
  )

  AvsB.QLF.test <- glmQLFTest(qlfit, contrast = contrasts[, "AvsB"])

  list(
    AvsB.DE = AvsB.QLF.test
  )
}
```

```{r get dte scaling}
filter_dge <- function(dge, min.count = 5) {
  keep <- filterByExpr(dge, group = dge$samples$group, min.count = min.count)
  dge[keep, , keep.lib.sizes = FALSE]
}

# Scale Illumina by overdispersion (from Salmon bootstraps); LR used as-is
ill_bulk.txi.dge_sequins_scaled <- filter_dge(ill_bulk.txi.dge_sequins)
ill_bulk.txi.dge_sequins_scaled$counts <- ill_bulk.txi.dge_sequins_scaled$counts / ill_bulk.txi.dge_sequins_scaled$genes$Overdispersion

ill_bulk.dte_sequins.list <- DTE.edgeR_sequins(ill_bulk.txi.dge_sequins_scaled)
ont_bulk.dte_sequins.list <- DTE.edgeR_sequins(filter_dge(ont_bulk.txi.dge_sequins))
drna_bulk.dte_sequins.list <- DTE.edgeR_sequins(filter_dge(drna_bulk.txi.dge_sequins))
pb_bulk.dte_sequins.list <- DTE.edgeR_sequins(filter_dge(pb_bulk.txi.dge_sequins))

ill_bulk.txi.dge_sirvercc_scaled <- filter_dge(rbind(ill_bulk.txi.dge_sirv, ill_bulk.txi.dge_ercc))
ill_bulk.txi.dge_sirvercc_scaled$counts <- ill_bulk.txi.dge_sirvercc_scaled$counts / ill_bulk.txi.dge_sirvercc_scaled$genes$Overdispersion

ill_bulk.dte_sirv.list <- DTE.edgeR_sequins(ill_bulk.txi.dge_sirvercc_scaled)
ont_bulk.dte_sirv.list <- DTE.edgeR_sequins(filter_dge(rbind(ont_bulk.txi.dge_sirv, ont_bulk.txi.dge_ercc)))
drna_bulk.dte_sirv.list <- DTE.edgeR_sequins(filter_dge(rbind(drna_bulk.txi.dge_sirv, drna_bulk.txi.dge_ercc)))
pb_bulk.dte_sirv.list <- DTE.edgeR_sequins(filter_dge(rbind(pb_bulk.txi.dge_sirv, pb_bulk.txi.dge_ercc)))
```

### Upset plot helper
```{r upset-helper}
complex_upset_plot <- function(lst) {
  DEGs <- lapply(lst, function(x) rownames(x)[decideTests(x) != 0])
  long_df <- stack(DEGs)

  if (nrow(long_df) == 0) {
    message("No significant DE features detected in any platform — skipping upset plot.")
    return(list(plot = ggplot() + annotate("text", x = 0.5, y = 0.5, label = "No significant DE detected", size = 6) + theme_void(), matrix = tibble(Gene = character())))
  }

  binary_matrix <- long_df %>%
    mutate(present = 1) %>%
    pivot_wider(names_from = ind, values_from = present, values_fill = 0)

  # ensure all platform columns exist (platforms with 0 DE get added as all-zero)
  for (nm in names(lst)) {
    if (!nm %in% colnames(binary_matrix)) binary_matrix[[nm]] <- 0L
  }

  p <- ComplexUpset::upset(
    binary_matrix %>% select(-values) %>% as.data.frame(),
    intersect = names(lst),
    base_annotations = list(
      'Intersection size' = ComplexUpset::intersection_size(
        counts = TRUE,
        text = list(element_text(size = 3, vjust = -0.2))
      )
    ),
    wrap = TRUE
  )
  list(plot = p, matrix = binary_matrix %>% rename(Gene = values))
}
```

### DTE Upset plot (sequins transcript)
```{r upset plot dte sequins edgeR scaling, fig.width=8, fig.height=6}
lst <- list(
  "Illumina" = ill_bulk.dte_sequins.list$AvsB.DE,
  "ONT cDNA" = ont_bulk.dte_sequins.list$AvsB.DE,
  "PacBio" = pb_bulk.dte_sequins.list$AvsB.DE,
  "ONT dRNA" = drna_bulk.dte_sequins.list$AvsB.DE
)
complex_upset_plot(lst)$plot + ggtitle("DTE Sequins MIX_A vs MIX_B (IsoQuant)") + theme(plot.title = element_text(size = 20))
```

### DTE Upset plot (SIRV + ERCC transcript)
```{r upset plot dte sirv edgeR scaling, fig.width=8, fig.height=6}
lst_sirv <- list(
  "Illumina" = ill_bulk.dte_sirv.list$AvsB.DE,
  "ONT cDNA" = ont_bulk.dte_sirv.list$AvsB.DE,
  "PacBio" = pb_bulk.dte_sirv.list$AvsB.DE,
  "ONT dRNA" = drna_bulk.dte_sirv.list$AvsB.DE
)
complex_upset_plot(lst_sirv)$plot + ggtitle("DTE SIRV+ERCC MIX_A vs MIX_B (IsoQuant, FP calls)") + theme(plot.title = element_text(size = 20))
```

# Spike in fold change analysis
```{r}
compare_sequins_txi_logFC <- function(glm.test.rst, fig.title){
  sequins.gt <- read_tsv(params$sequins_tsv)
  sequins.gt$gt.logFC <- log2(sequins.gt$MIX_A/sequins.gt$MIX_B)

  # glm.test result
  df <- glm.test.rst %>%
    topTags(adjust.method = 'BH', n = nrow(glm.test.rst))  %>%
    data.frame() %>% tibble::rownames_to_column(var="Transcript") %>%
    select(Transcript, logFC, FDR) %>%
    left_join(sequins.gt %>% select(NAME, gt.logFC, LENGTH), by = c("Transcript"="NAME"))

  # add a column for significant DE, "down" for downregulated, "up" for upregulated
  df$DE <- ifelse(df$logFC > 0 & df$FDR < 0.05, "up", ifelse(df$logFC < 0 & df$FDR < 0.05, "down", "not significant"))

  # plot gt.logFC vs logFC, add correlation
  p <- ggplot(df, aes(x = logFC, y = gt.logFC, color = DE, size = LENGTH, label=Transcript)) +
    geom_point() +
    # geom_text_repel(size = 3)    +
    scale_size(limits = c(min(sequins.gt$LENGTH), max(sequins.gt$LENGTH))) +
    geom_abline(intercept = 0, slope = 1, linetype = "dashed") +
    scale_color_viridis_d() +
    theme_minimal() +
    labs(x = "Estimated logFC", y = " Ground truth logFC", color = "DE") +
    annotate("text",
      x = min(df$logFC), y = max(df$gt.logFC),
      label = paste(
        "\nMAE: ", round(mean(abs(df$logFC - df$gt.logFC)), 2),
        "\nMSE: ", round(mean((df$logFC - df$gt.logFC)^2), 2)
      ),
      hjust = 0, vjust = 1,
      color = "black", size = 4,
      fontface = "italic"
    ) +
    ggtitle(fig.title)
  return(p)
}



```


```{r spike in logFC scater edgeR + scaling, fig.width=8, fig.height=6}
# Sequins
lst <- list(
  "Illumina" = ill_bulk.dte_sequins.list$AvsB.DE,
  "ONT cDNA" = ont_bulk.dte_sequins.list$AvsB.DE,
  "PacBio" = pb_bulk.dte_sequins.list$AvsB.DE,
  "ONT dRNA" = drna_bulk.dte_sequins.list$AvsB.DE
)
lfc_plot <- lapply(1:4, function(x) compare_sequins_txi_logFC(lst[[x]], fig.title = names(lst)[x]))
(lfc_plot[[1]] | lfc_plot[[2]]) / (lfc_plot[[3]] | lfc_plot[[4]]) + patchwork::plot_annotation(
  title = "Sequins MixA vs MixB (IsoQuant)",
  theme = theme(plot.title = element_text(size =25))
  )  + patchwork::plot_layout(
    guides = "collect"
)
```


```{r spike in logFC scater edgeR + scaling SIRV, fig.width=8, fig.height=6}
# SIRV (including ERCC)

compare_sirv_txi_logFC <- function(glm.test.rst.list){
  df <- do.call(rbind, lapply(
    names(glm.test.rst.list),
    function(x){
      glm.test.rst.list[[x]] %>%
        topTags(adjust.method = 'BH', n = nrow(glm.test.rst.list[[x]])) %>%
        data.frame() %>%
        tibble::rownames_to_column(var="NAME") %>%
        select(NAME, logFC, FDR) %>%
        mutate(gt.logFC = 0, Platform = x)
    }
  ))

  df$DE <- ifelse(df$logFC > 0 & df$FDR < 0.05, "up",
                  ifelse(df$logFC < 0 & df$FDR < 0.05, "down", "not_sig"))

  # Calculate MAE and MSE per platform
  metrics <- df %>%
    group_by(Platform) %>%
    summarise(
      MAE = mean(abs(logFC)),
      MSE = mean(logFC^2)
    )

  # Generate plot
  p <- ggplot(df, aes(x = logFC, fill = DE, y = Platform)) +
    geom_dotplot(binwidth = (max(df$logFC) - min(df$logFC))/100, stackdir = "up", color = NA, dotsize = 1.5) +
    scale_fill_viridis_d() +
    theme_minimal() +
    geom_vline(xintercept = 0, linetype = "dashed") +
    labs(x = "Estimated logFC", fill = "DE")

  # Add per-platform MAE and MSE as annotations
  y_positions <- seq_along(unique(df$Platform))
  for (i in seq_len(nrow(metrics))) {
    p <- p + annotate("text",
                      x = min(df$logFC),
                      y = metrics$Platform[i],
                      vjust = -4,
                      hjust = -1,
                      label = paste0("MAE=", round(metrics$MAE[i], 2),
                                     " | MSE=", round(metrics$MSE[i], 2)),
                      size = 3.5,
                      fontface = "italic")
  }

  return(p)
}

list(
  "Illumina" = ill_bulk.dte_sirv.list$AvsB.DE,
  "ONT cDNA" = ont_bulk.dte_sirv.list$AvsB.DE,
  "PacBio" = pb_bulk.dte_sirv.list$AvsB.DE,
  "ONT dRNA" = drna_bulk.dte_sirv.list$AvsB.DE
) %>% compare_sirv_txi_logFC + ggtitle("SIRV + ERCC logFC (IsoQuant)")
```

# DTE method comparison: IsoQuant vs Oarfish

> **Note:** Each platform's DTE is run independently using all features available in its own count matrix (no cross-platform overlap filter). Illumina and Oarfish LR counts are scaled by `$genes$Overdispersion` before DTE; IsoQuant LR counts are used unscaled.

```{r dte-method-comparison}
run_dte_count <- function(dge, filter_fn, scale_od = FALSE, min.count = 5) {
  sub <- dge[filter_fn(dge), ]
  keep <- filterByExpr(sub, group = sub$samples$group, min.count = min.count)
  sub <- sub[keep, , keep.lib.sizes = FALSE]
  if (nrow(sub) == 0) return(0L)
  if (scale_od && !is.null(sub$genes$Overdispersion)) {
    sub$counts <- sub$counts / sub$genes$Overdispersion
  }
  res <- DTE.edgeR_sequins(sub)
  sum(decideTests(res$AvsB.DE) != 0)
}

# Helper to normalise oarfish row names the same way as IsoQuant
normalize_rownames_oarfish <- function(dge) {
  rownames(dge) <- sub("\\|.*", "", rownames(dge))
  rownames(dge) <- sub("\\..*", "", rownames(dge))
  dge
}

oarfish_ont_n   <- normalize_rownames_oarfish(bulk_obj[["ont_bulk.tx.dge"]])
oarfish_pb_n    <- normalize_rownames_oarfish(bulk_obj[["pb_bulk.tx.dge"]])
oarfish_drna_n  <- normalize_rownames_oarfish(bulk_obj[["drna_bulk.tx.dge"]])

is.sirv_and_ercc <- function(dge) is.sirv_all(dge) | is.ercc(dge)

spike_filters <- list(
  Sequins    = is.sequins,
  `SIRV+ERCC` = is.sirv_and_ercc
)

# Illumina (Salmon) is the same for both columns; scale_od = TRUE because it has Overdispersion
ill_dte_count <- function(fn) run_dte_count(ill_bulk.tx.dge, fn, scale_od = TRUE)

platforms_lr <- list(
  list(label = "ONT cDNA",  isoquant = ont_bulk.tx.dge,  oarfish = oarfish_ont_n),
  list(label = "PacBio",    isoquant = pb_bulk.tx.dge,   oarfish = oarfish_pb_n),
  list(label = "ONT dRNA",  isoquant = drna_bulk.tx.dge, oarfish = oarfish_drna_n)
)

dte_rows <- do.call(rbind, lapply(names(spike_filters), function(spk) {
  fn <- spike_filters[[spk]]
  ill_n <- ill_dte_count(fn)
  lr_rows <- do.call(rbind, lapply(platforms_lr, function(p) {
    data.frame(
      `Spike-in`  = spk,
      Platform    = p$label,
      IsoQuant    = run_dte_count(p$isoquant, fn, scale_od = FALSE),
      Oarfish     = run_dte_count(p$oarfish,  fn, scale_od = TRUE),
      check.names = FALSE
    )
  }))
  rbind(
    data.frame(`Spike-in` = spk, Platform = "Illumina", IsoQuant = ill_n, Oarfish = ill_n, check.names = FALSE),
    lr_rows
  )
}))

kable(dte_rows, row.names = FALSE,
      caption = "Number of significant DETs (FDR < 0.05) per platform and spike-in type") %>%
  kable_styling(bootstrap_options = c("striped","hover","condensed"), full_width = FALSE) %>%
  collapse_rows(columns = 1, valign = "top")
```

```{r dte-comparison-figure, fig.width=10, fig.height=6}
run_dte_summary <- function(dge, filter_fn, scale_od = FALSE) {
  sub <- dge[filter_fn(dge), ]
  keep <- filterByExpr(sub, group = sub$samples$group, min.count = 5)
  sub <- sub[keep, , keep.lib.sizes = FALSE]
  n_detected <- nrow(sub)
  if (n_detected == 0) return(tibble(Detected = 0L, DE = 0L))
  if (scale_od && !is.null(sub$genes$Overdispersion)) {
    sub$counts <- sub$counts / sub$genes$Overdispersion
  }
  res <- DTE.edgeR_sequins(sub)
  tibble(Detected = n_detected, DE = sum(decideTests(res$AvsB.DE) != 0))
}

platforms_all <- list(
  list(label = "Illumina",  isoquant = ill_bulk.tx.dge,  oarfish = NULL,          scale_iq = TRUE),
  list(label = "ONT cDNA",  isoquant = ont_bulk.tx.dge,  oarfish = oarfish_ont_n, scale_iq = FALSE),
  list(label = "PacBio",    isoquant = pb_bulk.tx.dge,   oarfish = oarfish_pb_n,  scale_iq = FALSE),
  list(label = "ONT dRNA",  isoquant = drna_bulk.tx.dge, oarfish = oarfish_drna_n, scale_iq = FALSE)
)

fig_df <- do.call(rbind, lapply(names(spike_filters), function(spk) {
  fn <- spike_filters[[spk]]
  do.call(rbind, lapply(platforms_all, function(p) {
    iq <- run_dte_summary(p$isoquant, fn, scale_od = p$scale_iq)
    rows <- tibble(`Spike-in` = spk, Platform = p$label,
                   Method = if (p$label == "Illumina") "Salmon" else "IsoQuant",
                   Detected = iq$Detected, DE = iq$DE)
    if (!is.null(p$oarfish)) {
      oa <- run_dte_summary(p$oarfish, fn, scale_od = TRUE)
      rows <- bind_rows(rows, tibble(`Spike-in` = spk, Platform = p$label,
                                     Method = "Oarfish", Detected = oa$Detected, DE = oa$DE))
    }
    rows
  }))
}))

fig_df %>%
  pivot_longer(cols = c(Detected, DE), names_to = "Metric", values_to = "Count") %>%
  mutate(
    Platform = factor(Platform, levels = c("Illumina", "ONT cDNA", "PacBio", "ONT dRNA")),
    Method   = factor(Method,   levels = c("Salmon", "IsoQuant", "Oarfish")),
    Metric   = factor(Metric,   levels = c("Detected", "DE"))
  ) %>%
  ggplot(aes(x = Platform, y = Count, fill = Method)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  facet_grid(`Spike-in` ~ Metric, scales = "free_y") +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 30, hjust = 1),
    strip.text  = element_text(face = "bold")
  ) +
  labs(x = NULL, y = "Count", fill = "Method",
       title = "Spike-in transcript detection and DTE: IsoQuant vs Oarfish")
```
