---
title: "dRNA Run1 vs Run2 Consistency Check"
author: "Yupei"
date: '`r Sys.Date()`'
version: "v1.2"
output:
  html_document:
    code_folding: hide
    toc: yes
    toc_float: yes
    number_sections: yes
    toc_depth: 4
  pdf_document:
    toc: yes
    toc_depth: '4'
params:
  random_seed: 2024
  cache_dir: NULL
  fig.path: NULL
---

```{r setup_params_template, include=FALSE, eval=FALSE, message=FALSE, warning=FALSE}
params <- list(
  random_seed = 2024,
  cache_dir = NULL,
  fig.path = NULL
)
```

```{r setup_knitr_and_libraries, include=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = FALSE)
knitr::opts_chunk$set(
  cache = !is.null(params$cache_dir),
  fig.width = 12,
  fig.height = 7,
  message = FALSE,
  warning = FALSE,
  fig.path = params$fig.path,
  dev = if (!is.null(params$fig.path)) c("png", "svg") else NULL
)
if (!is.null(params$cache_dir)) {
  knitr::opts_chunk$set(cache.path = params$cache_dir)
}

library(tidyverse)
library(ggplot2)
library(patchwork)
library(ggrepel)

# Paths (hard-coded as per project convention — Rmd can be run independently)
dRNA_comparison_dir <- "/vast/projects/LongBench/analysis/dRNA_comparison/result"
oarfish_dir_run1    <- file.path(dRNA_comparison_dir, "oarfish_cov_output/dRNA_bulk_run1")
oarfish_dir_run2    <- file.path(dRNA_comparison_dir, "oarfish_cov_output/dRNA_bulk_run2")
read_count_dir      <- file.path(dRNA_comparison_dir, "qc/read_counts")

cell_lines <- c("H146", "H69", "H526", "H211", "SHP77", "H1975", "H2228", "HCC827")
```

# Read Counts per Cell Line

Comparison of sequencing depth between run1 (original) and run2 (top-up) across all eight cell lines.

```{r read_counts}
load_read_counts <- function(run_id, cell_lines, read_count_dir) {
  map_dfr(cell_lines, function(cl) {
    f <- file.path(read_count_dir, paste0(run_id, "_", cl, ".count"))
    tibble(
      run  = run_id,
      cell_line = cl,
      reads = as.numeric(readLines(f, n = 1))
    )
  })
}

rc_run1 <- load_read_counts("dRNA_bulk_run1", cell_lines, read_count_dir)
rc_run2 <- load_read_counts("dRNA_bulk_run2", cell_lines, read_count_dir)
rc_all  <- bind_rows(rc_run1, rc_run2) %>%
  mutate(run = factor(run, levels = c("dRNA_bulk_run1", "dRNA_bulk_run2"),
                      labels = c("Run1 (original)", "Run2 (top-up)")))

p_rc <- ggplot(rc_all, aes(x = cell_line, y = reads / 1e6, fill = run)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  labs(title = "Read counts per cell line",
       x = "Cell line", y = "Reads (millions)", fill = "Run") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
print(p_rc)

rc_wide <- rc_all %>%
  pivot_wider(names_from = run, values_from = reads)
knitr::kable(rc_wide, caption = "Read counts per cell line and run")
```

# Oarfish Quantification: Transcript-Level Comparison

For each cell line, load oarfish `.quant` files and compare estimated counts between run1 and run2.

```{r load_oarfish}
load_oarfish_quant <- function(run_dir, cell_lines) {
  map_dfr(cell_lines, function(cl) {
    cl_dir  <- file.path(run_dir, cl)
    qf      <- list.files(cl_dir, pattern = "\\.quant$", full.names = TRUE, all.files = TRUE)
    if (length(qf) == 0) return(NULL)
    read_tsv(qf[1], show_col_types = FALSE) %>%
      select(tname, num_reads) %>%
      rename(transcript_id = tname, count = num_reads) %>%
      mutate(cell_line = cl)
  })
}

quant_run1 <- load_oarfish_quant(oarfish_dir_run1, cell_lines) %>% rename(count_run1 = count)
quant_run2 <- load_oarfish_quant(oarfish_dir_run2, cell_lines) %>% rename(count_run2 = count)

quant_merged <- inner_join(quant_run1, quant_run2, by = c("transcript_id", "cell_line"))
```

## Detected Transcripts per Run

Number of transcripts with >0 estimated counts in each run.

```{r detected_transcripts}
det_summary <- quant_merged %>%
  group_by(cell_line) %>%
  summarise(
    n_run1     = sum(count_run1 > 0),
    n_run2     = sum(count_run2 > 0),
    n_both     = sum(count_run1 > 0 & count_run2 > 0),
    n_run1_only = sum(count_run1 > 0 & count_run2 == 0),
    n_run2_only = sum(count_run2 > 0 & count_run1 == 0),
    .groups = "drop"
  )
knitr::kable(det_summary, caption = "Transcript detection summary per cell line")

det_long <- det_summary %>%
  pivot_longer(c(n_run1, n_run2), names_to = "run", values_to = "n_detected") %>%
  mutate(run = factor(run, levels = c("n_run1", "n_run2"),
                      labels = c("Run1 (original)", "Run2 (top-up)")))

p_det <- ggplot(det_long, aes(x = cell_line, y = n_detected, fill = run)) +
  geom_col(position = position_dodge(0.8), width = 0.7) +
  labs(title = "Detected transcripts (count > 0) per cell line",
       x = "Cell line", y = "Number of transcripts", fill = "Run") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
print(p_det)
```

## Quantification Correlation

Scatter plots of log2(count + 1) between run1 and run2 per cell line.

```{r correlation_plots, fig.width=16, fig.height=12}
corr_stats <- quant_merged %>%
  filter(count_run1 > 0 & count_run2 > 0) %>%
  group_by(cell_line) %>%
  summarise(
    pearson  = cor(log2(count_run1 + 1), log2(count_run2 + 1), method = "pearson"),
    spearman = cor(count_run1, count_run2, method = "spearman"),
    .groups  = "drop"
  )
knitr::kable(corr_stats %>% mutate(across(c(pearson, spearman), ~round(.x, 4))),
             caption = "Correlation between run1 and run2 (commonly detected transcripts: count > 0 in both runs)")

plots <- map(cell_lines, function(cl) {
  d <- quant_merged %>%
    filter(cell_line == cl, count_run1 > 0 & count_run2 > 0) %>%
    mutate(log_r1 = log2(count_run1 + 1),
           log_r2 = log2(count_run2 + 1))
  r  <- filter(corr_stats, cell_line == cl)
  ggplot(d, aes(x = log_r1, y = log_r2)) +
    geom_point(alpha = 0.25, size = 0.6) +
    geom_abline(slope = 1, intercept = 0, colour = "red", linetype = "dashed") +
    annotate("text", x = Inf, y = -Inf, hjust = 1.1, vjust = -0.5,
             label = sprintf("r = %.3f\nρ = %.3f", r$pearson, r$spearman), size = 3.5) +
    labs(title = cl,
         x = "log2(count+1) Run1", y = "log2(count+1) Run2") +
    theme_bw(base_size = 11)
})
wrap_plots(plots, ncol = 4)
```

# Transcript Detection Consistency

Fraction of transcripts detected in both runs, run1-only, or run2-only.

```{r detection_consistency}
consist <- quant_merged %>%
  filter(count_run1 > 0 | count_run2 > 0) %>%
  mutate(category = case_when(
    count_run1 > 0 & count_run2 > 0 ~ "Both",
    count_run1 > 0 & count_run2 == 0 ~ "Run1 only",
    count_run1 == 0 & count_run2 > 0 ~ "Run2 only"
  )) %>%
  count(cell_line, category) %>%
  group_by(cell_line) %>%
  mutate(pct = n / sum(n) * 100)

p_consist <- ggplot(consist, aes(x = cell_line, y = pct, fill = category)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c("Both" = "#4393C3", "Run1 only" = "#D6604D", "Run2 only" = "#74ADD1")) +
  labs(title = "Transcript detection consistency across runs",
       x = "Cell line", y = "Percentage (%)", fill = "Detected in") +
  theme_bw(base_size = 13) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
print(p_consist)
```

# Summary

```{r summary_table}
summary_tbl <- rc_wide %>%
  left_join(det_summary, by = "cell_line") %>%
  left_join(corr_stats, by = "cell_line")
knitr::kable(summary_tbl %>% mutate(across(c(pearson, spearman), ~round(.x, 4))),
             caption = "Combined summary: read counts, detection, and correlation per cell line")
tsv_path <- if (!is.null(params$fig.path)) {
  sub("_figures[/]*$", "_summary.tsv", params$fig.path)
} else {
  file.path(dRNA_comparison_dir, "dRNA_run_comparison_summary.tsv")
}
write_tsv(summary_tbl, tsv_path)
message("Summary table written to: ", tsv_path)
```
